---
title: "Session 4 — Survey Design & Measurement Methods"
subtitle: "M731 Week 4 · September 30, 2026"
author: "Shay Abbasi"
date: "2026-09-30"
---

## Session 4 {background-color="#7A003C"}

Survey design and measurement methods

::: {.notes}
Collect any graded work due today per A2L logistics - keep it under 2
min. Then dive into content.
:::

## Today's agenda

1. From construct to item to questionnaire
2. Question wording — the small choices with big consequences
3. Response formats and scale choice
4. Questionnaire structure and flow
5. **Break (15 min)**
6. R lab: cleaning messy survey exports
7. Ethics: consent, sensitive items, respondent burden
8. Industry: Qualtrics demo
9. Team questionnaire draft workshop

::: {.notes}
Draft workshop is peer-review of team questionnaire drafts. Teams
should have brought a draft on their laptop. If not, they draft during
the workshop.
:::

## Learning objectives

By end of session 4 you will be able to:

- Move from a construct to items to a well-flowing questionnaire
- Identify sources of measurement error from wording, format, and order
- Load, clean, and reverse-code survey exports in R
- Peer-review a questionnaire against a design checklist

::: {.notes}
Peer-review is a graduate-level skill. Encourage constructive critique
- teams that give tough feedback get tough feedback back, and all
questionnaires improve.
:::

# Part 1 — Construct to item {background-color="#7A003C"}

## The construct-item-questionnaire chain

- **Construct** — the abstract thing you want to measure (brand attachment, satisfaction, price sensitivity)
- **Items** — specific questions that operationalise the construct
- **Questionnaire** — the ordered set of items delivered to respondents

Every item should trace back to a construct. Every construct should have multiple items.

::: {.notes}
Multiple items per construct is the reliability rationale (Wk 8).
Single-item measures are unreliable and unvalidated. Rule of thumb:
3-5 items per construct minimum, more for abstract constructs like
brand attachment.
:::

## Example: brand attachment for Maritime Roast

Construct: *how emotionally attached is a subscriber to Maritime Roast?*

Items (5-point agreement):

1. Maritime Roast is part of my morning routine
2. I feel connected to Maritime Roast as a brand
3. I would be upset if Maritime Roast disappeared
4. I recommend Maritime Roast to friends
5. My identity is tied to being a Maritime Roast customer

::: {.notes}
Draw the loading from Park et al. brand-attachment work if familiar.
Not perfect items - discuss with the class which is weakest. Item 5
is arguably measuring self-brand connection, not attachment - shows
how items can drift from the construct.
:::

## Item wording principles

- **Simple language.** Grade 6-8 reading level.
- **One idea per item.** No double-barrelled questions.
- **Neutral framing.** Not leading, not loaded.
- **Answerable.** Respondent has the information.
- **Universally applicable.** No assumed context.

::: {.notes}
Double-barrelled example: "Maritime Roast is high-quality and
affordable." What if it's high-quality but expensive? Respondent can't
answer without splitting. Teach students to sniff these out.
:::

## Question types

- **Closed-ended** (fixed options): easier to analyse, faster to answer, but constrained
- **Open-ended** (free text): richer answers, harder to analyse, longer to answer

Most quantitative surveys are 90% closed-ended, with 1-2 open-ended for depth.

::: {.notes}
Open-ended is now more tractable with LLM-based coding but still
adds time and cost. Keep to 1-2 in most surveys - "What is the most
important thing you'd change?" is a classic and always earns its
place.
:::

# Part 2 — Response formats {background-color="#7A003C"}

## Common closed-ended formats

- **Dichotomous:** yes/no, agree/disagree
- **Multiple choice:** pick one
- **Multiple select:** pick all that apply
- **Likert:** 5- or 7-point agreement
- **Semantic differential:** bipolar adjectives (Hot ← → Cold)
- **Numeric rating:** 0-10, NPS
- **Constant sum:** allocate 100 points
- **Rank order:** rank options 1 to k

::: {.notes}
Each has its use case. Likert is workhorse for attitudes. NPS is
workhorse for loyalty. Constant sum for trade-offs (budget
allocation). Rank order for preference among small option sets.
:::

## Likert design choices

- **Points:** 5 vs 7 vs 10
- **Labels:** all points labelled? or just endpoints?
- **Neutral point:** included or forced choice?
- **Reversed items:** to detect straight-lining

Best practice for MR: 5- or 7-point, all points labelled, include neutral, reverse ~20% of items.

::: {.notes}
Debate about neutral point: forcing a choice eliminates fence-sitting
but introduces measurement error (respondents fake a side). Include a
neutral option and note the proportion who use it.
:::

## Semantic differential

Adjective pairs:

- Boring ← 1 2 3 4 5 6 7 → Exciting
- Cheap ← 1 2 3 4 5 6 7 → Expensive
- Local ← 1 2 3 4 5 6 7 → Global

Good for brand image work.

::: {.notes}
Classic for brand perception. Osgood, Suci, Tannenbaum 1957. Still
widely used in brand tracking. Osgood found three universal
dimensions: evaluation, potency, activity.
:::

## NPS (Net Promoter Score)

"How likely are you to recommend [X] to a friend or colleague?" 0-10.

- Promoters: 9-10
- Passives: 7-8
- Detractors: 0-6
- NPS = %Promoters − %Detractors

Simple, standard, widely tracked. Researchers dislike the information loss.

::: {.notes}
NPS has been challenged academically (Keiningham et al. 2007) but
remains an industry standard because CEOs understand it and it tracks
consistently. For your project, use NPS if the client tracks NPS;
otherwise use a fuller satisfaction scale.
:::

# Part 3 — Questionnaire structure {background-color="#7A003C"}

## Order effects

- **Screening** first (do they qualify?)
- **Warm-up** questions (easy, engaging)
- **Main content** (grouped by topic)
- **Sensitive items** later (once trust built)
- **Demographics** last (unless needed for skip logic)

Random or fixed order for battery items? Random reduces order bias.

::: {.notes}
Order effects are real. Asking about brand awareness before brand
preference primes the awareness set. Randomise brand order within
awareness questions.
:::

## Length matters

- Under 5 min: high completion rates
- 5-10 min: standard, moderate drop-out
- 10-15 min: needs incentive
- Over 15 min: high drop-out, quality decay

Budget your questions. Every added question increases drop-out.

::: {.notes}
Real-world constraint. If a client wants a 25-min survey, push back -
data quality suffers. Better to run two shorter surveys with
overlapping samples.
:::

## Skip logic and piping

- **Skip logic:** show/hide questions based on prior answers
- **Piping:** insert prior answers into later question text ("You said you subscribe to [Signature]. How satisfied with [Signature]?")

Modern survey platforms (Qualtrics, SurveyMonkey) do these easily.

::: {.notes}
Show a Qualtrics example if possible. Skip logic reduces respondent
burden (non-subscribers don't see subscription-tier questions).
Piping increases relevance (personalised feel).
:::

## Straight-lining and speeder checks

Build in:

- **1-2 reverse-worded items** in every battery (detects straight-lining)
- **Attention check** halfway through ("Select 'Somewhat agree' for quality control")
- **Time-in-survey** captured for post-hoc speeder identification

Flag ~5-10% typical straight-line/speeder rate.

::: {.notes}
Attention checks are contested - some argue they annoy honest
respondents. Reasonable compromise: one attention check in a 15-min
survey, framed politely.
:::

## Break — 15 minutes {background-color="#FDBF57"}

::: {.notes}
Have `kaggle_survey_raw.csv` open on the projector. This is a real
messy dataset - students will feel the pain of dirty data quickly.
:::

# Part 4 — R lab: cleaning survey exports {background-color="#7A003C"}

## Loading a real messy export

```{r}
#| eval: false
library(tidyverse)
library(janitor)

raw <- read_csv("data/kaggle_survey_raw.csv")

dim(raw)
glimpse(raw)
```

`kaggle_survey_raw.csv` — Kaggle "Young People Survey," 1010 respondents, mixed dirty.

::: {.notes}
Live-load. Show glimpse. Column names are inconsistent (spaces,
capitals, punctuation). Values include "" for missing but sometimes
NA, and sometimes actual "NA" as a string. This is real-world messy.
:::

## Fix column names with `janitor`

```{r}
#| eval: false
raw <- raw |> clean_names()
names(raw)
```

`clean_names()` converts to snake_case, replaces punctuation, deduplicates.

::: {.notes}
This is the first thing every dirty-data workflow should do. Saves
hours of quoting weird column names later. `janitor` is a
Wickham-adjacent package with many small QoL functions.
:::

## Reverse-coding

```{r}
#| eval: false
# Suppose items "loneliness_1" and "loneliness_2" are reverse-worded
# 1 = strongly disagree, 5 = strongly agree
# Reverse: 6 - x

raw <- raw |>
  mutate(across(c(loneliness_1, loneliness_2), ~ 6 - .))
```

Reverse before computing composite scores.

::: {.notes}
Common bug: forgetting to reverse. Then Cronbach's alpha comes back
low or negative and students panic. Always document which items are
reverse-coded in a code book.
:::

## Handling missing values

```{r}
#| eval: false
# Count missing per column
raw |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "col", values_to = "n_missing") |>
  arrange(desc(n_missing)) |>
  head(20)
```

Decide per column: drop, impute, or keep NA.

::: {.notes}
Show output. Some columns will be nearly all missing (skip logic).
Others just have a few. The decision to impute vs drop is a research
call - impute for models that need complete cases, drop for
descriptive analysis. Never blanket-impute silently.
:::

## Straight-line detection

```{r}
#| eval: false
# Compute row-wise SD across a battery of Likert items
music_items <- raw |>
  select(starts_with("music_"))

raw$music_sd <- apply(music_items, 1, sd, na.rm = TRUE)

# Straight-liners have very low SD
raw |>
  arrange(music_sd) |>
  select(music_sd, starts_with("music_")) |>
  head(10)
```

::: {.notes}
Row-wise operations are slightly awkward in tidyverse - the `apply()`
call keeps it simple. Threshold for flagging depends on the battery -
typically SD < 0.3 for a 5-point scale is suspicious.
:::

## Composite scale construction

```{r}
#| eval: false
raw <- raw |>
  mutate(
    music_score = rowMeans(select(., starts_with("music_")), na.rm = TRUE)
  )

hist(raw$music_score)
```

Mean or sum? Use mean if items may have missing values; sum otherwise.

::: {.notes}
For your project questionnaire, plan the composites at design time.
Documenting which items compose which construct saves confusion at
analysis time.
:::

## In-lab exercise

Using `kaggle_survey_raw.csv`:

1. Clean column names
2. Identify and reverse-code any items that need reversing (check the data dictionary)
3. Flag straight-liners on the "movies_" battery
4. Compute a composite "movie enthusiasm" score and plot its distribution
5. Write your cleaning script as a `.qmd` that renders cleanly

A fuller worked version of this cleaning pipeline is in
**`solutions-04.qmd`** in this session's folder.

::: {.notes}
Give ~30 min. Circulate. This lab is a good test of R fluency -
students who struggle here should invest more time before later
graded work or the project.
:::

# Part 5 — Ethics thread {background-color="#7A003C"}

## Informed consent

Every survey must have:

- **Purpose** of the research
- **Who's collecting** (identifiable sponsor)
- **What data** is collected and how used
- **Voluntary participation** and right to withdraw
- **Data storage** and retention
- **Contact** for questions

Usually appears on the first page as a consent screen.

::: {.notes}
Show a real consent screen from a McMaster-approved study if
available. Length is a trade-off: complete disclosure vs. respondent
patience. Most panels require standard consent language.
:::

## Sensitive items

Handle carefully:

- Income, wealth, debt
- Health conditions, mental health
- Illegal or stigmatised behaviour
- Political views, religion
- Sexual orientation, gender identity

Techniques: bracketed ranges (not exact income), self-administered mode, "prefer not to answer" option, question near end.

::: {.notes}
Ask: what makes an item "sensitive"? Answer: potential harm from
disclosure. Different across cultures. Political affiliation is
low-sensitivity in the US, higher in some countries. Design with the
population in mind.
:::

## Respondent burden

- Every question costs the respondent time and attention
- Long surveys pay in drop-out, not just money
- Randomise item order in long batteries to spread fatigue

Design for the respondent, not just for your analysis wish-list.

::: {.notes}
Common failure mode: client stakeholders add "just one more question"
each. By day 5 the survey is 45 min. Push back. Every question needs
a plan for how the answer will be used.
:::

# Part 6 — Industry: Qualtrics {background-color="#7A003C"}

## McMaster has Qualtrics

- Site licence — free for students, staff, faculty
- Log in at `mcmaster.ca1.qualtrics.com`
- Full features (survey flow, block logic, distributions, exports)

Anyone who's used a survey tool in Canadian industry has probably used Qualtrics.

::: {.notes}
Have your Qualtrics account open. 15-min live tour: create survey,
add block, add question, set skip logic, generate distribution link,
export as CSV. Students appreciate seeing the tool they'll actually
use.
:::

## Qualtrics-to-R workflow

1. Design survey in Qualtrics
2. Launch (email, link, panel API)
3. Export data as CSV (with numeric coding, not labels)
4. Read into R with `read_csv()`
5. Clean, reverse-code, compute composites
6. Analyse

Alternate: `qualtRics` package for direct API access.

::: {.notes}
Numeric coding vs. labels: numeric is easier for analysis but you
lose the meaning. Best practice: export both, use numeric for
computation and labels for reporting. `qualtRics` package is nice for
recurring pulls (e.g., a monthly tracker).
:::

# Part 7 — Team workshop {background-color="#7A003C"}

## Peer-review your team questionnaire draft

Each team:

1. Trade questionnaire drafts with another team
2. Apply this checklist:
   - [ ] Every construct has ≥3 items
   - [ ] No double-barrelled questions
   - [ ] Reversed items included
   - [ ] Length ≤ 12 min
   - [ ] Consent language present
3. Return written critique

::: {.notes}
Time: 25-30 min. Circulate. Push teams to be honest, not polite.
Every draft has weaknesses. Catching them now saves rework at Wk 8
(scales revision) and Wk 9 (project deliverable).
:::

# Part 8 — Wrap-up {background-color="#7A003C"}

## Week 5 preview

- **Concept:** qualitative marketing research (focus groups, IDIs, projective techniques)
- **Test 1** in Week 5 (covers Lectures 1-4)
- **R lab compressed** — mostly test administration

Test 1 review sheet on A2L by Friday.

::: {.notes}
Give students the study advice: focus on process, design vocabulary,
sampling logic, measurement scales. Practice questions on A2L. Test 1
will be MC and/or written - exact format TBD.
:::

## What to study for Test 1

- MR process (7 steps)
- Research design families (exploratory / descriptive / causal)
- Scale types (nominal / ordinal / interval / ratio) + legal statistics
- Data-collection modes and trade-offs
- Sampling: probability vs non-probability, sample-size logic
- Questionnaire design principles

::: {.notes}
Emphasise: I don't test trivia. I test whether students can *apply*
concepts. Sample questions on A2L show the style.
:::

## Thank you — questions?

`abbass44@mcmaster.ca`

::: {.notes}
Stay for test-prep questions and questionnaire peer-review issues.
:::
