---
title: "Session 4 — Survey Design & Measurement Methods"
subtitle: "M731 Week 4 · September 30, 2026 · Comprehensive Reading Notes"
author: "Shay Abbasi"
date: "2026-09-30"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    self-contained: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

# Overview

Survey design is where research planning becomes measurement. In Sessions 1–3, we established *what* to measure (the research question), *who* to measure (the sampling plan), and *at what scale* (nominal/ordinal/interval/ratio). Session 4 asks *how* to measure: how to move from abstract constructs to specific questionnaire items, how to arrange those items into a flowing survey instrument, and how to clean and prepare the resulting data for analysis.

Measurement error — introduced at the instrument design stage — is one of the most consequential and least visible sources of error in marketing research. A poorly-worded question biases every downstream analysis that uses it, and no statistical technique can detect or correct for this bias. Getting the instrument right is therefore one of the highest-leverage investments in the research process.

---

# From Construct to Item to Questionnaire

## The Measurement Chain

The path from a vague research interest to a clean dataset follows a three-link chain:

**Construct → Items → Questionnaire**

A **construct** is the abstract concept you intend to measure: brand attachment, price sensitivity, satisfaction, perceived quality. Constructs are latent — they cannot be directly observed. They are inferred from observable responses to specific questions.

An **item** is a specific question or stimulus that operationalises some aspect of the construct. Multiple items are used for each construct because any single item is an imperfect and unreliable indicator of a latent concept. The collective evidence from 4–7 items provides a more stable, less error-prone measure than any single question.

A **questionnaire** is the ordered, formatted set of items delivered to respondents through a survey platform.

Every item should trace back to a construct. Every construct should have multiple items. An item that cannot be linked to a construct has no place in the questionnaire — it is either measuring something irrelevant or measuring something relevant that was never explicitly defined.

## The Three-Step Operationalisation Process

Moving from construct to items requires three sequential decisions (Park et al., 2010, adapted):

**Step 1 — Definition**: What does the construct mean, precisely? "Brand attachment" is not self-evident. Park et al. (2010) define it as the emotional bond between a person and a brand, characterised by identity, affection, and passion. The definition determines what domains the items must cover and, equally importantly, what the items should *not* measure (advocacy, which is behaviour, not attachment; awareness, which is cognitive recall, not emotional connection).

**Step 2 — Domain mapping**: What aspects of the construct must be covered by the items? Brand attachment has three components per Park et al.: identity (the brand is part of who I am), affection (I have warm feelings toward the brand), and passion (I feel deeply connected). A well-developed item set covers all three domains — an item set that covers only one domain underrepresents the construct (construct underrepresentation, a form of content invalidity).

**Step 3 — Item form**: How should each domain aspect be measured? Agreement statements ("I feel connected to Maritime Roast as a brand"), semantic differentials (Traditional ←→ Contemporary), frequency scales (How often?), or behavioural frequency counts (How many times in the past month?) are different formats with different strengths.

## Example: Operationalising Brand Attachment for Maritime Roast

**Construct definition**: the emotional bond that a subscriber feels toward Maritime Roast as a brand — encompassing identity (the brand reflects who I am), affection (I feel warmly toward the brand), and passion (I feel strongly connected).

**Draft items** (5-point agreement scale):

1. Maritime Roast is part of my morning routine *(draft — targets habit, not identity)*
2. I feel connected to Maritime Roast as a brand *(targets affection)*
3. I would be upset if Maritime Roast disappeared *(targets passion/loss aversion)*
4. I recommend Maritime Roast to friends *(targets advocacy — wrong domain)*
5. My identity is tied to being a Maritime Roast customer *(targets identity — or self-brand connection?)*

**Critique of the draft items**:

| Item | Issue | Fix |
|---|---|---|
| 1. "part of my morning routine" | Measures habit, not emotional bond | Revise to: "I look forward to my Maritime Roast coffee each morning" |
| 2. "I feel connected" | Close to construct; acceptable | Keep |
| 3. "I would be upset if MR disappeared" | Classic attachment/loss-aversion item | Keep |
| 4. "I recommend to friends" | Measures advocacy behaviour, not attachment emotion | Move to an NPS or advocacy scale |
| 5. "My identity is tied to..." | Closer to self-brand connection (a related but distinct construct in Park et al.) | Revise to be clearer about emotional identity, or move to a self-brand connection scale |

**Revised battery**: Items 2, 3, plus revised item 1, plus "I feel an emotional connection to Maritime Roast" and "Being a Maritime Roast subscriber says something about who I am" — giving 5 items covering affection, passion, and identity domains.

---

# Question Wording

## Core Wording Principles

Good survey questions obey five principles:

1. **Simplicity**: written at a grade 6–8 reading level. Respondents in an online panel include people across the full education spectrum; complex vocabulary and long sentences introduce miscomprehension that looks like random error in the data.

2. **Single focus**: each item measures exactly one thing. A double-barrelled question ("Maritime Roast is high-quality and affordable") cannot be answered if the respondent agrees with one clause and disagrees with the other.

3. **Neutral framing**: the question should not push respondents toward a particular answer. Leading questions ("Don't you think Maritime Roast offers excellent value?") and loaded language ("excessive pricing," "irresponsible data practices") systematically bias responses.

4. **Answerability**: respondents must actually have the information required to answer the question. Asking "How much did you spend on coffee in the past 12 months?" requires recall over an entire year — most people cannot do this accurately. Asking "Roughly how much do you spend on coffee per month?" is more answerably framed.

5. **Universal applicability within the target population**: the question must make sense for every qualifying respondent. "How satisfied are you with your Maritime Roast subscription renewal experience?" excludes people who have not yet gone through a renewal — for them, it is unanswerable.

## Common Question Wording Errors

| Error type | Bad example | Problem | Better version |
|---|---|---|---|
| Double-barrelled | "The product is high quality and good value" | Cannot agree with one without both | Split into two separate items |
| Leading question | "Don't you agree Maritime Roast is superior?" | Presupposes agreement | "Compared to other coffee subscriptions, Maritime Roast is..." |
| Loaded language | "Do you support excessive taxation on coffee?" | "Excessive" is evaluative | "Do you support an additional tax on coffee products?" |
| Unanswerable | "How much did you spend on coffee last year?" | Recall over 12 months is unreliable | "About how much do you spend on coffee per month?" |
| Assumed context | "How satisfied with your renewal experience?" | Non-renewers cannot answer | Add skip logic: only ask renewers |
| Jargon | "Rate our omnichannel customer experience" | "Omnichannel" is industry language | "Rate your experience across different ways of contacting us" |

## The Krosnick Principle: Small Wording Changes Shift Results

Jon Krosnick's decades of survey methodology research (Krosnick & Presser, 2010) demonstrate that apparently minor wording choices produce large, systematic differences in measured responses:

- Replacing "support" with "favour" or "endorse" in an attitude question can shift responses by 5–15 percentage points
- Acquiescence bias — the tendency to agree with agreement-type statements regardless of content — inflates positive attitude scores by 10–25% in surveys that lack reverse-worded items
- Social desirability bias on sensitive items (income, health behaviour, political attitudes) produces responses that skew toward socially acceptable answers
- Question order effects mean that prior questions prime particular considerations, shifting subsequent responses in predictable directions

The implication is uncomfortable: a researcher who writes questions quickly without pilot testing is almost certainly producing systematically wrong data, and the data itself cannot reveal this.

---

# Response Formats

## Closed-Ended Question Types

Closed-ended questions present a fixed set of response options. They are faster to complete, easier to analyse, and more consistent across respondents. The choice among closed-ended formats depends on the question objective:

| Question objective | Best format | Rationale |
|---|---|---|
| Does the respondent have characteristic X? | Dichotomous (Yes/No) | Simple, unambiguous |
| Which one option does the respondent prefer? | Single-choice | Forces a clear preference ranking |
| Which things apply simultaneously? | Multi-select | When multiple answers are genuinely possible |
| How much does respondent agree with statements? | Likert 5 or 7 | Workhorse for attitude measurement |
| How does the respondent perceive brand image? | Semantic differential | Good for bipolar attribute ratings |
| How likely to recommend? | NPS 0–10 | Industry standard; widely benchmarked |
| How should priority be allocated among attributes? | Constant sum | Forces trade-offs; reduces positivity bias |
| Which features are most preferred? | Rank order or MaxDiff | Better than ratings for relative preferences |

### Likert Scale Design Decisions

Four design decisions arise with every Likert scale:

**Number of points**: 5 vs. 7 vs. 10. More points provide more resolution but increase cognitive burden. The 5-point scale is standard for concrete evaluative items (satisfaction, quality). The 7-point scale is preferred for abstract attitudinal constructs (brand attachment, values alignment) where respondents can genuinely distinguish finer gradations. The 10-point scale is used mainly for NPS.

**Full labelling vs. endpoint-only labelling**: Research by Krosnick and Fabrigar (1997) shows that fully-labelled scales (all 5 or 7 points have text anchors) produce more reliable data than endpoint-only scales. Without intermediate labels, different respondents interpret the interior points differently, adding random error. This course recommends fully-labelled scales for all Likert items.

Recommended label sets:
- 5-point: Strongly disagree — Disagree — Neither agree nor disagree — Agree — Strongly agree
- 7-point: Strongly disagree — Disagree — Somewhat disagree — Neither — Somewhat agree — Agree — Strongly agree
- 5-point satisfaction: Very dissatisfied — Dissatisfied — Neither — Satisfied — Very satisfied

**Neutral midpoint**: include or force-choice? Including a neutral point ("Neither agree nor disagree") is generally preferred. It allows genuine ambivalence to be expressed and reduces the error introduced when respondents who are genuinely neutral are forced to pick a side. The proportion using the neutral option is itself analytically useful (high neutral use may signal an unclear question).

**Reverse-worded items**: at least 20% of items in any Likert battery should be reverse-worded (see Section 5 below).

### Semantic Differential

Semantic differential items present bipolar adjective pairs with a numbered scale between them:

> Traditional ← 1 2 3 4 5 6 7 → Contemporary

Developed by Osgood, Suci, and Tannenbaum (1957), semantic differentials are particularly useful for brand image research. Factor analysis of semantic differential batteries reliably recovers three universal dimensions across cultures: **Evaluation** (good–bad), **Potency** (strong–weak), and **Activity** (active–passive). Applied to Maritime Roast brand perception, the three dimensions might yield factors labeled Quality/Value, Brand Authority, and Brand Dynamism.

### Net Promoter Score

NPS asks one question: "How likely are you to recommend [X] to a friend or colleague?" on a 0–10 scale. Respondents are categorised as Promoters (9–10), Passives (7–8), or Detractors (0–6). NPS = %Promoters − %Detractors, which ranges from −100 to +100.

NPS is ubiquitous in industry because it is simple, universally understood, and benchmarkable across categories. Academically, it has been challenged: Keiningham et al. (2007) found that across 21 companies in five industries, NPS predicted growth no better than standard customer satisfaction measures. Collapsing the 0–10 scale into three categories also loses approximately 40% of the information in the original scale.

**The practical resolution**: analyse the full 0–10 distribution for internal research purposes; report NPS for external benchmarking and executive communication. If the client tracks NPS historically, maintain it for trend continuity; but do not treat the derived score as a richer measure than it is.

---

# Questionnaire Structure and Flow

## The Standard Ordering Principle

Survey questions should follow this general ordering:

1. **Screening questions** — establish that the respondent qualifies before asking anything substantive. If they do not qualify, exit them gracefully without asking questions they cannot answer.
2. **Warm-up questions** — easy, non-threatening, and engaging. General behaviour questions ("How often do you drink coffee?") work well. These build rapport and get respondents into the rhythm of the survey.
3. **Main content blocks** — the core measurement items, grouped by topic or construct. Move from easier/more objective items toward more abstract/attitudinal items.
4. **Sensitive items** — placed after engagement is established, when social trust in the survey context is higher. Income, political opinions, and health behaviours perform better near the end.
5. **Demographics** — always last, unless demographic characteristics are needed for skip logic. Demographics are functional (I need this for analysis) but feel intrusive when asked early.

## Context Effects: How Prior Questions Bias Responses

Tourangeau, Rips, and Rasinski (2000) document three reliable context effects that must be managed through questionnaire design:

**Assimilation**: a prior question makes the topic salient, and responses to the next question assimilate toward it. If you ask about excellent service experiences first, subsequent overall satisfaction ratings inflate. In the Maritime Roast context: asking about the best coffee the respondent has ever tasted before rating Maritime Roast quality will inflate those quality ratings.

**Contrast**: a prior question establishes a high standard, against which subsequent responses contrast downward. Asking about a memorable service failure before asking about overall satisfaction will deflate those satisfaction scores.

**Order primacy in lists**: options presented first in a multiple-choice question receive disproportionately more responses than options presented later (primacy effect for visually-presented lists) or last (recency effect for verbally-presented lists).

The primary mitigation is **randomisation**: randomise question order within topic blocks and randomise response option order for lists. Modern survey platforms (Qualtrics) implement this at the block level. The limit of randomisation is logical dependency: questions where B logically follows A cannot be randomised.

## Survey Length and Completion Rates

Expected completion rates as a function of survey length:

- Under 5 minutes: very high completion (85–95%)
- 5–10 minutes: standard commercial length; moderate drop-out (70–85%)
- 10–15 minutes: noticeable drop-out acceleration; incentive usually required
- Over 15 minutes: high drop-out, quality decay from respondent fatigue; strongly discouraged without substantial incentive

Estimating survey completion time before fielding:

| Item type | Avg. seconds |
|---|---|
| Screener question | 8–15 |
| Multiple-choice (pick one) | 8–12 |
| Likert item | 6–10 |
| Semantic differential pair | 8–12 |
| Open-ended (short answer) | 30–60 |
| Open-ended (full paragraph) | 60–120 |
| Demographic item | 5–8 |
| Survey instruction screen | 10–15 |

Build the time budget before finalising question count. Then pilot-test with 20–30 respondents and use the median completion time (not mean — outliers who take breaks inflate the mean) to calibrate.

## Skip Logic and Piping

**Skip logic** (also called display logic in Qualtrics) shows or hides questions based on prior answers. Non-subscribers should not see subscription satisfaction items. Respondents who have not renewed should not see renewal experience items. Proper skip logic both reduces respondent burden and eliminates invalid responses.

Sketch the complete survey flow on paper before programming:

```
Q1: "Are you a current Maritime Roast subscriber?"
  → YES → show subscription experience block (Q2–Q4)
  → NO  → show awareness and intent block (Q5–Q7)

Q2: "Which tier?" [Standard / Signature / Discovery]
  → each option routes to tier-specific satisfaction item

Q3: "How satisfied with your [tier name] subscription?" (1–7)
Q4: "What one thing would improve your [tier name] experience?" (open)
  → merge to Q8

Q5 (non-subscribers): "Have you heard of Maritime Roast?"
  → YES → brand image battery (Q6)
  → NO  → category needs only (Q7)

Q8: Demographics (all respondents)
```

**Piping** inserts prior answers into subsequent question text. "You mentioned subscribing to [Signature] — how satisfied are you with your Signature experience?" creates a personalised question at no additional programming cost.

---

# Reverse Coding

## Why Reverse Coding Matters

Without reverse-worded items, all items in a Likert battery point in the same direction. A respondent who circle all 5s ("Strongly agree") with every item appears highly positive — but may simply be choosing without reading (straight-lining). Reverse-coded items break this: a genuine high-attachment respondent should select a *low* score on a negatively-worded attachment item.

For a 5-item brand attachment battery, items 2 and 4 might be reverse-worded:

| Item | Direction | Expected response (high-attachment subscriber) |
|---|---|---|
| "I feel connected to Maritime Roast" | Positive | 5 (Strongly agree) |
| "Maritime Roast is not worth its price" | **Reverse** | 1 (Strongly disagree) |
| "I would be upset if Maritime Roast disappeared" | Positive | 5 (Strongly agree) |
| "I could easily switch to another subscription" | **Reverse** | 1–2 (Disagree) |
| "Maritime Roast reflects who I am" | Positive | 5 (Strongly agree) |

## Reversing in R Before Computing Composites

```{r}
#| label: reverse-coding
#| eval: false
# For a 5-point scale: reversed score = 6 - original
# For a 7-point scale: reversed score = 8 - original
# General formula: reversed = (max + min) - original = (5 + 1) - x = 6 - x

survey <- survey |>
  mutate(
    attach_2_r = 6 - brand_attach_2,   # "not worth its price"
    attach_4_r = 6 - brand_attach_4    # "could easily switch"
  )

# Verify: the correlation between the original and reversed item should be negative
cor(survey$brand_attach_2, survey$attach_2_r)   # should be -1.0

# Composite score: average of all items (both original and reverse-coded)
survey <- survey |>
  mutate(
    brand_attachment_score = rowMeans(
      select(., brand_attach_1, attach_2_r, brand_attach_3, attach_4_r, brand_attach_5),
      na.rm = TRUE
    )
  )
```

Always document reverse-coded items in your codebook. Failing to reverse before computing composites produces a Cronbach's alpha near zero or negative — a diagnostic sign of a forgotten reversal, not a validity problem.

---

# R Lab: Cleaning Survey Exports

## The Reality of Survey Export Files

Real survey export files from Qualtrics and similar platforms are rarely clean. The most common problems:

| Problem | Example | Fix |
|---|---|---|
| Inconsistent column names | `"Brand_Trust 1"`, `"brand trust2"` | `janitor::clean_names()` |
| Numeric stored as character | `"5"` instead of `5` | `as.numeric()` |
| Missing coded as blank string | `""` | `na_if(., "")` |
| Missing coded as literal "NA" string | `"NA"` | `na_if(., "NA")` |
| Qualtrics metadata rows | Row 1 = question text; Row 2 = import ID | `slice(-1:-2)` after loading |
| Trailing whitespace in text | `"Atlantic "` ≠ `"Atlantic"` | `str_trim()` |
| Duplicated respondent IDs | Same person submitted twice | `filter(!duplicated(respondent_id))` |

## Complete Cleaning Pipeline Template

```{r}
#| label: cleaning-pipeline
#| eval: false
library(tidyverse)
library(janitor)

# ---- 1. Load ----
# Skip = 2 skips the Qualtrics question-text and import-ID rows
raw <- read_csv("data/kaggle_survey_raw.csv", skip = 0)
# If using Qualtrics export: raw <- read_csv("file.csv", skip = 2)

# ---- 2. Clean column names ----
raw <- raw |> clean_names()
# clean_names() converts to snake_case, replaces spaces and punctuation

# ---- 3. Handle special missing values ----
raw <- raw |>
  mutate(across(where(is.character), ~ na_if(.x, ""))) |>
  mutate(across(where(is.character), ~ na_if(.x, "NA")))

# ---- 4. Coerce types ----
likert_cols <- c("music_1", "music_2", "music_3", "movies_1", "movies_2")
raw <- raw |>
  mutate(across(all_of(likert_cols), as.numeric))

# ---- 5. Reverse-code ----
# music_3 and movies_2 are reverse-worded (from codebook)
raw <- raw |>
  mutate(
    music_3_r  = 6 - music_3,
    movies_2_r = 6 - movies_2
  )

# ---- 6. Quality control flags ----
raw <- raw |>
  mutate(
    flag_speeder = survey_duration_sec < (median(survey_duration_sec, na.rm = TRUE) / 3),
    music_sd     = apply(select(., music_1, music_2, music_3, music_3_r), 1, sd, na.rm = TRUE),
    flag_sl      = music_sd < 0.5,
    flag_attn    = (attention_check != "Strongly agree")
  )

# ---- 7. Clean dataset ----
clean <- raw |>
  filter(!flag_speeder, !flag_sl, !flag_attn)

cat("Removed:", nrow(raw) - nrow(clean), "respondents\n")
cat("Retained:", nrow(clean), "(", round(100 * nrow(clean)/nrow(raw), 1), "%)\n")

# ---- 8. Composite scores ----
clean <- clean |>
  mutate(
    music_score  = rowMeans(select(., music_1, music_2, music_3_r), na.rm = TRUE),
    movies_score = rowMeans(select(., movies_1, movies_2_r), na.rm = TRUE)
  )
```

## Handling Missing Values

```{r}
#| label: missing-values
#| eval: false
# Count missing per column, sorted from most to least
raw |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  arrange(desc(n_missing)) |>
  filter(n_missing > 0)
```

The decision about how to handle missing values depends on the source:

- **Skip-logic missingness**: respondents who were not shown a question (e.g., non-subscribers skipped the subscription satisfaction items). This is **not** problem missing — it is by design. Code as NA and handle with `na.rm = TRUE` in analyses, or subset to the relevant group.
- **Item non-response**: respondents who were shown a question but did not answer. Examine the proportion by item; high rates suggest a problematic question. Handle with listwise deletion (complete-case analysis) for small amounts; consider imputation for larger amounts.
- **Speeder missingness**: respondents who dropped out partway. Typically handled by removing the entire record.

Never blanket-impute missing values without first understanding why they are missing.

---

# Ethics: Consent, Privacy, and Respondent Burden

## Informed Consent in Survey Research

Every survey — commercial or academic — requires a consent screen on the first page. The consent screen should clearly state:

- **Who is collecting the data** (organisation name; survey sponsor may be blinded but organisation should be identified)
- **Purpose of the research** — what the data will be used for
- **What data is collected** — which categories of personal information
- **Voluntary participation** — respondents may stop at any time without penalty
- **Data storage** — how data is secured, who has access, how long it is retained
- **Contact** for questions or concerns about the research

A model consent screen for a course project survey:

> *You are invited to participate in a research study conducted by [Team Name] as part of the DeGroote School of Business MBA program (BUSADMIN M731, McMaster University). This survey explores [topic]. Results will be used for academic purposes only. All responses are anonymous — no identifying information will be collected. Participation is completely voluntary; you may stop at any time. By clicking "I agree," you confirm that you are 18 years of age or older and consent to participate.*

## Sensitive Items

Items that touch on sensitive domains (income, health, political views, sexual orientation, religion, illegal behaviour) require additional care:

- Use **bracketed ranges** for income rather than exact figures: "$25K–$50K" rather than "What is your income?"
- Place sensitive items **near the end** of the survey, after rapport and engagement are established
- Always provide a **"Prefer not to answer"** option for sensitive items — forcing a response on a sensitive topic is both ethically problematic and methodologically counterproductive (respondents who do not wish to disclose will either guess or quit the survey)
- Use **self-administered mode** (online) rather than interviewer-administered mode for sensitive items — social desirability effects are lower without an interviewer present

## Respondent Burden: The Every-Question Test

Every question in a survey is a cost imposed on the respondent. The obligation is to impose only costs that are justified by the research value of the answer.

Before adding any question, ask:
1. Exactly what decision does this answer inform?
2. What analysis will use this variable?
3. Could we get a good enough answer from existing data?

If you cannot answer all three, remove the question. The "just one more thing while I have them" impulse produces surveys that are 5 minutes too long, which increases drop-out and data quality problems downstream. Under PIPEDA, the "limiting collection" principle provides a legal rationale to enforce this discipline: you may not collect personal information beyond what is needed for the stated research purpose.

---

# Qualtrics for Survey Design and Data Collection

## McMaster Qualtrics Access

McMaster has a site licence for Qualtrics. Students, staff, and faculty access it at `mcmaster.ca1.qualtrics.com`. Full features are included: survey flow with skip logic, question randomisation, response piping, multiple distribution channels (email link, anonymous link, embedded data), and CSV export.

## The Qualtrics-to-R Workflow

1. **Design**: build the survey in Qualtrics, including skip logic and randomisation
2. **Field**: distribute via email link, anonymous link, or panel API
3. **Export**: download responses as CSV with numeric value coding (not label coding) for analysis
4. **Import**: `read_csv()` in R
5. **Clean**: janitor + reverse coding + QC screens (see cleaning pipeline above)
6. **Analyse**: all subsequent sessions

### Direct API Access with the qualtRics Package

For studies requiring repeated data pulls (tracking studies, ongoing panels):

```{r}
#| label: qualtrics-api
#| eval: false
# install.packages("qualtRics")
library(qualtRics)

# One-time setup: store credentials securely
qualtrics_api_credentials(
  api_key  = "YOUR_API_KEY_HERE",
  base_url = "mcmaster.ca1.qualtrics.com",
  install  = TRUE,
  overwrite = TRUE
)

# List all surveys in your account
all_surveys()

# Pull responses from a specific survey
responses <- fetch_survey(surveyID = "SV_xxxxxxxxx", verbose = TRUE)
glimpse(responses)
```

For a typical course project, the CSV export path is simpler and fully sufficient. The API is useful for automated workflows — monthly tracking pulls, live dashboards — that are more common in industry contexts.

---

# The Questionnaire Peer-Review Checklist

Before finalising any questionnaire, apply this checklist. Strong questionnaires pass every item; most first drafts fail several.

**Construct coverage:**

- Each research question maps to at least one measured construct
- Each construct has an explicit definition in the codebook
- Each construct has at least three items (five is better for Cronbach's alpha)
- No important construct is unmeasured

**Item quality:**

- No double-barrelled items (each item measures exactly one thing)
- No leading or loaded language
- All items are answerable by the target population
- At least 20% of items in each Likert battery are reverse-worded
- Grade 6–8 reading level throughout

**Flow and length:**

- Screening questions appear first
- Warm-up questions are easy and engaging
- Sensitive items appear after engagement is established
- Demographics appear last
- Estimated completion time is 12 minutes or less
- Skip logic is complete and tested (no dead ends, no logical loops)

**Measurement:**

- Scale type is appropriate for each construct (Likert for attitudes, ratio for counts and dollar amounts)
- All Likert scale points are fully labelled
- Response option order is randomised where applicable

**Ethics:**

- Consent screen is present on the first page with all required elements
- "Prefer not to answer" option present for all sensitive items
- No personal information is collected beyond what is needed

---

# Preparing for Test 1

Test 1 (Session 5) covers Sessions 1–4. The test rewards application of concepts, not memorisation of definitions. Practice applying each concept to the Maritime Roast context:

- The seven-step MR process: what happens at each step?
- Research design families: when is each appropriate?
- Scales of measurement: what statistics are appropriate at each level?
- Data-collection modes: what are the trade-offs?
- Sampling: probability vs. non-probability; how to compute required n
- Questionnaire design: what makes a question good or bad?

---

# References

Chapman, C., & Feit, E. M. (2023). *R for marketing research and analytics* (2nd ed.). Springer.

Keiningham, T. L., Cooil, B., Andreassen, T. W., & Aksoy, L. (2007). A longitudinal examination of net promoter and firm revenue growth. *Journal of Marketing, 71*(3), 39–51.

Krosnick, J. A., & Fabrigar, L. R. (1997). Designing rating scales for effective measurement in surveys. In L. Lyberg et al. (Eds.), *Survey measurement and process quality*. Wiley.

Krosnick, J. A., & Presser, S. (2010). Question and questionnaire design. In P. V. Marsden & J. D. Wright (Eds.), *Handbook of survey research* (2nd ed., pp. 263–314). Emerald Group.

Osgood, C. E., Suci, G. J., & Tannenbaum, P. H. (1957). *The measurement of meaning*. University of Illinois Press.

Park, C. W., MacInnis, D. J., Priester, J., Eisingerich, A. B., & Iacobucci, D. (2010). Brand attachment and brand attitude strength: Conceptual and empirical differentiation of two critical brand equity drivers. *Journal of Marketing, 74*(6), 1–17.

Silver, L., Stevens, R., Wrenn, B., & Loudon, D. (2021). *The essentials of marketing research*. SAGE Publications.

Tourangeau, R., Rips, L. J., & Rasinski, K. (2000). *The psychology of survey response*. Cambridge University Press.
