---
title: "Solutions — Session 4: Cleaning a Messy Survey Export"
subtitle: "M731 · Full worked walkthrough (companion to session-04.qmd)"
author: "Shay Abbasi"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Purpose

Full worked version of the Session 4 in-lab exercise: cleaning column
names, reverse-coding, straight-line detection, and composite-score
construction on a deliberately messy survey export.

**Dataset provenance.** The original slide deck referenced Kaggle's
["Young People Survey"](https://www.kaggle.com/datasets/miroslavsabo/young-people-survey)
(Sabo, 2016). That dataset is third-party and not ours to redistribute
in a course repo, so `kaggle_survey_raw.csv` here is an
**instructor-simulated stand-in** with the same messy structure
(inconsistent column names, mixed NA encodings, a reverse-worded item
pair, and two multi-item batteries: an 8-item `Music` battery and a
6-item `Movies_` battery, plus `loneliness_1`/`loneliness_2`). If your
own teaching context has access to the real dataset, swap it in — the
cleaning code below is unaffected either way.

## Setup

```{r}
#| label: setup
library(tidyverse)
library(janitor)

raw <- read_csv("data/kaggle_survey_raw.csv")
dim(raw)
```

## Step 1 — Clean column names

```{r}
#| label: clean-names
raw <- raw |> clean_names()
names(raw)
```

**Interpretation.** `clean_names()` converts `"Age "`, `"Village -
town"`, `"Music 1"` etc. into consistent `snake_case` (`age`,
`village_town`, `music_1`...). This single step eliminates most
quoting headaches for the rest of the pipeline.

## Step 2 — Reverse-code the reversed item

`loneliness_2` was deliberately generated as the reverse-worded
partner of `loneliness_1` (see `data-raw/generate_datasets.R`).

```{r}
#| label: reverse-code
raw <- raw |>
  mutate(loneliness_2_r = 6 - loneliness_2)

cor(raw$loneliness_1, raw$loneliness_2, use = "complete.obs")
cor(raw$loneliness_1, raw$loneliness_2_r, use = "complete.obs")
```

**Interpretation.** Before reversing, `loneliness_1` and
`loneliness_2` correlate negatively (as expected for a reverse-worded
pair). After reversing `loneliness_2`, the correlation flips positive
— confirmation the reversal was applied correctly. **Always check this
correlation before and after** — it's the fastest way to catch a
reverse-coding bug.

## Step 3 — Missing-value audit

```{r}
#| label: missing-audit
raw |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "col", values_to = "n_missing") |>
  arrange(desc(n_missing)) |>
  head(10)
```

**Interpretation.** Missingness is scattered at roughly 3-6% per
column by design — realistic for a real export, not catastrophic.
Decide per column: for the music/movies batteries we'll keep `NA` and
use `na.rm = TRUE` in composite scoring below; for `age` (a screening
variable) a stricter case-wise drop might be warranted for downstream
models.

## Step 4 — Straight-line detection

```{r}
#| label: straightline
music_items <- raw |> select(starts_with("music_"))

raw$music_sd <- apply(music_items, 1, sd, na.rm = TRUE)

raw |>
  arrange(music_sd) |>
  select(music_sd, starts_with("music_")) |>
  head(10)
```

**Interpretation.** Rows with `music_sd` at or near 0 answered every
item in the battery identically — the classic straight-lining
signature. `data-raw/generate_datasets.R` deliberately planted 25 such
rows so this diagnostic has something real to catch; in your own data,
flag rows below roughly SD < 0.3 on a 5-point battery for review or
exclusion.

## Step 5 — Composite score + distribution

```{r}
#| label: fig-composite
#| fig-cap: "Distribution of the composite music-enthusiasm score"
raw <- raw |>
  mutate(music_score = rowMeans(select(raw, starts_with("music_")), na.rm = TRUE))

raw |>
  ggplot(aes(x = music_score)) +
  geom_histogram(binwidth = 0.25, fill = "#7A003C", colour = "white") +
  labs(title = "Composite music-enthusiasm score", x = "Score (1-5)", y = "Count") +
  theme_minimal()
```

**Interpretation.** The composite is roughly centred with a mild
concentration near the straight-liner value (3) driven by the planted
rows — worth excluding the flagged straight-liners before treating this
as a clean analytic variable, e.g. `filter(music_sd > 0.3)`.

## Take-away for students

Every messy-data pipeline follows the same order: **clean names →
check/reverse-code → audit missingness → detect quality problems →
build composites → visualise before you trust anything.** Skipping a
step (especially the reverse-code check) is the single most common
source of "why is my alpha negative" panic in Week 8.
