---
title: "Session 9 — Factor Analysis: Uncovering Latent Structure"
subtitle: "M731 Marketing Research · Comprehensive Lecture Notes"
author: "Shay Abbasi"
date: "2026-11-04"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    self-contained: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

# Why Reduce Complexity? The Case for Factor Analysis

## The Problem of Too Many Variables

Consider a marketing survey with a 12-item battery asking subscribers about their coffee subscription needs: ease of ordering, fast delivery, auto-renewal, premium beans, single-origin sourcing, and so on. If you want to use these items in a downstream regression to understand which needs drive satisfaction, you face a problem. Entering all 12 items creates severe multicollinearity — the items are correlated with each other, making individual coefficient estimates unstable and difficult to interpret. More fundamentally, 12 separate items are not 12 independent constructs. They reflect a smaller number of underlying motivations or need states.

Factor analysis solves this problem by identifying the latent structure underlying a set of observed variables. Instead of working with 12 correlated items, you work with 3 or 4 latent factors — each factor representing a meaningful cluster of items that share common variance. The factors are more interpretable than the individual items, less collinear with each other, and theoretically grounded in the constructs you care about.

## PCA vs. EFA: A Fundamental Choice

The terms "factor analysis" and "principal component analysis" are often used interchangeably, but they are methodologically distinct and the choice between them matters for marketing research.

**Principal Component Analysis (PCA)** decomposes the total variance in the correlation matrix. Every unit of variance — including item-specific variance and measurement error — goes into the components. PCA is purely a data-reduction tool: it produces components that account for as much total variance as possible. There is no model of what generates the data; components are mathematical constructs, not theoretical entities.

The mathematical formulation of PCA decomposes the correlation matrix R as:

$$\mathbf{R} = \mathbf{V}\mathbf{\Lambda}\mathbf{V}'$$

where V is the matrix of eigenvectors and Λ is the diagonal matrix of eigenvalues. The 1s on the diagonal of R mean PCA treats all item variance as signal.

**Exploratory Factor Analysis (EFA)** models only the common variance — the variance shared among items, which presumably reflects the latent constructs. Item-specific variance and measurement error are partitioned into a separate term (the uniqueness). The model is:

$$\mathbf{R} = \mathbf{\Lambda}\mathbf{\Phi}\mathbf{\Lambda}' + \mathbf{\Psi}$$

where Λ is the loading matrix, Φ is the factor correlation matrix, and Ψ is a diagonal matrix of uniquenesses (item-specific variance plus error). EFA replaces the 1s on the diagonal of R with communality estimates — each item's estimated proportion of shared variance.

**When to use which:**

- Use EFA when you want to uncover or test latent constructs — the theoretically motivated choice for marketing scale development and needs segmentation
- Use PCA when you purely want to compress variables for predictive modelling and make no theoretical claim about what generates the data

For the Maritime Roast needs battery, EFA is the right choice: you are trying to understand what underlying need states drive subscription behaviour, not just compress 12 variables into fewer dimensions.

## The Factor Model Explained

In EFA, each observed item is modelled as a weighted linear combination of latent factors plus a unique component:

$$X_i = \lambda_{i1}F_1 + \lambda_{i2}F_2 + \cdots + \lambda_{ik}F_k + U_i$$

where:

- $X_i$ is the standardised score on item i
- $\lambda_{ij}$ is the factor loading of item i on factor j (the relationship strength)
- $F_j$ is the score on latent factor j
- $U_i$ is the unique factor for item i (item-specific variance plus random error)

The factor loadings λ are the central outputs of factor analysis. A loading of 0.75 on Factor 1 means that item i shares substantial variance with the other items loading on Factor 1, and this shared variance is attributed to the latent construct Factor 1 represents. A loading near zero means the item has little relationship to that factor.

Each item's **communality** (h²) is the proportion of its total variance explained by all factors combined: $h_i^2 = \sum_j \lambda_{ij}^2$. Items with low communality (below 0.30) are poorly captured by the factor solution and are candidates for removal.

# Assessing Factorability

## Should You Run Factor Analysis on This Data?

Before extracting factors, confirm that the correlation structure of your items justifies factor analysis. Two tests address this:

**Bartlett's test of sphericity** tests the null hypothesis that the population correlation matrix is an identity matrix — that is, that items are completely uncorrelated and share no common variance. If the null is true, there is nothing for factor analysis to extract. You want to reject the null (p < .05), which confirms that items share enough correlation to support factor analysis.

Bartlett's test rejects easily with large samples even when correlations are trivially small. It is a necessary but not sufficient check.

**Kaiser-Meyer-Olkin (KMO) measure of sampling adequacy** is more informative. It measures the proportion of total variance among items that is common variance — the proportion attributable to underlying factors rather than unique variance. The formula compares actual pairwise correlations to partial correlations (correlations with the other items controlled). When partial correlations are small relative to pairwise correlations, items have high common variance and KMO is high.

Kaiser's labels for KMO values: above 0.90 is "marvelous," above 0.80 is "meritorious," above 0.70 is "middling," above 0.60 is "mediocre," 0.50–0.60 is "miserable," below 0.50 is "unacceptable." The minimum acceptable overall KMO is 0.60. Per-item KMO values below 0.50 identify specific items that are pulling the overall KMO down and should be considered for removal.

```{r}
#| label: factorability-checks

library(psych)
library(tidyverse)

survey  <- read_csv("data/maritime_survey.csv")
battery <- survey |> select(needs_1:needs_12)

# KMO measure — overall and per item
kmo_result <- KMO(battery)
kmo_result$MSA    # overall KMO
kmo_result$MSAi   # per-item KMO

# Bartlett's sphericity test
cortest.bartlett(cor(battery, use = "complete.obs"),
                 n = sum(complete.cases(battery)))
```

If overall KMO is below 0.60, factor analysis is not appropriate for this data. Investigate the inter-item correlation matrix to identify items that are not well-integrated with the rest of the battery.

# Determining the Number of Factors

## Why This Is the Central Decision

The number of factors to extract is the most consequential decision in exploratory factor analysis. Extract too few and you miss meaningful distinctions; extract too many and you end up with factors driven by noise rather than signal. Three complementary methods triangulate toward the right number.

## The Kaiser Criterion (Eigenvalue > 1)

The eigenvalues of the correlation matrix measure how much total variance each factor accounts for. A factor with eigenvalue greater than 1 accounts for more variance than a single item — the logic being that a factor should at least justify its own existence by explaining as much as one of the variables it came from.

The Kaiser criterion (retain factors with eigenvalue > 1) is ubiquitous because it is automatic and requires no simulation. But it systematically over-extracts, especially with many items. With a 12-item battery, random data regularly produces 3–4 eigenvalues above 1 through sampling variation alone. Kaiser may suggest 5 factors when the true structure has 3.

## The Scree Plot

The scree plot graphs eigenvalues in descending order against factor number. The "elbow" — the point where the curve bends from steep decline to a flat plateau — indicates the number of factors to retain. Factors before the elbow reflect genuine structure; factors on the flat portion reflect noise.

```{r}
#| label: scree-plot

eig_values <- eigen(cor(battery, use = "complete.obs"))$values

plot(eig_values, type = "b",
     main = "Scree plot — needs battery",
     xlab = "Factor number",
     ylab = "Eigenvalue")
abline(h = 1, lty = 2, col = "red")
```

The limitation of the scree plot is subjectivity: two competent analysts may disagree about where the elbow falls. It works best as a corroborating tool alongside parallel analysis.

## Parallel Analysis (Recommended)

Parallel analysis (Horn, 1965) is the best-supported factor-number criterion in simulation studies. It generates many random datasets with the same n and number of items as your sample, extracts eigenvalues from each, and computes the distribution of random eigenvalues. Retain only factors where your sample eigenvalue exceeds the random eigenvalue at the chosen percentile (typically the 95th, which is conservative — your factor has to beat 95% of what random data produces).

```{r}
#| label: parallel-analysis

set.seed(42)
pa_result <- fa.parallel(
  battery,
  fm     = "ml",
  fa     = "fa",
  n.iter = 200,
  quant  = 0.95   # use 95th percentile of random eigenvalues
)

# Suggested number of factors
pa_result$nfact
```

The output plot overlays your sample eigenvalues (solid line) with the simulated random eigenvalues at the chosen percentile (dashed line). Retain factors where the solid line is above the dashed line. This is the correct stopping rule.

Using the 95th percentile (Glorfeld, 1995) rather than the mean of simulated eigenvalues is more conservative and reduces over-extraction. The default in `fa.parallel()` uses the mean; specify `quant = 0.95` for the more rigorous criterion.

## The MAP (Minimum Average Partial) Test

Velicer's MAP test (1976) is an algorithmic alternative. It successively extracts components, computes the partial correlation matrix after each extraction, and tracks the average squared partial correlation. The optimal number of factors is where this average is minimised — the point at which subsequent extractions remove more construct variance than error. MAP tends to be conservative (extracts fewer factors than parallel analysis) and provides a useful second opinion when methods disagree.

```{r}
#| label: map-test

library(psych)
# vss() also runs MAP; n specifies maximum factors to test
vss(battery, n = 6)
```

## When Criteria Disagree

No single criterion is always correct. The practical approach:

1. Run parallel analysis and note its suggestion
2. Confirm or question with the scree plot and MAP
3. Extract solutions for 2–3 candidate numbers of factors
4. Choose the solution with the clearest, most interpretable factor structure
5. Document your reasoning — the decision is yours to justify, not the algorithm's

If parallel analysis suggests 4 factors but the 4-factor solution produces one factor with only two items and no clear substantive meaning, prefer the 3-factor solution and acknowledge this in the report.

## ML Fit Statistics for Comparing Solutions

When using maximum likelihood extraction, model fit statistics provide a principled basis for comparing solutions of different dimensionality:

```{r}
#| label: model-fit-comparison

# Fit 2, 3, and 4 factor solutions for comparison
fa_2 <- fa(battery, nfactors = 2, rotate = "oblimin", fm = "ml")
fa_3 <- fa(battery, nfactors = 3, rotate = "oblimin", fm = "ml")
fa_4 <- fa(battery, nfactors = 4, rotate = "oblimin", fm = "ml")

# Compare fit statistics
data.frame(
  nfactors = c(2, 3, 4),
  RMSEA    = c(fa_2$RMSEA[1], fa_3$RMSEA[1], fa_4$RMSEA[1]),
  TLI      = c(fa_2$TLI,      fa_3$TLI,      fa_4$TLI),
  BIC      = c(fa_2$BIC,      fa_3$BIC,      fa_4$BIC)
)
```

Interpret fit indices using the same benchmarks as CFA: RMSEA < 0.06 and TLI > 0.95 indicate good fit; RMSEA < 0.08 is acceptable. Lower BIC values favour the more parsimonious solution (penalises additional factors). A solution where adding one more factor produces minimal improvement in RMSEA but increases BIC has gone far enough.

# Extraction Methods

Three extraction methods are commonly available:

| Method | Assumption | When to use |
|--------|-----------|-------------|
| ML (maximum likelihood) | Normal distribution | Preferred when items are approximately normal; required for fit statistics |
| PAF (principal axis factoring) | No distributional assumption | Robust to non-normality; use when ML fails to converge |
| OLS (ordinary least squares) | Minimises residual correlations | Fallback when both ML and PAF have issues |

For marketing survey data on 5–7 point scales with n > 200, ML usually works and is preferred because it enables the fit-statistics comparison described above. PAF is the backup when ML produces convergence warnings.

# Rotation

## Why Unrotated Solutions Are Uninterpretable

The initial, unrotated factor solution is mathematically valid but typically uninterpretable. The first factor captures the largest amount of variance and usually has substantial loadings on most items — there is no clean structure where items load primarily on one factor.

Rotation transforms the loading matrix into a form where each item has a high loading on one factor and near-zero loadings on others — called **simple structure**. Critically, rotation does not change the total variance explained by the factors or the model fit; it only redistributes the variance across factors to improve interpretability. Think of the factors as axes in a multi-dimensional space: rotation spins the axes to be closer to the natural clusters of items, without moving the items themselves.

## Orthogonal vs. Oblique Rotation

**Orthogonal rotation** (varimax, quartimax) constrains rotated factors to be uncorrelated. Varimax is the most common: it finds the rotation that maximises the variance of the squared loadings within each factor, producing clean "high or near-zero" loading patterns.

**Oblique rotation** (oblimin, promax) allows factors to correlate with each other. This is more realistic for marketing constructs, which rarely are truly orthogonal. Convenience needs and quality needs likely correlate positively — respondents who care about one tend to care about the other. Fabrigar et al. (1999) recommended oblique rotation as the default for most social science applications.

The practical decision: start with oblimin rotation. Examine the factor correlation matrix (Φ): if factor correlations are all below 0.30, the orthogonal and oblique solutions will be nearly identical and you can report either. If correlations exceed 0.30–0.40, oblique rotation better represents the true structure.

```{r}
#| label: rotation-comparison

# Oblique rotation (recommended default)
fa_oblimin <- fa(battery, nfactors = 3, rotate = "oblimin", fm = "ml")

# Orthogonal rotation (for comparison)
fa_varimax <- fa(battery, nfactors = 3, rotate = "varimax", fm = "ml")

# Factor correlation matrix from oblimin — is it large?
fa_oblimin$Phi
```

## The Pattern Matrix vs. the Structure Matrix

With oblique rotation, you receive two loading matrices:

**The pattern matrix** shows the unique contribution of each factor to each item, partialling out the correlation between factors. These are analogous to partial regression coefficients. When factors correlate, a factor's total correlation with an item (the structure matrix) includes a spurious portion from the other factors.

**The structure matrix** shows the total (bivariate) correlation between each factor and each item. Because factors correlate with each other, the structure matrix entries will be larger and more uniformly elevated than the pattern matrix entries.

Interpret the pattern matrix — it shows each factor's unique effect. Most software (including `psych::fa()`) prints the pattern matrix by default, but verify which one you are looking at.

```{r}
#| label: pattern-structure-matrices

# Pattern matrix (unique loadings — interpret this)
print(fa_oblimin$loadings, cutoff = 0.30, sort = TRUE)

# Structure matrix (total correlations — for context)
fa_oblimin$Structure
```

# Interpreting the Loading Matrix

## What Simple Structure Looks Like

A well-behaved factor solution exhibits simple structure: each item has one substantially high loading (above 0.40 is a common threshold) on one factor, and near-zero loadings (below 0.30) on all others. When this pattern holds, interpretation is straightforward.

Consider this hypothetical pattern for the 12-item Maritime Roast needs battery after oblimin rotation:

| Item | Convenience (F1) | Quality (F2) | Social (F3) | Communality |
|------|-----------------|--------------|-------------|-------------|
| needs_1: "Easy to order" | **0.81** | 0.09 | 0.06 | 0.67 |
| needs_2: "Fast delivery" | **0.76** | 0.11 | 0.02 | 0.60 |
| needs_3: "Auto-renew" | **0.69** | 0.14 | 0.08 | 0.52 |
| needs_4: "Skip-a-month" | **0.62** | 0.19 | 0.11 | 0.47 |
| needs_5: "Premium beans" | 0.07 | **0.84** | 0.10 | 0.72 |
| needs_6: "Single-origin" | 0.12 | **0.79** | 0.07 | 0.65 |
| needs_7: "Roast freshness" | 0.18 | **0.71** | 0.13 | 0.57 |
| needs_8: "Curated selection" | 0.22 | **0.65** | 0.19 | 0.52 |
| needs_9: "Share with friends" | 0.05 | 0.13 | **0.78** | 0.63 |
| needs_10: "Gift subscription" | 0.10 | 0.08 | **0.74** | 0.57 |
| needs_11: "Community events" | 0.15 | 0.17 | **0.67** | 0.51 |
| needs_12: "Recommend to others" | 0.09 | 0.21 | **0.61** | 0.43 |

This is clean simple structure: each item loads primarily on one factor with near-zero cross-loadings. Communalities range from 0.43 to 0.72, indicating the factors capture 43–72% of each item's variance. The three factors are named by examining the items that load highest on each.

## Naming the Factors

Factor naming is the interpretive heart of EFA. It requires bringing theoretical knowledge to the table rather than mechanically reading off numbers.

**Factor 1 — Convenience:** Items 1–4 all concern reducing effort in the subscription relationship: easy ordering, fast delivery, auto-renewal, and the ability to skip. These items share a common theme of minimising friction and cognitive load.

**Factor 2 — Quality Seeking:** Items 5–8 concern the coffee itself: premium beans, single-origin sourcing, freshness, and curated selection. These items share a theme of intrinsic product quality and the experiential dimension of specialty coffee.

**Factor 3 — Social/Gift:** Items 9–12 concern the social dimension of coffee subscription: sharing with friends, gifting subscriptions, community participation, and recommending the service. These items share a theme of using the subscription as a vehicle for social connection and expression.

Document the naming rationale. Different reasonable analysts might call Factor 3 "Community Orientation" or "Social Expression" — both are defensible. What matters is that the name is derived from the content of the items, not chosen to match a theoretical preconception.

## Cross-Loading Items

A cross-loading item loads substantially (above 0.30) on two or more factors. Cross-loaders are problematic because they cannot be cleanly assigned to one construct and undermine the distinctiveness of the factor solution.

Consider an item worded "Fresh quality coffee delivered to my door." This item mentions freshness and quality (suggesting Factor 2) but also delivery (suggesting Factor 1). It may well load 0.48 on Convenience and 0.41 on Quality — a cross-loader. Three options exist:

1. **Drop the item** if it has low communality and its removal does not substantially reduce reliability of either factor
2. **Revise the wording** to make the item more specifically about one dimension (if revising the questionnaire is still possible)
3. **Assign to the primary factor** and report the cross-loading transparently

Cross-loadings should be reported, not hidden. They are informative — they tell you that some items resist neat categorisation, and that the factors themselves may have genuine conceptual overlap.

# Factor Scores and Their Uses

## Three Methods for Computing Factor Scores

Once you have a factor solution, you can compute per-respondent factor scores — each respondent's estimated position on each latent factor. These scores can then be used in downstream analyses (regression, clustering, segmentation).

| Method | Description | Best for |
|--------|-------------|---------|
| Regression (default) | Weighted combination using factor loadings and item intercorrelations | Most uses; scores are standardised |
| Bartlett | Maximises validity, minimises bias | Predictive modelling where factor score accuracy matters |
| Anderson-Rubin | Forces orthogonality of factor scores | Downstream regression requiring independent predictors |

```{r}
#| label: factor-scores-extraction

fa_result <- fa(battery, nfactors = 3, rotate = "oblimin",
               fm = "ml", scores = "regression")

# Scores: one row per respondent, one column per factor
scores_df <- as.data.frame(fa_result$scores)
colnames(scores_df) <- c("f_convenience", "f_quality", "f_social")

# Bind back to the survey data
survey <- bind_cols(survey, scores_df)

# Sanity check: convenience factor should correlate with needs_1
cor(survey$f_convenience, survey$needs_1, use = "complete.obs")
# Should be positive and substantial
```

## Sum Scores as an Alternative

A simpler alternative to factor scores is sum scores: the average of the items that loaded on each factor. Sum scores are less statistically efficient but more transparent, easier to explain to non-technical stakeholders, and reproducible without re-running the factor analysis.

```{r}
#| label: sum-scores

survey <- survey |>
  rowwise() |>
  mutate(
    convenience_sum = mean(c(needs_1, needs_2, needs_3, needs_4),
                           na.rm = TRUE),
    quality_sum     = mean(c(needs_5, needs_6, needs_7, needs_8),
                           na.rm = TRUE),
    social_sum      = mean(c(needs_9, needs_10, needs_11, needs_12),
                           na.rm = TRUE)
  ) |>
  ungroup()
```

Grice (2001) compared factor score methods in simulation studies and found little practical difference for most applications. For M731 project reports, either factor scores or sum scores are acceptable — choose based on which is easier to justify and communicate to your audience.

# Complete EFA Workflow

## Step 1 — Prepare and Inspect

```{r}
#| label: efa-step1

library(tidyverse)
library(psych)

survey  <- read_csv("data/maritime_survey.csv")
battery <- survey |> select(needs_1:needs_12)

# Descriptive statistics per item
describe(battery) |>
  select(n, mean, sd, skew, kurtosis, min, max) |>
  round(2)

# Missing value check
colSums(is.na(battery))
```

Factor analysis requires complete cases or imputed data — it cannot handle missing values without a strategy. With n = 800 and missing rates below 5%, listwise deletion (complete cases only) is acceptable.

## Step 2 — Check Factorability

```{r}
#| label: efa-step2

kmo_result <- KMO(battery)
cat("Overall KMO:", round(kmo_result$MSA, 3), "\n")
print(kmo_result$MSAi)   # per-item KMO

cortest.bartlett(cor(battery, use = "complete.obs"),
                 n = sum(complete.cases(battery)))
```

## Step 3 — Determine Number of Factors

```{r}
#| label: efa-step3

set.seed(42)
pa_result <- fa.parallel(battery, fm = "ml", fa = "fa",
                         n.iter = 200, quant = 0.95)
cat("Parallel analysis suggests:", pa_result$nfact, "factors\n")

# Scree for visual confirmation
eig_values <- eigen(cor(battery, use = "complete.obs"))$values
plot(eig_values, type = "b",
     xlab = "Factor", ylab = "Eigenvalue",
     main = "Scree plot — needs battery")
abline(h = 1, lty = 2, col = "red")

# MAP test as second opinion
vss(battery, n = 6)
```

## Step 4 — Extract and Rotate

```{r}
#| label: efa-step4

n_factors <- 3   # based on Step 3 decision

fa_result <- fa(
  r        = battery,
  nfactors = n_factors,
  rotate   = "oblimin",
  fm       = "ml",
  use      = "complete.obs",
  scores   = "regression"
)

# Pattern matrix — suppress loadings below 0.30 for clarity
print(fa_result$loadings, cutoff = 0.30, sort = TRUE)

# Factor correlations
fa_result$Phi

# Fit statistics
cat("RMSEA:", fa_result$RMSEA[1], "\n")
cat("TLI:  ", fa_result$TLI,      "\n")
cat("BIC:  ", fa_result$BIC,      "\n")
```

## Step 5 — Build the Loading Table

```{r}
#| label: efa-step5

library(gt)

# Extract loadings into a clean data frame
load_mat <- fa_result$loadings
load_df <- as.data.frame(unclass(load_mat)) |>
  tibble::rownames_to_column("Item") |>
  rename(Convenience = MR1, Quality = MR2, Social = MR3) |>
  mutate(
    Communality = fa_result$communalities,
    across(where(is.numeric), ~ round(., 2))
  )

# Suppress small loadings for readability
load_display <- load_df |>
  mutate(across(c(Convenience, Quality, Social),
                ~ if_else(abs(.) < 0.30, NA_real_, .)))

load_display |>
  gt() |>
  tab_header(
    title    = "Factor loadings — Maritime Roast needs battery",
    subtitle = "Oblimin rotation; ML extraction; loadings < .30 suppressed"
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 2) |>
  sub_missing(missing_text = "—")
```

## Step 6 — Attach Factor Scores and Validate

```{r}
#| label: efa-step6

# Attach factor scores
scores_df <- as.data.frame(fa_result$scores)
colnames(scores_df) <- c("f_convenience", "f_quality", "f_social")
survey <- bind_cols(survey, scores_df)

# Reliability per factor (items should hang together)
alpha(select(survey, needs_1:needs_4))$total$raw_alpha   # convenience
alpha(select(survey, needs_5:needs_8))$total$raw_alpha   # quality
alpha(select(survey, needs_9:needs_12))$total$raw_alpha  # social

# Correlation sanity check on factor scores
cor(scores_df, use = "complete.obs") |> round(2)
```

## Reporting the EFA Results

A complete EFA write-up contains six elements: sample and factorability statistics, extraction and rotation method, number-of-factors criterion and rationale, factor names and item assignments, reliability per factor, and factor intercorrelations.

Template: "Exploratory factor analysis on the 12-item needs battery (n = 800, KMO = 0.85, Bartlett's χ²(66) = 4,823, p < .001) supported a three-factor solution based on parallel analysis (200 iterations, 95th percentile criterion). Factors were extracted using maximum likelihood with oblimin rotation (RMSEA = 0.048, TLI = 0.96). After rotation, factors were labelled Convenience (items 1–4, α = 0.83), Quality Seeking (items 5–8, α = 0.79), and Social/Gift (items 9–12, α = 0.75), accounting for 62% of total variance. Factor intercorrelations ranged from r = 0.18 to r = 0.35."

# From EFA to CFA: The Two-Study Research Cycle

## The Gold Standard Design

Best practice for scale validation distinguishes the exploratory from the confirmatory phase:

**Study 1 (EFA):** Explore factor structure on an initial sample, drop problematic items, identify a candidate factor solution. This is the discovery phase.

**Study 2 (CFA):** Test the factor structure from Study 1 on a new, independent sample using confirmatory factor analysis. CFA tests whether the pre-specified factor structure fits the new data. This is the confirmation phase.

Using the same data for EFA and CFA capitalises on chance: the EFA solution is already optimised for that sample, so CFA on the same data will fit better than it deserves. Commercial marketing research rarely follows this two-study cycle because it doubles fieldwork costs. For academic scale development published in top journals, it is expected. For M731 projects, EFA on the full sample is acceptable — acknowledge the limitation.

## A CFA Preview with lavaan

The `lavaan` package implements structural equation modelling, of which CFA is a special case. You specify the model based on the EFA results, then test its fit on new data:

```{r}
#| label: cfa-preview

library(lavaan)

# Model specification based on EFA results
cfa_model <- "
  convenience =~ needs_1 + needs_2 + needs_3 + needs_4
  quality     =~ needs_5 + needs_6 + needs_7 + needs_8
  social      =~ needs_9 + needs_10 + needs_11 + needs_12
"

# Fit the CFA (using a new holdout sample if available)
fit_cfa <- cfa(cfa_model, data = survey, estimator = "ML")

# Model fit summary
summary(fit_cfa, fit.measures = TRUE, standardized = TRUE)
fitMeasures(fit_cfa, c("cfi", "tli", "rmsea", "srmr", "bic"))
```

Good CFA fit: CFI and TLI above 0.95, RMSEA below 0.06, SRMR below 0.08. Poor fit signals that either the factor structure needs revision or that the items do not replicate well in the new sample.

# Ethical Considerations in Factor Analysis

## The Interpretation Problem

Factor analysis involves substantial researcher judgment at every stage: choosing the number of factors, selecting the rotation, dropping items, and naming factors. Each of these decisions can be influenced by what the researcher hopes to find, creating a form of analytic flexibility that is analogous to — and as problematic as — the p-hacking and HARKing discussed in Sessions 6 and 7.

**Selective extraction:** trying solutions with 3, 4, and 5 factors, then reporting the one whose factors were easiest to name, is equivalent to trying multiple model specifications and reporting only the significant one. This practice inflates the probability of finding a spurious interpretable structure.

**Cherry-picking rotation:** trying varimax and oblimin, then reporting whichever produced cleaner loadings, adds another layer of multiplicity.

**Suppressing cross-loadings:** omitting cross-loading items from the loading table because they "don't make sense" misrepresents the factor solution.

The remedy is transparency: report your complete decision process, including which factor numbers you considered and why you chose the reported solution. Report all cross-loadings, even the inconvenient ones. If you tried both orthogonal and oblique rotation, report that and explain the choice.

# References and Further Reading

**Fabrigar, L.R., Wegener, D.T., MacCallum, R.C., & Strahan, E.J. (1999).** Evaluating the use of exploratory factor analysis in psychological research. *Psychological Methods*, 4(3), 272–299. The definitive methodological review; argues for oblique rotation and ML extraction.

**Horn, J.L. (1965).** A rationale and test for the number of factors in factor analysis. *Psychometrika*, 30(2), 179–185. The original parallel analysis paper.

**Glorfeld, L.W. (1995).** An improvement on Horn's parallel analysis methodology for selecting the correct number of factors to retain. *Educational and Psychological Measurement*, 55(3), 377–393. Recommends the 95th percentile criterion over the mean.

**Velicer, W.F. (1976).** Determining the number of components from the matrix of partial correlations. *Psychometrika*, 41(3), 321–327. The MAP test.

**Grice, J.W. (2001).** Computing and evaluating factor scores. *Psychological Methods*, 6(4), 430–450. Comparison of factor score methods.

**Zwick, W.R. & Velicer, W.F. (1986).** Comparison of five rules for determining the number of components to retain. *Psychological Bulletin*, 99(3), 432–442. Classic comparison showing Kaiser's rule over-extracts.
