---
title: "Session 9 — Factor Analysis"
subtitle: "M731 Week 9 · November 4, 2026"
author: "Shay Abbasi"
date: "2026-11-04"
---

## Session 9 {background-color="#7A003C"}

Factor analysis: reducing complexity

**Project proposal + questionnaire due at start of class today.**

::: {.notes}
Collect proposals first. Confirm A2L submissions. Then start content.
Factor analysis is the trickiest concept in this course - budget extra
time for the concept block, less for the R lab.
:::

## Today's agenda

1. Project proposals collected
2. Why reduce complexity? PCA vs. EFA
3. Assessing factorability (KMO, Bartlett)
4. Choosing the number of factors
5. Rotation (varimax, oblimin)
6. Loading interpretation and factor scores
7. **Break (15 min)**
8. R lab: EFA on `maritime_survey.csv`
9. Wrap-up

::: {.notes}
Ethics thread today is light - the deep ethics of factor
interpretation (naming factors, cherry-picking structures) was covered
under Wk 6-7 p-hacking. Mention only briefly.
:::

## Learning objectives

By end of session 9 you will be able to:

- Distinguish PCA from EFA and know when each is appropriate
- Test factorability (KMO, Bartlett's sphericity)
- Choose the number of factors using scree, parallel analysis, and eigenvalues
- Rotate factors and interpret loadings
- Extract factor scores for downstream use

::: {.notes}
Objective 4 is the hardest. Rotation makes results interpretable but
can seem arbitrary. Walk through why we rotate.
:::

# Part 1 — Why reduce complexity? {background-color="#7A003C"}

## The problem

Your survey has 30 items. You want to:

- Understand the underlying structure
- Reduce 30 variables to a manageable few
- Use factor scores in downstream analysis (regression, segmentation)

Factor analysis does all three.

::: {.notes}
Frame the payoff first. Motivation: 30-item satisfaction batteries
are common. Are there really 30 distinct constructs? Usually 3-6
latent factors span the items.
:::

## PCA vs. EFA — the practical difference

- **PCA (Principal Component Analysis):** describes total variance; components are exact combinations of items
- **EFA (Exploratory Factor Analysis):** models common (shared) variance; factors are latent constructs

PCA is data-reduction. EFA is theory-building.

::: {.notes}
Distinction matters. PCA doesn't distinguish measurement error from
signal; EFA does. For MR construct-measurement work, EFA is
theoretically preferred. For pure data-reduction (compressing
predictors for a regression), PCA is fine.
:::

## When to use which

- **Use PCA when:** you just want fewer variables; no theoretical structure claim
- **Use EFA when:** you want to uncover / test latent constructs

For your Maritime Roast needs battery: EFA (we want to identify need-based segments).

For pure data-reduction of demographic variables: PCA.

::: {.notes}
Most MR work uses EFA. PCA is more common in machine-learning
pipelines (feature engineering). Same math family, different framing
and interpretation.
:::

## The factor model

Observed item = shared factors + unique factor

xᵢ = λᵢ₁F₁ + λᵢ₂F₂ + ... + Uᵢ

- λ = loading of item i on factor j
- F = factor scores (latent)
- U = unique variance (item-specific + error)

::: {.notes}
Don't dwell on math. The key intuition: each item has some shared
signal (factors) and some unique noise. Factor analysis estimates the
shared signal.
:::

# Part 2 — Factorability {background-color="#7A003C"}

## Is factor analysis appropriate for this data?

Two tests:

- **Bartlett's test of sphericity** — H₀: correlation matrix is identity (no factors)
  - Want p < .05 (reject H₀)
- **KMO (Kaiser-Meyer-Olkin) measure** — how much variance is shared vs unique
  - Overall KMO > 0.6 recommended
  - Item KMO < 0.5 → consider dropping the item

::: {.notes}
Both should be checked before running FA. Bartlett rejects easily
with large n - not very informative. KMO is more useful. If KMO < 0.5,
FA is not appropriate.
:::

## KMO in R

```{r}
#| eval: false
library(psych)

battery <- survey |> select(needs_1:needs_12)

KMO(battery)
cortest.bartlett(cor(battery), n = nrow(battery))
```

::: {.notes}
Live-code. Interpret output. Overall KMO of .85 = "meritorious." Below
.6 = don't factor. Show item-level KMOs - flag any below .5 for
possible removal.
:::

# Part 3 — How many factors? {background-color="#7A003C"}

## Three common criteria

1. **Kaiser criterion:** retain factors with eigenvalue > 1
2. **Scree plot:** look for the "elbow"
3. **Parallel analysis:** compare to random-data eigenvalues (recommended)

Parallel analysis is the best-supported by simulation studies.

::: {.notes}
Kaiser is convenient but over-extracts. Scree is subjective. Parallel
analysis (Horn 1965) compares your eigenvalues to those from random
data with the same n and k. Retain factors that exceed random.
:::

## Parallel analysis in R

```{r}
#| eval: false
fa.parallel(battery, fm = "ml", fa = "fa")
```

Output plot: your eigenvalues vs. simulated. Retain factors above the simulated line.

::: {.notes}
Live-run. Interpret plot. Typically 3-6 factors for a 12-item MR
battery. Discuss what the simulation is doing conceptually.
:::

## Scree plot manually

```{r}
#| eval: false
eig <- eigen(cor(battery))$values
plot(eig, type = "b", main = "Scree plot",
     xlab = "Factor", ylab = "Eigenvalue")
abline(h = 1, lty = 2, col = "red")
```

Look for the elbow where eigenvalues level off.

::: {.notes}
Scree is subjective. Reasonable people disagree about where the elbow
is. Combine with parallel analysis for a better decision.
:::

## What to do when criteria disagree

- Extract candidate solutions (3, 4, 5 factors)
- Look at which one gives interpretable structure
- Prioritise interpretability + theoretical fit

Factor analysis is exploratory — judgement matters.

::: {.notes}
This is where FA becomes an art. If parallel analysis says 4 factors
but the 4-factor solution has a hard-to-name factor and the 3-factor
is clean, pick 3. Document your reasoning.
:::

# Part 4 — Rotation {background-color="#7A003C"}

## Why rotate?

- Initial (unrotated) factor solution is mathematically valid but often uninterpretable
- Rotation transforms the solution into one where each item loads strongly on one factor
- **Rotation doesn't change fit** — it changes interpretability

::: {.notes}
Rotation feels arbitrary to students. Emphasise: it doesn't cheat -
the total variance explained doesn't change. Rotation just realigns
axes for readability.
:::

## Orthogonal vs. oblique

- **Orthogonal** (varimax, quartimax): factors uncorrelated
- **Oblique** (oblimin, promax): factors can correlate

Real-world constructs usually correlate → oblique is more realistic.

Try both — pick the one that produces clearer loadings.

::: {.notes}
Varimax is historically dominant because it was easiest to compute
in the 1960s. Modern software makes oblique easy. Fabrigar et al.
1999 recommends oblique as the default.
:::

## Interpreting a loading matrix

|         | Factor 1 | Factor 2 | Factor 3 |
|---------|----------|----------|----------|
| item_1  | **.72**  | .12      | .08      |
| item_2  | **.68**  | .05      | .11      |
| item_3  | .09      | **.65**  | .21      |
| item_4  | .14      | **.71**  | .09      |
| item_5  | .06      | .18      | **.63**  |

Bold = "primary loading" (usually > .4).

::: {.notes}
Show a sample loading matrix. A clean structure has each item
loading > .4 on one factor and < .3 on others (called "simple
structure"). Messy loadings mean items are cross-loading - candidates
for revision.
:::

## Naming the factors

- Look at items with the highest loadings on each factor
- Ask: what do these items have in common?
- Give the factor a short, evocative label

Naming is interpretation, not computation. Different researchers may name the same factors differently.

::: {.notes}
Naming is where researcher bias enters. "Convenience," "Habit," and
"Automaticity" might all describe the same factor - the choice affects
downstream framing. Document your reasoning; be honest about
alternatives.
:::

# Part 5 — Factor scores {background-color="#7A003C"}

## Extracting factor scores

Once you have a factor solution, compute per-respondent factor scores:

```{r}
#| eval: false
fa_result <- fa(battery, nfactors = 3, rotate = "oblimin", fm = "ml")

# Regression method scores
scores <- fa_result$scores
head(scores)

# Or add back to the survey data
survey$conv <- scores[, 1]
survey$qual <- scores[, 2]
survey$soc  <- scores[, 3]
```

::: {.notes}
Factor scores let you use factors in downstream models. Regression
scores are standard. Note that different methods (regression,
Bartlett) give slightly different values - convention is regression
method.
:::

## Payoff: use in regression (Wk 11)

Instead of 12 correlated items:

```{r}
#| eval: false
lm(satisfaction ~ needs_1 + needs_2 + ... + needs_12, data = survey)
```

You use 3 orthogonal-ish factors:

```{r}
#| eval: false
lm(satisfaction ~ conv + qual + soc, data = survey)
```

Cleaner, less collinearity, more interpretable.

::: {.notes}
Preview the Wk 11 payoff. This is why students should invest in
factor analysis now - it makes Wk 11 regression much cleaner. Foreshadow.
:::

## Break — 15 minutes {background-color="#FDBF57"}

::: {.notes}
Have needs_1..12 columns visible in maritime_survey.csv. Prep the
lab script.
:::

# Part 6 — R lab: EFA on Maritime Roast {background-color="#7A003C"}

## Lab task

Using the needs battery in `maritime_survey.csv`:

1. Check factorability (KMO, Bartlett)
2. Determine number of factors (scree, parallel analysis)
3. Extract with oblimin rotation
4. Interpret and name the factors
5. Produce a publication-style loading table
6. Extract factor scores for downstream use

A fuller worked version of this EFA — including the loading table and
a template write-up — is in **`solutions-09.qmd`** in this session's
folder.

::: {.notes}
Give 45 min. This is a long lab. Provide the script skeleton so
students focus on interpretation, not typing. Circulate for name-
the-factor conversations - these are the pedagogical gold.
:::

## Loading table with `gt`

```{r}
#| eval: false
library(gt)

# Extract loadings as tibble
loadings_df <- as_tibble(unclass(fa_result$loadings)) |>
  mutate(item = rownames(fa_result$loadings)) |>
  relocate(item) |>
  mutate(across(where(is.numeric), ~ round(., 2)))

loadings_df |>
  gt() |>
  tab_header(title = "Factor loadings (oblimin rotation)")
```

::: {.notes}
gt is Posit's modern table package. Beautiful output. Alternative:
kableExtra. For simple loading tables, both work. Encourage
publication-quality output for the project report.
:::

## Reporting factor analysis

"Exploratory factor analysis on the 12-item needs battery (n=800, KMO=.85, Bartlett's χ² p<.001) supported a three-factor solution (parallel analysis; total variance explained = 62%). After oblimin rotation, factors were labelled **Convenience** (items 1-4, α=.83), **Quality** (items 5-8, α=.79), and **Social** (items 9-12, α=.75). Factor correlations ranged from .18 to .35."

::: {.notes}
Template sentence. Every EFA in every report should read approximately
like this. Six elements: sample + factorability, number of factors +
extraction, rotation + naming + reliability, factor correlations.
:::

# Part 7 — Ethics thread {background-color="#7A003C"}

## Interpretive ethics in factor analysis

- **Naming factors** is inherently interpretive — different names change framing
- **Cherry-picking factor structures** (trying multiple extractions until you get one you like) is a form of p-hacking
- **Cross-loading items** should be flagged, not hidden

Report your process, not just your final solution.

::: {.notes}
Brief - Wk 6-7 covered p-hacking in general. Factor analysis is
especially prone because there's so much researcher judgement (number
of factors, rotation, item removal). Transparency = credibility.
:::

# Part 8 — For your project {background-color="#7A003C"}

## If your project questionnaire has multi-item batteries

- Run EFA before regression (Wk 11)
- Report the factor structure in your Wk 13 report
- Use factor scores, not individual items, in downstream analyses

If your questionnaire has no multi-item batteries, EFA isn't relevant to your project.

::: {.notes}
Not every project uses factor analysis - that's fine. But for projects
with survey data and multi-item constructs, factor analysis is the
right move.
:::

## Week 10 preview

- **Concept:** Experimental research (lab, field, quasi-experiments)
- **R lab:** Analyse `maritime_abtest.csv` (A/B test on subscription upgrade offer)
- Prep reading: CJ Ch. 7

::: {.notes}
Wk 10 marks the shift to causal research. Very different frame from
Wks 6-9 (association / description). Foreshadow: "next week we ask
whether X *causes* Y, not just whether they co-vary."
:::

## Thank you — questions?

`abbass44@mcmaster.ca`

::: {.notes}
Stay for questions on the project proposal and EFA interpretation.
:::
