---
title: "Solutions — Session 9: Exploratory Factor Analysis on the Needs Battery"
subtitle: "M731 · Full worked walkthrough (companion to session-09.qmd)"
author: "Shay Abbasi"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Purpose

Full worked version of the Session 9 in-lab exercise: factorability
checks, choosing the number of factors, oblimin rotation, and naming
+ scoring the resulting factors.

## Setup

```{r}
#| label: setup
library(tidyverse)
library(psych)
library(gt)

survey <- read_csv("data/maritime_survey.csv")
battery <- survey |> select(needs_1:needs_12)
```

## Step 1 — Factorability

```{r}
#| label: factorability
KMO(battery)
cortest.bartlett(cor(battery), n = nrow(battery))
```

**Interpretation.** Overall KMO typically lands around 0.80 —
"meritorious," well above the 0.6 minimum. Bartlett's test rejects the
identity-matrix null (p < .001), as expected with n=800. Together:
factor analysis is appropriate here.

## Step 2 — How many factors?

```{r}
#| label: fig-parallel
#| fig-cap: "Parallel analysis: needs battery"
fa.parallel(battery, fm = "ml", fa = "fa")
```

**Interpretation.** Parallel analysis should point to 3 factors,
matching the battery's built-in design (4 items each on Convenience,
Quality, and Social factors — see
`Sources/slides/data-raw/generate_datasets.R`).

## Step 3 — Extract with oblimin rotation

```{r}
#| label: fa-extract
fa_result <- fa(battery, nfactors = 3, rotate = "oblimin", fm = "ml")
print(fa_result$loadings, cutoff = 0.3)
```

**Interpretation.** You should see a clean simple structure: items
1-4 load together, items 5-8 load together, items 9-12 load together,
each on a different factor with small cross-loadings — exactly the
"one item, one primary factor" pattern that makes naming straightforward.

## Step 4 — Publication-style loading table

```{r}
#| label: tbl-loadings
loadings_df <- as_tibble(unclass(fa_result$loadings)) |>
  mutate(item = rownames(fa_result$loadings)) |>
  relocate(item) |>
  mutate(across(where(is.numeric), ~ round(., 2)))

loadings_df |>
  gt() |>
  tab_header(title = "Factor loadings (oblimin rotation)") |>
  cols_label(item = "Item")
```

## Step 5 — Naming and reliability by factor

Based on which items load together (items 1-4, 5-8, 9-12 by
construction):

```{r}
#| label: subscale-alpha
cat("Convenience (needs_1-4):\n")
alpha(battery |> select(needs_1:needs_4))$total$std.alpha

cat("Quality (needs_5-8):\n")
alpha(battery |> select(needs_5:needs_8))$total$std.alpha

cat("Social (needs_9-12):\n")
alpha(battery |> select(needs_9:needs_12))$total$std.alpha
```

**Interpretation.** Each 4-item subscale should show reasonable
internal consistency (roughly 0.6-0.75 for a 4-item scale) — lower
than the 5-item brand-attachment battery in Week 8 simply because
fewer items are averaged, not because the items are worse.

## Step 6 — Extract factor scores for downstream use

```{r}
#| label: factor-scores
survey$conv <- fa_result$scores[, 1]
survey$qual <- fa_result$scores[, 2]
survey$soc  <- fa_result$scores[, 3]

survey |> select(respondent_id, conv, qual, soc) |> head()
```

## Reporting sentence

> Exploratory factor analysis on the 12-item needs battery (n=800,
> KMO ≈ 0.80, Bartlett's χ² p < .001) supported a three-factor
> solution (parallel analysis; ML extraction). After oblimin rotation,
> factors were labelled **Convenience** (items 1-4), **Quality** (items
> 5-8), and **Social** (items 9-12), each showing a clean simple
> structure with minimal cross-loading.

## Take-away for students

Factor analysis reduces 12 correlated items to 3 interpretable scores
you can carry into Week 11's regression — much cleaner than throwing
12 raw items into `lm()` and fighting multicollinearity.
