---
title: "Session 11 — Regression: Identifying Drivers of Outcomes"
subtitle: "M731 Week 11 · November 18, 2026"
author: "Shay Abbasi"
date: "2026-11-18"
---

## Session 11 {background-color="#7A003C"}

Regression: identifying drivers of marketing outcomes

**Test 2 today (covers Lectures 5-10)**

::: {.notes}
Compressed session. Test 2 first, then a fast walkthrough of
regression content. Students should have done the CF Ch. 4, 7, 9
reading before class - the concept content is compressed.
:::

## Today's shape (differs from usual)

1. Test 2 administration (~75 min)
2. **Break (15 min)** — test collected
3. Regression concept + R walkthrough (~75 min)
4. Wrap-up (~15 min)

::: {.notes}
Signal the shape at the start. This session leans on pre-reading -
if students haven't read CF Ch. 4, 7, 9, they'll be lost during the
compressed content block.
:::

# Part 1 — Test 2 {background-color="#7A003C"}

## Test 2 logistics

- Individual, closed-book (McMaster standard calculator permitted)
- Covers Lectures 5-10: qualitative, hypothesis testing, ANOVA, scales, factor analysis, experiments
- Multiple choice and/or short-answer
- ~75 minutes

Any last logistical questions?

::: {.notes}
Take 2-3 min of logistics Q&A. Distribute tests. Timer on projector.
Invigilate actively during the test.
:::

## Break (15 min) — after test {background-color="#FDBF57"}

Collect papers. Prepare for regression walkthrough.

::: {.notes}
Collect and count. During the break, load `maritime_survey.csv` and
prep the regression demo. Have `broom`, `car`, `sandwich`,
`lm.beta` loaded.
:::

# Part 2 — Regression (compressed) {background-color="#7A003C"}

## Learning objectives (post-test)

By end of the regression segment you will be able to:

- Specify, estimate, and interpret a linear regression
- Diagnose OLS assumption violations
- Use standardised coefficients for driver-importance
- Introduce logistic regression for binary outcomes

Pre-reading (CF Ch. 4, 7, 9) is assumed.

::: {.notes}
Fast pace. Skip mechanics that CF covers well. Focus on
interpretation and diagnostics - the hard part.
:::

## Regression: what it does

Given predictors X₁, X₂, ..., Xₖ and outcome Y:

Y = β₀ + β₁X₁ + β₂X₂ + ... + βₖXₖ + ε

- βⱼ is the change in Y for a 1-unit change in Xⱼ, **holding other predictors constant**
- Fits the model that minimises squared residuals (OLS)

::: {.notes}
The "holding other predictors constant" phrase is the whole ballgame.
Regression coefficients are conditional on the other Xs. Marginal
interpretation is different.
:::

## Fitting a linear model

```{r}
#| eval: false
library(tidyverse)
library(broom)

survey <- read_csv("data/maritime_survey.csv")

fit <- lm(satisfaction ~ tenure_months + tier + region + age,
          data = survey)

summary(fit)
tidy(fit, conf.int = TRUE)
glance(fit)
```

::: {.notes}
Live-code. broom::tidy gives tidy coefficient output; broom::glance
gives model-level stats (R², F, sigma). Cleaner than summary().
:::

## Interpreting coefficients

- **Continuous X:** β = change in Y for 1-unit change in X
- **Binary / dummy X:** β = difference in Y between the two levels
- **Categorical X (3+ levels):** k-1 dummies, each vs. reference level

Always report β with a CI. Sign matters; magnitude matters more.

::: {.notes}
"For each additional month of tenure, satisfaction increases by 0.02
points, holding tier, region, and age constant." That's the sentence.
Not "tenure predicts satisfaction (p<.05)" - that's the p-value trap.
:::

## R² and adjusted R²

- **R²** = proportion of variance in Y explained by the model
- **Adjusted R²** = penalises adding predictors

More predictors always raises R²; adjusted R² can fall if the added predictor is weak.

::: {.notes}
Neither is a measure of causal effect. R²=0.05 can still be an
important effect if the outcome variance is huge. Don't fetishise
R² - focus on effect estimates.
:::

# Part 3 — Diagnostics {background-color="#7A003C"}

## OLS assumptions

1. **Linearity** in the parameters
2. **Independence** of errors
3. **Homoscedasticity** (constant error variance)
4. **Normality** of errors (for inference, not estimation)
5. **No perfect multicollinearity**

Robust to mild violations. Wild violations = biased estimates.

::: {.notes}
Independence is often violated in MR (repeated measures, clustered
sampling) - use mixed models. Heteroscedasticity is common - use
robust SEs. Normality matters less with big samples (CLT).
:::

## Diagnostic plots

```{r}
#| eval: false
par(mfrow = c(2, 2))
plot(fit)
```

Same four plots as ANOVA:

- Residuals vs fitted (linearity, heteroscedasticity)
- Q-Q plot (normality)
- Scale-location (heteroscedasticity)
- Residuals vs leverage (influential points)

::: {.notes}
Live-run. Interpret each. Well-behaved residuals hug the horizontal
in the first plot, the diagonal in Q-Q, and don't have any high-
leverage/high-residual points in the fourth.
:::

## Multicollinearity

Predictors that are highly correlated make coefficients unstable.

```{r}
#| eval: false
library(car)
vif(fit)
```

- VIF < 5: usually fine
- VIF 5-10: borderline
- VIF > 10: multicollinearity problem

::: {.notes}
VIF = variance inflation factor. If two predictors are near-copies of
each other (e.g., age and tenure_months for young subscribers),
their individual effects can't be untangled. Drop one, combine them,
or use factor scores.
:::

## Heteroscedasticity

Error variance changes across X.

```{r}
#| eval: false
library(lmtest)
bptest(fit)   # Breusch-Pagan test
```

If significant → use robust standard errors.

::: {.notes}
Heteroscedasticity biases standard errors (not coefficients). Fix
with robust SEs; coefficients don't change.
:::

## Robust standard errors

```{r}
#| eval: false
library(sandwich)
coeftest(fit, vcov = vcovHC(fit, type = "HC3"))
```

HC3 is a good default. Coefficient estimates unchanged; SEs adjusted for heteroscedasticity.

::: {.notes}
Robust SEs are standard practice in econometrics; less so in
psychology / marketing but increasingly. If BP test rejects, use
robust SEs. Same coefficients, different SEs and p-values.
:::

# Part 4 — Standardised coefficients {background-color="#7A003C"}

## Comparing predictor importance

Raw coefficients are on the scale of X — not directly comparable across predictors on different scales.

**Standardised coefficients** (β̂ᵢ × SD(Xᵢ) / SD(Y)) put predictors on a common scale.

```{r}
#| eval: false
library(lm.beta)
lm.beta(fit)
```

::: {.notes}
Standardised = "effect on Y in SD units per SD unit of X." Lets you
say "tier is the most important driver." Some methodologists
disagree - standardisation depends on the SD of each predictor,
which depends on your sample.
:::

## Alternative: relative-weights analysis

For strong collinearity, standardised coefficients can mislead. Relative-weights analysis (or dominance analysis) does better.

```{r}
#| eval: false
# library(dominanceanalysis)  # if available
# dominanceAnalysis(fit)
```

Beyond the scope of this session — flag for interested students.

::: {.notes}
Optional. Package availability varies. Point interested students to
Johnson & LeBreton 2004 for background on relative importance methods.
:::

# Part 5 — Interactions {background-color="#7A003C"}

## Modelling interactions

Interaction = the effect of one predictor depends on another.

```{r}
#| eval: false
fit_int <- lm(satisfaction ~ tenure_months * tier + region + age,
              data = survey)
tidy(fit_int, conf.int = TRUE)
```

`*` includes main effects + interaction.

::: {.notes}
"Does tenure matter more for Discovery subscribers than Standard?" -
that's the interaction question. Interpret carefully - main effects
change meaning when interactions are present.
:::

## Interpreting interactions

- Main effect of tenure = effect at the *reference* level of tier
- Interaction coefficient = additional effect at the non-reference level

Use emmeans or interaction plots to visualise:

```{r}
#| eval: false
library(emmeans)
emm <- emmeans(fit_int, ~ tenure_months | tier,
               at = list(tenure_months = c(3, 12, 24)))
plot(emm)
```

::: {.notes}
Emmeans handles the interpretive complexity. Interaction plots
often show the story more clearly than tables.
:::

# Part 6 — Logistic regression {background-color="#7A003C"}

## For binary outcomes

- Linear regression is wrong for binary Y (predictions can be < 0 or > 1)
- Logistic regression models log-odds

```{r}
#| eval: false
fit_log <- glm(upgraded ~ tenure_months + tier + age,
               family = binomial, data = ab_data)
tidy(fit_log, exponentiate = TRUE, conf.int = TRUE)
```

`exponentiate = TRUE` gives odds ratios instead of log-odds.

::: {.notes}
Odds ratios are more interpretable than log-odds. OR = 1.5 means
"1.5× higher odds." Not "1.5× higher probability" - a common student
error.
:::

## Interpreting odds ratios

- OR = 1: no effect
- OR > 1: positive effect (increases odds)
- OR < 1: negative effect (decreases odds)
- OR = 2 doubles odds; OR = 0.5 halves odds

Log-odds are on a symmetric scale; odds ratios are not.

::: {.notes}
Odds and probabilities aren't the same. OR=2 doesn't mean "twice as
likely" - it means "twice the odds." If base rate is 20%, doubling
odds gives 33%, not 40%. Non-linear conversion.
:::

# Part 7 — R lab: Maritime Roast regression {background-color="#7A003C"}

## Compressed lab task

Using `maritime_survey.csv`:

1. Regress satisfaction on tenure + tier + region + age + Wk-9 factor scores
2. Check assumptions (residual plots, VIF)
3. Standardise coefficients
4. Identify top 3 drivers of satisfaction
5. Write a 100-word interpretation for a general manager including both statistical and practical significance

A fuller worked version of this regression (plus the logistic
regression demo) is in **`solutions-11.qmd`** in this session's folder.

::: {.notes}
Give 30-40 min. Emphasise interpretation - most of the coefficients
will be significant with n=800. The "which is most important" and
"what should the CEO do" questions are the value-add.
:::

## Ethics thread (5 min)

Causal language in regression:

- **Say:** "tenure is associated with satisfaction"
- **Don't say:** "tenure causes satisfaction"
- Regression on observational data can't identify causation
- If you saw a client report using causal language on observational regression, push back

::: {.notes}
This is the ethical maturity of a graduate MR analyst. Coefficient
size tells you association strength. Causation requires either a
randomised experiment (Wk 10) or a strong quasi-experimental
identification strategy (beyond this course).
:::

# Part 8 — For your project {background-color="#7A003C"}

## Regression is often the project centrepiece

- Descriptive: "who are our subscribers?" (Wk 6)
- Comparative: "how do subscribers differ?" (Wk 6-7)
- Reduced: "what latent needs drive them?" (Wk 9 factor scores)
- Modelled: "which drivers matter most?" (Wk 11 regression) ← project report often lives here

::: {.notes}
Frame regression as the analytical crescendo. Descriptives + tests
give you clues; regression puts drivers in one framework.
:::

## Week 12 preview

- **Concept:** Research reports and presentations
- **Guest speaker** (industry MR practitioner)
- **R lab:** Quarto reporting workshop
- **Report writing clinic** — bring drafts

Prep: draft your project report before Wk 12 so the clinic can help.

::: {.notes}
Push draft-in-hand attendance. Wk 12 clinic is only useful if teams
have real drafts to work with, not blank pages.
:::

## Thank you — questions?

`abbass44@mcmaster.ca`

::: {.notes}
Stay for questions on project analysis and Test 2 (though don't
discuss test content - grading in progress).
:::
