---
title: "Solutions — Session 11: Regression Drivers of Satisfaction"
subtitle: "M731 · Full worked walkthrough (companion to session-11.qmd)"
author: "Shay Abbasi"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Purpose

Full worked version of the compressed Session 11 regression
walkthrough: linear regression, diagnostics, standardised
coefficients, interactions, and a logistic-regression coda reusing
the Week 10 A/B dataset.

## Setup

```{r}
#| label: setup
library(tidyverse)
library(broom)
library(car)
library(lmtest)
library(sandwich)
library(lm.beta)
library(emmeans)

survey <- read_csv("data/maritime_survey.csv") |>
  mutate(tier = factor(tier, levels = c("Standard", "Signature", "Discovery")))
```

## Step 1 — Fit the model

```{r}
#| label: fit
fit <- lm(satisfaction ~ tenure_months + tier + region + age, data = survey)
summary(fit)
tidy(fit, conf.int = TRUE)
glance(fit)
```

**Interpretation.** Read each coefficient as "holding the other
predictors constant." Tier dummies compare Signature and Discovery to
the reference level (Standard); a positive, significant Discovery
coefficient means Discovery subscribers report higher satisfaction
than Standard subscribers, net of tenure, region, and age — consistent
with the Week 7 ANOVA finding, now controlling for other factors.

## Step 2 — Diagnostics

```{r}
#| label: fig-diagnostics
#| fig-cap: "Regression diagnostic panel"
par(mfrow = c(2, 2))
plot(fit)
par(mfrow = c(1, 1))
```

```{r}
#| label: vif-bp
vif(fit)
bptest(fit)
```

**Interpretation.** VIFs should sit comfortably under 5 — tenure,
tier, region, and age aren't redundant with each other here. If the
Breusch-Pagan test rejects (p < .05), switch to robust standard
errors:

```{r}
#| label: robust-se
coeftest(fit, vcov = vcovHC(fit, type = "HC3"))
```

## Step 3 — Standardised coefficients (driver importance)

```{r}
#| label: standardised
lm.beta(fit)
```

**Interpretation.** Standardised coefficients put tenure, tier, region,
and age on a common (SD-unit) scale, letting you rank them by relative
importance — useful for a slide that says "the three biggest drivers of
satisfaction are X, Y, Z," which raw coefficients on different scales
can't support directly.

## Step 4 — Interaction: does tenure matter differently by tier?

```{r}
#| label: interaction
fit_int <- lm(satisfaction ~ tenure_months * tier + region + age, data = survey)
tidy(fit_int, conf.int = TRUE)
```

```{r}
#| label: fig-interaction
#| fig-cap: "Estimated satisfaction by tenure and tier"
emm <- emmeans(fit_int, ~ tenure_months | tier, at = list(tenure_months = c(3, 12, 24)))
plot(emm)
```

**Interpretation.** If the interaction term is not significant, the
tenure effect is essentially parallel across tiers — the simpler
additive model (Step 1) is preferred on grounds of parsimony.

## Step 5 — Logistic regression coda (reusing the Week 10 A/B data)

```{r}
#| label: logistic
ab <- read_csv("data/maritime_abtest.csv")

fit_log <- glm(upgraded ~ tenure_months + treatment + age, family = binomial, data = ab)
tidy(fit_log, exponentiate = TRUE, conf.int = TRUE)
```

**Interpretation.** Odds ratios above 1 increase the odds of
upgrading; below 1 decrease it. Remember: an odds ratio of 2 does not
mean "twice as likely" in probability terms — always convert back to
predicted probabilities (e.g., with `emmeans` or `predict(type =
"response")`) before putting a number in front of a non-technical
audience.

## Step 6 — Interpretation for a general manager (100 words)

> Satisfaction is highest among Discovery subscribers and rises
> modestly with tenure, holding other factors constant; region and age
> show smaller effects. Standardised coefficients indicate tier
> membership is the single largest driver of satisfaction we can
> measure directly. **Practically**: investments in moving Standard
> subscribers toward Signature or Discovery (via bundling, trial
> upgrades, or targeted offers) are likely to move satisfaction more
> than tenure-based loyalty programs alone — though tenure still helps
> at the margin.

## Take-away for students

Say "associated with," never "causes," when reporting observational
regression results — a rule this deck's own ethics thread insists on.
The A/B test in Week 10 is what actually identifies a causal effect;
this regression identifies associations conditional on the observed
covariates.
