---
title: "Session 7 — ANOVA and Comparing Multiple Groups"
subtitle: "M731 Marketing Research · Comprehensive Lecture Notes"
author: "Shay Abbasi"
date: "2026-10-21"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    self-contained: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

# Why ANOVA? The Problem with Multiple t-Tests

## The Type I Error Inflation Problem

In Session 6 we covered the t-test for comparing two group means. A natural question arises: what do you do when you have three or more groups? The tempting answer is to run multiple t-tests — one for each pair of groups — but this is statistically dangerous.

Consider a study comparing satisfaction across three subscription tiers (Standard, Signature, Discovery). There are three possible pairwise comparisons: Standard vs. Signature, Standard vs. Discovery, and Signature vs. Discovery. If you run each at α = 0.05, you are not controlling your error rate at 5%. You are actually accepting a much higher risk of finding at least one false positive.

The math follows from the complement rule. The probability of not making a Type I error on a single test is 1 − 0.05 = 0.95. Across three independent tests, the probability of making no errors is 0.95³ = 0.857. Therefore the probability of at least one false positive across the three tests is 1 − 0.857 = 0.143 — roughly 14%, not 5%. With ten comparisons it rises to about 40%; with twenty, to about 64%.

This inflation of the familywise error rate is why ANOVA exists. Analysis of Variance (ANOVA) provides a single omnibus test at the intended α that simultaneously addresses all groups. When the omnibus test is significant, you then proceed to post-hoc comparisons using correction methods designed to control the familywise error rate.

## The Core Intuition

ANOVA works by comparing two sources of variance:

**Between-group variance** captures how much the group means differ from the overall grand mean. If the means are far apart, between-group variance is large.

**Within-group variance** captures the spread of individual observations around their respective group means. This represents natural sampling variability and measurement error.

The F-ratio is the quotient of these two quantities:

$$F = \frac{MS_{between}}{MS_{within}}$$

When the null hypothesis is true (all group means are equal), both the numerator and denominator estimate the same population variance σ², so F hovers around 1. When the null is false (at least one group mean differs), the numerator is inflated by the real group differences while the denominator is unaffected, pushing F above 1. Large F values — relative to their sampling distribution — lead to rejection of the null.

The visual intuition is equally powerful: imagine three overlapping distributions plotted on the same axis. If the distributions are nearly identical (close means, wide spreads), the group signal is buried in the noise and F is small. If the distributions are well-separated (distant means, narrow spreads), the group signal is clear and F is large.

# The ANOVA Model

## One-Way ANOVA: The Statistical Model

A one-way ANOVA assumes that every observation $Y_{ij}$ (the $i$-th observation in group $j$) can be decomposed as:

$$Y_{ij} = \mu + \alpha_j + \varepsilon_{ij}$$

where:

- $\mu$ is the grand mean across all observations
- $\alpha_j$ is the deviation of group $j$ from the grand mean (the "treatment effect" for group $j$)
- $\varepsilon_{ij}$ is the individual-level error, assumed to be normally distributed with mean 0 and constant variance σ²

The constraint $\sum_j \alpha_j = 0$ ensures that the treatment effects are deviations that sum to zero around the grand mean.

The null hypothesis is $H_0: \alpha_1 = \alpha_2 = \cdots = \alpha_k = 0$ — that is, no group deviates from the grand mean. The alternative is that at least one $\alpha_j \neq 0$.

## The F-Distribution

The F-statistic follows an F-distribution with two degrees of freedom parameters: $(k-1)$ in the numerator (where $k$ is the number of groups) and $(N-k)$ in the denominator (where $N$ is the total sample size). These degrees of freedom determine the shape of the F-distribution and thus the critical value at any chosen α level.

A simulation makes the F-distribution concrete. If you repeatedly sample data where all group means are truly equal (the null is true) and compute the F-statistic each time, the resulting histogram of F values follows the theoretical F(k−1, N−k) distribution. Your critical value is the point in the right tail that cuts off exactly α proportion of the area. An observed F exceeding this critical value leads to rejection.

```{r}
#| label: f-distribution-sim

# Simulate the F-distribution under the null hypothesis
# Three groups with equal means — what F values do we expect?
set.seed(42)
n_sim <- 10000

f_null <- replicate(n_sim, {
  group <- factor(rep(1:3, each = 100))
  y <- rnorm(300, mean = 0, sd = 1)   # all groups have the same mean
  summary(aov(y ~ group))[[1]]["group", "F value"]
})

# Plot the null distribution with the 5% critical value marked
hist(f_null, breaks = 100, probability = TRUE,
     main = "F-distribution under H0: df(2, 297)",
     xlab = "F value", col = "lightblue", border = "white")
abline(v = qf(0.95, 2, 297), col = "red", lwd = 2, lty = 2)
legend("topright", legend = "Critical value (α = .05)",
       col = "red", lty = 2, lwd = 2)
```

Values to the right of the red line lead to rejection of the null hypothesis. Under the null, only 5% of F values fall there by chance — that is the definition of a 5% Type I error rate.

## Sum of Squares Decomposition

The total variability in the data (SS_total) decomposes exactly into two non-overlapping parts:

$$SS_{total} = SS_{between} + SS_{within}$$

**SS_total** = $\sum_{ij}(Y_{ij} - \bar{Y})^2$, the sum of squared deviations of every observation from the grand mean.

**SS_between** = $n_j \sum_j (\bar{Y}_j - \bar{Y})^2$, the sum of squared deviations of group means from the grand mean, weighted by group size. This is the signal — how much variation is accounted for by group membership.

**SS_within** = $\sum_{ij}(Y_{ij} - \bar{Y}_j)^2$, the sum of squared deviations of individual observations from their own group mean. This is the noise.

The mean squares are the SS divided by their respective degrees of freedom:
- $MS_{between} = SS_{between} / (k-1)$
- $MS_{within} = SS_{within} / (N-k)$

And the F-statistic is the ratio $F = MS_{between} / MS_{within}$.

Working through this decomposition manually builds genuine understanding of what ANOVA is testing:

```{r}
#| label: ss-decomposition

# Manually compute the ANOVA decomposition for a Maritime Roast example
set.seed(7)
n_per <- 50
group_means <- c(3.0, 3.4, 3.9)   # Standard, Signature, Discovery

data_sim <- data.frame(
  tier = rep(c("Standard", "Signature", "Discovery"), each = n_per),
  satisfaction = c(
    rnorm(n_per, group_means[1], 0.8),
    rnorm(n_per, group_means[2], 0.8),
    rnorm(n_per, group_means[3], 0.8)
  )
)

grand_mean <- mean(data_sim$satisfaction)

# SS_between: how different are group means from the grand mean?
group_means_obs <- tapply(data_sim$satisfaction, data_sim$tier, mean)
SS_between <- n_per * sum((group_means_obs - grand_mean)^2)

# SS_total: total variability around the grand mean
SS_total <- sum((data_sim$satisfaction - grand_mean)^2)

# SS_within: variability inside each group
SS_within <- SS_total - SS_between

# Mean squares and F
MS_between <- SS_between / (3 - 1)
MS_within  <- SS_within  / (nrow(data_sim) - 3)
F_stat     <- MS_between / MS_within

cat("SS_between:", round(SS_between, 2), "\n")
cat("SS_within: ", round(SS_within,  2), "\n")
cat("F-statistic:", round(F_stat, 3),    "\n")
cat("p-value:   ", round(pf(F_stat, 2, 147, lower.tail = FALSE), 4), "\n")
```

Compare this manual calculation with the output from `aov()` — they will agree exactly. Understanding the decomposition is what lets you interpret ANOVA output intelligently rather than simply reading off a p-value.

# Running One-Way ANOVA in R

## Basic Implementation

The Maritime Roast research question in Session 7 is: *Does customer satisfaction differ systematically across the three subscription tiers?* This is a natural ANOVA question — three groups, a continuous outcome (satisfaction measured on a 1–7 scale), and a theoretical expectation that higher tiers may attract or produce more satisfied customers.

```{r}
#| label: anova-basic

library(tidyverse)
library(car)

# Load and prepare data — tier must be a factor
survey <- read_csv("data/maritime_survey.csv") |>
  mutate(
    tier   = factor(tier, levels = c("Standard", "Signature", "Discovery")),
    region = factor(region)
  )

# One-way ANOVA
fit <- aov(satisfaction ~ tier, data = survey)
summary(fit)
```

The `summary()` output presents the ANOVA table with columns for degrees of freedom (Df), sum of squares (Sum Sq), mean squares (Mean Sq), F value, and Pr(>F). Read the tier row: if F(2, 797) = 14.3, p < .001, you conclude that at least one tier mean differs from the others.

## Type I, II, and III Sums of Squares

For a simple one-way ANOVA with equal cell sizes, the choice of SS type does not matter. But for multi-factor designs — particularly unbalanced ones — it matters considerably.

**Type I (sequential):** Each effect is tested after adjusting for all previously entered effects. The result depends on the order you specify terms in the model formula. This is almost never what you want in observational marketing research, where predictor order is arbitrary.

**Type II:** Each effect is tested after adjusting for all other main effects, but not for interaction terms involving that effect. This is generally the appropriate choice for main effects in marketing research.

**Type III:** Each effect is tested after adjusting for all other terms in the model, including interactions. This is what SPSS reports by default. It can be misleading when interactions are present (because it tests main effects that are only interpretable when all interactions are zero — an unusual situation).

For most M731 analyses, Type II from `car::Anova()` is the recommended default:

```{r}
#| label: anova-types

# Type II SS — adjust for other main effects, not interactions
car::Anova(fit, type = 2)

# For two-way ANOVA with interaction
fit2 <- aov(satisfaction ~ tier * region, data = survey)
car::Anova(fit2, type = 2)
```

Use `car::Anova()` (capital A) rather than base R's `anova()` (lowercase a) for cleaner output and the ability to specify the SS type.

# Effect Size for ANOVA

## Why the F-Test Alone Is Insufficient

A statistically significant F-test tells you only that the group means are unlikely to be equal. It says nothing about how large the differences are or whether they matter in practice. With n = 800, a difference of 0.1 scale points between tiers will be highly significant; whether it is worth acting on is a separate question.

ANOVA effect sizes quantify the proportion of total variability explained by group membership. They are analogous to R² in regression.

## Eta-Squared and Omega-Squared

The two most common measures are η² (eta-squared) and ω² (omega-squared):

| Measure | Formula | Notes |
|---------|---------|-------|
| η² | SS_between / SS_total | Simple; overestimates true effect in small samples |
| ω² | (SS_between − df_between × MS_within) / (SS_total + MS_within) | Bias-corrected; preferred for reporting |
| Partial η² | SS_effect / (SS_effect + SS_error) | Used in multi-factor designs |

Cohen's (1988) benchmarks for η²: small = 0.01, medium = 0.06, large = 0.14. The same thresholds apply approximately to ω².

Importantly, these benchmarks come from the social sciences broadly and should be treated as rough guides. In commercial marketing research, even a small effect (η² = 0.02) may represent a meaningful business opportunity if the outcome is revenue, churn, or customer lifetime value.

```{r}
#| label: effect-sizes-anova

library(effectsize)

# Eta-squared with confidence interval
eta_squared(fit, partial = FALSE)

# Omega-squared — less biased (preferred)
omega_squared(fit, partial = FALSE)

# Partial effect sizes for multi-factor design
eta_squared(fit2, partial = TRUE)

# The effectsize package also provides interpretation
interpret_eta_squared(0.12)   # returns "large"
```

Always report effect sizes with confidence intervals. An effect size point estimate without uncertainty is incomplete. The `effectsize` package computes CIs using the non-central F distribution (Steiger 2004), which is the correct method.

# Post-Hoc Tests: Finding Which Groups Differ

## The Purpose of Post-Hoc Testing

A significant ANOVA F-test establishes that at least one group mean differs from the others, but does not identify which pairs. Post-hoc tests address this by running pairwise comparisons with appropriate multiplicity corrections.

The correction is not optional. Running raw pairwise t-tests after a significant ANOVA — what is sometimes called the "protected t-test" — does not adequately control the familywise error rate. You need an explicit multiplicity correction applied to the correct set of comparisons.

## The Post-Hoc Landscape

| Method | Controls | When to use |
|--------|---------|-------------|
| Tukey HSD | Familywise α, all pairwise | Default for balanced one-way ANOVA |
| Bonferroni | Familywise α | Very conservative; use for small k |
| Sidak | Familywise α | Slightly less conservative than Bonferroni |
| Scheffé | Familywise α, any contrast | Most liberal; applies to any comparison, not just pairwise |
| Dunnett | Familywise α vs. one control | Compare all groups to a single reference |
| Benjamini-Hochberg (FDR) | Expected proportion of false discoveries | Large k, exploratory analyses |

Tukey's Honest Significant Difference (HSD) is the workhorse. It uses the Studentized range distribution (q distribution) rather than the t-distribution, and it controls the familywise error rate exactly for all pairwise comparisons.

## Tukey HSD in R

```{r}
#| label: tukey-basic

# Base R approach
tukey_result <- TukeyHSD(fit)
tukey_result

# Visualise with confidence intervals
plot(tukey_result)
```

The output shows, for each pair: the estimated difference in means, the lower and upper 95% confidence interval for the difference, and the adjusted p-value. Pairs whose confidence interval does not include zero are significantly different at the corrected level.

## The Modern Approach: emmeans

The `emmeans` package (estimated marginal means, also known as least-squares means) is the modern standard for post-hoc analysis. It handles unbalanced designs correctly, integrates cleanly with multi-factor models, and provides a consistent interface for any contrast type.

```{r}
#| label: emmeans-posthoc

library(emmeans)

# Estimated marginal means
emm <- emmeans(fit, ~ tier)
emm   # shows each group mean with CI

# All pairwise contrasts with Tukey correction
pairs(emm, adjust = "tukey")

# Bonferroni correction instead
pairs(emm, adjust = "bonferroni")

# Raw p-values (for applying your own FDR correction downstream)
pairs(emm, adjust = "none")
```

Prefer `emmeans` over `TukeyHSD` for any analysis beyond simple balanced one-way ANOVA.

## Planned vs. Unplanned Contrasts

A critical distinction exists between contrasts specified before seeing the data (planned or a priori contrasts) and those conducted after observing results (unplanned or post-hoc contrasts):

**Planned contrasts** are derived from theory. If your theoretical model predicted, before data collection, that Discovery subscribers would be more satisfied than the average of Standard and Signature subscribers, you can test that specific comparison at nominal α without multiplicity correction — provided it was genuinely specified in advance.

**Unplanned contrasts** are exploratory. They require full multiplicity correction because you are searching a space of possible comparisons rather than confirming a pre-specified hypothesis.

The discipline of specifying contrasts in advance is another form of informal preregistration. Teams that write their hypotheses before running analyses avoid accidental HARKing (Hypothesising After Results are Known).

Custom contrasts in emmeans use contrast vectors where values sum to zero:

```{r}
#| label: custom-contrasts

library(emmeans)
emm <- emmeans(fit, ~ tier)

# Test: Discovery vs. the average of Standard and Signature
# Levels are in alphabetical order by default; check with levels(survey$tier)
# Discovery=1, Signature=-0.5, Standard=-0.5
my_contrast <- list(
  "Discovery vs. others" = c(1, -0.5, -0.5)
)

contrast(emm, my_contrast)
```

This tests exactly the comparison your theory predicted, at the full power of a planned test rather than the reduced power of a post-hoc comparison.

# Assumption Checking

## The Three ANOVA Assumptions

One-way ANOVA rests on three assumptions:

**Independence of observations.** Each data point must come from a different, independent individual. This is violated by repeated measures on the same person, clustered sampling (multiple respondents from the same household or location), or any panel structure where the same individual appears multiple times. Independence is the most serious violation — there is no "robust to mild violations" clause here.

**Normality of residuals within each group.** The errors $\varepsilon_{ij}$ should be approximately normally distributed. Thanks to the Central Limit Theorem, ANOVA is robust to moderate non-normality when groups are reasonably large (n > 30 per group is a common rule of thumb). The normality assumption matters most when samples are small.

**Homogeneity of variances (homoscedasticity).** The population variance σ² should be the same in each group. When variances differ markedly, the F-test can be anti-conservative (too many false positives) or overly conservative, depending on which groups have the larger variance and which are the largest. The solution is Welch's ANOVA, discussed below.

## Checking Normality

The Q-Q plot (quantile-quantile plot) of residuals is more informative than a formal normality test, especially at large sample sizes where Shapiro-Wilk will reject trivially small deviations:

```{r}
#| label: normality-check

# Q-Q plot of residuals
plot(fit, which = 2)

# Shapiro-Wilk test (reliable for n < 5000)
shapiro.test(residuals(fit))
```

In the Q-Q plot, residuals that hug the diagonal line are approximately normal. S-shaped patterns indicate heavy tails; curved departures from the diagonal indicate skewness. Minor wobble along the diagonal is expected and acceptable; severe departures warrant action (transformation or non-parametric alternative).

## Checking Homoscedasticity

Levene's test is the standard check for equal variances. It is more robust than Bartlett's test to non-normality of residuals:

```{r}
#| label: levene-test

library(car)

# Levene's test for equal variances
leveneTest(satisfaction ~ tier, data = survey)
```

A significant Levene test (p < .05) indicates that population variances differ across groups, and you should switch to Welch's ANOVA rather than the standard equal-variance version.

## The Full Diagnostic Panel

The `plot()` method for an `aov` object produces four diagnostic plots:

```{r}
#| label: diagnostic-plots

par(mfrow = c(2, 2))
plot(fit)
par(mfrow = c(1, 1))
```

The four panels are:

1. **Residuals vs. fitted values:** Check for non-linearity (curved pattern) and heteroscedasticity (funnel shape). For ANOVA, fitted values are just group means, so you should see three vertical columns of points; each column should have roughly equal spread.

2. **Normal Q-Q plot:** Diagnose normality as described above.

3. **Scale-location plot (sqrt of standardised residuals vs. fitted):** Another view of homoscedasticity. A horizontal smoother line with uniform spread is ideal.

4. **Residuals vs. leverage:** Identify influential observations. Points with high leverage and large residuals (Cook's distance > 1) may be pulling the model fit unduly.

Make this panel part of every ANOVA analysis, not an afterthought.

# Welch's ANOVA: A Safe Default

## The Case for Welch's

Just as Welch's t-test is now recommended over Student's t-test for comparing two groups (because it does not assume equal variances), Welch's ANOVA is increasingly recommended as the default for one-way designs. Simulation studies (Clinch & Keselman 1982; Tomarken & Serlin 1986) demonstrate that Welch's ANOVA controls Type I error better than standard ANOVA when variances are unequal, while losing very little power when variances happen to be equal.

The practical recommendation is to use Welch's ANOVA unless you have a specific reason not to:

```{r}
#| label: welch-anova

# Welch's ANOVA — does not assume equal variances
oneway.test(satisfaction ~ tier, data = survey, var.equal = FALSE)

# Standard ANOVA — assumes equal variances (shown for comparison only)
oneway.test(satisfaction ~ tier, data = survey, var.equal = TRUE)
```

The degrees of freedom in Welch's ANOVA are adjusted using the Welch-Satterthwaite approximation, which is why the denominator df is often a non-integer. This is expected and correct.

## Games-Howell Post-Hoc for Welch's ANOVA

If you use Welch's ANOVA, the standard Tukey HSD is not appropriate — it assumes equal variances, which contradicts your choice to use Welch's. The matching post-hoc procedure is Games-Howell (1976), available in the `rstatix` package:

```{r}
#| label: games-howell

library(rstatix)

# Games-Howell: post-hoc that does not assume equal variances
games_howell_test(survey, satisfaction ~ tier)
```

Use Games-Howell whenever Levene's test is significant or whenever you have elected Welch's ANOVA as your omnibus test.

# Two-Way ANOVA and Interactions

## Adding a Second Factor

Two-way ANOVA extends the model to include two categorical predictors simultaneously. In the Maritime Roast context, you might ask not only whether satisfaction differs by tier, but also whether it differs by region — and whether the tier effect is the same across regions.

```{r}
#| label: two-way-anova

# Two-way ANOVA with interaction
fit2 <- aov(satisfaction ~ tier * region, data = survey)

# The * operator expands to: tier + region + tier:region
# Type II SS for clean hypothesis tests
car::Anova(fit2, type = 2)
```

The output now shows three tests: the main effect of tier (adjusting for region), the main effect of region (adjusting for tier), and the tier × region interaction.

## What an Interaction Means

An interaction exists when the effect of one factor depends on the level of the other. In the Maritime Roast example, an interaction between tier and region would mean that the satisfaction difference between tiers is not the same in every region. Perhaps Discovery subscribers in Atlantic Canada are dramatically more satisfied than Standard subscribers, while in Ontario the gap is minimal.

Interactions are conceptually important because they signal that a single, simple recommendation ("invest in Discovery tier") may not apply uniformly across the business. A disordinal interaction — where the rank ordering of tiers changes across regions — is particularly important strategically: it means different regions may need different tier strategies.

The visual diagnosis is the interaction plot:

```{r}
#| label: interaction-plot

library(emmeans)
library(ggplot2)

# Estimated marginal means for each tier × region combination
emm_int <- emmeans(fit2, ~ tier | region)
emm_df  <- as.data.frame(emm_int)

# Interaction plot: parallel lines = no interaction; crossing = interaction
ggplot(emm_df, aes(x = tier, y = emmean,
                   colour = region, group = region)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  labs(
    title    = "Satisfaction by tier and region",
    subtitle = "Estimated marginal means ± 95% CI",
    x        = "Subscription tier",
    y        = "Mean satisfaction (1-7)",
    colour   = "Region"
  ) +
  theme_minimal() +
  scale_colour_manual(values = c("#7A003C", "#FDBF57", "#333333"))
```

Three interaction patterns matter for interpretation:

**No interaction:** The lines connecting group means across levels of one factor are parallel. Tier effects are the same in every region. Main effects are interpretable on their own.

**Ordinal interaction:** Lines are not parallel but do not cross. One group is always higher, but the gap varies across levels of the other factor. Main effects are still directionally informative, but the interaction adds nuance.

**Disordinal (crossing) interaction:** Lines cross. The rank order of groups changes across levels of the other factor. Main effects can be actively misleading — reporting "Discovery is better" is wrong if Discovery is only better in some regions.

## Following Up a Significant Interaction

When the interaction term is statistically significant, do not interpret main effects in isolation. Instead, examine conditional effects (the effect of tier separately within each region):

```{r}
#| label: interaction-followup

# Test interaction first
car::Anova(fit2, type = 2)

# If interaction is significant: simple effects (tier within each region)
emm2 <- emmeans(fit2, ~ tier | region)

# Pairwise contrasts within each region, Tukey-corrected
pairs(emm2, by = "region", adjust = "tukey")
```

The `by = "region"` argument tests all tier pairwise comparisons separately within each region. The Tukey correction applies within each region's set of comparisons.

# Non-Parametric Alternatives

## When to Use Non-Parametric Tests

Non-parametric tests should be used when the assumptions of ANOVA are genuinely untenable — not as a precautionary default. They are indicated when:

- Sample size is small (fewer than 15–20 per group) and the distribution is clearly non-normal
- The outcome is truly ordinal and interval-level assumptions are not defensible
- Extreme outliers are present that cannot be justified for removal
- Levene's test is highly significant and the groups are also very unequal in size

The cost of using non-parametric tests unnecessarily is reduced statistical power. When the ANOVA assumptions approximately hold, parametric tests are more efficient — they extract more information from the same data.

## Kruskal-Wallis: The Non-Parametric ANOVA

The Kruskal-Wallis test is the non-parametric analogue of one-way ANOVA. It works by replacing raw values with their ranks across all observations and testing whether the average ranks differ significantly across groups.

A common misconception is that Kruskal-Wallis tests medians. It technically tests for stochastic dominance — whether one group tends to produce higher values than another — which coincides with a median difference only under specific distributional assumptions. For practical purposes in marketing research, you can interpret a significant Kruskal-Wallis as indicating that at least one group tends to have higher values.

```{r}
#| label: kruskal-wallis

# Kruskal-Wallis test
kruskal.test(satisfaction ~ tier, data = survey)
```

When the Kruskal-Wallis test is significant, use Dunn's test with Benjamini-Hochberg correction for post-hoc comparisons (the FDR approach is common here because Kruskal-Wallis contexts often involve smaller samples where Bonferroni is overly conservative):

```{r}
#| label: dunn-test

library(FSA)

# Dunn's test: post-hoc for Kruskal-Wallis
dunnTest(satisfaction ~ tier, data = survey, method = "bh")
```

## Power of Non-Parametric vs. Parametric Tests

When data are truly normally distributed, the Kruskal-Wallis test has about 95% of the power of one-way ANOVA (the asymptotic relative efficiency is 3/π ≈ 0.955). This means you lose only about 5% of your power by using the non-parametric test in the best case for ANOVA.

However, when the data are heavily skewed or have heavy tails, the non-parametric test can be substantially more powerful. With satisfaction measured on a 5–7 point ordinal scale, both tests typically give similar results when group sizes exceed 30, and reporting disagreements between them is informative about the robustness of the finding.

## Friedman Test: Repeated Measures

The Friedman test is the non-parametric analogue of one-way repeated-measures ANOVA. It applies when the same subjects are measured multiple times (e.g., satisfaction measured at three time points — before a loyalty programme, six months in, and one year in):

```{r}
#| label: friedman-test

# Friedman test for repeated measures (data must be in long format)
friedman.test(satisfaction ~ time | subject_id, data = long_data)
```

The `time` variable identifies the repeated measurement occasion; `subject_id` identifies each unique respondent. This design is relatively rare in standard cross-sectional marketing surveys but becomes relevant for longitudinal or panel research.

# The Multiple-Comparisons Ethics Thread

## Multiple Testing and Scientific Integrity

The multiple-comparison corrections discussed in this session — Tukey, Bonferroni, FDR — are mathematical tools, but they carry an ethical dimension. The corrections only work honestly when applied to the full set of tests that were actually performed.

Consider the following scenario: a researcher runs 30 group comparisons, finds 5 that reach significance at α = .05, and reports only those 5 — correctly noting that they applied Bonferroni correction to the 5 reported tests. This is dishonest. The correction must apply to all 30 tests that were conducted, not just the ones selected for reporting. A Bonferroni correction across 5 tests sounds rigorous but is entirely cosmetic if the true search space was 30 tests.

## Familywise Error Rate vs. False Discovery Rate

Two different frameworks for multiplicity control serve different research purposes:

**Familywise error rate (FWER) control** — achieved by Bonferroni, Sidak, Tukey — ensures that the probability of making even one false positive across the entire family of tests is at most α. This is the appropriate standard for confirmatory research where any single false positive is costly.

**False discovery rate (FDR) control** — achieved by the Benjamini-Hochberg procedure — ensures that the expected proportion of false positives among the significant findings is at most α. If you report 20 significant tests with FDR controlled at 5%, you expect at most 1 of those 20 to be a false positive. This is more lenient than FWER control, and is appropriate for exploratory analyses where a few false leads are tolerable and you want to maximise the number of true findings detected.

In standard M731 work (5–30 comparisons), Tukey or Bonferroni is appropriate. FDR becomes attractive for large-scale analyses such as text mining, item analysis of lengthy survey instruments, or any context with hundreds of simultaneous tests.

## Pre-Specification as the Structural Solution

The cleanest resolution to the multiple-comparisons problem is pre-specification: decide which tests you will run before looking at the data, write them down, and stick to the plan. Analyses that deviate from the pre-specified plan should be flagged as exploratory and not presented as confirmatory evidence.

This discipline does not eliminate the need for correction (exploratory analyses still require it), but it makes the correction honest — the space of pre-specified tests is the family to which the correction applies.

In team projects, the practical implementation is a "hypothesis document" — a brief written record of the tests the team plans to run, dated before the analysis phase begins. This is an informal version of the preregistration practices that are becoming standard in academic marketing research.

# Complete Five-Step ANOVA Workflow

The following workflow integrates all the elements covered in this session into a reproducible analysis sequence.

## Step 1 — Prepare the Data

```{r}
#| label: anova-step1

library(tidyverse)
library(car)
library(effectsize)
library(emmeans)

# Load and prepare — ensure tier is a factor with explicit level order
survey <- read_csv("data/maritime_survey.csv") |>
  mutate(
    tier   = factor(tier, levels = c("Standard", "Signature", "Discovery")),
    region = factor(region)
  )

# Descriptive statistics by tier — always start here
survey |>
  group_by(tier) |>
  summarise(
    n      = n(),
    mean   = mean(satisfaction, na.rm = TRUE),
    sd     = sd(satisfaction,   na.rm = TRUE),
    median = median(satisfaction, na.rm = TRUE),
    q25    = quantile(satisfaction, 0.25, na.rm = TRUE),
    q75    = quantile(satisfaction, 0.75, na.rm = TRUE)
  )
```

Means, SDs, and medians together tell you whether the groups look different and whether distributions are skewed within groups. Convert to factor explicitly — `aov()` will work with character vectors but factor is unambiguous about level ordering.

## Step 2 — Visualise

```{r}
#| label: anova-step2

library(ggplot2)

# Boxplot + jitter: shows both summary and raw data density
ggplot(survey, aes(x = tier, y = satisfaction, fill = tier)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1.5) +
  scale_fill_manual(values = c("#7A003C", "#FDBF57", "#888888")) +
  labs(
    title    = "Satisfaction by subscription tier",
    subtitle = "Maritime Roast Coffee Co. · n = 800",
    x        = "Subscription tier",
    y        = "Satisfaction (1-7)",
    caption  = "Jittered points; whiskers = 1.5 × IQR"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The combination of boxplot and jitter is more informative than either alone. The boxplot shows the distribution summary; the jitter reveals data density. If you observe points clustered at integer values (because satisfaction is measured on a 5-point scale), the jitter makes this visible in a way a boxplot alone does not.

## Step 3 — Run ANOVA and Check Assumptions

```{r}
#| label: anova-step3

# Fit both Welch and standard ANOVA for comparison
fit_welch    <- oneway.test(satisfaction ~ tier, data = survey, var.equal = FALSE)
fit_standard <- aov(satisfaction ~ tier, data = survey)

# Normality of residuals (use Q-Q plot as primary; Shapiro as secondary)
shapiro.test(residuals(fit_standard))   # reliable if n < 5000
plot(fit_standard, which = 2)           # Q-Q plot

# Homoscedasticity
leveneTest(satisfaction ~ tier, data = survey)

# Full diagnostic panel
par(mfrow = c(2, 2))
plot(fit_standard)
par(mfrow = c(1, 1))
```

With n = 800, the Shapiro-Wilk test will almost certainly return a significant result for even minor departures from normality. Rely primarily on the Q-Q plot. If residuals approximately follow the diagonal, normality is not a practical concern.

## Step 4 — Post-Hoc Tests and Effect Size

```{r}
#| label: anova-step4

# Post-hoc via emmeans (handles balanced and unbalanced designs)
emm <- emmeans(fit_standard, ~ tier)
pairs(emm, adjust = "tukey")

# Effect sizes with confidence intervals
eta_squared(fit_standard)
omega_squared(fit_standard)   # preferred — less biased

# If Levene's was significant: use Games-Howell instead
library(rstatix)
games_howell_test(survey, satisfaction ~ tier)
```

Match your post-hoc procedure to your omnibus test. If you used Welch's ANOVA, use Games-Howell. If you used standard ANOVA, use Tukey HSD. Mixing them (Welch's omnibus + Tukey post-hoc) is inconsistent.

## Step 5 — Write the Result

Every ANOVA write-up for M731 purposes should contain six elements:

1. Which ANOVA variant was used and why
2. The omnibus F-statistic with degrees of freedom and p-value
3. Which post-hoc method was applied
4. Which pairs differ, with mean differences and confidence intervals
5. Effect size with confidence interval
6. A managerial interpretation sentence

**Example write-up:**

> "A one-way Welch's ANOVA (no equal-variance assumption) revealed that mean satisfaction differed significantly across subscription tiers, F(2, 486.3) = 14.2, p < .001. Post-hoc comparisons using the Games-Howell procedure showed that Discovery subscribers reported higher satisfaction than Standard subscribers (Δ = 0.7, 95% CI [0.5, 0.9], p < .001) and higher than Signature subscribers (Δ = 0.3, 95% CI [0.1, 0.6], p = .004). The effect was medium in magnitude (ω² = 0.09, 95% CI [0.05, 0.14]). The satisfaction deficit in the Standard tier represents a potential retention risk and warrants investigation into whether it reflects product quality, customer expectations at enrolment, or a mismatch between tier features and subscriber needs."

The managerial sentence transforms a statistical finding into a research insight. Without it, the analysis is technically complete but not finished.

# Power Analysis for ANOVA

## Why Power Matters Before Data Collection

A common mistake in marketing research is to determine sample size based on budget or convenience rather than statistical power. If the study has insufficient power to detect effects of the size that actually exist, it will produce inconclusive results — not because the effects are absent, but because the study was underpowered to find them.

The relevant question is: given an expected effect size and desired power (typically 80%), how many respondents are needed per group?

Cohen's f is the standardised effect size measure for ANOVA:
$$f = \sqrt{\frac{\eta^2}{1 - \eta^2}}$$

Benchmarks: f = 0.10 (small), f = 0.25 (medium), f = 0.40 (large).

```{r}
#| label: power-analysis-anova

library(pwr)

# Required n per group for one-way ANOVA
# Targeting medium effect (η² ≈ 0.06, f ≈ 0.25)
pwr.anova.test(
  k         = 3,       # number of groups
  f         = 0.25,    # Cohen's f for medium effect
  sig.level = 0.05,
  power     = 0.80
)
```

For a medium effect and three groups, you need approximately 52 per group (156 total). Maritime Roast's n = 800 (roughly 267 per tier if balanced) provides well over 99% power for a medium effect — effectively guaranteeing that if a medium tier difference exists, the study will detect it.

## Simulation-Based Power Illustration

The following simulation makes the power-sample size relationship tangible. It runs 500 ANOVA analyses at each of several sample sizes, with a fixed small-to-medium effect, and reports the proportion that reach significance — that is, the empirical power:

```{r}
#| label: power-simulation

set.seed(42)

# True group means produce η² ≈ 0.04 (small-medium effect)
run_anova <- function(n_per_group) {
  y <- c(
    rnorm(n_per_group, 3.5, 1.0),
    rnorm(n_per_group, 3.7, 1.0),
    rnorm(n_per_group, 3.9, 1.0)
  )
  g <- factor(rep(c("Standard", "Signature", "Discovery"), each = n_per_group))
  p <- summary(aov(y ~ g))[[1]][1, "Pr(>F)"]
  p
}

# Empirical power at each sample size
ns <- c(20, 50, 100, 200, 400)
power_results <- sapply(ns, function(n) {
  mean(replicate(500, run_anova(n)) < 0.05)
})

data.frame(n_per_group = ns, empirical_power = round(power_results, 3))
```

A study with only 20 respondents per group has roughly 25% power for this effect size — it misses the real difference 75% of the time. A study with 200 per group has over 95% power. This is why power analysis belongs in the research design phase, before data collection begins.

# Common Questions and Pitfalls

## "My ANOVA is significant — which groups differ?"

The ANOVA F-test only establishes that at least one pair of group means differs. It does not tell you which pair. You must follow up with post-hoc tests (`TukeyHSD` or `emmeans::pairs`). Stopping at the omnibus F is a common reporting error.

## "I have a 5-point Likert item — should I use ANOVA?"

This is one of the most frequently debated questions in applied marketing research. Technically, a 5-point Likert item is ordinal, and ANOVA assumes interval-level measurement. In practice:

- With 5 or more scale points and group sizes over 30, ANOVA is typically robust. The violation of the interval assumption has minimal practical impact.
- The safest approach is to run both ANOVA and Kruskal-Wallis. If they agree (they usually do), report ANOVA (it is more familiar and communicates effect size more naturally). If they disagree, report both and discuss.
- Do not average Likert items from a multi-item scale and then worry about the interval assumption — the averaging process itself moves the distribution toward normality.

## "My residuals are not normal — is my ANOVA invalid?"

Not necessarily. With group sizes above 30, the Central Limit Theorem makes the F-statistic robust to moderate non-normality. Examine the Q-Q plot: minor wobble around the diagonal is expected and acceptable. Severe departures (heavy tails, large skew visible in the Q-Q plot) are the cases that warrant the Kruskal-Wallis alternative.

Formal normality tests (Shapiro-Wilk) will reject for trivial deviations when sample sizes are large. At n = 800, even normally distributed data will occasionally produce a significant Shapiro-Wilk result due to sampling variation. Do not use the p-value from Shapiro-Wilk mechanically — use your eyes on the Q-Q plot.

## "I have very unequal group sizes — what do I do?"

Use Welch's ANOVA (`oneway.test(..., var.equal = FALSE)`) and Games-Howell post-hoc. They handle unbalanced designs and unequal variances simultaneously. When cell sizes are very unequal and you are using a multi-factor design, specify Type II or Type III SS explicitly with `car::Anova()` rather than relying on base R's `anova()`.

## "How do I report ANOVA in a table?"

The superscript convention is common in commercial research reports: group means that share the same superscript letter are not significantly different from each other (at the Tukey-corrected level). For example:

| Tier | Mean satisfaction | SD | n |
|------|------------------|----|---|
| Standard | 3.9^a^ | 0.9 | 271 |
| Signature | 4.2^ab^ | 0.8 | 265 |
| Discovery | 4.6^b^ | 0.7 | 264 |

Here, Standard and Discovery differ (different superscripts); Signature does not differ significantly from either (shares superscript with both). This is more compact than listing every pairwise comparison.

# References and Further Reading

The foundational treatment of ANOVA remains Kirk (1995) *Experimental Design: Procedures for the Behavioral Sciences*, which covers the full range from one-way to nested and repeated-measures designs. For the practical case against the equal-variance assumption, Tomarken & Serlin (1986) and Clinch & Keselman (1982) demonstrate the superiority of Welch's approach through simulation.

Cohen (1988) *Statistical Power Analysis for the Behavioral Sciences* (2nd ed.) establishes the effect size conventions used throughout this session. The `effectsize` package (Ben-Shachar et al., 2020) implements these computations with proper confidence intervals based on the non-central F distribution, following Steiger (2004).

For the `emmeans` package and the general framework of estimated marginal means for post-hoc inference, Lenth (2022) "emmeans: Estimated Marginal Means, aka Least-Squares Means" is the authoritative reference.

Benjamini & Hochberg (1995) "Controlling the false discovery rate: a practical and powerful approach to multiple testing" (*Journal of the Royal Statistical Society B*, 57(1), 289–300) is the original FDR paper and worth reading for the conceptual framing of what "controlling the proportion of false discoveries" means.
