---
title: "Session 7 — Data Analysis & Hypothesis Testing (continued)"
subtitle: "M731 Week 7 · October 21, 2026"
author: "Shay Abbasi"
date: "2026-10-21"
---

## Session 7 {background-color="#7A003C"}

Data analysis and hypothesis testing (cont.)

Focus: ANOVA and post-hoc, assumption checks, robust alternatives.

::: {.notes}
This continues Wk 6 without a hard break. Recap the framework quickly
and move on. Students should already be comfortable with t-tests and
chi-square from Wk 6.
:::

## Today's agenda

1. Recap of Wk 6
2. Beyond two groups — one-way ANOVA
3. Post-hoc tests (Tukey HSD, etc.)
4. Two-way ANOVA and interactions
5. Assumption checks
6. Non-parametric alternatives
7. **Break (15 min)**
8. R lab: ANOVA on `maritime_survey.csv`
9. Ethics: multiple-comparison corrections
10. Wrap-up

::: {.notes}
Multiple-comparison ethics fits naturally right after post-hoc
discussion.
:::

## Learning objectives

By end of session 7 you will be able to:

- Run one-way and two-way ANOVA in R
- Apply appropriate post-hoc contrasts and understand p-value adjustment
- Diagnose assumption violations and choose robust alternatives
- Interpret interaction effects

::: {.notes}
Interactions are the concept most students find slippery. Budget
extra time and use a visual example.
:::

# Part 1 — ANOVA {background-color="#7A003C"}

## Why ANOVA?

When you have more than two groups:

- Running multiple t-tests inflates Type I error
- ANOVA gives one global test at the intended α

If 3 groups → 3 pairwise t-tests → familywise α ≈ 14%, not 5%.

::: {.notes}
Motivate ANOVA. Students often ask "why not just do multiple t-tests?"
The answer is Type I inflation. Whiteboard the math briefly if
useful: 1 - (1-.05)³ ≈ .143.
:::

## The intuition

ANOVA compares:

- **Between-group** variance (differences among group means)
- **Within-group** variance (spread inside each group)

Big between + small within → F is large → likely real effect.

::: {.notes}
Draw the picture on the whiteboard: three distributions, close means
+ wide spreads = small F; distant means + narrow spreads = large F.
Students remember pictures better than formulas.
:::

## One-way ANOVA in R

```{r}
#| eval: false
library(car)

fit <- aov(satisfaction ~ tier, data = survey)
summary(fit)
```

Output: F-statistic, df, p-value.

::: {.notes}
Live-code. Show output. Interpret: F(2, 797)=X.X, p=P. Reject H0 that
all tier means are equal. Then follow with post-hoc.
:::

## Post-hoc: which groups differ?

```{r}
#| eval: false
TukeyHSD(fit)
```

Tukey's HSD adjusts for the number of comparisons.

Output: pairwise differences with adjusted p-values.

::: {.notes}
Tukey is the default post-hoc for one-way ANOVA. Bonferroni is more
conservative; Sidak, Scheffé are alternatives. Show output - which
pair(s) drive the ANOVA rejection?
:::

## Type I / Type II / Type III sums of squares

For unbalanced designs, choice of SS type matters.

```{r}
#| eval: false
Anova(fit, type = 2)   # car::Anova
```

- **Type I:** sequential, order-dependent (rarely what you want)
- **Type II:** each effect adjusted for main effects only
- **Type III:** each effect adjusted for all others (SPSS default)

::: {.notes}
For balanced designs, all three give same answer. For unbalanced,
Type II is the sensible default in most cases. Use car::Anova() not
base anova() for cleaner output.
:::

## Two-way ANOVA

Two factors at once:

```{r}
#| eval: false
fit2 <- aov(satisfaction ~ tier * region, data = survey)
Anova(fit2, type = 2)
```

`*` includes main effects + interaction.

::: {.notes}
Tier * region = tier + region + tier:region. Interaction is the
"do the tier effects differ across regions?" question. Often the
most interesting finding.
:::

## Interpreting interactions

An **interaction** means the effect of one factor depends on the level of the other.

- No interaction: parallel lines when you plot group means
- Interaction: non-parallel lines

Visualise with an **interaction plot**.

::: {.notes}
Draw parallel-lines vs. crossing-lines on the whiteboard. Interaction
= "it depends" - the effect of region depends on tier, or vice versa.
Interesting substantively but complicates communication.
:::

## Interaction plot

```{r}
#| eval: false
library(emmeans)

emm <- emmeans(fit2, ~ tier | region)
plot(emm)

# Or base plot:
interaction.plot(survey$tier, survey$region, survey$satisfaction)
```

::: {.notes}
`emmeans` is the modern package for estimated marginal means (aka
"least-squares means"). Handles unbalanced designs correctly. Plot
gives publication-ready output.
:::

# Part 2 — Assumption checks {background-color="#7A003C"}

## ANOVA assumptions

1. **Independence** of observations
2. **Normality** of residuals within each group
3. **Homogeneity** of variances (homoscedasticity)

Robust to mild violations if samples are large-ish.

::: {.notes}
Independence is the killer assumption - repeated measures, clustered
sampling, panel data all violate it. Normality matters less as n
grows (CLT). Homoscedasticity matters more.
:::

## Check normality

```{r}
#| eval: false
# Q-Q plot of residuals
plot(fit, which = 2)

# Shapiro test
shapiro.test(residuals(fit))
```

QQ plot is more informative than the test — the test rejects at large n even for trivial deviations.

::: {.notes}
This is a recurring theme: with large samples, normality tests
reject anything. Look at the plot. If the residuals hug the diagonal,
you're fine.
:::

## Check homoscedasticity

```{r}
#| eval: false
leveneTest(satisfaction ~ tier, data = survey)
```

Levene's test is more robust than Bartlett's.

If significant → variances differ → consider Welch's ANOVA (`oneway.test`).

::: {.notes}
Welch's ANOVA doesn't assume equal variances and is a safe default.
`oneway.test()` in base R runs Welch by default. Some statisticians
argue Welch should be the default over standard ANOVA.
:::

## Diagnostic plots

```{r}
#| eval: false
par(mfrow = c(2, 2))
plot(fit)
```

Four plots:

- Residuals vs fitted (check linearity, homoscedasticity)
- Q-Q plot (check normality)
- Scale-location (check homoscedasticity)
- Residuals vs leverage (check influential points)

::: {.notes}
This 2x2 diagnostic panel is essential. Every ANOVA or regression
should include a look at these. Live-code the panel. Interpret each
quadrant briefly.
:::

# Part 3 — Non-parametric alternatives {background-color="#7A003C"}

## Kruskal-Wallis (non-parametric ANOVA)

When assumptions violated or data ordinal:

```{r}
#| eval: false
kruskal.test(satisfaction ~ tier, data = survey)
```

Tests medians rather than means.

::: {.notes}
Kruskal-Wallis is the ANOVA equivalent of Wilcoxon. Less powerful than
ANOVA when assumptions hold; more valid when they don't. Post-hoc for
Kruskal is Dunn's test (`FSA::dunnTest`).
:::

## Friedman (non-parametric repeated-measures)

For repeated measures on the same subjects:

```{r}
#| eval: false
friedman.test(y ~ time | subject, data = long_data)
```

Non-parametric equivalent of repeated-measures ANOVA.

::: {.notes}
Rare in commercial MR unless doing longitudinal work. Mention
briefly. If your study has pre/post/follow-up measures on the same
customers, this is your test.
:::

## When to use non-parametric

- Sample small (n < 15/group) AND distribution non-normal
- Ordinal outcome variable
- Extreme outliers you can't justify removing
- Homoscedasticity badly violated

Otherwise: parametric is more powerful, robust to mild violations.

::: {.notes}
Do not default to non-parametric out of caution. Parametric tests
handle mild violations. Reserve non-parametric for real assumption
failures.
:::

## Break — 15 minutes {background-color="#FDBF57"}

::: {.notes}
Have satisfaction × tier ANOVA ready to demo. Refresh `maritime_survey.csv`
if lab machines have stale data.
:::

# Part 4 — R lab: ANOVA on Maritime Roast {background-color="#7A003C"}

## Research question

**Does satisfaction differ across the three subscription tiers?**

If yes: which tiers differ from which? Are the differences big enough to matter?

::: {.notes}
Frame as a manager's question. The CEO wants to know: "should I be
worried about satisfaction in one tier?" Statistical test answers
"is there a difference." Effect size + practical judgement answers
"should you care."
:::

## Lab task

Using `maritime_survey.csv`:

1. Descriptive statistics by tier (n, mean, SD, median)
2. Boxplot of satisfaction by tier
3. One-way ANOVA
4. Diagnostic plots
5. Tukey HSD post-hoc
6. Effect size (η² or ω²)
7. Short interpretation (100 words) for the CEO

A fuller worked version of this ANOVA walkthrough is in
**`solutions-07.qmd`** in this session's folder.

::: {.notes}
Give 40 min. Circulate. Common issues: forgetting to convert tier to
factor, missing na.rm=TRUE, panicking at diagnostic plots that look
weird (usually fine).
:::

## Effect size for ANOVA

```{r}
#| eval: false
library(effectsize)
eta_squared(fit)      # or omega_squared() for less-biased
```

- η² < 0.06: small
- η² 0.06-0.14: medium
- η² > 0.14: large

::: {.notes}
η² is proportion of variance explained. ω² is preferred (less biased
for small samples). Report either with CI.
:::

## Reporting one-way ANOVA

"Satisfaction differed significantly across subscription tiers, F(2, 797) = X.X, p < .001, η² = 0.12. Tukey HSD contrasts showed that Discovery subscribers were more satisfied than Standard (Δ = 0.4, p < .001) but not Signature (Δ = 0.1, p = .32). **Managerially**: Standard-tier satisfaction is a real gap worth investigating."

::: {.notes}
Show this template. Every ANOVA in every report should read
approximately like this. Six elements: descriptives, F, p, effect
size, contrasts, managerial interpretation.
:::

# Part 5 — Ethics thread {background-color="#7A003C"}

## Multiple comparisons — the honest version

When you run k post-hoc comparisons, familywise α inflates:

- k=3 → familywise α ≈ 14% (at nominal α=.05)
- k=10 → familywise α ≈ 40%
- k=20 → familywise α ≈ 64%

Corrections (Tukey, Bonferroni, FDR) restore control.

::: {.notes}
This is the mathematical reality. Post-hoc without correction inflates
false-positive rates. Tukey HSD does it correctly. If your software
gives you naked pairwise t-tests, you must correct.
:::

## Multiple comparisons — the dishonest version

The same correction, applied selectively, hides p-hacking.

- Ran 30 tests, reported 3 significant, corrected for those 3? **Dishonest.**
- Ran 30 tests, reported all 30, corrected for all 30? **Honest.**

The correction is about the space of tests you searched, not the tests you report.

::: {.notes}
Subtle but critical. Correction only works if applied to the true set
of tests done. Selective correction is worse than no correction - it
gives a false sense of rigour. This is a common form of p-hacking.
:::

## FDR (False Discovery Rate)

Alternative to controlling Type I familywise:

- **FWER** (Bonferroni, Tukey): control probability of *any* false positive
- **FDR** (Benjamini-Hochberg): control expected *proportion* of false positives

FDR is less conservative — appropriate for exploratory analyses with many tests.

::: {.notes}
Genomics popularised FDR when researchers were running 20,000 tests.
For most MR work (5-30 tests), Tukey or Bonferroni is fine. FDR
becomes attractive when you have hundreds of tests (e.g., text mining,
survey item analysis).
:::

## What to actually do

- Pre-specify the tests you'll run (before looking at data)
- Report all tests, not just significant ones
- Apply correction appropriate to the total number of tests
- If you post-hoc explore, flag it: "exploratory, not confirmatory"

::: {.notes}
This is the discipline. Not perfect - real analysis is iterative -
but flagging exploratory vs. confirmatory keeps you honest.
:::

# Part 6 — For your project {background-color="#7A003C"}

## Team analytical work begins

By end of this week:

- Data-cleaning script complete
- Hypotheses locked (in writing, dated)
- Test choices documented
- Preliminary results computed

Project proposal due in 2 weeks (Wk 9, Nov. 4).

::: {.notes}
"Hypotheses locked" is the informal preregistration. Teams that write
their hypotheses before running tests avoid HARKing accidentally. Push
them to do it in writing, dated.
:::

# Part 7 — Wrap-up {background-color="#7A003C"}

## Week 8 preview

- **Concept:** marketing scales — Likert, semantic differential, reliability, validity
- **R lab:** Cronbach's alpha on `maritime_survey.csv`

Prep reading: CJ Ch. 10; classic papers Churchill 1979 and Rossiter 2002 (optional but recommended).

::: {.notes}
Preview: Wk 8 shifts from analysis to measurement (going backwards
in the process, but that's the outline sequence). Frame this as "we
know how to test - now let's make sure the things we're testing are
measured reliably."
:::

## Thank you — questions?

`abbass44@mcmaster.ca`

::: {.notes}
Stay for any follow-up questions.
:::
