---
title: "Solutions — Session 7: ANOVA on Subscription Tiers"
subtitle: "M731 · Full worked walkthrough (companion to session-07.qmd)"
author: "Shay Abbasi"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Purpose

Full worked version of the Session 7 in-lab exercise: one-way ANOVA,
diagnostics, Tukey post-hoc, and effect size for satisfaction across
subscription tiers.

## Setup

```{r}
#| label: setup
library(tidyverse)
library(car)
library(effectsize)

survey <- read_csv("data/maritime_survey.csv") |>
  mutate(tier = factor(tier, levels = c("Standard", "Signature", "Discovery")))
```

## Step 1 — Descriptives by tier

```{r}
#| label: descriptives
survey |>
  group_by(tier) |>
  summarise(n = n(), mean_sat = mean(satisfaction), sd_sat = sd(satisfaction),
            median_sat = median(satisfaction), .groups = "drop")
```

## Step 2 — Boxplot

```{r}
#| label: fig-boxplot
#| fig-cap: "Satisfaction by subscription tier"
ggplot(survey, aes(x = tier, y = satisfaction, fill = tier)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("#B8B8B8", "#FDBF57", "#7A003C")) +
  labs(title = "Satisfaction by subscription tier", x = NULL, y = "Satisfaction (1-10)") +
  theme_minimal()
```

## Step 3 — One-way ANOVA

```{r}
#| label: anova
fit <- aov(satisfaction ~ tier, data = survey)
summary(fit)
```

**Interpretation.** The tier effect is significant — satisfaction is
not identical across Standard, Signature, and Discovery subscribers.

## Step 4 — Diagnostics

```{r}
#| label: fig-diagnostics
#| fig-cap: "ANOVA diagnostic panel"
par(mfrow = c(2, 2))
plot(fit)
par(mfrow = c(1, 1))
```

```{r}
#| label: assumption-checks
shapiro.test(residuals(fit))
leveneTest(satisfaction ~ tier, data = survey)
```

**Interpretation.** Residuals hug the Q-Q diagonal reasonably well;
Levene's test checks homoscedasticity. With n=800, Shapiro's test can
reject on trivial deviations — trust the Q-Q plot over the p-value
here, as the deck's ethics-of-testing discussion in Week 6 emphasised.

## Step 5 — Tukey HSD post-hoc

```{r}
#| label: tukey
TukeyHSD(fit)
```

**Interpretation.** Look at which pairwise comparisons have adjusted
p < .05. Typically the Standard-vs-Discovery gap is the clearest and
largest; Signature often sits in between and may not differ
significantly from one or both neighbours — read your own output
rather than assuming this pattern.

## Step 6 — Effect size

```{r}
#| label: effect-size
eta_squared(fit)
```

**Interpretation.** η² in the 0.06-0.14 range would indicate a medium
practical effect — tier membership explains a non-trivial share of the
variance in satisfaction, even though it's far from the only driver
(see Week 11 regression for the fuller multi-predictor picture).

## Step 7 — CEO interpretation (100 words)

> Satisfaction differs significantly across subscription tiers,
> F(2, 797) ≈ [value], p < .001, η² ≈ [value]. Tukey contrasts show
> Discovery subscribers report meaningfully higher satisfaction than
> Standard subscribers; Signature sits in between. **Managerially**,
> Standard-tier satisfaction is the segment most worth investigating —
> whether through product improvements, communication of value, or a
> guided upgrade path toward Signature. The gap is large enough to act
> on, not just statistically noteworthy.

## Take-away for students

ANOVA answers "does group matter at all?" Tukey answers "which groups
differ?" Effect size answers "how much should I care?" A report that
gives only the first is an incomplete report.
