---
title: "Session 2 — Secondary Data & Research Designs"
subtitle: "M731 Week 2 · September 16, 2026"
author: "Shay Abbasi"
date: "2026-09-16"
---

## Welcome back {background-color="#7A003C"}

Session 2 — Secondary data and research designs

::: {.notes}
Quick temperature check: how many got R + Quarto working since last
week? Show of hands. Anyone stuck - stay after for setup help. If
several are stuck, spend 5 min live-debugging before starting content.
:::

## Today's agenda

1. Recap of Wk 1 + housekeeping
2. Secondary data: sources, uses, limits
3. Research designs: exploratory / descriptive / causal
4. Choosing among designs
5. **Break (15 min)**
6. R lab: data visualisation with ggplot2
7. Ethics: attribution and terms of use
8. Industry: the commercial secondary-data landscape
9. Wrap-up

::: {.notes}
Keep pace brisk through the concept block so the lab has full time.
:::

## Recap: the MR process

Where are we today?

1. Define the problem
2. **Choose the research design** ← today
3. Design the sampling plan
4. Design the data-collection instrument
5. Collect the data (**secondary data** short-circuits some of this ← today)
6. Analyse the data
7. Report and recommend

::: {.notes}
Anchor every session in the seven-step process. Students should
internalise the arc. Say "we're in step 2 today" so they see the map.
:::

## Learning objectives

By the end of session 2 you will be able to:

- Identify major sources of secondary data and evaluate their fitness for purpose
- Distinguish exploratory, descriptive, and causal research designs
- Choose among designs for a given decision problem
- Produce publication-quality figures with ggplot2

::: {.notes}
Read out. Objective 4 (ggplot2) is the R lab; objectives 1-3 are the
concept block.
:::

# Part 1 — Secondary data {background-color="#7A003C"}

## Primary vs. secondary data

- **Primary data:** you collect it yourself for this specific research question
- **Secondary data:** already collected (by someone else, for some other purpose)

Rule: always ask *"does secondary data already answer this?"* before designing primary data collection.

::: {.notes}
Simple distinction but critical. Every year, students design a survey
when a Statistics Canada table already answers the question. Cheaper,
faster, often better.
:::

## Types of secondary data

- **Public / open:** Statistics Canada, government open data, World Bank, Bank of Canada
- **Commercial / syndicated:** Nielsen, IRI/Circana, Statista, Euromonitor, Mintel, Kantar
- **Internal:** CRM, POS, loyalty, web analytics
- **Academic / research:** ICPSR, published datasets, journal supplements
- **Web:** scraped data, social media, review sites

::: {.notes}
Give one Canadian example for each: StatsCan Table 11-10-0125 for
household spending. Nielsen Canada retail measurement. Loblaws PC
Optimum. ICPSR Michigan. Trustpilot for reviews. Match examples to
your audience.
:::

## Evaluating secondary data — the fitness-for-purpose test

Five questions to ask of any secondary source:

1. **Coverage:** does the population match yours?
2. **Timing:** is the data current enough?
3. **Definition:** are variables measured the way you need?
4. **Method:** how was the data collected? Any bias?
5. **Reliability:** is the source credible? Independent?

::: {.notes}
Emphasise coverage and definition. Students often grab a US dataset
and try to use it for Canadian decisions. Or use "customer" data that
excludes returns, or "revenue" data that includes taxes. Definition
mismatches kill analyses silently.
:::

## Strengths of secondary data

- **Fast** — no fieldwork
- **Cheap** — often free or subscription already paid
- **Longitudinal** — many series go back decades
- **Comparative** — same measure across regions or years

::: {.notes}
The comparative angle is under-appreciated. If you want to know how
Canada compares to the US on coffee spending, StatsCan + BEA get you
there in an afternoon. No survey could do that in a term.
:::

## Weaknesses of secondary data

- **Fit:** measured for someone else's question, not yours
- **Age:** often stale (annual releases)
- **Depth:** rarely goes into the specifics you need
- **Bias:** who's included? Who's excluded?

Rule of thumb: use secondary to *size* the market and *frame* the question. Use primary to *answer* it.

::: {.notes}
Good heuristic. Secondary tells you the industry is $2B, growing 3%,
with 4 major players. Primary tells you why your specific customers
are churning. Both matter; they play different roles.
:::

# Part 2 — Research designs {background-color="#7A003C"}

## The three families

- **Exploratory** — you don't know what you don't know
- **Descriptive** — you know the question, need to quantify the answer
- **Causal** — you need to know if X causes Y

These aren't rungs on a ladder — most research programmes cycle through them.

::: {.notes}
Common misconception: students think you must start exploratory and
"progress" to causal. Not true - many descriptive studies stand alone.
Some causal studies are best without a preceding exploratory phase.
Design serves the question, not vice versa.
:::

## Exploratory research

**When?** Problem is fuzzy, hypotheses unclear, need to understand *what's going on*.

**Methods:**

- Qualitative: interviews, focus groups, ethnography
- Secondary-data scans
- Expert interviews
- Pilot studies

**Output:** hypotheses, refined research questions, sharper problem definition.

::: {.notes}
Wk 5 goes deep on qualitative methods. Preview: qualitative is not
just "asking people what they think" - it's systematic, with coding
schemes and analytic rigour. Students often underestimate this.
:::

## Descriptive research

**When?** You know the questions; you need to quantify or profile.

**Methods:**

- Cross-sectional surveys
- Panel / longitudinal surveys
- Observational studies
- Sales / audit data

**Output:** frequencies, means, cross-tabs, trends, segmentation, tracking.

::: {.notes}
Most commercial MR is descriptive. Brand tracking, market sizing,
customer profiling, satisfaction studies. Not glamorous but pays the
bills. Give an example: a monthly brand-health tracker for a bank.
:::

## Causal research

**When?** You need to prove X causes Y, not just that they co-vary.

**Methods:**

- Randomised experiments (A/B tests)
- Field experiments (in real-world settings)
- Quasi-experiments (natural experiments)
- Structural econometrics

**Output:** causal estimates + confidence in those estimates.

::: {.notes}
Wk 10 dives in. Foreshadow: causal claims are harder than students
think. Correlation is not causation - this old saw becomes very real
when a marketing manager acts on a spurious correlation and loses
money.
:::

## Choosing among designs

The design should follow the decision, not the researcher's preference.

| Decision | Best design |
|----------|-------------|
| "Should we enter market X?" | Descriptive (secondary + survey) |
| "Why are customers leaving?" | Exploratory (qual) then descriptive |
| "Does new packaging lift sales?" | Causal (experiment) |
| "What segments are there?" | Descriptive (segmentation on survey) |

::: {.notes}
Table is a decision aid. Not exhaustive - real problems often need
multi-phase research programmes. Ask students: for their group
project, which design fits? Most will land on descriptive.
:::

## Time and money

Rough cost/time trade-offs (Canadian market, 2025):

- Focus group (10 participants, 1 city): $6-15K, 3-6 weeks
- Online survey (n=500): $10-25K, 3-4 weeks
- A/B test (in-house, existing website): $2-8K, 2-4 weeks
- Ethnographic study: $20-80K, 6-12 weeks
- Secondary syndicated data: subscription (often paid at org level)

::: {.notes}
Real numbers ground the design conversation. Cost matters. When a
client says "just do a survey," it's a $15K decision minimum. Focus
groups feel cheap but recruiting good participants is expensive.
:::

## Break — 15 minutes {background-color="#FDBF57"}

Back at [time].

::: {.notes}
Circulate. Check that everyone has ggplot2 (part of tidyverse)
installed. Have `statscan_coffee.csv` open on the projector ready for
the lab.
:::

# Part 3 — R lab: data visualisation with ggplot2 {background-color="#7A003C"}

## The grammar of graphics

ggplot2 builds plots in layers:

1. **Data** — a data frame
2. **Aesthetics** (`aes()`) — map columns to visual channels (x, y, colour, size)
3. **Geoms** (`geom_*()`) — the actual marks (points, bars, lines)
4. **Scales, facets, themes** — polish

Each layer added with `+`.

::: {.notes}
Explain the "grammar" idea - Leland Wilkinson's book, Wickham's
implementation. Once students see the pattern, ggplot becomes composable
rather than something to memorise.
:::

## Loading the data

```{r}
#| eval: false
library(tidyverse)
coffee <- read_csv("data/statscan_coffee.csv")
glimpse(coffee)
```

`statscan_coffee.csv` — household coffee spending by province, 2019-2024.

::: {.notes}
Live-code this. Show glimpse output. Point out that read_csv returns
a tibble - fancy data frame with better printing. Note the column
types: chr for province, int for year, dbl for spending.
:::

## Your first ggplot

```{r}
#| eval: false
ggplot(coffee, aes(x = year, y = spending, colour = province)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Household coffee spending by province, 2019-2024",
    x = "Year",
    y = "Annual spending ($ per household)",
    colour = "Province",
    caption = "Source: Statistics Canada"
  )
```

::: {.notes}
Live-code. Show the plot appearing in the Viewer pane. Interpret:
which province spends most? Any obvious trend? Are the Maritimes
comparable to Ontario? This maps back to Maritime Roast's Ontario
question.
:::

## Bar plots for categorical comparisons

```{r}
#| eval: false
coffee |>
  filter(year == 2024) |>
  ggplot(aes(x = fct_reorder(province, spending), y = spending)) +
  geom_col(fill = "#7A003C") +
  coord_flip() +
  labs(x = NULL, y = "2024 spending per household ($)")
```

`fct_reorder()` sorts the axis by spending — much easier to read.

::: {.notes}
`fct_reorder` is a common trick. Without it, provinces appear
alphabetically - useless for comparison. With it, the plot tells you
the ranking instantly. Small thing, big impact.
:::

## Faceting: small multiples

```{r}
#| eval: false
ggplot(coffee, aes(x = year, y = spending)) +
  geom_line() +
  facet_wrap(~ province, ncol = 4) +
  labs(y = "Spending per household ($)")
```

`facet_wrap()` splits the plot into panels — one per province.

::: {.notes}
Tufte would approve. Small multiples let you compare shapes across
groups. This is often more insightful than one plot with 10 coloured
lines. Show the facet output - students who've only used Excel will
have an "oh" moment.
:::

## Themes

```{r}
#| eval: false
last_plot() +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

`theme_minimal()`, `theme_bw()`, `theme_classic()` are the workhorse presets.

::: {.notes}
Themes are a rabbit hole - resist. `theme_minimal()` is fine for 90%
of course work. Publication-quality reports may need custom themes; the
group project doesn't.
:::

## In-lab exercise

**Task.** Using `statscan_coffee.csv`, produce:

1. A time-series line plot showing all provinces from 2019-2024
2. A bar chart of 2024 spending sorted from highest to lowest
3. A one-paragraph memo answering: **is StatsCan coffee spending data useful for a Maritime Roast Ontario-entry decision?**

A fuller worked version of this exercise — including additional plots
and a sample memo answer — is in **`solutions-02.qmd`** in this
session's folder.

::: {.notes}
Give students ~30-40 min. Circulate. Common gotchas: forgetting the
pipe, wrong quotes around column names, using `=` instead of `<-`
inside functions. Debug patiently. Ask 2-3 students to share their
memo answer at the end.
:::

# Part 4 — Ethics thread {background-color="#7A003C"}

## Ethical use of secondary data

- **Attribution:** always cite the source, at the granularity that lets a reader retrieve it
- **Terms of use:** many "free" datasets have licence terms (share-alike, non-commercial, attribution)
- **Aggregation:** even public data can identify individuals when cross-tabulated too finely
- **Currency:** using stale data without flagging it can mislead

::: {.notes}
Give a concrete example: StatsCan is CC-BY, cite as such. Scraped
Twitter data has ToS restrictions - Elon Musk's API changes make this
messy. Kaggle datasets vary by dataset. When in doubt, read the
licence.
:::

## Scraped data — special case

- Robots.txt is a starting point, not a ceiling
- Terms of Service usually prohibit commercial scraping
- Even legal scraping can cross ethical lines (rate-limiting, user privacy)
- **For your project:** stick to publicly-published data or open APIs

::: {.notes}
If a student is enthusiastic about scraping, redirect them to the
Twitter Academic API, Reddit's Pushshift (rest in peace), or an open
government dataset. Scraping is a headache for a course project.
:::

# Part 5 — Industry context {background-color="#7A003C"}

## The Canadian secondary-data landscape

**Government (free):**

- Statistics Canada — `www.statcan.gc.ca`
- Bank of Canada — `www.bankofcanada.ca`
- Open Government — `open.canada.ca`

**Commercial (subscription):**

- Nielsen Canada (retail measurement)
- Circana (formerly IRI + NPD; scanner + panel)
- Environics Analytics (geodemographics, PRIZM segments)
- Kantar, Ipsos (syndicated brand health)

::: {.notes}
Environics is worth mentioning - PRIZM segmentation is uniquely
Canadian and widely used in retail and telecom. Nielsen and Circana
are the two big scanner-data players; most CPG marketers touch one or
both. Ask which students have used any of these.
:::

## Statista as a starting point

- `www.statista.com` — aggregates market-size and consumer statistics
- McMaster Library likely provides student access
- Good for **framing** ("the Canadian coffee market is $X, growing Y%")
- Not good for **decision-grade** analysis (aggregated, no methodology detail)

::: {.notes}
Statista is a good starting point but not the answer. Push students to
look upstream at the original sources Statista cites. That's where
the real methodology is.
:::

# Part 6 — For your project {background-color="#7A003C"}

## What to do this week

- **Team formation:** talk to classmates about pairing up
- Start noticing marketing decisions in your own workplace / life
- The Maritime Roast secondary-data walkthrough is your template for auditing sources on your own project
- Complete TCPS-2 CORE registration (if not done)

::: {.notes}
Push team formation - if students dawdle, teams don't finalise until
Wk 3, which is late. If they want instructor help forming teams,
that's fine.
:::

## Week 3 preview

- **Concept:** measurement scales (nominal / ordinal / interval / ratio), data-collection modes, sampling plans
- **R lab:** sampling and power calculations with `pwr`
- Prep reading: SIL Ch. 6-7-9 or CJ Ch. 6-8-9

::: {.notes}
Wk 3 combines a lot of content. Warn students that the reading load is
heavier. Encourage skimming to core concepts, not deep chapter-mining.
:::

## Thank you — questions?

`abbass44@mcmaster.ca`

::: {.notes}
Stay for 5-10 min. Common questions after Wk 2: about team formation;
about R install issues. Pre-empt if you can.
:::
