---
title: "Solutions — Session 2: Secondary Data Visualisation"
subtitle: "M731 · Full worked walkthrough (companion to session-02.qmd)"
author: "Shay Abbasi"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Purpose

Full worked version of the Session 2 in-lab exercise: building
publication-quality charts from `statscan_coffee.csv` with ggplot2,
and answering the assigned memo question.

**Dataset provenance.** `statscan_coffee.csv` is instructor-simulated
household coffee spending by province, 2019-2024 — illustrative
figures modelled loosely on published Statistics Canada household
spending patterns, **not** a direct extract of a specific StatsCan
table. If you use this in a real deliverable, replace it with an
actual StatsCan CANSIM/Table extract and cite the table number.

## Setup

```{r}
#| label: setup
library(tidyverse)

coffee <- read_csv("data/statscan_coffee.csv")
glimpse(coffee)
```

## Task 1 — Time-series line plot, all provinces

```{r}
#| label: fig-timeseries
#| fig-cap: "Household coffee spending by province, 2019-2024"
ggplot(coffee, aes(x = year, y = spending, colour = province)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  scale_colour_brewer(palette = "Paired") +
  labs(
    title = "Household coffee spending by province, 2019-2024",
    x = "Year", y = "Annual spending ($ per household)",
    colour = "Province",
    caption = "Source: instructor-simulated, modelled on published StatsCan patterns"
  ) +
  theme_minimal()
```

**Interpretation.** Spending trends gently upward across every
province over the period, consistent with general price inflation on
coffee. British Columbia and Ontario sit at the top of the range;
Newfoundland sits at the bottom — useful context for sizing a national
vs. regional launch.

## Task 2 — 2024 bar chart, sorted

```{r}
#| label: fig-bar2024
#| fig-cap: "2024 household coffee spending by province, sorted"
coffee |>
  filter(year == 2024) |>
  ggplot(aes(x = fct_reorder(province, spending), y = spending)) +
  geom_col(fill = "#7A003C") +
  coord_flip() +
  labs(
    x = NULL, y = "2024 spending per household ($)",
    title = "2024 coffee spending by province",
    caption = "Source: instructor-simulated, modelled on published StatsCan patterns"
  ) +
  theme_minimal()
```

**Interpretation.** Sorting (via `fct_reorder`) turns this from a
lookup table into a ranking students can read at a glance — the single
biggest ggplot habit worth building early.

## Task 3 — Memo: is this data useful for the Ontario-entry decision?

> **Memo — Secondary-data fitness check, Ontario market entry**
>
> StatsCan-style household coffee-spending data is useful for **sizing
> and framing** Maritime Roast's Ontario-entry decision but not for
> **answering** it. It tells us Ontario households spend roughly in
> line with the national average and slightly above the Atlantic
> region, which supports a market-size argument for entry. It does
> not tell us anything about brand awareness, willingness to switch to
> a subscription model, or price sensitivity specific to a Halifax
> challenger brand — those require primary data (see the Week 3-4
> sampling and survey-design labs, and `maritime_survey.csv` from Week
> 6 onward, which includes an `Ontario_prospect` region flag for
> exactly this purpose). Recommendation: use this secondary data in the
> secondary-data-audit section of the research plan to establish
> market size, then commission the primary survey to answer the
> harder awareness/attachment/WTP questions.

## Take-away for students

Secondary data is for **sizing and framing**; primary data is for
**answering**. This memo pattern — one paragraph, explicit about what
the data can and can't tell you — is the shape every secondary-data
audit in this course should take.
