---
title: "Session 1 — Marketing Research Process & Getting Started with R"
subtitle: "M731 Week 1 · September 9, 2026"
author: "Shay Abbasi"
date: "2026-09-09"
---

## Welcome to M731 {background-color="#7A003C"}

Marketing Research at DeGroote

::: {.notes}
Open by greeting students by name where possible. Take 2-3 minutes on
introductions - who you are (research background, industry experience,
why you teach this course). Set the tone: this is a graduate-level
applied course. Emphasise that most of what they'll learn is used every
day in Canadian marketing organisations. Ask students to introduce
themselves quickly - name, work function, why they signed up. Keep it
tight - 30 seconds each. This is worth the time; it drives team
formation later.
:::

## Today's agenda

1. Course logistics
2. Introduction to marketing research
3. The marketing research process
4. Our running case: Maritime Roast Coffee Co.
5. **Break (15 min)**
6. Getting started with R, RStudio, and Quarto
7. Live-code: Chapman & Feit Chapter 2
8. Ethics thread + industry context
9. Wrap-up + Week 2 preview

::: {.notes}
Show this agenda for ~1 minute. It doubles as the "session shape" they
should expect every week: 75-min concept, 15-min break, 75-min R lab,
15-min wrap-up. Point at the wall clock so they know when the break is.
:::

## Learning objectives for today

By the end of session 1 you will be able to:

- Locate marketing research within the broader marketing decision process
- Walk through the marketing research process end-to-end
- Set up a working R + RStudio + Quarto environment
- Render a first `.qmd` report from a real dataset
- About TCPS-2 CORE ethics certification

::: {.notes}
Read this out loud. LOs matter - they're the promise the session makes.
If you don't hit one, call it out at the end and defer to next week
rather than pretending it happened.
:::

# Part 1 — Course logistics {background-color="#7A003C"}

## Who and where

- **Instructor.** Shay Abbasi · `abbass44@mcmaster.ca`
- **TA.** Ms. Saba Khorasani · `sabaakhr@mcmaster.ca`
- **Where.** RJC 5100 (Burlington campus), Wednesdays, 3 hours
- **Office hours.** Zoom, by appointment (book by email)
- **Course site.** Avenue to Learn ("BUSADMIN M731: Marketing Research")

::: {.notes}
Point to the room map if there's a whiteboard. Emphasise that all
communication uses their McMaster email, per outline. Anonymous
concerns channel exists via the DeGroote MBA site - mention this once
so students who need it know it exists.
:::

## Prerequisites (checked at registration)

- **M650** (or equivalent) — statistics / quantitative methods
- **I602** (or equivalent) — introductory quantitative methods

We will review and learn more about statistical tests such as t-tests, chi-square, or basic regression as we go. 

::: {.notes}
This is where you catch students who are underprepared. Some students
in the MBA program may have taken these years ago; some may have taken
equivalents at other institutions. Offer a 15-min diagnostic during
next week's office hours for anyone worried. Do NOT tell students to
drop the course on Day 1.
:::

## Course structure at a glance

- **13 weeks**: 12 lecture weeks + 1 presentation week
- **All materials as HTML files & Quatro Files** — you'll be reading, writing, and rendering these
- **A running case** (Maritime Roast Coffee Co.) threads every week

::: {.notes}
Slide is dense - walk through each bullet. Emphasise Quarto - many
students haven't touched it, some have. Reassure both groups: we'll
build fluency together. The running case is a big deal - it's what
makes concepts stick. Preview: Maritime Roast is a Halifax coffee
subscription business we'll use every week.
:::

## Assessment (100%)

| Component                              | Weight | When            |
|----------------------------------------|--------|-----------------|
| Assignment 1 (individual)              | 10%    | Due Wk 6        |
| Assignment 2 (individual)              | 10%    | Due Wk 9        |
| Test 1 (covers Lectures 1-4)           | 20%    | Wk 5            |
| Test 2 (covers Lectures 5-10)          | 20%    | Wk 11           |
| Project proposal + questionnaire (grp) | 10%    | Due Wk 4 & 9        |
| Written report (group)                 | 20%    | Due Wk 13       |
| Presentation (group)                   | 10%    | Wk 13           |
| **Total**                              | **100%** | **No final exam**|

::: {.notes}
No final exam is unusual and worth flagging - students will exhale.
Two tests, two individual assignments, plus a big group project.
Groups of 5-6, formed in Weeks 1-2. Test dates are firm; assignment
due dates are firm too. Assignments are developed and distributed
separately (see A2L) - this table is just the weights and dates.
:::

## Group project overview

- **Groups of 5-6** — we'll form these by end of Week 2
- **Company of your choice** (A B2C Canadian Business)
- **Three deliverables:** proposal (Wk 4) + questionnaire (Wk 9), written report (Wk 13), presentation (Wk 13)

::: {.notes}
The REB restriction is critical. If a student wants to run a real
survey they can do it post-course, but during the term the project
uses secondary or simulated data. This prevents the REB timeline from
blowing up the course. If asked why, explain: REB approval can take
6-8 weeks, our term is 13 weeks. Doesn't fit.
:::

## Group Project Structure
- **Nature of Marketing Research Project**:  through consultation with the client and the profess
- **Adequate Participation**: There will be optional peer evaluation.
- **Select a Team Leader**
- **Find a Client by Sep 25**
- **Submit Proposal by Sep 30**
- **All needed forms and instructions on A2L**

## Generative AI policy

- Generative AI (ChatGPT, Claude, Copilot, etc.) is **allowed** for:
  - Learning course material
  - Studying for tests
  - Explaining error messages
- Generative AI is **prohibited** for graded assessments:
  - Tests
  - Any part of the group project (except as an explicitly disclosed learning aid)

::: {.notes}
Read this out. This is the McMaster provost policy. Give one concrete
example: "using ChatGPT to explain why a t-test failed to reject" is
fine; "using ChatGPT to write a graded submission for you" is
prohibited. Students always ask about the grey area - be explicit that
when in doubt, disclose.
:::

## Tests and missed work

- **Tests:** Multiple choice and/or written. Sample questions will be provided beforehand.
- **Missed test:** Do NOT use MSAF (MBA program doesn't accept it). Documented reasons via DSSAO within 5 working days.
- **Alternate test:** Administered in Wk 6 (for Test 1) or December exam window (for Test 2).
- **SAS accommodations:** Coordinate via `DSBSAS@mcmaster.ca` at least two weeks before each test.

::: {.notes}
The MSAF-not-accepted point is unusual and confuses students. Say it
twice. Point them to the DSSAO email if they think they'll miss.
SAS-registered students should activate accommodations in the first
three weeks - remind them this is their responsibility.
:::

## Attendance & Class Participation

- **Attendance matters** — you'll get more out of the course by being present
- **Class Participation**: Still not evaluated, but valuable to me personally.

::: {.notes}
Don't lecture about attendance - graduate students know. Mention Top
Hat neutrally. The alternate check-in matters for anyone without a
smartphone; some international students may not have data plans yet.
:::

## What is Marketing?

> Marketing is **the process of planning and executing the conception, pricing, promotion, and distribution of ideas, goods, and services to create exchanges that satisfy individual and organizational needs.**
>
> — American Marketing Association (1985)

Note the four levers: **conception** (product), **pricing**, **promotion**, **distribution** — this is the marketing mix, and it's the frame for everything that follows.

::: {.notes}
This is the classic, broad definition of marketing itself - not marketing
research specifically. Read it slowly; it packs a lot in. "Exchanges"
is the key word - marketing exists because two parties trade value.
"Individual and organizational needs" signals this applies to B2C and
B2B alike. This sets up the narrower "what is marketing research"
discussion on the next slide.
:::

## Marketing management philosophies

Five philosophies describe how an organisation decides *what* to produce and sell:

| Philosophy                    | Core belief                                                                 | Managerial focus                                          |
|--------------------------------|------------------------------------------------------------------------------|-------------------------------------------------------------|
| **Production Concept**         | Consumers prefer products that are widely available and inexpensive          | Production efficiency, low cost, mass distribution           |
| **Product Concept**            | Consumers favour products offering the best quality, performance, features   | Continuous product improvement and innovation                |
| **Selling Concept**            | Consumers won't buy enough unless the firm sells and promotes aggressively   | Aggressive selling and promotion                              |
| **Marketing Concept**          | Achieving organisational goals depends on knowing target-market needs and delivering value better than competitors | Customer needs first, integrated marketing effort |
| **Societal Marketing Concept** | Marketing decisions should balance company profits, consumer wants, and society's long-run well-being | Consumer welfare + societal interest alongside profit |

*Source: adapted from Kotler, P., & Keller, K. L., Marketing Management.*

::: {.notes}
Walk the room through the historical arc: production → product → selling
→ marketing → societal marketing roughly tracks how firms evolved from
scarcity-era manufacturing to today's customer- and society-centred
thinking. Ask: "which of these philosophies have you seen in your own
company?" Most will recognise Selling Concept somewhere in their org.
The Marketing Concept row is the one we build on for the rest of this
slide sequence.
:::

## The Marketing Concept: three orientations

The **Marketing Concept** (previous slide) isn't just an attitude — it shows up as three concrete orientations inside a market-oriented organisation:

- **Consumer orientation** — understand and respond to customer needs
- **System orientation** — every function coordinates to deliver value consistently
- **Goal orientation** — customer satisfaction must convert into sustainable profit

The next three slides unpack each one.

::: {.notes}
This is the bridge slide between "philosophies" (historical/conceptual)
and "market orientation" (operational). Emphasise: a company can *say*
it follows the marketing concept and still not be market-oriented in
practice. Market orientation is what the marketing concept looks like
when it's actually implemented.
:::

## Consumer orientation: customer focus & market intelligence

- **Customer focus** — continuously understand and respond to customer needs, through feedback loops, responsive service, and regular engagement
- **Market intelligence** — actively collect and analyse information on customers, competitors, and the broader market (competitive analysis, trend monitoring, ongoing research)
- **Distributing intelligence** — insight only matters if it's shared: cross-functional teams, shared dashboards, regular briefings
- Together, these help the organisation spot opportunities, anticipate demand shifts, and stay ahead of competitors

::: {.notes}
This is the orientation students will recognise most easily - it's the
"listen to the customer" piece. The distribution point is the one
people forget: market intelligence sitting in one department's
dashboard that nobody else sees isn't market orientation, it's a
report. Ask: "where does customer feedback go in your organisation,
and who actually sees it?"
:::

## System orientation: coordinated marketing & responsiveness

- **Coordinated marketing** — every department works together to create, communicate, and deliver value consistently, via integrated marketing communications and cross-functional collaboration, so the customer gets one unified experience
- **Responsiveness** — the ability to adapt quickly when customer needs or market conditions change (agile marketing, rapid prototyping)
- Both require clear communication, shared goals, and a willingness to change course when the data says so

::: {.notes}
Give a concrete failure example: a customer complains to support, but
product and marketing never hear about it, so the same complaint
recurs for a year. That's a coordination failure, not a data failure -
the intelligence existed, it just didn't move. Responsiveness is the
speed half of the same coin.
:::

## Goal orientation: profitability & long-term perspective

Customer satisfaction is the priority, but it has to be **sustainable**:

- **Profitability** — satisfying customers is the goal, but the organisation still needs to be financially viable; customer lifetime value (CLV) and return on marketing investment (ROMI) measure whether customer-focused efforts pay off
- **Long-term perspective** — the focus is on building lasting relationships, not maximising next quarter's numbers (loyalty programs, strategic partnerships)
- Sometimes this means short-term sacrifices (e.g., investing in service improvements that don't immediately lift revenue) in exchange for sustainable growth

::: {.notes}
This is the orientation students most often forget - they hear "market
orientation" and think it's purely about pleasing the customer. Push
back gently: an organisation that satisfies customers into bankruptcy
isn't market-oriented, it's just generous. CLV is worth a 30-second
definition if this is the first time the term comes up in the course -
it returns in Wk 11 regression content on churn drivers.
:::

# Part 2 — What is marketing research? {background-color="#7A003C"}

## Marketing research, defined

> Marketing research is **the systematic gathering, recording, and analysis of data** about problems relating to the marketing of goods and services.
>
> — American Marketing Association

Three words matter: **systematic**, **data**, **problems**.

::: {.notes}
Emphasise "systematic" - this distinguishes MR from journalism or
opinion. "Data" - not gut feel. "Problems" - MR exists to solve a
business problem, not to satisfy curiosity. If a research request
doesn't map to a decision, push back.
:::

## Marketing research vs. market research

- **Market research:** studying a specific market (size, structure, competitors)
- **Marketing research:** studying anything that informs marketing decisions — pricing, product, promotion, place, plus customer research, brand research, etc.

The terms are often used interchangeably. This course covers the broader "marketing research."

::: {.notes}
Small distinction, worth 30 seconds. Most textbooks use "marketing
research"; most Canadian job postings say "market research" or "market
insights." Both terms work.
:::

## Where does MR sit in an organisation?

- **Client-side:** In-house Insights team at a brand (Loblaws, Bell, TD)
- **Supplier-side:** MR agencies (Nielsen, IRI/Circana, Ipsos, Environics)
- **Consulting:** Strategy firms with insights practices (McKinsey, BCG)
- **Increasingly:** Product analytics teams (Shopify, Amplitude clients)

::: {.notes}
Ask the room: how many are client-side vs. supplier-side vs.
consulting? Get a sense of the audience - it shapes how you frame
examples. If mostly client-side, emphasise the "hiring a supplier"
angle. If mostly supplier-side, emphasise the "reading a brief" angle.
:::

## What questions does MR answer?

- **Descriptive:** Who buys, what they buy, when, how much?
- **Diagnostic:** Why did share drop? Why are customers churning?
- **Predictive:** Will this new product succeed? What is the price elasticity?
- **Prescriptive:** Which segment should we target? What should the pack say?

Every research project should answer one of these — clearly.

::: {.notes}
This is a useful framework students will return to. When someone
proposes a study, ask "which of these four is it?" If they can't
answer, the research question isn't sharp enough. This is a good
question to steal for their project's problem statement.
:::

## The four Ps and MR

MR informs each of the marketing mix decisions:

- **Product:** concept tests, product-in-use tests, package tests
- **Price:** price sensitivity, willingness-to-pay, van Westendorp
- **Place:** channel performance, distribution audits
- **Promotion:** ad tests, brand tracking, media effectiveness

Plus: segmentation, positioning, brand equity, customer satisfaction.

::: {.notes}
Ask students: which of these do you touch in your day job? Get 3-4
hands up. This connects the course to their existing work - they've
already been consuming MR outputs (dashboards, tracking studies)
without knowing it.
:::

## Quick check: descriptive, diagnostic, predictive, or prescriptive?

Ten real Maritime Roast research questions, one at a time — each tied
to one of the 4 Ps. Guess the question **type** out loud, then press
**space** to reveal the answer.

::: {.notes}
This is a fast, interactive check that exercises two frameworks at
once: the 4 Ps (what part of the marketing mix) and descriptive /
diagnostic / predictive / prescriptive (what kind of question). Don't
let any one example eat more than 20-30 seconds - if the room is
stuck, give a hint rather than lingering.
:::

## Example 1 — Product

> "How many subscribers have tried the Discovery tier since launch?"

::: {.fragment}
**Descriptive** — counting who did what, how much.
:::

## Example 2 — Product

> "Why are Discovery-tier subscribers cancelling within their first two months?"

::: {.fragment}
**Diagnostic** — explaining a problem (early churn).
:::

## Example 3 — Product

> "If we add a decaf option to the Discovery tier, how many current subscribers would upgrade?"

::: {.fragment}
**Predictive** — forecasting the effect of a product change.
:::

## Example 4 — Product

> "Which new roast should we add to the Discovery tier lineup?"

::: {.fragment}
**Prescriptive** — recommending a specific action.
:::

## Example 5 — Price

> "What is the average monthly spend per subscriber across our three tiers?"

::: {.fragment}
**Descriptive** — quantifying current behaviour.
:::

## Example 6 — Price

> "Why did the Standard-tier renewal rate drop after last month's price increase?"

::: {.fragment}
**Diagnostic** — explaining a change in outcome.
:::

## Example 7 — Price

> "What is the price elasticity of demand for the Signature tier?"

::: {.fragment}
**Predictive** — the classic price-elasticity example from a few slides back.
:::

## Example 8 — Place

> "What percentage of Atlantic Canada grocers currently carry Maritime Roast retail bags?"

::: {.fragment}
**Descriptive** — current distribution coverage.
:::

## Example 9 — Place

> "Which independent grocers in Ontario should we approach first for retail placement?"

::: {.fragment}
**Prescriptive** — recommending where to act.
:::

## Example 10 — Promotion

> "Will a referral-bonus campaign increase new subscriber sign-ups next quarter?"

::: {.fragment}
**Predictive** — forecasting a promotion's effect.
:::

::: {.notes}
After example 10, point out the pattern deliberately: all 4 Ps
appeared, and so did all 4 question types — Product got all four types
in a row (2-4), which is a nice teaching moment that any P can be
asked about descriptively, diagnostically, predictively, or
prescriptively. This connects directly back to "What questions does MR
answer?" and "The four Ps and MR" a few slides earlier.
:::

# Part 3 — The marketing research process {background-color="#7A003C"}

## The seven-step process

1. **Define the problem** (translate a business decision into a research question)
2. **Choose the research design** (exploratory, descriptive, causal)
3. **Design the sampling plan**
4. **Design the data-collection instrument** (questionnaire, guide, protocol)
5. **Collect the data** (field the study)
6. **Analyse the data**
7. **Report and recommend**

Every session in this course corresponds to a step (or part of one).

::: {.notes}
Steps 1-3 are the design phase (Wks 1-3). Step 4 is Wk 4-8. Step 5
happens between weeks or in the project. Step 6 is Wks 6-11. Step 7 is
Wk 12. Point at the wall calendar as you say this so the arc is
visible.
:::

## Step 1: Problem definition (the hard step)

The **business problem** ≠ the **research problem**.

- Business problem: "Our subscription churn is up 3 pp."
- Research problem: "Identify the leading drivers of 6-month churn among Maritime Roast subscribers, and quantify the impact of each."

The research problem is **specific, measurable, and answerable with data**.

::: {.notes}
This is where most bad MR starts. Client says "sales are down, find
out why." Research problem needs to be much sharper. Give an example
from your own experience if you have one. Emphasise that "why are
sales down" is not a research question - it's the beginning of a
conversation with the client to arrive at a research question.
:::

## Step 2: Research design

Three families:

- **Exploratory:** hypotheses unclear, need to understand the phenomenon → qualitative, secondary data
- **Descriptive:** you know the questions, need to quantify → surveys, panel data
- **Causal:** need to know if X causes Y → experiments, quasi-experiments

Most projects are descriptive. All three families are on our syllabus.

::: {.notes}
Wk 2 develops this in detail. Foreshadow: "next week we'll spend the
whole session on this - so remember these three words." Ask if anyone
knows what "quasi-experiment" means - collect the answer and defer
until Wk 10.
:::

## Step 3: Sampling plan

- **Frame:** who is eligible to be in the sample?
- **Method:** probability (random) or non-probability (convenience, quota)?
- **Size:** how many? (calculated from precision and effect size)

Not: "we'll get 400 responses." Yes: "we'll get 400 responses because that gives us ±5% precision at 95% confidence for a proportion around 50%."

::: {.notes}
The "not/yes" contrast is important. A number without justification
is a red flag in any research proposal. Wk 3 covers the calculation.
Don't get into it here - just plant the flag that sample size is
justified, not guessed.
:::

## Step 4: Instrument design

- **Surveys** use questionnaires (Wk 4, refined in Wk 8)
- **Focus groups** use discussion guides (Wk 5)
- **Observational studies** use coding protocols (Wk 5)
- **Experiments** use treatment scripts (Wk 10)

Instruments are where measurement error sneaks in — bad wording, bad order, leading questions.

::: {.notes}
Instrument design is under-rated by students. They think "just write
good questions." In reality, small changes in wording can shift
results by 10 percentage points. Krosnick 1999 is the classic paper
here - mention as optional reading for Wk 4.
:::

## Step 5: Data collection

- **Modes:** in-person, phone, mail, online, mobile, panel
- **Trade-offs:** cost, speed, coverage, response bias, incidence rate
- **Quality:** attention checks, straight-lining detection, IP dedup

For your group project, you'll use **secondary or simulated data** — but you should be able to design a collection plan as if you were fielding.

::: {.notes}
Even though group projects don't field, the plan must be defensible.
Students design a sampling plan as if fielding, using Maritime Roast's
Ontario entry as the worked example, but never actually field it.
:::

## Step 6: Analysis

- **Describe** the data (Wk 6)
- **Test** hypotheses (Wks 6-7)
- **Reduce** complexity (Wks 8 scales, 9 factor)
- **Model** drivers (Wk 11 regression)
- **Design experiments** (Wk 10)

R is our tool for every one of these. That's why we start setting up R today.

::: {.notes}
Signpost: this is 6 of our 13 sessions. Analysis is the meat of the
course. Reassure students who are R-anxious - we build slowly.
:::

## Step 7: Report and recommend

- **Structure:** executive summary → findings → evidence → limitations → recommendations
- **Audience:** the decision-maker, not the analyst
- **Format:** Quarto, HTML for reading, PDF for archiving, slides for presenting

The best analysis in the world is worthless if the recommendation doesn't land. Wk 12 covers reporting.

::: {.notes}
This is often what industry hires you for - not the analysis but the
communication. Emphasise: a mediocre analysis well-communicated will
beat a brilliant analysis poorly communicated, every time.
:::

# Part 4 — The running case: Maritime Roast Coffee Co. {background-color="#7A003C"}

## Meet Maritime Roast

- Halifax-based direct-to-consumer specialty coffee subscription
- Founded 2019, ~15,000 subscribers
- Small retail presence across Atlantic Canada (independent grocer partnerships)
- Three subscription tiers: **Standard, Signature, Discovery**

*(Fictional company - but based loosely on real Canadian coffee subscription businesses.)*

::: {.notes}
Introduce the case warmly - this is our shared world for the next 13
weeks. Ask if anyone lives in Halifax or subscribes to a coffee
service. Personal connection to the case makes labs stick.
:::

## The decisions on Maritime Roast's plate

Four strategic questions we'll explore this term:

1. **Should we enter the Ontario market?** (Wks 2-4)
2. **How do we reduce 6-month subscriber churn?** (Wks 5, 11)
3. **Should we launch a "Discovery" premium tier?** (Wk 10 A/B test)
4. **How should we price subscriptions?** (Wks 3, 6)

Every technique this term is motivated by one of these decisions.

::: {.notes}
This is the spine. Every week you'll say "this technique lets us
answer question X for Maritime Roast." That's what makes the arc
coherent. Students will build muscle memory: whenever they see a new
technique, first question is "what business decision does this
inform?"
:::

## What the case gives us

- **A recurring dataset spine** — `maritime_survey.csv`, `maritime_transactions.csv`
- **A worked example every week** — see how one company applies each technique
- **A template for your project** — you'll do the equivalent for your chosen company

The group project uses a **different** company (Maritime Roast is the teacher's example).

::: {.notes}
Reassure the room: Maritime Roast is not the project. It's the
worked example. The project is theirs. This is important - some
students would happily just do Maritime Roast for their project.
Off-limits.
:::

## Break — 15 minutes {background-color="#FDBF57"}

Back at [time].

Meanwhile: install R (`https://cran.rstudio.com/`) and RStudio
(`https://posit.co/download/rstudio-desktop/`) if you haven't yet.

::: {.notes}
This is critical - install failures on Day 1 eat lab time. Circulate
during the break and help anyone stuck. Windows users usually smooth;
Mac users occasionally need Xcode command-line tools; Linux users
sort themselves out. If someone can't get R installed in the break,
pair them with a working neighbour during the lab.
:::

# Part 5 — Getting started with R, RStudio, Quarto {background-color="#7A003C"}

## What are we installing?

- **R** — the language and statistical engine (`https://cran.rstudio.com/`)
- **RStudio (Posit)** — the IDE we'll use to write R code (`https://posit.co/download/`)
- **Quarto** — the publishing system we'll use for all documents (`https://quarto.org`)
- **TinyTeX** — a lightweight LaTeX install so we can render to PDF (`quarto install tinytex`)

Everything is free and cross-platform.

::: {.notes}
Show installers in sequence on the projector if network allows. Talk
through what each does: R is the engine, RStudio is the workshop,
Quarto is the printing press. TinyTeX is optional if we stick to HTML,
but PDF for submissions is common - install it once and forget it.
:::

## RStudio anatomy

::: {.incremental}
- **Top-left:** editor pane (where you write .qmd and .R files)
- **Bottom-left:** console (where R evaluates commands)
- **Top-right:** environment (variables, data frames)
- **Bottom-right:** files, plots, packages, help
:::

::: {.notes}
Actually project RStudio and click through the panes. Students learn
this by watching, not from a slide. Show them how to reset the pane
layout if they mess it up (Tools → Global Options → Pane Layout).
:::

## Anatomy of a .qmd file

`````yaml
---
title: "My first Quarto document"
format: html
---

# A section header

Some prose.

```{r}
1 + 1
```
`````

Three parts: **YAML header** (metadata), **prose** (markdown), **code chunks** (R).

::: {.notes}
Explain each part slowly. YAML controls output format - html here, but
could be pdf, docx, revealjs. Markdown is easy - headers, bold, lists,
tables. Code chunks are the magic - R runs and results embed into the
document.
:::

## Installing packages

```{r}
#| eval: false
install.packages(c(
  "tidyverse", "readr", "readxl", "haven", "jsonlite",
  "knitr", "gt", "kableExtra", "janitor", "broom",
  "renv", "psych", "car", "lmtest", "sandwich",
  "effectsize", "pwr", "corrplot", "emmeans", "lm.beta",
  "tidytext", "readtext", "MASS", "RColorBrewer"
))
```

Run this **once** during the lab. Takes 5-10 minutes.

::: {.notes}
Have this ready in a script file to email out or paste from the
chalkboard. On slow WiFi at RJC, downloads can take a while - start
this at the beginning of the lab and let it run while you talk. If a
package fails, most common cause on Windows is antivirus scanning -
add R's library path to the exclusion list.
:::

## The `renv` reproducibility promise

```{r}
#| eval: false
renv::init()      # first time: creates project-level package library
renv::snapshot()  # before submitting: lock package versions
renv::restore()   # someone else opens the project: restore the lock
```

Every group project must include a working `renv.lock` file.

::: {.notes}
renv is worth 10 minutes now to save 10 hours later. If groups don't
use it, one member updates a package and breaks the analysis for the
rest. Demonstrate initialising renv in a scratch project during the
lab.
:::

# Part 6 — Live-code: CF Chapter 2 {background-color="#7A003C"}

## What we'll cover in the lab

- Vectors, data frames, indexing
- Reading a CSV file
- Basic descriptive statistics
- Your first Quarto render

Working dataset: `maritime_transactions.csv` (~15,000 rows).

::: {.notes}
This is the meat of the R lab. Have the dataset pre-loaded on your
machine. Live-code slowly - narrate every command. Students at the
back should be following along on their own machines. Ask "did that
work for everyone?" after each block.
:::

## Vectors

```{r}
#| eval: false
prices <- c(15.99, 22.50, 18.75, 25.00, 19.99)
mean(prices)
sd(prices)
median(prices)
```

A vector is a sequence of values of the same type.

::: {.notes}
Start with numeric vectors. Then show character `c("Standard",
"Signature", "Discovery")` and logical `c(TRUE, FALSE, TRUE)`. Point
out that R vectorises operations - `prices * 1.13` scales every price
without a loop.
:::

## Data frames

```{r}
#| eval: false
library(tidyverse)
tx <- read_csv("data/maritime_transactions.csv")
glimpse(tx)
dim(tx)
```

A data frame is a table where each column is a vector of the same length.

::: {.notes}
Explain that `read_csv()` from readr is fast and infers column types
sensibly. `glimpse()` from dplyr is your friend - it shows the first
few values of every column vertically. Contrast with `head()` which
shows rows horizontally.
:::

## Indexing

```{r}
#| eval: false
tx$tier            # one column
tx[1, ]            # first row
tx[1:5, "amount"]  # first five values of the amount column

tx |> filter(tier == "Discovery") |> summarise(mean_amount = mean(amount))
```

The pipe `|>` reads left-to-right, "and then".

::: {.notes}
Native pipe `|>` (base R since 4.1) preferred to magrittr `%>%` for
new work - simpler, no package required. But both work. If students
prefer `%>%` they can use it; consistency within a document is what
matters.
:::

## Descriptives with dplyr

```{r}
#| eval: false
tx |>
  group_by(tier, region) |>
  summarise(
    n = n(),
    mean_amount = mean(amount),
    median_amount = median(amount),
    total = sum(amount)
  )
```

Read this as: "take tx, group by tier and region, then compute these summaries per group."

::: {.notes}
This is the workhorse pattern for exploratory analysis. group_by +
summarise appears in every session from now on. Have students run
this on their own machines. When they get output, ask them to
interpret: which tier has highest mean order value? Which region?
:::

## Full worked solution

The in-class snippets above are the live-coding version.

A complete, expanded walkthrough of `maritime_transactions.csv` — every
step above plus additional descriptives, output, and interpretation —
is in **`solutions-01.qmd`** in this session's folder.

::: {.notes}
Point students to the solutions file if they want to see a fuller
worked version outside of class, or if they fall behind during
live-coding and need to catch up afterward.
:::

## Your first Quarto render

```{r}
#| eval: false
# In RStudio: File > New File > Quarto Document
# Save as: `session-01-lab.qmd`
# Click "Render" button (or Ctrl/Cmd + Shift + K)
```

You should see:

- HTML file rendered in the Viewer pane
- File saved next to your .qmd
- Any errors shown in the Render tab

::: {.notes}
Ceremony matters here. First render is a milestone. Congratulate the
class. If someone's render fails, most common cause is missing TinyTeX
(if PDF) or a package not installed. Debug live for the room - it's
educational.
:::

## Before next week

- **Not graded**, but essential: verify your R/RStudio/Quarto setup works
- Render the lab .qmd on your own machine before Wk 2
- Post any install errors in the A2L discussion so we can debug collectively

::: {.notes}
Frame this as insurance, not homework. If someone can't render by
Wk 2, they'll fall behind fast. Announce office hours for setup help.
:::

# Part 7 — Ethics thread {background-color="#7A003C"}

## Why ethics is a course thread, not a lecture

- Every session has an ethics moment (10-15 min) baked in
- Because ethics isn't a topic - it's a way of doing research
- We don't wait until the ethics lecture to worry about consent

::: {.notes}
Explain the deliberate design: ethics threads make ethics real, not
abstract. When a p-hacking issue comes up in Wk 6, we discuss it
there, not in a Week-14 ethics lecture that doesn't exist. This is a
graduate research course - ethics is professional practice.
:::

## TCPS-2 CORE — your first ethics deliverable

- **Tri-Council Policy Statement, 2nd edition, CORE tutorial**
- Free, online, ~2 hours
- Register at `https://tcps2core.ca`
- Complete by end of Week 4
- Upload certificate to A2L

Every marketing researcher who touches human data in Canada should have this.

::: {.notes}
This is genuinely useful, not busywork. Students who work in
consulting or agency roles will encounter TCPS when clients ask about
data practices. Some employers require it. Give it 2 minutes to
register during the break next week if they haven't yet.
:::

## PIPEDA in a marketing-research context

- **Personal Information Protection and Electronic Documents Act** (Canadian federal)
- Applies whenever you collect / use / disclose personal information for commercial purposes
- Governs: consent, purpose limitation, retention, security, individual access
- Even secondary data - if it identifies people, PIPEDA applies

::: {.notes}
Provincial equivalents exist too (Quebec Law 25, Alberta PIPA). This
is deep water for a Wk 1 session - keep it high-level. If students
want depth, point at the Privacy Commissioner's website. The key
takeaway: personal data is not neutral - it comes with legal
obligations.
:::

# Part 8 — Industry context {background-color="#7A003C"}

## What an MR analyst actually does

Typical workflow at a client-side Insights team:

1. **Brief in** (from marketing manager or product team)
2. **Scope the study** (design, sample, budget, timeline)
3. **Field** (in-house or via supplier)
4. **Analyse** (in R, SPSS, Q, or Excel-plus-tears)
5. **Report** (deck for stakeholders + one-page executive summary)
6. **Follow up** (dashboards, tracking, next research question)

::: {.notes}
This is where you help students visualise what employment looks like.
Most MR analyst job postings ask for 2-3 of: SQL, R or Python, one MR
platform (Qualtrics, Confirmit), presentation skills. Emphasise that
communication is often the differentiator between hire and no-hire.
:::

## Two industries in one profession

- **Client-side.** Insights function inside a brand — mainly consumers of MR + occasional in-house studies
- **Supplier-side.** MR agencies — design, field, analyse, and deliver studies for clients

Career paths differ. Client-side rewards domain expertise; supplier-side rewards methodological range.

::: {.notes}
Ask again which side students are (or aspire to be) on. Some are neither
- they work in marketing but not insights. Point out that these people
still benefit from MR literacy: they consume MR outputs, they hire MR
suppliers, they evaluate MR proposals.
:::

## Professional bodies (all free to join as students)

- **ESOMAR** (global, EU-centred): `www.esomar.org`
- **Insights Association** (US): `www.insightsassociation.org`
- **MRIA-ARIM** (Canada): `www.mria-arim.ca`

Codes of conduct, career resources, conferences, mentorship.

::: {.notes}
Encourage joining at least one - MRIA is the most Canada-relevant.
Student memberships are usually free. Local events at MRIA Toronto
chapter are worth attending.
:::

# Part 9 — For your project {background-color="#7A003C"}

## Team formation begins today

- Talk to 2-3 classmates during the wrap-up
- By end of Wk 2: submit preferred teammates via A2L form
- I'll finalise teams by Wk 3 — mixing background, industry, and R experience

If you don't have a preference, I'll assign you.

::: {.notes}
Explicitly encourage mixed teams - students without a network
disadvantage themselves in self-selection. Mention: teams that mix R
comfort levels do better than teams where everyone is R-scared. Same
for domain diversity.
:::

## What to think about for your project

- What **decision** would you like to inform?
- Whose decision is it? (Real or plausible client)
- What data would you need? Is secondary data available?
- Rough timeline of what you'll design vs. analyse

Don't pick a topic yet. Just start noticing decisions in the world that could use research.

::: {.notes}
Give one example: "you're browsing Amazon and see a new brand of
protein powder priced 30% below competitors - is that a viable
positioning?" That's a research question. Push students to notice
everyday marketing decisions.
:::

# Part 10 — Wrap-up & Week 2 preview {background-color="#7A003C"}

## What we did today

- Located marketing research in the marketing decision process
- Walked through the seven-step MR process
- Met Maritime Roast, our running case
- Installed R, RStudio, Quarto; rendered our first `.qmd`
- Kicked off the TCPS-2 CORE certification
- Started team formation

::: {.notes}
Recap is worth 60 seconds. It cements the arc of the session. Students
who checked out for parts of the class will re-engage here.
:::

## Next week: Secondary data and research designs

- **Concept:** where to find secondary data, when to trust it, choosing among exploratory / descriptive / causal designs
- **R lab:** data visualisation with ggplot2 (StatsCan coffee data)

Come prepared: R environment working, TCPS-2 CORE registration underway.

::: {.notes}
Ask if anyone has questions. Announce office hours. Give the room a
minute to network - team formation for the project starts now, so
they should be talking to each other. Leave the projector on the
Maritime Roast slide so they associate the room with the case.
:::

## References

Silver, L., Stevens, R., Wrenn, B., & Loudon, D. (2021). The essentials of marketing research. SAGE Publications.

## Thank you — see you next Wednesday

Questions? Reach me at **abbass44@mcmaster.ca** or book office hours.

::: {.notes}
End on a positive note. Repeat email address. Stay for 5-10 min for
individual questions - that's where you catch students who won't ask
in front of the class.
:::


