---
title: "Session 1 — Marketing Research Process & Getting Started with R"
subtitle: "M731 Week 1 · September 9, 2026 · Comprehensive Reading Notes"
author: "Shay Abbasi"
date: "2026-09-09"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    self-contained: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

# Introduction and Course Overview

## What This Course Is About

Marketing Research (M731) is a graduate course in systematic inquiry: how to design studies, collect data, apply statistical methods, and translate findings into decisions. The course runs for 13 weeks and follows a single running case — Maritime Roast Coffee Co., a fictional Halifax-based coffee subscription business — so that every analytic technique you learn is grounded in a concrete business problem rather than abstract exercises.

By the end of the course, you should be able to: commission a research study, evaluate a supplier's methodology, analyse survey data in R, and present findings to a decision-making audience. These are the skills that distinguish a marketing professional who *consumes* research from one who *understands and can produce it*.

## Course Logistics

The course meets on Wednesdays for three hours at RJC 5100 (Burlington campus). The instructor is Shay Abbasi (`abbass44@mcmaster.ca`); the teaching assistant is Saba Khorasani (`sabaakhr@mcmaster.ca`). Office hours are by appointment via Zoom. All course materials, assignments, and announcements are on Avenue to Learn under "BUSADMIN M731: Marketing Research."

### Assessment Structure

The course has no final exam. Assessment is distributed across two individual tests, two individual assignments, and a substantial group project:

| Component | Weight | Timing |
|---|---|---|
| Assignment 1 (individual) | 10% | Due Week 6 |
| Assignment 2 (individual) | 10% | Due Week 9 |
| Test 1 (Lectures 1–4) | 20% | Week 5 |
| Test 2 (Lectures 5–10) | 20% | Week 11 |
| Project proposal + questionnaire (group) | 10% | Due Weeks 4 & 9 |
| Written report (group) | 20% | Due Week 13 |
| Presentation (group) | 10% | Week 13 |

The group project runs in parallel with the weekly content. Groups of 5–6 students select a real B2C Canadian company and conduct a complete marketing research study: a scoping proposal, a questionnaire, a written analysis, and a live presentation. The project is not Maritime Roast — that company is the instructor's worked example throughout the term. You find your own client.

### Generative AI Policy

Generative AI tools (ChatGPT, Claude, Copilot, etc.) are permitted for learning course material, studying for tests, and understanding error messages. They are prohibited on graded submissions: tests, assignments, and any part of the group project (except as an explicitly disclosed learning aid). When in doubt, disclose.

### Missed Tests

The MBA program does not accept MSAF. If you miss a test, contact the instructor through DSSAO within five working days with documentation. An alternate test is available in Week 6 (for Test 1) and during the December exam window (for Test 2). Students with SAS accommodations should activate them through `DSBSAS@mcmaster.ca` at least two weeks before each test.

---

# What Is Marketing?

## The AMA Definition

Marketing has been defined many ways, but the American Marketing Association's classic formulation remains the clearest foundation:

> "Marketing is the process of planning and executing the conception, pricing, promotion, and distribution of ideas, goods, and services to create exchanges that satisfy individual and organizational needs." — AMA, 1985

Every word in this definition is load-bearing. "Planning and executing" means marketing is both strategy (what to make, at what price, for whom) and operations (getting it made, priced, distributed, and communicated). "Conception" — not production — signals that the marketing function begins at the idea stage, before engineering writes a specification. "Exchanges" is the core concept: marketing exists because two parties trade value, and it breaks down when one or both parties perceive no value in the trade. "Individual and organizational needs" reminds us that B2C and B2B marketing share the same conceptual logic even when their tactics differ enormously.

The AMA updated the definition in 2007 to emphasize a broader stakeholder view: "Marketing is the activity, set of institutions, and processes for creating, communicating, delivering, and exchanging offerings that have value for customers, clients, partners, and society at large." The 2007 version is more expansive but the 1985 four-P formulation remains more useful as a teaching structure.

## Marketing Management Philosophies

An organisation does not just practise marketing — it operates under an implicit or explicit philosophy about *what* it should produce and sell and for whom. Five philosophies describe this evolution:

**The Production Concept** holds that consumers prefer products that are widely available and inexpensive. Managerial attention focuses on production efficiency, cost reduction, and mass distribution. This made sense in eras of scarcity; it describes low-cost airline strategy (WestJet in its early years) and commodity manufacturing.

**The Product Concept** holds that consumers favour products offering the best quality, performance, and features. Firms operating under this philosophy invest heavily in R&D and product improvement. Apple's "technology + liberal arts" philosophy is a contemporary example — though Apple also understands market orientation deeply, making it a hybrid case.

**The Selling Concept** holds that consumers will not buy enough without aggressive selling and promotion. It is the philosophy behind door-to-door sales, telemarketing, and many insurance products. The selling concept treats the sale as the endpoint; customer satisfaction afterward is secondary.

**The Marketing Concept** reverses the orientation: start with customer needs, then design products and communications that satisfy those needs better than competitors do. Achieving organisational goals *depends* on satisfying customers. Procter & Gamble's insight-led R&D and Shopify's merchant-first strategy reflect this orientation.

**The Societal Marketing Concept** extends the marketing concept by requiring that the firm balance profit, consumer satisfaction, and society's long-run welfare. Patagonia's "Don't buy this jacket" campaign — which honestly communicated the environmental cost of a new garment purchase — is the canonical example.

In practice, most firms blend philosophies. A company can be Production-oriented in its factory and Marketing-oriented in its brand strategy. What matters for this course is recognising that the Marketing Concept is the philosophical home of marketing research: if decisions are supposed to start from customer needs, then *research into those needs* is not optional overhead — it is the prerequisite for sound strategy.

## From Marketing Concept to Market Orientation

Saying your firm follows the Marketing Concept is easy. Kohli and Jaworski (1990, *Journal of Marketing*) formalised what it looks like when a firm actually *does it*:

> "Market orientation is the organisation-wide generation of market intelligence pertaining to current and future customer needs, dissemination of intelligence across departments, and organisation-wide responsiveness to it."

Three components:

1. **Intelligence generation** — systematic collection of information about customer needs, competitive moves, and environmental trends. Surveys, focus groups, sales-force feedback, and secondary data are all generation mechanisms.
2. **Intelligence dissemination** — structured sharing of that information across functional boundaries. Market intelligence sitting in one department's dashboard that no one else sees is not market orientation; it is a report.
3. **Responsiveness** — actually changing products, prices, communications, or distribution in response to intelligence. Generation and dissemination without responsiveness are activity without outcomes.

Marketing research *operationalises* market orientation. An insights team that generates, disseminates, and drives action on customer intelligence is the institutional embodiment of the Marketing Concept.

---

# What Is Marketing Research?

## The Definition and Its Key Words

> "Marketing research is the systematic gathering, recording, and analysis of data about problems relating to the marketing of goods and services." — American Marketing Association

Three words anchor this definition:

**Systematic** — the most important word. Systematic means the study has a defined procedure that another researcher could replicate. It follows an agreed data-collection protocol (same questions, same order, same mode). It analyses data using pre-specified methods, not methods chosen after looking at results. It reports findings with stated confidence levels, not vague impressions. Systematicity is what separates marketing research from anecdote, intuition, and rumour.

To see why this matters, contrast two statements: "I spoke to 10 customers at the trade show and they loved the concept" versus "We surveyed 400 randomly-selected prospects from our Ontario prospect list and 73% rated the concept as 'likely to purchase' (95% CI: 68%–78%)." Both are data. Only the second is systematic — and only the second supports a decision to invest in a product launch.

**Data** — not gut feel, not narrative, not single cases. Data means measurements — numbers, categories, text — collected from a defined group under defined conditions.

**Problems** — marketing research exists to solve a business problem, not to satisfy curiosity. If a research request does not map to a decision, the research should not be commissioned. "What do customers think of us?" is curiosity. "Should we launch a premium tier at $54/month, and if so, which features should it include?" is a problem.

## Marketing Research versus Market Research

The terms are often used interchangeably. Technically, *market research* refers to the study of a specific market — its size, structure, competitors, and trends. *Marketing research* is broader — it encompasses any study that informs a marketing decision, including product research, pricing research, advertising testing, brand equity measurement, customer satisfaction tracking, and more. This course uses the broader "marketing research" throughout, but you will encounter "market research" in most job postings and industry conversations, where it means the same thing.

## Where Marketing Research Sits in an Organisation

Marketing researchers work in three main settings:

**Client-side** (also called "in-house" or "corporate side"): an insights or consumer intelligence team embedded within a brand. Client-side researchers commission studies from agencies, translate findings into recommendations for brand and product teams, build and maintain tracking dashboards, and represent the voice of the customer in strategy meetings. Examples: the insights teams at Loblaws, Bell Canada, or TD Bank.

**Supplier-side** (also called "agency side"): marketing research agencies that design, field, analyse, and deliver studies to clients. Supplier-side researchers see a wide range of categories and clients, giving them deep methodological expertise. Examples: Ipsos, Nielsen, Environics Analytics, or boutique qualitative agencies.

**Consulting**: strategy firms with embedded insights practices (McKinsey's consumer analytics teams, BCG's consumer and marketing analytics) where research is embedded in broader strategic assignments.

Both client-side and supplier-side roles require understanding the full research process. A client-side researcher who does not understand methodology cannot evaluate their supplier's work or push back on a flawed study design. A supplier-side researcher who does not understand business decisions cannot write a finding that anyone will act on.

---

# The Types of Marketing Research Questions

## The DDPP Framework

Every marketing research question falls into one of four types, which form a natural decision hierarchy:

**Descriptive questions** ask: "What is happening?" Descriptive research counts, profiles, and quantifies. How many subscribers do we have in the Discovery tier? What is the mean monthly spend per subscriber? What proportion of Ontario coffee drinkers have heard of Maritime Roast?

**Diagnostic questions** ask: "Why is it happening?" Diagnostic research explains a pattern or problem. Why did the Standard-tier renewal rate drop after last month's price increase? Why are Discovery subscribers cancelling within their first two months?

**Predictive questions** ask: "What will happen if we act?" Predictive research forecasts outcomes under hypothetical conditions. What is the price elasticity of demand for the Signature tier? If we add a decaf option, how many subscribers would upgrade?

**Prescriptive questions** ask: "What should we do?" Prescriptive research recommends a specific action. Which new roast should we add to the Discovery tier lineup? Which Ontario grocers should we approach first for retail placement?

These four types form a hierarchy: prescriptive recommendations without descriptive foundations are guesses dressed up as research. A client who asks "which market should we enter?" before doing the descriptive work — establishing market size, competitive structure, and customer needs — is asking an unanswerable question. A good researcher helps the client see that they need descriptive research first.

**A common failure mode**: many clients ask prescriptive questions before doing the descriptive work. Part of a researcher's job is to scope the question to the right level of the hierarchy before choosing a method.

---

# The Marketing Research Process

## A Seven-Step Overview

The MR process is sequential and iterative. Every session in this course corresponds to one or more steps:

1. Define the problem
2. Choose the research design
3. Design the sampling plan
4. Design the data-collection instrument
5. Collect the data
6. Analyse the data
7. Report and recommend

The process is sequential because errors propagate downstream and become increasingly expensive to fix. A poorly-worded questionnaire item (Step 4) produces systematically wrong data that cannot be fixed in analysis (Step 6). A biased sample (Step 3) cannot be corrected by better statistics (Step 6). Investing heavily in problem definition (Step 1) before committing to a method is the single most cost-effective thing a researcher can do.

## Step 1: Problem Definition — The Hard Step

The business problem is not the research problem. This distinction matters enormously.

A manager comes to you and says: "Our subscription churn is up 3 percentage points this quarter." That is the *business* problem — a bad outcome that someone needs to fix. The *research* problem is what you need to know in order to help them fix it: "Identify the leading drivers of 6-month churn among Maritime Roast subscribers, and quantify the relative impact of each driver, distinguishing price-sensitive from engagement-sensitive churners."

The research problem is specific, measurable, and answerable with data. The business problem is not.

The **Situation Brief** technique (Silver et al., 2021) provides a structured method for translating business problems into research questions:

| Stage | Content |
|---|---|
| Situation | What is happening in the market that prompted this request? |
| Decision | What specific decision will the research findings support? |
| Alternatives | What options are on the table? |
| Key uncertainties | What don't we know that would change the decision? |
| Research question | What specific data would resolve those uncertainties? |

Applied to Maritime Roast's churn problem: the situation is a 3 pp increase in 6-month churn; the decision is whether to invest in retention programs and which ones; the alternatives are price discounts, product improvements, better onboarding, and community features; the key uncertainties are whether churners are price-sensitive, dissatisfied with quality, or simply never-engaged; and the research question becomes "What factors most strongly predict 6-month churn, and which are actionable within our budget?"

## Step 2: Research Design

Three research design families are available, and the choice among them depends on the research question:

**Exploratory designs** are used when hypotheses are unclear and the researcher needs to understand a phenomenon before quantifying it. Methods include qualitative interviews, focus groups, ethnographic observation, secondary data scans, and expert interviews. The output is a refined research question and a set of testable hypotheses, not a definitive answer.

**Descriptive designs** are used when the question is known and the goal is to quantify or profile. Methods include cross-sectional surveys, panel surveys, observational studies, and sales/audit data. The output is frequencies, means, cross-tabs, trends, and profiles.

**Causal designs** are used when the goal is to establish that X causes Y, not merely that they co-vary. Methods include randomised experiments (A/B tests), field experiments, and quasi-experiments. The output is a causal estimate with quantified uncertainty.

Most commercial marketing research is descriptive. Causal designs are rarer but increasingly valued as organisations invest in experimentation capabilities.

## Steps 3–7: A Brief Preview

**Step 3 (Sampling)** involves three decisions: who is in the eligible population (the frame), how members of the frame are selected (probability or non-probability methods), and how many are needed (calculated from desired precision and effect size). Sample size is not "as big as budget allows" — it is justified by the analytic requirements.

**Step 4 (Instrument design)** produces the questionnaire, discussion guide, or observation protocol used to collect data. Instrument design is where most measurement error enters the research — bad question wording, leading questions, inappropriate scale types, and order effects all introduce bias that cannot be removed in analysis.

**Step 5 (Data collection)** fields the study. Trade-offs among modes (online, phone, in-person, mail) involve cost, speed, coverage, and response bias.

**Step 6 (Analysis)** is the six-week analytical block of this course (Sessions 6–11): descriptives, hypothesis testing, ANOVA, scale validation, factor analysis, experiments, and regression.

**Step 7 (Reporting)** delivers the findings to the decision-maker. The best analysis in the world is worthless if the recommendation does not land. This step is covered in Session 12.

---

# The Running Case: Maritime Roast Coffee Co.

## Company Background

Maritime Roast is a fictional Halifax-based direct-to-consumer specialty coffee subscription company, founded in 2019. It has approximately 15,000 subscribers and a small retail presence across Atlantic Canada through independent grocer partnerships. The company offers three subscription tiers:

- **Standard** ($24/month) — curated selection of house roasts
- **Signature** ($39/month) — single-origin and small-batch roasts
- **Discovery** ($54/month) — rotating experimental and specialty roasts

Maritime Roast is a subscription business, which means its core financial metric is Monthly Recurring Revenue (MRR), and its most dangerous operating problem is churn. In a subscription business, Customer Lifetime Value (CLV) = (monthly contribution margin) / (monthly churn rate). A 3 percentage-point increase in churn — from, say, 12% to 15% — cuts CLV by roughly 25%. This is why the company's research agenda is dominated by two themes: acquisition (entering the Ontario market) and retention (reducing 6-month churn).

## The Four Strategic Questions

Four decisions drive the research programme across all 13 sessions:

1. **Should Maritime Roast enter the Ontario market?** (Sessions 2–4)
2. **How can Maritime Roast reduce 6-month subscriber churn?** (Sessions 5, 11)
3. **Should they launch a Discovery premium tier?** (Session 10 — A/B test)
4. **How should they price subscriptions?** (Sessions 3, 6)

Every technique in the course is motivated by one of these decisions. When you encounter a new statistical method, the first question should always be: what business decision does this method enable?

## The Weekly Map

| Session | Topic | Maritime Roast application |
|---|---|---|
| 1 | Process intro | Define the Ontario-entry problem |
| 2 | Secondary data + designs | StatsCan coffee data for Ontario sizing |
| 3 | Sampling | Design a 400-person Ontario prospect survey |
| 4 | Survey design | Draft the Maritime Roast questionnaire |
| 5 | Qualitative | Focus group guide for churned subscribers |
| 6–7 | Hypothesis testing | Compare WTP and satisfaction across groups |
| 8 | Scales | Validate the brand-attachment battery |
| 9 | Factor analysis | Identify need-based segments |
| 10 | Experiments | A/B test of the Discovery tier upgrade offer |
| 11 | Regression | Model drivers of satisfaction and churn |
| 12 | Reporting | Build the Ontario-entry recommendation report |
| 13 | Presentations | Project presentations |

---

# Getting Started with R, RStudio, and Quarto

## Why R?

R is the lingua franca of academic statistics. Every major statistical method in this course — factor analysis, structural equation modelling, experimental design, mixed-effects models — has a mature, peer-reviewed R package. The marketing research package ecosystem is especially strong: `psych`, `lavaan`, `mlogit`, `cluster`, `broom`, `emmeans`, `conjoint`. Python is increasingly used in product analytics and machine learning contexts, and it is a legitimate choice for many data problems. We use R because the course textbook (Chapman & Feit, 2023) is R-centric and because the statistical packages for survey-based marketing research are stronger in R than in any other environment.

R also produces publication-quality graphics (via `ggplot2`) that embed directly into Quarto documents, and it is free — no institutional licence required after graduation.

## Installation

Install the following, in this order:

1. **R** from `https://cran.rstudio.com/` — the language and computational engine
2. **RStudio** from `https://posit.co/download/rstudio-desktop/` — the integrated development environment
3. **Quarto** from `https://quarto.org` — the document authoring system (separate from RStudio)
4. **TinyTeX** — a lightweight LaTeX installation for PDF rendering: run `quarto install tinytex` in a terminal after installing Quarto

## RStudio Anatomy

RStudio has four panes, each with a specific function:

- **Top-left (Editor)**: where you write `.qmd` and `.R` files. Think of this as your manuscript.
- **Bottom-left (Console)**: where R evaluates commands interactively. You can type here, or send code from the editor with Ctrl+Enter.
- **Top-right (Environment)**: shows every object (vector, data frame, model) currently in R's memory, with its type and size.
- **Bottom-right (Files / Plots / Packages / Help)**: a multi-tab pane for navigating files, viewing plots, managing packages, and reading documentation.

### Keyboard Shortcuts That Save Hours

| Shortcut | Action |
|---|---|
| `Ctrl+Enter` | Run current line or selection in console |
| `Ctrl+Shift+K` | Render (knit) the current `.qmd` |
| `Tab` | Autocomplete function names and column names |
| `?functionname` | Open the help page for a function |
| `Ctrl+R` | Search command history |
| `Ctrl+Shift+M` | Insert the pipe operator `|>` |

**The most important workflow habit**: always use RStudio Projects (`File → New Project`). A Project keeps your working directory, history, and R profile together. Never use `setwd()` in a script — it breaks portability. When you send a `.qmd` file to a collaborator, they should be able to render it by opening the Project, not by hunting for file paths.

## Anatomy of a Quarto Document

A Quarto document (`.qmd`) has three components:

**YAML header** — metadata and output format instructions, surrounded by `---` delimiters.

**Prose** — ordinary markdown text: headings with `#`, bold with `**`, italics with `*`, links with `[text](url)`.

**Code chunks** — R (or Python, Julia) code enclosed in triple backticks with `{r}` language identifier.

```yaml
---
title: "My first Quarto document"
format: html
---

# A section heading

Some prose explaining what follows.

```{r}
#| label: first-chunk
#| echo: true
#| eval: true
1 + 1
```
```

When you render this document (Ctrl+Shift+K), Quarto runs the R code, captures the output, and weaves it together with the prose into an HTML file. The output updates automatically when the data or code changes — this is the "living document" property that makes Quarto far more useful for research reporting than copy-pasting from Excel.

## Chunk Options — The Modern Way

Modern Quarto uses `#|` ("hash-pipe") options inside the chunk, rather than legacy knitr comma-separated options in the chunk header. The most important options:

```{r}
#| label: chunk-options-demo
#| echo: true        # show the code in the rendered output
#| eval: true        # actually run the code
#| warning: false    # suppress R warnings from appearing
#| message: false    # suppress package-loading messages
#| fig-width: 7      # figure width in inches
#| fig-height: 4.5   # figure height in inches
#| fig-cap: "Caption for the figure"
```

For course work, the most frequently used options are `echo`, `eval`, `warning`, `message`, and the `fig-*` options for any chunk that produces a plot.

## Installing the Course Packages

Run this once during the first lab session. It installs every package used across all 13 sessions:

```{r}
#| label: install-packages
#| eval: false
install.packages(c(
  "tidyverse",    # core data wrangling and visualisation
  "readr",        # fast CSV reading (part of tidyverse)
  "readxl",       # read Excel files
  "haven",        # read SPSS/Stata/SAS files
  "jsonlite",     # read JSON data
  "knitr",        # document rendering engine
  "gt",           # publication-quality tables
  "kableExtra",   # HTML/PDF table styling
  "janitor",      # column-name cleaning, cross-tabs
  "broom",        # tidy model output
  "renv",         # reproducibility: package version locking
  "psych",        # factor analysis, reliability, descriptives
  "car",          # Anova() Type II/III, vif(), leveneTest()
  "lmtest",       # Breusch-Pagan, Durbin-Watson tests
  "sandwich",     # robust (HC3) standard errors
  "effectsize",   # cohens_d(), eta_squared(), cramers_v()
  "pwr",          # sample size and power calculations
  "corrplot",     # correlation heatmaps
  "emmeans",      # estimated marginal means, post-hoc contrasts
  "lm.beta",      # standardised regression coefficients
  "tidytext",     # text tokenisation for qualitative analysis
  "MASS"          # mvrnorm() for multivariate simulation
))
```

This takes 5–10 minutes. If a package fails on Windows, the most common cause is antivirus software scanning the installation directory. Adding R's library path to the antivirus exclusion list usually fixes it.

## The Five Core dplyr Verbs

Almost all data manipulation in this course is built from five functions from the `dplyr` package (part of `tidyverse`):

```{r}
#| label: dplyr-verbs
#| eval: false
library(tidyverse)

# Load the Maritime Roast transaction data
tx <- read_csv("data/maritime_transactions.csv")

# 1. filter() — keep rows matching a condition
tx |> filter(amount > 30, tier == "Signature")

# 2. select() — keep specific columns
tx |> select(subscriber_id, tier, amount, transaction_date)

# 3. mutate() — add or modify columns
tx |> mutate(amount_with_tax = amount * 1.13)

# 4. arrange() — sort rows
tx |> arrange(desc(amount))   # largest first

# 5. summarise() + group_by() — aggregate by group
tx |>
  group_by(tier) |>
  summarise(
    n         = n(),
    total_rev = sum(amount),
    mean_amt  = mean(amount),
    med_amt   = median(amount)
  )
```

These five verbs, composed with the pipe operator `|>` ("and then"), answer almost any data question you will encounter. The pipe operator makes code read left-to-right: take `tx`, *and then* group by tier, *and then* summarise. This is more readable and less error-prone than nested function calls.

A useful pattern for proportion calculations — which appears constantly in marketing research summaries — exploits the fact that `TRUE` coerces to 1 and `FALSE` to 0:

```{r}
#| label: proportion-trick
#| eval: false
# What proportion of transactions are from Atlantic-region subscribers?
tx |>
  summarise(prop_atlantic = mean(region == "Atlantic"))
```

This returns the fraction of rows where `region == "Atlantic"` — a clean, one-line proportion calculation that generalises to any binary condition.

## Reproducibility with renv

Every group project submission must include a working `renv.lock` file. The `renv` package creates a project-level package library so that every collaborator runs the same package versions:

```{r}
#| label: renv-workflow
#| eval: false
renv::init()       # first time: create the project-level library
renv::snapshot()   # before submitting: lock all package versions
renv::restore()    # collaborator opens the project: restore everything
```

Without `renv`, a project that works on one machine may fail on another because package updates break function arguments or change default behaviours. This is a common cause of last-minute submission failures that `renv` eliminates entirely.

## Diagnosing Render Errors

When a Quarto document fails to render, the error message appears in the Render tab at the bottom of RStudio. The most common first-render failures and their fixes:

| Error | Likely cause | Fix |
|---|---|---|
| `there is no package called 'X'` | Package not installed | `install.packages("X")` |
| `object 'X' not found` | Chunk runs before the object is created | Check chunk order; restart R and run all chunks |
| `YAML parse error` | Bad indentation or missing quote in the header | Inspect the YAML header carefully |
| Blank output / all literal text | Mismatched backtick fences | Verify outer fences have more backticks than inner fences |
| `cannot find input file` | File path wrong | Use relative paths from the `.qmd` file location |

When in doubt: restart R (`Session → Restart R and Run All Chunks`) before concluding there is an error. A stale R environment from a previous session is the source of roughly a third of all apparent render errors.

---

# Research Ethics in Marketing

## Why Ethics Is a Course Thread, Not a Single Lecture

Ethics is integrated into every session of this course — 10–15 minutes per session — because ethics in research practice is not a topic you visit once and then set aside. It is a way of doing research. When a sampling concern arises in Session 3, we discuss it there. When p-hacking comes up in Session 6, we discuss it there. Graduate researchers should build ethical reflexes, not just ethical knowledge.

## TCPS-2 CORE Certification

The Tri-Council Policy Statement, 2nd Edition (TCPS-2) is Canada's national ethics framework for research involving human participants. Every marketing researcher who touches human data in Canada should be familiar with it. The CORE online tutorial (`https://tcps2core.ca`) takes approximately 2 hours, is free, and produces a certificate. Complete it by the end of Week 4 and upload the certificate to Avenue to Learn.

The three core principles of TCPS-2 are:

1. **Respect for Persons** — participants have the right to make their own informed decisions about whether to participate. Informed consent is the operational manifestation of this principle: respondents must understand what data is being collected, for what purpose, by whom, and stored how — before they participate.

2. **Concern for Welfare** — research should minimise harms and maximise benefits. In the marketing context, this means protecting privacy, not over-collecting personal information, and treating respondents' time with respect (not designing 45-minute surveys and calling them 10-minute surveys).

3. **Justice** — the benefits and burdens of research should be distributed fairly. Vulnerable populations should not bear disproportionate burdens of research participation.

## PIPEDA in a Marketing Research Context

The *Personal Information Protection and Electronic Documents Act* (PIPEDA) is Canada's federal privacy law for commercial organisations. It applies whenever a company collects, uses, or discloses personal information for commercial purposes. Its 10 fair information principles translate directly into survey research obligations:

| Principle | What it means for survey research |
|---|---|
| Accountability | Someone in your org is responsible for PIPEDA compliance |
| Identifying purposes | Tell respondents *before* collection why you're asking |
| Consent | Explicit (opt-in) preferred |
| Limiting collection | Collect only what you need for the stated purpose |
| Limiting use | Don't use data for secondary purposes without new consent |
| Accuracy | Keep data current; correct errors |
| Safeguards | Encrypt, restrict access, secure storage |
| Openness | Privacy policy must be accessible |
| Individual access | Respondents can request to see their data |
| Challenging compliance | A process for privacy complaints must exist |

The "limiting collection" principle is the most commonly violated in commercial surveys — stakeholders add questions "just in case." PIPEDA provides a principled reason to push back: "We cannot collect this data without a specific stated purpose." The practical rule before adding any survey question: "Do I actually need this? How will I use it?" If you cannot answer both questions concretely, remove the question.

---

# The Marketing Research Industry

## What an MR Analyst Actually Does

A typical workflow for a client-side insights analyst:

1. **Brief in** — receive a request from a marketing manager or product team describing a decision they face
2. **Scope the study** — translate the brief into a research design, sampling plan, budget, and timeline
3. **Field** — collect data (in-house or via a supplier agency)
4. **Analyse** — in R, SPSS, Q, or Excel, depending on the organisation
5. **Report** — produce a deck for stakeholders and a one-page executive summary
6. **Follow up** — dashboards, tracking, and the next research question

The research brief is the critical document that initiates this workflow. A strong brief contains: background (what is the business context?), objectives (what specific questions must the research answer?), audience (who will use these findings, for what decision?), scope (what is in and out of scope?), budget and timing (hard constraints), and deliverable format (report, dashboard, presentation). Bad briefs produce bad research — not because the researcher is incompetent, but because without clarity on objectives and decision context, the wrong questions get answered rigorously.

## Professional Bodies

Three professional associations are worth joining as a student (all free or low-cost for student membership):

- **MRIA-ARIM** (Marketing Research and Intelligence Association — Canada): `www.mria-arim.ca` — the most Canada-relevant; local chapter events in Toronto
- **ESOMAR** (global, EU-centred): `www.esomar.org` — international MR standards, conferences
- **Insights Association** (US): `www.insightsassociation.org` — strong on professional development resources

---

# The Group Project

## Finding a Good Topic

The group project asks you to conduct a marketing research study for a real (or realistic) B2C Canadian company. Strong topics satisfy four criteria:

1. **A genuine business decision** — not "study the market," but "Company X is deciding whether to do Y."
2. **A real target audience** whose preferences or behaviour you can study.
3. **Secondary data availability** to frame the market context (StatsCan, industry reports).
4. **Operationalisable constructs** — the key measurements fit in a questionnaire of 12 minutes or less.

Topics that fail the filter: historical analysis (no decision), purely theoretical questions (no client), topics requiring REB-approved primary research on sensitive populations.

A strong example: "Should a Canadian DTC skincare brand enter the Quebec market? Characterise Québécois consumer preferences for sustainable beauty products."

Teams of 5–6 are formed by the end of Week 2. If you do not submit a preference, the instructor will assign you to a team. The strongest teams mix R comfort levels and industry backgrounds.

---

# References

Chapman, C., & Feit, E. M. (2023). *R for marketing research and analytics* (2nd ed.). Springer.

Cohen, J. (1988). *Statistical power analysis for the behavioral sciences* (2nd ed.). Lawrence Erlbaum.

Kohli, A. K., & Jaworski, B. J. (1990). Market orientation: The construct, research propositions, and managerial implications. *Journal of Marketing, 54*(2), 1–18.

Kotler, P., & Keller, K. L. (2016). *Marketing management* (15th ed.). Pearson.

Krosnick, J. A., & Presser, S. (2010). Question and questionnaire design. In P. V. Marsden & J. D. Wright (Eds.), *Handbook of survey research* (2nd ed., pp. 263–314). Emerald Group.

Silver, L., Stevens, R., Wrenn, B., & Loudon, D. (2021). *The essentials of marketing research*. SAGE Publications.
