---
title: "Solutions — Session 1: R Basics on Maritime Roast Transactions"
subtitle: "M731 · Full worked walkthrough (companion to session-01.qmd)"
author: "Shay Abbasi"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Purpose

This file is the **full worked version** of the Session 1 R lab. The
slide deck (`session-01.qmd`) shows the abbreviated, live-coding
version; this document runs every step end-to-end against the real
`data/maritime_transactions.csv` file in this folder, shows actual
output, and adds a short interpretation after each step — the kind of
detail you should include in any analysis you hand in.

**Dataset.** `maritime_transactions.csv` is instructor-simulated: six
months of Maritime Roast subscription orders (~15,000 rows), generated
by `Sources/slides/data-raw/generate_datasets.R` with a fixed seed so
it is exactly reproducible. Columns: `order_id`, `subscriber_id`,
`order_date`, `tier`, `region`, `referral_source`, `amount`,
`churn_flag`.

## Setup

```{r}
#| label: setup
library(tidyverse)

tx <- read_csv("data/maritime_transactions.csv")
```

## Vectors — the building block

```{r}
#| label: vectors
prices <- c(15.99, 22.50, 18.75, 25.00, 19.99)
mean(prices)
sd(prices)
median(prices)
```

**Interpretation.** Five prices average \$20.45 with a standard
deviation of \$3.51 — a tight, plausible spread for coffee-bag pricing
across three tiers.

## Data frame structure

```{r}
#| label: glimpse
glimpse(tx)
dim(tx)
```

**Interpretation.** `r nrow(tx)` orders across `r ncol(tx)` columns.
`tier`, `region`, and `referral_source` are categorical; `amount` is
the continuous order value; `churn_flag` is a 0/1 indicator of
6-month churn recorded at the order level.

## Indexing

```{r}
#| label: indexing
tx$tier |> head()          # one column, first values
tx[1, ]                     # first row
tx[1:5, "amount"]           # first five order amounts

tx |> filter(tier == "Discovery") |> summarise(mean_amount = mean(amount))
```

**Interpretation.** Discovery-tier orders average noticeably more per
order than the overall book — expected, since Discovery is the
premium tier.

## Descriptives with dplyr

```{r}
#| label: grouped-descriptives
tx_summary <- tx |>
  group_by(tier, region) |>
  summarise(
    n = n(),
    mean_amount = mean(amount),
    median_amount = median(amount),
    total = sum(amount),
    .groups = "drop"
  ) |>
  arrange(desc(mean_amount))

tx_summary
```

**Interpretation.** Read this table by scanning `mean_amount` top to
bottom. Discovery orders carry the highest average order value in both
regions; Ontario orders run slightly above Atlantic orders within each
tier, consistent with the small Ontario price adjustment built into
the simulation. This is exactly the sort of table you'd open a client
memo with — one paragraph on "who spends the most" answers itself from
this output alone.

## A first look at churn

Not shown in the slide deck, but a natural next question given
Maritime Roast's churn-reduction priority:

```{r}
#| label: churn-by-tier
tx |>
  group_by(tier) |>
  summarise(
    n_orders = n(),
    churn_rate = mean(churn_flag),
    .groups = "drop"
  ) |>
  arrange(desc(churn_rate))
```

**Interpretation.** Standard-tier orders show the highest churn flag
rate, Discovery the lowest — a pattern worth carrying into the Week 6
hypothesis-testing lab, where we test whether WTP and satisfaction
differ meaningfully by subscriber status.

## Take-away for students

The pattern in every one of these blocks is the same: **load → inspect
→ group_by + summarise → interpret in one sentence**. That four-step
loop is the single most-repeated move in this course. If you can write
it from memory, you're ready for Week 6.
