---
title: "Session 5 — Qualitative Marketing Research"
subtitle: "M731 Week 5 · October 7, 2026 · Comprehensive Reading Notes"
author: "Shay Abbasi"
date: "2026-10-07"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    self-contained: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

# Overview

Session 5 is the qualitative research session. It covers the theory, methods, and practice of qualitative inquiry in marketing research — why qualitative methods exist, what they can and cannot do, and how to design and conduct major qualitative approaches: focus groups, in-depth interviews, projective techniques, and observational methods.

A note on the session structure: Test 1 (covering Sessions 1–4) takes place in class on October 7. This reading document covers the qualitative content from the second half of Session 5 in depth, since qualitative methods receive only a fraction of the classroom time a topic of this scope warrants. Read it before class to prepare for the focus-group workshop that follows the test.

---

# Qualitative versus Quantitative: Not a Hierarchy

## The Persistent Misconception

A deeply embedded bias in MBA and business programmes treats qualitative research as the "soft" option — exploratory, informal, suitable for forming ideas before "real" (quantitative) research begins. This mischaracterises both types of research and leads to poor study design decisions.

The correct framing is that qualitative and quantitative methods answer different kinds of questions:

| Dimension | Qualitative | Quantitative |
|---|---|---|
| Sample size | Small (8–30 typically) | Large (100–2,000+) |
| Data type | Text, images, observation, video | Numbers, ratings, counts |
| Inference | Interpretive; meaning-generating | Statistical; magnitude-estimating |
| Appropriate for | Understanding why, how, and what it means | Estimating how many, how much, and whether X causes Y |
| Rigour | High when done well; low when done poorly | High when done well; low when done poorly |

Both methods can be rigorous or sloppy. The academic marketing literature (Journal of Consumer Research, Journal of Marketing Research) publishes high-quality ethnographic, phenomenological, and grounded-theory work alongside quantitative experiments and structural equation models. The choice between them should be driven by the research question, not by a preference hierarchy.

## When Qualitative Is the Right Choice

Qualitative research is appropriate when:

- The phenomenon is poorly understood and the researcher needs to discover *what questions to ask* before designing a quantitative study
- The goal is to understand motivations, feelings, meanings, and social processes that surveys cannot capture
- The population is rare, vulnerable, or difficult to reach through standard panel methods
- Concept development is needed — new product ideas, new positioning, new communications — where consumer reactions require depth rather than frequency counts
- A theoretical or explanatory framework needs to be built from the data rather than tested against it

Qualitative research is *not* appropriate for quantifying, generalising to a population, or making causal claims. A focus group of 8 participants cannot tell you what proportion of Ontario coffee drinkers would subscribe to Maritime Roast. A series of depth interviews cannot prove that better onboarding causes lower churn. These require quantitative methods.

## Mixed-Methods Research Design

The most powerful research programmes combine both approaches. Creswell and Plano Clark (2018) identify three main mixed-methods architectures:

**Exploratory sequential design**: qualitative phase first to discover themes and generate hypotheses, followed by a quantitative phase to test those hypotheses at scale. This is the most common commercial MR sequence: qualitative to develop the survey instrument, quantitative to measure prevalence.

For Maritime Roast's churn problem: (1) 12 in-depth interviews with churned subscribers to discover the top 5 reasons people leave; (2) a survey of 500 current and lapsed subscribers to quantify how often each reason applies and which predicts churn most strongly.

**Concurrent triangulation design**: qualitative and quantitative data collected simultaneously, with findings cross-validated. Quantitative patterns are explained and contextualised by qualitative insights.

**Embedded design**: a primary method (usually quantitative) with qualitative data embedded within — for example, a large-scale survey with open-ended questions analysed thematically alongside the closed-ended scale data.

---

# Focus Groups

## What Focus Groups Are and Are Not

A focus group is a moderated discussion among 6–10 participants selected to represent the target population, lasting 60–90 minutes. The key distinguishing feature is the group dynamic: participants react to and build on each other's comments, surfacing reactions that would not emerge in a one-on-one interview.

Focus groups are not mini-surveys. The goal is not to count how many of 8 participants hold a view, but to understand the range of views, the reasoning behind them, and how they interact socially. The analytical output is themes and hypotheses, not percentages.

## Strengths and Weaknesses

**Strengths**: the group dynamic can surface social norms (how people talk about a topic in public) and reveal when one person's view shifts another's. They are cost-effective relative to in-depth interviews: 8 participants in 90 minutes yields more qualitative insight than 8 separate 15-minute phone calls. Client observation (from behind a one-way mirror, or remotely via video platform) is feasible.

**Weaknesses**: dominant participants can skew the discussion. Social desirability effects are stronger in group settings than in private self-administered research (though weaker than in interviewer-administered surveys). They are poorly suited to sensitive topics — discussing financial anxiety, relationship difficulties, or health conditions in front of 7 strangers inhibits honest disclosure.

## The Art of Focus Group Moderation

A focus group is only as good as its moderator. The three core moderating skills are:

**Probing**: moving from surface-level answers to depth. Effective probes include:

- *Clarifying probes*: "Can you say more about what you mean by 'convenient'?"
- *Elaborating probes*: "You mentioned the packaging — what specifically about it bothered you?"
- *Contrast probes*: "You said you wouldn't pay more — what if the quality were noticeably better?"
- *Third-party probes*: "Why do you think other people might feel that way?" (This lets respondents express genuine views while attributing them to a hypothetical third party — a useful technique for socially sensitive topics.)

**Neutrality**: moderators must not signal agreement or disagreement with participants. Affirmations ("That's great!" "I totally agree") bias the group toward that view. Neutral acknowledgements ("I see," "That's interesting," "Mm-hmm") keep the discussion flowing without directing it.

**Managing group dynamics**: quieter participants should be gently invited in ("We haven't heard from you yet — what's your take?"). Dominant participants should be redirected without embarrassing them ("Good point — let's hear what others think"). Conflict between participants should be surfaced as a research opportunity, not resolved by the moderator.

---

# In-Depth Interviews (IDIs)

## IDIs versus Focus Groups

An in-depth interview (IDI) is a one-on-one conversation between a researcher and a single participant, typically lasting 45–90 minutes. IDIs eliminate group-dynamic effects — the participant speaks freely without social pressure from other group members. This makes them better for:

- Sensitive topics (financial stress, health behaviour, relationship dynamics)
- B2B research with senior professionals who will not participate in groups with their peers or competitors
- Topics requiring detailed personal narrative (life history, career trajectory, long purchase journeys)
- Situations where the researcher suspects social desirability would strongly suppress honest group discussion

The tradeoff: IDIs are more expensive per participant (typically $1,000–$2,500 per interview including recruitment, incentive, and analysis time) and produce more data per participant that must be individually analysed.

## The Funnel Interview Structure

IDIs follow a funnel structure, moving from broad and general to specific and focused:

**Opening (5–10 minutes)**: build rapport; broad life-stage questions that have nothing to do with the research topic. "Tell me about how you typically start your morning" or "Walk me through your coffee routine on a typical workday." This serves two purposes: establishing comfort and obtaining unaided associations before priming the participant with the research topic.

**Middle (30–50 minutes)**: navigate gradually toward the research topic. For Maritime Roast churn research, this might move from "Tell me about your experience with coffee subscriptions generally" to "You mentioned you cancelled a subscription — walk me through that experience" to "What was the moment you first started thinking about cancelling?" The key moderating technique here is the **follow-up probe** on specific words the participant uses: "You used the word 'disappointed' — tell me more about what disappointed you."

**Projective exercise (10–15 minutes)**: embedded near the middle or toward the end (once rapport is established), a projective technique often surfaces motivational content that direct questioning cannot reach.

**Close (5–10 minutes)**: open invitation to add anything the guide did not cover ("Is there anything I haven't asked that you think is important?") and a closing question that gives the participant a sense of agency ("If you could give the CEO of Maritime Roast one piece of advice, what would it be?").

---

# Projective Techniques

## Why Direct Questioning Has Limits

Direct questions about brand perceptions, motivations, and purchase drivers are answered by the respondent's conscious, rational System 2 mind (Kahneman, 2011). But a substantial proportion of consumer cognition occurs below conscious awareness — in the fast, intuitive, emotionally-driven System 1 system. Direct questions bypass System 1 entirely.

Projective techniques are designed to bypass the rational filter and access underlying motivations that respondents cannot or will not articulate directly. They typically involve ambiguous stimuli (incomplete sentences, images, stories) that the respondent completes or interprets — and the completion or interpretation reveals underlying structures of meaning.

## Common Projective Techniques

**Word association**: present a brand name or category word and ask for the first word that comes to mind. Fast, simple, and surprisingly revealing. "When I say 'Maritime Roast,' what's the first word that comes to mind?" Fast-moving word associations reveal automatic, affect-laden responses.

**Sentence completion**: present an incomplete sentence and ask the participant to complete it without pausing to think. "People who drink Maritime Roast are ___." "I signed up for a coffee subscription because ___." "A month after cancelling, I felt ___."

**Third-person projection**: "Your best friend has just cancelled their coffee subscription. What would they say about why they cancelled?" This allows respondents to attribute socially inconvenient motivations (price sensitivity, perceived pretentiousness of the product) to a hypothetical third party rather than admitting them directly.

**Brand personification**: "If Maritime Roast were a person, what would they be like? How old? What do they do on weekends? What kind of car do they drive?" Personification rapidly reveals the brand archetype (is Maritime Roast a sophisticated urban professional? a rugged outdoorsman? a friendly farmer's market vendor?) and reveals misalignments between intended positioning and consumer perception.

**Photo elicitation and visual collage**: participants select images from a set (or from magazines, or social media) that represent their feelings about the brand or category. Discussed in moderation, the images often reveal emotional associations that participants cannot put into words.

## ZMET: The Zaltman Metaphor Elicitation Technique

The most elaborate and theoretically grounded projective method is the Zaltman Metaphor Elicitation Technique (ZMET), developed by Gerald Zaltman at Harvard Business School (Zaltman, 2003). ZMET is based on Zaltman's claim that 95% of consumer cognition is non-conscious, and that metaphors are the primary vehicle through which non-conscious thought becomes accessible to language.

**The ZMET protocol**: participants spend 1–2 weeks before the interview collecting 8–12 images that capture their feelings about the brand or topic (using magazines, photographs, the internet). In the interview, the moderator uses a laddering technique ("Why did you choose this image? What does it represent? How does that connect to the brand?") to explore the imagery. Over many interviews, recurring imagery is categorised into **deep metaphors** — universal, unconscious structures that organise consumer thought.

Zaltman identified seven deep metaphors: Balance, Transformation, Journey, Container, Connection, Resource, and Control. For a coffee subscription, "Journey" might manifest in images of travel, exploration, and discovery. "Connection" might appear in images of community, warmth, and belonging. These metaphorical structures then inform positioning and communications strategy.

ZMET is patented by Harvard and requires licensed practitioners for commercial use. Its underlying principle — that images and metaphors access deeper emotional content than words alone — is broadly applied in adapted forms across the qualitative industry.

---

# Observational and Ethnographic Methods

## The Core Problem: Stated versus Revealed Behaviour

One of the most robust findings in consumer research is the gap between what people say they do and what they actually do:

- People say they recycle; behavioural tracking shows much lower rates
- People say they read nutrition labels; eye-tracking shows most shoppers look at the product, not the label
- Coffee drinkers say they would pay a premium for certified sustainable beans; revealed WTP in actual purchase is typically 20–40% below stated WTP

Survey research captures **stated behaviour** and **stated intentions**. Observational research captures **revealed behaviour** — what people actually do when they are not reporting to a researcher. Both are important; neither alone is complete.

## Observational Methods

**Non-participant observation**: the researcher watches participants in their natural environment without interacting with them. Shop-along studies observe shopper behaviour in a retail environment; home visits observe product use in the home context; video-mining of recorded point-of-sale footage quantifies shelf interaction without requiring any participant awareness.

**Participant observation / ethnography**: the researcher immerses in the participant's social world and participates in their activities. This yields rich contextual understanding of how products and brands are embedded in everyday life — how coffee drinking is embedded in morning routines, social rituals, and identity expression — that no survey can capture.

**Netnography** (Kozinets, 2020): systematic observation and analysis of online communities — Reddit forums, brand communities, review platforms, social media — to understand how consumers discuss a product category or brand. Netnography applies ethnographic analytical methods (interpretation, coding, saturation) to digital community data.

## Digital Qualitative Platforms

Covid-19 accelerated the adoption of digital qualitative methods, which are now standard:

- **Online focus groups**: Discuss.io, Recollective, Qualtrics Panels — video-based moderated group discussions, often asynchronous (participants respond on their own schedule over 3–5 days)
- **Mobile ethnography**: dScout, Indeemo — participants use smartphone apps to capture photos, video, and diary entries in their everyday environment
- **Community platforms**: Alida, Verto Analytics, Qualtrics Communities — online communities of 50–500 participants who engage in ongoing discussion over weeks or months

Digital qualitative methods are faster and cheaper than in-person alternatives and enable geographic diversity. They sacrifice some depth of in-person observation and group dynamics.

---

# Analysing Qualitative Data

## Thematic Analysis: A Step-by-Step Workflow

Thematic analysis (Braun & Clarke, 2006) is the most widely used approach to qualitative data analysis in marketing research. It is applicable across all qualitative data types (interview transcripts, focus group recordings, open-ended survey responses, observational field notes) and does not require a specific epistemological commitment.

**Phase 1 — Familiarise**: read all transcripts multiple times, making notes of initial ideas. The goal is immersion in the data before beginning formal coding. This phase is often skipped by time-pressured researchers; skipping it produces thinner analysis.

**Phase 2 — Generate initial codes**: systematically label meaningful segments of text with short descriptive codes. "price sensitivity," "service failure," "habit disruption," "product fatigue," "comparison shopping." Codes should be grounded in the data, not imported from theory (unless you are doing deductive thematic analysis, which starts from theory and codes against it).

**Phase 3 — Search for themes**: organise initial codes into potential themes. Multiple codes may converge on a single theme: "cancelled because too expensive," "couldn't justify the cost," and "found a cheaper alternative" might all converge on the theme "price-driven churn." Look for overarching patterns across the code list.

**Phase 4 — Review themes**: check candidate themes against the actual data. Do the themes hold when you re-read the transcripts? Are there segments that don't fit any theme (these may require a new theme or a revision of existing ones)? Are there themes that look coherent in the abstract but fall apart when you examine the quotes? This phase is the quality-control step that distinguishes rigorous thematic analysis from impressionistic summarisation.

**Phase 5 — Define and name themes**: write a clear, specific definition of what each theme captures (and what it does not capture). The theme name should convey its essence to someone who has not read the transcripts.

**Phase 6 — Write up**: produce the analysis with representative quotes as evidence and interpretive narrative as explanation. Quotes without interpretation is journalism; interpretation without quotes is assertion.

## Thematic Saturation: When Is Enough Enough?

The qualitative equivalent of sample size is **thematic saturation** — the point at which new data (interviews, observation sessions) add no new themes to the analysis. Guest, Bunce, and Johnson (2006) tested saturation empirically by examining when all major themes had appeared across a sequence of interviews:

- After 6 interviews: 92% of all themes that eventually emerged were already present
- After 12 interviews: 94% were present
- The remaining 6% emerged across interviews 13–60

**Practical implication**: for a focused research question with a reasonably homogeneous population, 8–12 IDIs typically reach saturation. For broader questions or heterogeneous populations, more are needed. Saturation is determined empirically by tracking themes across interviews (keep a running thematic log and note when the log stops growing) rather than by a predetermined n.

## Quotes and Inference: Getting the Balance Right

Qualitative reports should contain both quotes and interpretation. The right balance is: quotes as evidence, interpretation as synthesis.

A quote alone ("I cancelled because it just wasn't worth it anymore") requires interpretation to become a finding. Why wasn't it worth it? Worth it relative to what alternative? At what price point? The interpretation draws on multiple quotes from multiple participants to construct a finding that goes beyond any individual respondent's experience.

Interpretation without quotes, on the other hand, is just assertion. A finding that "subscribers cancel primarily because of perceived value decline" needs at least 3–5 quotes from different participants illustrating different facets of that value decline experience.

---

# R for Qualitative Research: Text Analytics

R is not the primary tool for qualitative analysis — CAQDAS (Computer-Aided Qualitative Data Analysis Software) packages like NVivo, ATLAS.ti, and Dedoose are better suited for the systematic coding and theme development that qualitative analysis requires. However, R is useful for supplementary text analytics on qualitative data, particularly when the corpus is large enough to warrant automated analysis.

## Basic Word Frequency Analysis

```{r}
#| label: word-frequency
#| eval: false
library(readtext)
library(tidytext)
library(tidyverse)

# Load focus group transcripts
transcripts <- readtext("data/maritime_focus.txt")

# Tokenise and remove stop words
word_counts <- transcripts |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  count(word, sort = TRUE)

# Visualise top 20 words
word_counts |>
  head(20) |>
  ggplot(aes(x = fct_reorder(word, n), y = n)) +
  geom_col(fill = "#7A003C") +
  coord_flip() +
  labs(
    title = "Most frequent words in focus group transcripts",
    x     = NULL,
    y     = "Count",
    caption = "Stop words removed"
  ) +
  theme_minimal()
```

## Sentiment Analysis

```{r}
#| label: sentiment-analysis
#| eval: false
# Sentiment using AFINN lexicon (-5 very negative, +5 very positive)
transcripts |>
  unnest_tokens(word, text) |>
  inner_join(get_sentiments("afinn"), by = "word") |>
  summarise(
    total_sentiment = sum(value),
    mean_sentiment  = mean(value),
    n_words         = n()
  )
```

**What automated text analytics adds**: fast frequency analysis across large corpora; automated sentiment scoring that is reproducible and consistent; pattern detection that may surface non-obvious themes.

**What it cannot replace**: nuanced interpretation of context, irony, and implicit meaning. "The coffee is not bad" receives negative sentiment from the AFINN lexicon (because "not" + "bad") but is clearly a positive evaluation from the speaker. "Sharp" is positive for attention and negative for pain — the lexicon has no way to distinguish. "Acidic" is negative in ordinary language but highly positive among specialty coffee enthusiasts. Domain-specific language and context-dependence limit the validity of automated sentiment analysis for qualitative MR data.

The appropriate role for text analytics in qualitative research: supplementary pattern detection that guides the human analyst toward promising areas, not a replacement for systematic human coding.

---

# Designing a Focus Group Discussion Guide

## The Guide Structure

A focus group discussion guide is not a survey — it is a structured but flexible conversation plan. It should specify the sequence of major topics and the key questions within each topic, but moderators should adapt in response to what participants say rather than rigidly executing a script.

A well-designed guide for a 90-minute Maritime Roast churn focus group:

**Introduction (5 minutes)**

The moderator reads a scripted introduction: "Welcome, and thank you for joining us. We're here to understand experiences with coffee subscriptions. There are no right or wrong answers — we want your honest opinions. Please feel free to disagree with each other or build on what others say. And one more thing — this discussion is being recorded for research purposes. All responses will be kept confidential and used only in aggregate. Any questions before we begin?"

**Warm-up (5 minutes)**

1. "Let's go around quickly — tell me your name and one word to describe your relationship with coffee."
2. "When did you first start drinking specialty coffee, and what drew you to it?"

**Section 1: Subscription experience (15 minutes)**

3. "Tell me about your experience with [Maritime Roast] — from when you first signed up to when you decided to cancel. Walk me through it in your own words."
4. (Probe): "What was the moment you first started thinking about cancelling?"
5. (Probe): "What finally made you decide to cancel rather than continue?"

**Section 2: Value and alternatives (10 minutes)**

6. "After you cancelled, what did you do instead for coffee? Where did you go?"
7. "What would Maritime Roast have needed to do differently to keep you as a subscriber?"

**Section 3: Brand relationship (10 minutes)**

8. "How would you describe Maritime Roast as a brand? What words or images come to mind?"
9. (Probe): "Did your feelings toward the brand change over the time you were a subscriber?"

**Projective exercise (10 minutes)**

10. "I'd like to try something a bit different. Complete this sentence in whatever way feels right to you: 'My relationship with Maritime Roast was like ___ because ___.' Take a minute to think, and there's no right answer."

OR: "If Maritime Roast were a person, what would they be like? How old? What do they do? What kind of person are they?"

**Wrap-up (5 minutes)**

11. "If you could give the CEO of Maritime Roast one piece of advice about keeping subscribers, what would it be?"
12. "Is there anything we haven't talked about today that you think is important for us to understand?"

## Discussion-Guide Principles

Several principles distinguish a well-designed discussion guide from a list of questions:

**Broad to narrow**: each section starts with an open, non-leading question and narrows through probes. Never start with the most sensitive or specific question.

**Explicit probe scaffolding**: write probes for each main question. Probes ensure that moderators obtain depth even if the participant's initial response is brief. "What makes you say that?" and "Can you give me an example?" are generic probes that work everywhere.

**Strategic projective placement**: projective exercises work best after rapport is established (not at the beginning) and before participants are fatigued (not at the very end). The midpoint of a 90-minute session is typically ideal.

**Genuine closing questions**: the wrap-up should include both a targeted question (CEO advice — which gives participants agency and often yields candid recommendations) and an open invitation (anything else — which consistently surfaces information the guide missed).

---

# Qualitative Ethics

## Participant Vulnerability in Group Settings

Focus groups create specific ethical risks that individual surveys do not. The group dynamic can pressure participants to disclose more than they intended. Socially dominant participants may encourage others to share embarrassing or sensitive information in ways that feel coercive. Sensitive research topics (financial stress, health, relationship difficulties, grief) should use IDIs rather than focus groups, precisely because the one-on-one setting is less coercive.

## Anonymisation in Qualitative Reports

Survey data anonymisation is procedural — remove identifying variables and the data is anonymous. Qualitative anonymisation is harder because the quote itself can identify. A participant's distinctive phrasing, their specific job title, or a unique experience ("I was the only subscriber in my building") can be sufficient for identification by people who know them.

Practical anonymisation techniques:

- Remove or alter specific identifying details in quotes (job title, company name, specific location)
- Use composite vignettes: combine experiences from multiple participants into one illustrative case (disclose that this is a composite)
- Member-check: share draft quotes with participants to confirm they are comfortable with how they have been represented
- Never use actual names in any report

## Consent for Group Discussion and Recording

Every focus group participant must consent to two things that ordinary survey participants do not face: (1) participating in a group discussion where other participants can hear their responses, and (2) being audio or video recorded. Both require explicit consent before the session begins. The consent form should specify who will access the recordings, how long recordings will be retained, and whether clients or other observers will be watching the session live.

---

# References

Braun, V., & Clarke, V. (2006). Using thematic analysis in psychology. *Qualitative Research in Psychology, 3*(2), 77–101.

Creswell, J. W., & Plano Clark, V. L. (2018). *Designing and conducting mixed methods research* (3rd ed.). SAGE.

Guest, G., Bunce, A., & Johnson, L. (2006). How many interviews are enough? *Field Methods, 18*(1), 59–82.

Kahneman, D. (2011). *Thinking, fast and slow*. Farrar, Straus and Giroux.

Kozinets, R. V. (2020). *Netnography: The essential guide to qualitative social media research* (3rd ed.). SAGE.

Silver, L., Stevens, R., Wrenn, B., & Loudon, D. (2021). *The essentials of marketing research*. SAGE Publications.

Zaltman, G. (2003). *How customers think: Essential insights into the mind of the market*. Harvard Business School Press.
