---
title: "Solutions — Session 5: Text Mining a Focus-Group Transcript"
subtitle: "M731 · Full worked walkthrough (companion to session-05.qmd)"
author: "Shay Abbasi"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Purpose

Session 5 is mostly a concept-and-workshop session (qualitative
methods, a discussion-guide workshop) rather than a coded analysis, so
this solutions file is shorter than most. It expands the deck's small
R text-mining demo and adds a short note on what such a demo can and
cannot substitute for.

**Dataset.** `maritime_focus.txt` contains four instructor-simulated
focus-group transcripts on Maritime Roast subscriber churn (lapsed
subscribers in Halifax and Moncton, current subscribers considering
downgrading, and Ontario prospects). These are illustrative, not real
fieldwork.

## Setup

```{r}
#| label: setup
library(tidyverse)
library(readtext)
library(tidytext)

transcripts <- readtext("data/maritime_focus.txt")
```

## Word frequencies

```{r}
#| label: fig-wordfreq
#| fig-cap: "Most frequent content words across the four transcripts"
tokens <- transcripts |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  filter(!word %in% c("p1","p2","p3","p4","p5","p6","p7","p8",
                       "p9","p10","p11","p12","p13","p14","p15","p16")) |>
  count(word, sort = TRUE)

tokens |>
  slice_head(n = 15) |>
  ggplot(aes(x = fct_reorder(word, n), y = n)) +
  geom_col(fill = "#7A003C") +
  coord_flip() +
  labs(title = "Most frequent words across focus groups", x = NULL, y = "Count") +
  theme_minimal()
```

**Interpretation.** Words like "subscription," "price," "shipping,"
and "coffee" dominate — consistent with the churn themes participants
raised: price increases, delivery reliability, and the coffee ritual
itself. This is a useful **first pass**, not a finding — it tells you
what to look for when you actually code the transcripts, not what the
transcripts mean.

## A simple bigram pass (what pairs of words co-occur?)

```{r}
#| label: bigrams
bigrams <- transcripts |>
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |>
  separate(bigram, into = c("w1", "w2"), sep = " ") |>
  filter(!w1 %in% stop_words$word, !w2 %in% stop_words$word) |>
  count(w1, w2, sort = TRUE) |>
  slice_head(n = 10)

bigrams
```

**Interpretation.** Bigrams like "price increase" and "delivery
window" surface the *specific* complaints behind the single-word
frequencies above — closer to the level of detail a real coding scheme
would capture.

## What this demo does and doesn't replace

- **Does:** give you a fast, rough sense of vocabulary and salient
  topics before you start coding by hand; useful for triaging which
  transcripts to prioritise in a large corpus.
- **Doesn't:** replace systematic coding. It can't tell you *why*
  someone mentioned price, whether the mention was positive or
  negative, or which theme a given quote supports. Word counts flatten
  context — "I didn't mind the price increase" and "the price increase
  is why I left" count identically here.

For a real project, use this kind of pass as a **starting map**, then
code manually (or with CAQDAS software) for anything you intend to
report as a finding.

## Take-away for students

Qualitative rigour comes from systematic coding and inference, not
from a word cloud. Use text mining to orient yourself in a large
corpus quickly — never as a substitute for reading and coding the
material.
