---
title: "Session 6 — Data Analysis & Hypothesis Testing"
subtitle: "M731 Week 6 · October 14, 2026"
author: "Shay Abbasi"
date: "2026-10-14"
---

## Session 6 {background-color="#7A003C"}

Data analysis and hypothesis testing

::: {.notes}
Big session - kicks off the analytical block. Two weeks (6 and 7) on
hypothesis testing.
:::

## Today's agenda

1. Test 1 debrief (5 min)
2. Describing distributions
3. From description to inference
4. Two-sample tests
5. Proportion tests, chi-square
6. Effect sizes
7. **Break (15 min)**
8. R lab: descriptives + hypothesis tests on `maritime_survey.csv`
9. Ethics: p-hacking and HARKing
10. Wrap-up

::: {.notes}
Ethics thread is 15 min today - matters this week. Post-Test-1 attention
will be lower; keep pace up and questions frequent.
:::

## Learning objectives

By end of session 6 you will be able to:

- Describe distributions in R (centre, spread, skewness, outliers)
- Choose the right hypothesis test for common marketing questions
- Interpret p-values, confidence intervals, and effect sizes together
- Report results without over-claiming

::: {.notes}
"Without over-claiming" is a deliberate LO - most students over-claim
from a single p-value. Effect size framing addresses this.
:::

## Test 1 debrief (5 min)

- Mean: TBD, distribution TBD
- Common errors (aggregated):
  - Confusion between exploratory and descriptive design
  - Sample-size calculations without justification
  - Missing ethics elements

::: {.notes}
Fill in real numbers on the day. Don't identify individual students.
Frame errors constructively - "these are the concepts to revisit
before Test 2."
:::

# Part 1 — Describing distributions {background-color="#7A003C"}

## Central tendency

- **Mean** = arithmetic average
- **Median** = middle value
- **Mode** = most frequent

When they differ, the distribution is skewed.

For marketing: income and spend are usually right-skewed. Median often more informative than mean.

::: {.notes}
Give an example: median household income in Canada ~$85K, mean higher
(~$100K+) because top-end incomes pull. For MR reporting, use median
unless there's a specific reason to use mean.
:::

## Spread

- **Range** = max - min
- **IQR** (interquartile range) = Q3 - Q1
- **Standard deviation** = average distance from mean
- **CV** (coefficient of variation) = SD/mean (unit-free)

CV is useful when comparing spread across different-scale variables.

::: {.notes}
CV lets you say "spending varies more than age" even though they're
in different units. Only defined for ratio-scale variables with
positive mean.
:::

## Shape

- **Skewness:** left- or right-tail
- **Kurtosis:** peakedness / fat tails
- **Outliers:** points beyond ±1.5×IQR or ±3σ

Skewness and outliers change your test choice.

::: {.notes}
Right-skewed variables (income, purchase amount) often benefit from
log transformation before parametric tests. Or use non-parametric
alternatives.
:::

## Visualising a distribution

- **Histogram:** binned frequency
- **Density plot:** smoothed histogram
- **Boxplot:** median, IQR, whiskers, outliers
- **QQ plot:** compare to a theoretical distribution (normal?)

Always plot before you test.

::: {.notes}
"Always plot" is one of the strongest MR habits. A weird distribution
that would break a t-test is obvious in a histogram, invisible in a
mean. Plot first, then test.
:::

## Descriptives in R

```{r}
#| eval: false
library(tidyverse)
library(psych)

survey <- read_csv("data/maritime_survey.csv")

summary(survey$wtp)          # base R quick summary
describe(survey$wtp)         # psych — includes SD, skew, kurt, se
```

`psych::describe()` is more informative than base `summary()`.

::: {.notes}
Live-code. Show output. Compare the two - psych::describe gives you
SD, skew, kurtosis, standard error of the mean. Useful defaults.
:::

## Grouped descriptives

```{r}
#| eval: false
survey |>
  group_by(region) |>
  summarise(
    n = n(),
    mean_wtp = mean(wtp, na.rm = TRUE),
    median_wtp = median(wtp, na.rm = TRUE),
    sd_wtp = sd(wtp, na.rm = TRUE)
  )
```

Almost every analysis starts with a `group_by + summarise` like this.

::: {.notes}
Workhorse pattern. Students should be able to write this from memory
by Wk 8. If they can't, they're behind - flag for office hours.
:::

# Part 2 — From description to inference {background-color="#7A003C"}

## Sample vs. population

- **Sample:** the units you have data for
- **Population:** the units you want to draw conclusions about

Descriptives describe the sample. Inference generalises to the population.

::: {.notes}
Distinguish descriptive analysis from inferential analysis. A mean is
a description of the sample. A confidence interval is an inference
about the population. Both are legitimate; you must know which
you're producing.
:::

## The hypothesis-testing framework

1. State a **null hypothesis** (H₀) — the "no effect" position
2. State an **alternative** (H₁)
3. Compute a **test statistic** from the sample
4. Compute a **p-value** — probability of seeing this data if H₀ were true
5. **Decide** — reject H₀ if p < α (usually 0.05)

Not a rigorous truth procedure — a decision rule.

::: {.notes}
Frame as decision-making under uncertainty. Null-hypothesis
significance testing (NHST) has known problems (which you'll cover in
the ethics thread). Students should understand what p-values do and
don't mean.
:::

## What p-values are not

- **NOT** the probability H₀ is true
- **NOT** the probability the effect is real
- **NOT** an effect size
- **NOT** evidence of practical importance

They are: the probability of the observed data (or more extreme) under H₀.

::: {.notes}
Common misinterpretations. Cohen 1994 "The Earth Is Round (p < .05)"
is the classic essay - point at it. p-values are a small piece of the
inferential picture, not the whole thing.
:::

## Type I and Type II errors

|              | H₀ true    | H₀ false   |
|--------------|------------|------------|
| Reject H₀    | Type I (α) | Correct    |
| Retain H₀    | Correct    | Type II (β)|

Power = 1 - β.

Bigger samples reduce Type II. Stricter α reduces Type I.

::: {.notes}
Both errors matter. Marketing decisions rarely need α=0.001; α=0.05
is fine most of the time. Power (avoiding Type II) is often more
important than tiny α - you want to detect real effects.
:::

# Part 3 — Two-sample tests {background-color="#7A003C"}

## Two-sample t-test

Compare means of two groups.

```{r}
#| eval: false
t.test(wtp ~ region, data = survey)
```

Output: mean by group, t-statistic, df, p-value, CI, alternative.

::: {.notes}
Live-code with real Maritime Roast data - "does WTP differ by
region?" Interpret output: which region has higher mean, is
difference significant, how big is the CI.
:::

## Paired t-test

For repeated measures on the same subjects.

```{r}
#| eval: false
t.test(pre_score, post_score, paired = TRUE, data = survey)
```

Use when the two measurements are linked (same person, before/after).

::: {.notes}
Common example: pre-post advertising exposure effect. Same subjects
measured twice. Paired test is more powerful than independent-samples
test when pairing matters.
:::

## Non-parametric alternative: Wilcoxon

When t-test assumptions violated (small n, non-normal, ordinal data).

```{r}
#| eval: false
wilcox.test(wtp ~ region, data = survey)
```

Tests medians rather than means. Less powerful, more robust.

::: {.notes}
Rule of thumb: n < 30 per group, or heavy skew, use Wilcoxon. For
larger samples with any skew, t-test is robust and usually fine.
:::

## Two-proportion test

Compare proportions across two groups.

```{r}
#| eval: false
prop.test(
  x = c(180, 210),   # successes
  n = c(400, 400)    # trials
)
```

E.g., "proportion of subscribers who recommend, subscribed vs. lapsed."

::: {.notes}
prop.test is base R. Equivalent to chi-square for 2x2. Useful when
your data is already summarised (counts, not raw rows).
:::

## Chi-square test of independence

For association between two categorical variables.

```{r}
#| eval: false
tbl <- table(survey$region, survey$prefers_dark_roast)
chisq.test(tbl)
```

Tests whether the two variables are independent.

::: {.notes}
Two questions: is there an association, and if so, where. Test
answers the first. Residuals (`chisq.test(tbl)$residuals`) answer the
second - large positive residuals mark cells with more observations
than expected.
:::

## Break — 15 minutes {background-color="#FDBF57"}

::: {.notes}
Have `maritime_survey.csv` loaded and ready for the lab. Check
`region`, `wtp`, `brand_attachment_*` columns exist.
:::

# Part 4 — Effect sizes {background-color="#7A003C"}

## Why effect sizes matter

- p-values shrink with sample size — huge samples find trivial effects "significant"
- Effect sizes measure the **magnitude** of the effect
- Combine both: "is the effect there, and is it big enough to care about?"

::: {.notes}
Give example: with n=10K, a 0.1% difference is significant. Nobody
cares. Effect size (Cohen's d = 0.02) says "the effect is tiny."
Report both.
:::

## Cohen's d (for means)

d = (M₁ - M₂) / SD

- 0.2 = small
- 0.5 = medium
- 0.8 = large

```{r}
#| eval: false
library(effectsize)
cohens_d(wtp ~ region, data = survey)
```

::: {.notes}
Cohen's rules-of-thumb are approximate but useful. In MR contexts
sometimes d=0.1 is meaningful (large-market, small margins). Context
matters. Always report d plus its CI.
:::

## Cramér's V (for chi-square)

Effect size for contingency tables.

```{r}
#| eval: false
cramers_v(tbl)
```

- 0.1 = small
- 0.3 = medium
- 0.5 = large

::: {.notes}
Chi-square p-value tells you if the association is likely non-zero;
Cramér's V tells you how strong the association is. Report both.
:::

## Statistical vs. economic significance

- **Statistical significance:** unlikely to be due to chance
- **Economic (managerial) significance:** big enough to change the decision

Reports should tell both stories.

::: {.notes}
This is the maturity marker in MR writing. "The difference is
statistically significant (p<.001) but small in magnitude (d=0.15).
For the CEO's decision, the effect is unlikely to move the needle."
That's the sentence to teach.
:::

# Part 5 — R lab: hypothesis tests {background-color="#7A003C"}

## Lab task

Using `maritime_survey.csv`:

1. Descriptives on subscriber demographics (mean age, gender split, tenure)
2. Test whether stated WTP differs between current subscribers and lapsed subscribers
3. Report effect size (Cohen's d)
4. Test whether preference for dark roast is associated with region (chi-square)
5. Write a plain-language recommendation for the CEO in ≤ 150 words

A fuller worked version of this exercise is in **`solutions-06.qmd`**
in this session's folder.

::: {.notes}
Give 40 min. Circulate. Common issues: wrong test choice, missing
na.rm=TRUE, forgetting effect size. Emphasise that the 150-word
recommendation is what the CEO actually reads - most students will
write 400 words. Push back.
:::

## Reporting a test — the sentence

"Current subscribers reported higher WTP (M=X.X, SD=Y.Y) than lapsed subscribers (M=X.X, SD=Y.Y), t(df) = Z, p = P, Cohen's d = D, 95% CI [L, U]. The effect is [small / moderate / large]; **managerially this means** [interpretation]."

Six elements: descriptives, test statistic, p, effect size, CI, interpretation.

::: {.notes}
This is the sentence template. Every hypothesis test in every report
should read like this. Students who master this sentence pattern
write clearer reports.
:::

# Part 6 — Ethics thread (15 min) {background-color="#7A003C"}

## P-hacking

Running many tests until one is "significant" and reporting only that one.

Examples:

- Testing many DVs; reporting only the significant one
- Excluding outliers "post-hoc" to move p across 0.05
- Adding covariates until the effect appears

**Consequence:** most "significant" findings in the literature don't replicate.

::: {.notes}
Simmons Nelson Simonsohn 2011 "False-Positive Psychology" is the
foundational paper. They show that with routine researcher DoF, false
positive rates are ~60%, not 5%. Standard practice in academic
psych has changed; commercial MR is still catching up.
:::

## HARKing

**H**ypothesising **A**fter **R**esults are **K**nown.

Running the analysis, seeing what's significant, writing hypotheses that "predicted" it.

Turns exploratory into confirmatory. Undermines inference.

::: {.notes}
HARKing is subtle because it feels honest ("I did find this
relationship!"). The problem: p-values assume the hypothesis was
specified before looking at data. Post-hoc "hypotheses" don't
control error rates.
:::

## Selective reporting

- Reporting only the analyses that "worked"
- Not reporting analyses that failed
- Not reporting the total number of tests performed

**Fix:** report everything, or report a preregistered subset.

::: {.notes}
For a client report this creates a bind - clients want a clean story,
not "we ran 20 tests and here's the 3 significant ones." Manage
expectations up front about what will be reported.
:::

## Preregistration

- Write hypotheses, DVs, tests **before** collecting data
- Timestamp in a public registry (OSF, AsPredicted)
- Report deviations transparently

Common in academic research. Rare in commercial MR — worth thinking about.

::: {.notes}
For your project, informal preregistration = writing your hypotheses
and test plans in the proposal (due Wk 9) before doing the analysis.
That's a preregistration in spirit.
:::

## The replication frame

Ask: if someone ran your study again, would they get the same result?

If your answer depends on cherry-picked analyses, tighten up before reporting.

::: {.notes}
Replication crisis has reshaped academic psych, medicine, and
increasingly MR. Marketing academic journals now require preregistration
for causal claims. Practical MR should adopt the same discipline.
:::

# Part 7 — For your project {background-color="#7A003C"}

## Use today's tools on your project data

- If you have your project data ready, run descriptives and one hypothesis test
- If you're still designing, use Wk 6-7 to think about what tests your final report will run
- Project proposal + questionnaire due Wk 9 (Nov. 4)

::: {.notes}
Push teams to be running R on their project data by Wk 8 - waiting
until after the proposal deadline is too late. The proposal should
name your intended analyses, so the analytical plan must exist by Wk
9.
:::

## Week 7 preview

- **Concept:** ANOVA (more than two groups), post-hoc tests, assumption checks
- **R lab:** ANOVA on satisfaction across subscription tiers
- Same dataset (`maritime_survey.csv`), same style

::: {.notes}
Continuous with today's session. Reading: CF Ch. 5-6 (revisit).
:::

## Thank you — questions?

`abbass44@mcmaster.ca`

::: {.notes}
Stay after for any follow-up questions.
:::
