---
title: "Solutions — Session 6: Descriptives and Hypothesis Tests"
subtitle: "M731 · Full worked walkthrough (companion to session-06.qmd)"
author: "Shay Abbasi"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Purpose

Full worked version of the Session 6 in-lab exercise: descriptives,
a two-sample t-test with effect size, and a chi-square test on
`maritime_survey.csv`.

## Setup

```{r}
#| label: setup
library(tidyverse)
library(psych)
library(effectsize)

survey <- read_csv("data/maritime_survey.csv")
```

## Step 1 — Subscriber demographics

```{r}
#| label: demographics
describe(survey$age)

survey |>
  summarise(
    n = n(),
    prop_female = mean(gender == "F"),
    mean_tenure = mean(tenure_months),
    median_tenure = median(tenure_months)
  )
```

**Interpretation.** The subscriber base skews slightly female
(~52%), averages 39 years old, with tenure distributed as expected for
a subscription business (a long right tail of loyal long-tenure
subscribers on top of a larger recent-signup base).

## Step 2 — WTP: current vs. lapsed subscribers

```{r}
#| label: wtp-test
t.test(wtp ~ subscriber_status, data = survey)
```

```{r}
#| label: wtp-effect-size
cohens_d(wtp ~ subscriber_status, data = survey)
```

**Reporting sentence.** Current subscribers reported higher WTP
(M ≈ \$22.07) than lapsed subscribers (M ≈ \$19.51), a statistically
significant difference (p < .001) with a moderate-to-large effect
(Cohen's d ≈ 0.6-0.7 depending on the draw). **Managerially**: lapsed
subscribers were already price-sensitive before they churned — pricing
and perceived value, not just delivery logistics, are worth
investigating as churn drivers.

## Step 3 — WTP by region (the slide-deck demo, expanded)

```{r}
#| label: wtp-region
t.test(wtp ~ region, data = survey)
cohens_d(wtp ~ region, data = survey)
```

**Interpretation.** Atlantic subscribers show higher stated WTP than
Ontario prospects, with a small-to-moderate effect. This matters for
the Ontario-entry pricing conversation: an entry-tier price point may
be worth testing in Ontario rather than importing Atlantic pricing
directly.

## Step 4 — Dark-roast preference by region (chi-square)

```{r}
#| label: chisq
tbl <- table(survey$region, survey$prefers_dark_roast)
tbl
chisq.test(tbl)
cramers_v(tbl)
```

**Interpretation.** Report both the chi-square p-value (is there an
association at all?) and Cramér's V (how strong is it?). A
statistically significant but small-V association means region
explains some, but not most, of the variation in roast preference —
useful context before recommending a region-specific product mix.

## Step 5 — CEO recommendation (≤150 words)

> Current subscribers report meaningfully higher willingness-to-pay
> than lapsed subscribers (\$22 vs. \$20, p < .001, moderate effect),
> suggesting price sensitivity plays a real role in the churn we're
> trying to reduce. Separately, Ontario prospects show slightly lower
> stated WTP than our existing Atlantic base — a signal to test a
> lower entry-tier price point rather than launch at Atlantic pricing.
> Roast preference differs modestly by region but not enough to justify
> a region-specific product line on this evidence alone. Recommendation:
> prioritize a churn-focused pricing/value review before the Ontario
> launch, and pilot an Ontario entry-tier price point rather than a
> new product line.

## Take-away for students

Every hypothesis test in this course should end with the six-part
sentence: **descriptives → test statistic → p → effect size → CI →
managerial interpretation.** The CEO reads the last clause; the rest
is your evidence for it.
