---
title: "Session 8 — Marketing Scales"
subtitle: "M731 Week 8 · October 28, 2026"
author: "Shay Abbasi"
date: "2026-10-28"
---

## Session 8 {background-color="#7A003C"}

Marketing scales — measurement reliability and validity

::: {.notes}
Collect any graded work due today per A2L logistics. Then dive into
scales content. Wk 8 revisits measurement (which was touched in
Wk 3) at greater depth, now that students have seen how scales feed
analysis.
:::

## Today's agenda

1. What is a scale? Types of rating scales
2. Reliability: what it is, how we measure it
3. Validity: content, construct, criterion
4. Item-level diagnostics
5. **Break (15 min)**
6. R lab: Cronbach's alpha on `maritime_survey.csv`
7. Ethics: culturally appropriate scales
8. Project questionnaire revision workshop
9. Wrap-up

::: {.notes}
The revision workshop is critical - teams must have their draft
questionnaire from Wk 4 and revise it using today's scale content
before the Wk 9 project deadline.
:::

## Learning objectives

By end of session 8 you will be able to:

- Choose among common rating-scale formats
- Assess reliability using Cronbach's alpha and item-total correlations
- Distinguish content, construct, and criterion validity
- Diagnose weak scale items and recommend drops or revisions
- Revise a multi-item scale for your project questionnaire

::: {.notes}
Practical framing - by end, students should be able to look at their
own questionnaire's multi-item batteries and say "this item is weak,
this is strong."
:::

# Part 1 — Rating scales {background-color="#7A003C"}

## Common rating scales

- **Likert:** 5- or 7-point agreement ("Strongly Disagree — Strongly Agree")
- **Semantic differential:** bipolar adjectives (Boring ← → Exciting)
- **Stapel:** single adjective with -/+ numeric range
- **Numeric rating:** 0-10, NPS
- **Constant sum:** allocate 100 points across items
- **MaxDiff / Best-Worst Scaling:** repeated choice of best and worst

::: {.notes}
Each has a place. Likert dominates commercial MR. Semantic differential
for brand-image. MaxDiff for feature prioritisation (increasingly
popular replacement for rating batteries).
:::

## Likert design refresher

- 5 or 7 points (research trades off precision vs. cognitive load)
- All points labelled (not just endpoints)
- Neutral point included (unless forcing a side)
- Reverse ~20% of items in a battery (detects straight-lining)

::: {.notes}
Wk 4 covered this. Recap briefly. The reverse-item point matters -
students often forget and end up with un-detectable straight-liners.
:::

## MaxDiff — the industry rising star

For preference among many items:

- Show 4-6 items at a time
- Ask "best" and "worst"
- Repeat with different sets
- Model preferences via hierarchical Bayes or logistic regression

More discriminating than rating scales; less prone to positive bias.

::: {.notes}
MaxDiff has largely replaced 10-item ratings in commercial preference
research. Suppliers like Sawtooth Software specialise. Overkill for
most course projects but worth naming.
:::

## Constant sum

"Allocate 100 points across these five features to reflect their importance to you."

- Forces trade-offs
- Reveals relative not just absolute preferences
- Cognitively demanding for respondents (limit to 4-6 items)

::: {.notes}
Good when a rating scale would produce "everything is important."
Constant sum forces choice. Cognitive load matters - respondents
can't easily allocate 100 across 12 items.
:::

# Part 2 — Reliability {background-color="#7A003C"}

## What is reliability?

Reliability = **consistency** of measurement.

If you measured the same construct twice on the same person, you should get similar results.

Reliability ≠ Validity. A reliable measurement can still be measuring the wrong thing.

::: {.notes}
Analogy: a bathroom scale that's consistently 5 lb off is reliable
but not valid. A scale that gives different readings each time is
neither. You want both.
:::

## Types of reliability

- **Test-retest:** same instrument, same people, two occasions
- **Parallel-forms:** two versions of the instrument, same people, same time
- **Split-half:** split items in half, correlate two halves
- **Internal consistency:** how much items in a battery correlate with each other

Internal consistency (Cronbach's alpha) is the dominant type in MR.

::: {.notes}
Test-retest is rare in commercial MR because it requires re-fielding.
Split-half is arbitrary depending on split. Cronbach's alpha
generalises split-half and is the standard.
:::

## Cronbach's alpha

α ranges from 0 to 1.

- **α > 0.9:** excellent (may indicate redundant items)
- **α 0.8-0.9:** good
- **α 0.7-0.8:** acceptable for research
- **α 0.6-0.7:** questionable
- **α < 0.6:** poor — revise items

Rules of thumb, not laws.

::: {.notes}
Nunnally 1978 gave the 0.7 threshold that everyone cites. It's a
convention, not a law. For exploratory research 0.6 may be fine; for
high-stakes clinical measurement 0.9+ is expected.
:::

## What alpha measures

Formally: α = (k / (k-1)) × (1 - Σvarᵢ / vartotal)

Intuition:

- Increases with average inter-item correlation
- Increases with number of items
- More items ≠ better if they're redundant

::: {.notes}
Alpha is inflated by adding more items even if quality is the same.
This is why you see 20-item batteries with α=.95 in academic papers -
that number is partly artefact.
:::

## Item-total correlation

For each item, correlation between that item and the rest of the battery.

- Items with low corrected item-total correlation (< 0.3) are candidates for removal
- Items with negative correlation are wrong (probably reversed incorrectly)

::: {.notes}
This is the diagnostic tool - alpha tells you overall reliability,
item-total tells you which items are pulling down. In R, `psych::alpha`
gives both.
:::

## "Alpha if item deleted"

For each item: what would alpha be if you dropped this item?

- If dropping an item **raises** alpha meaningfully → drop it
- If dropping an item lowers alpha → keep it

::: {.notes}
This is why psych::alpha is so useful - it computes drop-one alphas.
Iterate: drop worst item, re-run, see if new worst is worth dropping.
Stop when no drops meaningfully improve alpha.
:::

# Part 3 — Validity {background-color="#7A003C"}

## Validity families

- **Content validity:** items cover the construct's meaning
- **Construct validity:** the scale measures the theoretical construct
- **Criterion validity:** scale predicts an external criterion

Each requires different evidence.

::: {.notes}
Validity is harder than reliability because there's no single number.
It's an accumulation of evidence. In practice, MR shortcuts to
construct validity via factor analysis (Wk 9) and content via
expert review.
:::

## Content validity

- Do the items span the domain of the construct?
- Are there important aspects missing?
- Are there extraneous items?

Assessed by **expert review** — subject-matter experts rate item relevance.

::: {.notes}
Content validity is judged, not computed. In academic MR, you'd
convene 5-10 experts and compute a Content Validity Index (CVI). In
commercial MR, it's usually less formal - internal review, client
sign-off.
:::

## Construct validity

Two subtypes:

- **Convergent:** the scale correlates with other measures of the same construct
- **Discriminant:** the scale doesn't correlate with measures of different constructs

Assessed with factor analysis, multi-trait multi-method matrices, structural equation modelling.

::: {.notes}
Fornell & Larcker 1981 is the canonical treatment. Wk 9 covers factor
analysis, which is the first step. Full construct validation requires
data on multiple constructs from the same respondents.
:::

## Criterion validity

- **Concurrent:** scale predicts a current criterion (satisfaction predicts current spend)
- **Predictive:** scale predicts a future criterion (satisfaction predicts next-year churn)

Useful when your scale is a proxy for something behavioural.

::: {.notes}
NPS is often defended on criterion-validity grounds - "it predicts
future growth." The evidence is mixed (Keiningham et al. 2007).
Predictive validity is the strongest form because it demonstrates
practical use.
:::

## Composite scores

Once you're confident in reliability + validity:

```{r}
#| eval: false
survey <- survey |>
  mutate(
    attachment = rowMeans(
      select(., brand_attach_1:brand_attach_5),
      na.rm = TRUE
    )
  )
```

Analyse the composite, not individual items.

::: {.notes}
Composite is the point of multi-item measurement - reduce measurement
error by averaging. Analysing individual items in downstream models
throws away the benefit.
:::

## Break — 15 minutes {background-color="#FDBF57"}

::: {.notes}
Have `maritime_survey.csv` loaded, columns brand_attach_1..5 visible.
:::

# Part 4 — R lab: Cronbach's alpha {background-color="#7A003C"}

## The `psych::alpha` function

```{r}
#| eval: false
library(psych)

battery <- survey |>
  select(brand_attach_1, brand_attach_2, brand_attach_3,
         brand_attach_4, brand_attach_5)

alpha_result <- alpha(battery)
alpha_result
```

Output: overall alpha, alpha-if-deleted per item, item-total correlations.

::: {.notes}
Live-code. Show output. Walk through each section:
- $total: overall alpha (raw and standardised)
- $alpha.drop: what alpha would be without each item
- $item.stats: item statistics
Interpret which items are pulling down.
:::

## Interpreting the output

Look for:

- **Overall α** — is it above 0.7?
- **Alpha if deleted** — any item's deletion raises alpha?
- **Item-total correlation** — any items below 0.3?
- **Item-total correlation (corrected)** — this excludes the item itself

::: {.notes}
Corrected item-total is more meaningful than raw. Raw is inflated
because the item is in the total. Show both if useful, but interpret
corrected.
:::

## Simulating known-structure data

```{r}
#| eval: false
library(MASS)

# Design a covariance matrix for 6 correlated items
n <- 500
mu <- rep(0, 6)
sigma <- matrix(0.5, 6, 6)  # inter-item cor = 0.5
diag(sigma) <- 1

sim_data <- as.data.frame(mvrnorm(n, mu, sigma))
alpha(sim_data)
```

Useful for demonstrating what "high" and "low" alpha look like.

::: {.notes}
Live-simulate. Show alpha for r=0.5 (about 0.85) vs. r=0.2 (about
0.60). Students see the mechanics behind the number.
:::

## Correlation heatmap

```{r}
#| eval: false
library(corrplot)

battery |>
  cor(use = "pairwise.complete.obs") |>
  corrplot(method = "color", type = "upper",
           addCoef.col = "black")
```

Visualise the inter-item correlation structure.

::: {.notes}
Heatmap makes the redundancy visible. Items that are near-identical
show up as near-1 correlations - candidates for removal.
:::

## In-lab exercise

Using `brand_attach_1..5` from `maritime_survey.csv`:

1. Compute Cronbach's alpha
2. Report item-total correlations
3. Identify any items worth dropping
4. Compare with the simulated 8-item battery (script provided)
5. Write a 3-sentence recommendation for Maritime Roast's brand-attachment measure

A fuller worked version of this reliability analysis is in
**`solutions-08.qmd`** in this session's folder.

::: {.notes}
Give 30 min. Circulate. Students should get α ~0.83-0.88 depending
on how the data was simulated. Push them to interpret the number
managerially, not just report it.
:::

# Part 5 — Ethics thread {background-color="#7A003C"}

## Culturally appropriate scales

- Scales developed in the US on college students may not travel
- Translation introduces measurement drift ("satisfied" doesn't mean the same in Cantonese)
- Meaning of neutral point varies across cultures (Japanese respondents avoid extremes)

For studies in multi-cultural Canada, plan for it.

::: {.notes}
Response-style differences are real. East-Asian populations tend to
use middle-of-scale; Latin populations use extremes. Comparing raw
means across cultures without adjustment is misleading.
:::

## Item translation

- **Forward translation:** original → target language
- **Back translation:** target → original (by a different translator)
- **Reconciliation:** compare, adjust items where meaning drifted

Standard practice for cross-cultural work.

::: {.notes}
Time and cost matters - forward-back-reconcile easily doubles
questionnaire development time. For your project, note the limit if
your target audience is multilingual and you're only using English.
:::

## Response-style adjustment

Techniques for handling response-style differences:

- **Standardisation:** subtract each respondent's mean, divide by SD (ipsatisation)
- **Anchor-vignette methods:** calibrate against a fixed scenario
- **MaxDiff:** forced-choice avoids scale-use variation

::: {.notes}
Ipsatisation loses absolute-level information but preserves
within-person patterns. MaxDiff sidesteps the issue entirely.
Multi-national studies increasingly use MaxDiff for this reason.
:::

# Part 6 — Project questionnaire workshop {background-color="#7A003C"}

## Revise your questionnaire

Teams (30 min):

1. Pull up your Wk 4 draft
2. Identify all multi-item batteries
3. Apply today's checklist:
   - Are there ≥3 items per construct?
   - Are 20%+ reverse-coded?
   - Do items sound reliable (all measuring the same thing)?
   - Would content-validity experts approve?
4. Revise items where needed

::: {.notes}
Circulate. Push students to actually rewrite items, not just note
"could improve." The revised questionnaire is due Wk 9.
:::

## What "done" looks like

- Every construct: 3-5 items, reverse-coded item present
- Item wording: simple, one-idea, neutral
- Length: still under 12 min
- Consent language: present and complete
- Rendered as a Word / PDF / HTML file ready to submit

::: {.notes}
The Wk 9 submission is the questionnaire artefact, not just a research
plan. Teams should treat it as a fieldable document.
:::

# Part 7 — For your project {background-color="#7A003C"}

## Deliverable: Wk 9 (next Wednesday)

- Group proposal + questionnaire (10% of grade)
- Due at start of class Nov. 4

Structure of the proposal:

1. Business problem (1 paragraph)
2. Research objectives (3-5 bullets)
3. Design justification (1 paragraph)
4. Sampling plan (1 paragraph)
5. Draft questionnaire (attached)
6. Analysis plan (bullets: what tests, what visualisations)

::: {.notes}
Give students a rubric. Analysis plan is important - it forces teams
to think about how their questionnaire feeds their analysis.
:::

## Week 9 preview

- **Concept:** Factor analysis (PCA, EFA)
- **R lab:** EFA on the needs battery in `maritime_survey.csv`
- **Project proposal + questionnaire due at start of class**

Prep reading: CF Ch. 8.

::: {.notes}
Factor analysis follows scales naturally - reliability tells you a
battery hangs together; factor analysis tells you what structure the
batteries have.
:::

## Thank you — questions?

`abbass44@mcmaster.ca`

::: {.notes}
Stay for project questionnaire questions. This is when teams have
their most pressing questions.
:::
