---
title: "Solutions — Session 8: Reliability of the Brand-Attachment Battery"
subtitle: "M731 · Full worked walkthrough (companion to session-08.qmd)"
author: "Shay Abbasi"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Purpose

Full worked version of the Session 8 in-lab exercise: Cronbach's
alpha, item-total correlations, and a simulated comparison battery.

## Setup

```{r}
#| label: setup
library(tidyverse)
library(psych)
library(MASS)   # NOTE: MASS::select() masks dplyr::select() - use dplyr::select() explicitly below
library(corrplot)

survey <- read_csv("data/maritime_survey.csv")

battery <- survey |>
  dplyr::select(brand_attach_1, brand_attach_2, brand_attach_3, brand_attach_4, brand_attach_5)
```

## Step 1 — Cronbach's alpha

```{r}
#| label: alpha
alpha_result <- alpha(battery)
alpha_result
```

**Interpretation.** Overall standardised alpha lands around 0.85-0.86
— "good" reliability by the Nunnally convention (0.8-0.9), meaning the
five items are measuring the same underlying construct consistently
without being so redundant that we'd worry about over-length.

## Step 2 — Item-total correlations and alpha-if-deleted

```{r}
#| label: item-diagnostics
alpha_result$item.stats
alpha_result$alpha.drop
```

**Interpretation.** Check two things: (1) does any item have a
corrected item-total correlation below 0.3? (2) does dropping any item
*raise* alpha? In this simulated battery, all five items were
constructed with moderate-to-strong loadings (0.68-0.80), so you
should see no obvious candidate for removal — a "clean" battery,
useful as a contrast case against a weaker one.

## Step 3 — Correlation heatmap

```{r}
#| label: fig-heatmap
#| fig-cap: "Inter-item correlations, brand-attachment battery"
battery |>
  cor(use = "pairwise.complete.obs") |>
  corrplot(method = "color", type = "upper", addCoef.col = "black", tl.col = "black")
```

**Interpretation.** Moderate, fairly even correlations across all
item pairs (roughly 0.3-0.5) is the signature of a well-behaved
single-factor battery — no pair is near-identical (which would signal
redundancy) and none is near-zero (which would signal an off-construct
item).

## Step 4 — Simulated comparison: what does "low" alpha look like?

```{r}
#| label: sim-comparison
set.seed(731)
n <- 500

sim_high <- as.data.frame(mvrnorm(n, rep(0, 6), matrix(0.5, 6, 6) + diag(6) * 0.5))
sim_low  <- as.data.frame(mvrnorm(n, rep(0, 6), matrix(0.2, 6, 6) + diag(6) * 0.8))

cat("Inter-item r = 0.5 -> alpha =", round(alpha(sim_high)$total$raw_alpha, 2), "\n")
cat("Inter-item r = 0.2 -> alpha =", round(alpha(sim_low)$total$raw_alpha, 2), "\n")
```

**Interpretation.** This is the mechanism behind the number: alpha
rises with average inter-item correlation (holding item count fixed).
The Maritime Roast brand-attachment battery, at ~0.85, sits closer to
the r=0.5 simulation than the r=0.2 one — real, moderately-strong
shared signal across items.

## Step 5 — Recommendation for Maritime Roast

> The 5-item brand-attachment battery is reliable (α ≈ 0.85) with no
> weak items to drop. It's ready to use as a composite score in
> downstream analysis (e.g., as a regression predictor in Week 11)
> without further revision. If the team wants a shorter battery for
> survey-length reasons, dropping the weakest-loading item would cost
> only marginal reliability — but there's no evidence that's necessary
> here.

## Take-away for students

Alpha alone is a headline number; item-total correlations and
alpha-if-deleted are the diagnostic tools that tell you *what to do*
about a weak battery. Report all three, not just the top-line alpha.
