---
title: "Session 12 — Research Reports & Presentations"
subtitle: "M731 Week 12 · November 25, 2026"
author: "Shay Abbasi"
date: "2026-11-25"
---

## Session 12 {background-color="#7A003C"}

Research reports and presentations

**Guest speaker today.**

::: {.notes}
Wk 12 mixes lecture, guest speaker, and report clinic. Confirm guest
speaker arrival time. Have their slides / topic set. Introduce them
warmly - they're volunteering.
:::

## Today's agenda

1. Report structure — what a decision-oriented report looks like
2. Chart design and data storytelling
3. Quarto reporting features
4. **Break (15 min)**
5. **Guest speaker** (60 min including Q&A)
6. Project report clinic — teams work + get feedback (~30 min)
7. Wrap-up

::: {.notes}
Guest slot is 60 min - keep tight so the clinic has time. Report
clinic is the practical value for teams. Wk 13 report is due in 7
days.
:::

## Learning objectives

By end of session 12 you will be able to:

- Structure a decision-oriented research report
- Design charts that support the story
- Produce reproducible reports in Quarto (HTML, PDF, slides from one source)
- Present findings to a non-technical audience

::: {.notes}
Objective 3 is the R skill. Objectives 1 and 4 are the softer skills
that determine report quality.
:::

# Part 1 — Report structure {background-color="#7A003C"}

## The decision-oriented report

Structure for a marketing research report:

1. **Title + author + date**
2. **Executive summary** (1 page, plain-language, decision + evidence)
3. **Background and objectives** (context, research questions)
4. **Methodology** (design, sample, data, tests used)
5. **Findings** (the evidence, ordered by importance)
6. **Recommendations** (what to do)
7. **Limitations** (what your findings don't cover)
8. **Appendices** (detailed tables, code, questionnaire)

::: {.notes}
Executive summary is the most-read section. Some readers stop there.
Write it last, revise it three times. Recommendation → decision, not
just "here's what we found."
:::

## Executive summary — the format

Three paragraphs, one page:

1. **Situation:** what was the business question?
2. **Findings:** what did we learn? (2-3 headline points, quantified)
3. **Recommendation:** what should the client do?

If a busy executive reads only this page, they should be able to act.

::: {.notes}
Give an example: "Maritime Roast asked whether to enter Ontario.
Analysis of a 500-respondent survey finds 35% aided awareness in
Ontario vs. 68% in Atlantic; WTP is $2 lower per pack. We recommend
entering with a lower-priced entry SKU and a $150K first-year
awareness campaign." That's the shape. One page.
:::

## Findings — evidence-first

For each finding:

- **Headline** (one sentence in bold)
- **Evidence** (chart or table + numbers)
- **Interpretation** (what it means in plain language)
- **Confidence** (how sure are we?)

Repeat for 3-8 key findings, ordered by importance.

::: {.notes}
Number of findings depends on scope - a term project has 3-6; a
$100K commercial study might have 15. Order by importance, not
questionnaire order. Lead with the most decision-relevant.
:::

## Limitations — the credibility move

Every study has limitations. Naming them:

- Increases credibility (you sound honest)
- Prevents over-claiming
- Sets expectations for follow-up research

Never say "no limitations." Always name at least 2-3.

::: {.notes}
Weak reports hide limitations. Strong reports name them and explain
why the finding still holds despite them. This is the maturity
marker.
:::

## What makes a bad report

- No executive summary; buries the lede
- All methodology, no interpretation
- Charts without labels or units
- Recommendations that don't map to findings
- Statistical jargon without translation
- Length ≠ quality

::: {.notes}
Everyone has read reports like this. Point to your own painful
experiences. Push students to write for the CEO, not the professor
- the CEO doesn't want to see your work, they want to know what to
do.
:::

# Part 2 — Chart design {background-color="#7A003C"}

## Tufte's principles

- **Data-ink ratio:** maximise the ratio of data ink to total ink
- **Avoid chartjunk:** decorative elements without informational value
- **Small multiples:** repeat the same chart with different data
- **Sparklines:** small, in-line charts

Simple, clear, informationally dense.

::: {.notes}
Tufte 2001 is the classic. Show a bad example (3D pie chart with
gradients and drop shadows) vs. a good example (simple bar chart
with sorted categories, labelled directly).
:::

## Chart choice by data type

| Question              | Chart                          |
|-----------------------|--------------------------------|
| One value             | Big number or text             |
| Distribution          | Histogram, density, boxplot    |
| Category comparison   | Bar chart (sorted)             |
| Trend over time       | Line chart                     |
| Two continuous vars   | Scatter                        |
| Composition           | Stacked bar (rarely pie)       |
| Geographic            | Map                            |

::: {.notes}
Pie charts are almost always suboptimal. Bar charts are almost
always better. Data-driven design choice, not "let's use a pie for
variety." Show examples.
:::

## Communication tips for charts

- **Label directly** — put labels near the data, not in a legend
- **Sort by value** — not alphabetically
- **Use colour intentionally** — highlight, don't decorate
- **Include units** — always
- **Cite source** — always
- **Include the reference line** — where relevant

::: {.notes}
Show ggplot examples before/after each fix. Students often mimic
Excel defaults; help them see better. `fct_reorder`, `direct.labels`,
`labs(caption=...)` are their friends.
:::

# Part 3 — Quarto reporting {background-color="#7A003C"}

## Beyond the basics

You know how to render `.qmd` to HTML and PDF. New today:

- Tables with `gt` and `kableExtra`
- Cross-references (figure numbers, section links)
- Parameterised reports (same template, different data)
- Multiple outputs from one source (HTML + PDF + slides)

::: {.notes}
Focus on gt and cross-references - the two biggest polish wins.
Parameterisation is advanced; mention as reference.
:::

## Tables with `gt`

```{r}
#| eval: false
library(gt)

survey |>
  group_by(tier) |>
  summarise(
    n = n(),
    mean_sat = mean(satisfaction, na.rm = TRUE),
    sd_sat = sd(satisfaction, na.rm = TRUE)
  ) |>
  gt() |>
  tab_header(title = "Satisfaction by subscription tier") |>
  fmt_number(columns = c(mean_sat, sd_sat), decimals = 2)
```

::: {.notes}
Live-demo. gt produces publication-quality output. kableExtra is
older but still useful especially for LaTeX/PDF output.
:::

## Cross-references

`````markdown
See @fig-satisfaction and @tbl-descriptives.

```{r}
#| label: fig-satisfaction
#| fig-cap: "Satisfaction distribution by tier"
plot(...)
```
`````

Chunk labels starting with `fig-` or `tbl-` enable cross-refs.

::: {.notes}
Quarto's cross-ref system is nicer than R Markdown's. Auto-numbers
figures and tables; auto-links in HTML. Just prefix the label
correctly.
:::

## Full worked example

A short end-to-end example report — executive summary, one finding
with a `gt` table and a cross-referenced figure, built from
`maritime_survey.csv` — is in **`solutions-12.qmd`** in this session's
folder. Use it as a starting template for your own project report.

::: {.notes}
Point teams here if they want a concrete skeleton to adapt rather than
starting from a blank document.
:::

## Rendering to multiple formats

Your project needs both a written report (PDF or HTML) and slides.

Option 1: Two separate `.qmd` files, share a common data-loading `.R`.

Option 2: One `.qmd` with format-specific chunks:

```yaml
format:
  html: default
  pdf: default
  revealjs: default
```

::: {.notes}
Option 1 is simpler for teams. Option 2 is elegant but requires
careful chunk-writing that plays well with all formats. Recommend
Option 1 for the project.
:::

## Parameterised reports

For repeat reports (monthly tracker, per-region breakdown):

```yaml
params:
  region: "Atlantic"
```

Then in R chunks: `params$region` is accessible.

Render for different regions from a bash / R script.

::: {.notes}
Advanced. Not needed for the project but worth showing. In industry,
parameterised reports save hours on repeated deliverables (monthly
brand tracker, per-brand pack, per-region KPI report).
:::

## Break — 15 minutes {background-color="#FDBF57"}

Guest speaker arriving.

::: {.notes}
Meet guest speaker at the door, brief on room layout, mic, and
projector. Introduce them at the start of the guest slot.
:::

# Part 4 — Guest speaker (60 min) {background-color="#7A003C"}

## Guest speaker

Industry MR practitioner (or DeGroote faculty) on communicating research to decision-makers.

Q&A after presentation.

::: {.notes}
Introduce guest by name, title, and organisation. Give 2 sentences
of context on why they were invited. Sit at the back during their
talk. Manage the Q&A - ensure students get to ask, and follow up if
questions are thin.
:::

## Guest topics (typical)

Depending on the speaker:

- Real project walkthrough (a study they ran, what they learned)
- Career path advice
- "Things I wish I'd known coming out of school"
- The state of the MR industry / trends

::: {.notes}
Prep the guest with these possible framings. Encourage them to
include actual project artefacts (redacted where needed) - students
learn from concrete examples.
:::

## Q&A prompts (if the room is quiet)

- "What's the biggest mistake early-career MR analysts make?"
- "How do you handle a client who wants a different conclusion?"
- "What R / Python / SQL skills are most in demand?"
- "How is AI changing the industry?"
- "What's a project you're proud of?"

::: {.notes}
Have these on standby. Students may not have questions immediately -
prime the pump if 30 seconds pass silently.
:::

# Part 5 — Report clinic (30 min) {background-color="#7A003C"}

## Bring your drafts

Teams work on their drafts. Instructor + TA circulate.

Priorities:

- Executive summary present + one page?
- Findings ordered by importance?
- Recommendations map to findings?
- Charts clear + labelled?
- Limitations named?

::: {.notes}
Lightning reviews - 5-10 min per team. Ask each team: "what's the
top-line recommendation?" If they can't say it in one sentence, they
haven't finished thinking. Push for clarity.
:::

## Common report weaknesses

- **No executive summary** or a bloated one
- **Methodology dumping** — 3 pages of sample-size formulas
- **Findings without interpretation** — "The mean is X" instead of "This means X"
- **Recommendations that don't follow** from findings
- **Limitations section too short** (or absent)

::: {.notes}
Catch these live during the clinic. Every team has at least two of
these. Prioritise the two biggest for each team.
:::

# Part 6 — For your project {background-color="#7A003C"}

## Week 13 deliverables

- **Written report** (20%) due start of class Dec. 2
- **Presentation** (10%) delivered Dec. 2
- All group members present
- 15-20 min per team + Q&A

Both submitted / delivered on the same day.

::: {.notes}
Do NOT accept late reports on presentation day - the timing gets
messy. Encourage teams to submit report the evening before.
:::

## Presentation format

- 15-20 min total per team (roughly 15 slides)
- Slide 1: title, team members
- Slides 2-3: problem + objectives
- Slides 4-5: methodology
- Slides 6-12: findings (one per finding)
- Slides 13-14: recommendations
- Slide 15: limitations, thanks, Q&A

Peer + instructor Q&A: 5-7 min per team.

::: {.notes}
Give the format explicitly. Encourage teams to rehearse timing -
running long or short both cost points.
:::

## Wrap-up + Week 13 preview

**Next Wednesday:** presentation day.

- Arrive 10 min early
- Load your slides on the classroom machine (or bring a laptop with HDMI)
- Report submitted by start of class

::: {.notes}
Give presentation order in advance (email by Monday). Random or
sign-up? Announce your policy.
:::

## Thank you — questions?

`abbass44@mcmaster.ca`

::: {.notes}
Stay for report clinic questions. Thank guest speaker again. Confirm
Wk 13 logistics.
:::
