---
title: "Session 12 — Research Reports and Professional Communication"
subtitle: "M731 Marketing Research · Comprehensive Lecture Notes"
author: "Shay Abbasi"
date: "2026-11-25"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    self-contained: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

# The Decision-Oriented Research Report

## Why Report Structure Is a Methodological Choice

The way a research report is structured is not merely a stylistic question — it determines whether the evidence it contains will actually influence decisions. A report that buries its recommendation on page 47 after three pages of methodology and twelve pages of frequency tables may be technically complete but practically useless: the executives who need to act on it will not read far enough to find the actionable content.

Decision-oriented report structure reflects how busy executives process information: they want the answer first, followed by the evidence that supports it, not the reverse. This is the opposite of how academic papers are written (theory → method → results → discussion → implications). Marketing researchers who have trained academically often write executive reports in the academic style, to their professional detriment.

## The Standard Report Structure

A marketing research report for a business audience follows this structure:

**1. Title, author, and date.** Identifies the study, the team, and when the data was collected. The date of data collection, not just the report delivery date, should be visible — data has a shelf life.

**2. Executive summary (maximum one page).** Written last but read first. Should allow a busy executive to understand the business problem, the headline findings, and the recommended action without reading anything else. If your executive summary exceeds one page, you have not finished editing.

**3. Background and objectives.** One to two paragraphs establishing the business context and the specific research questions the study addresses. Connects the research to a real business decision.

**4. Methodology.** One to two pages covering: research design, sample description (who, how many, how recruited), data collection dates, instruments used, and analysis methods. Written in plain language, not statistical jargon. The appropriate level of detail allows a knowledgeable colleague to replicate the study — not a regression equation with every assumption stated.

**5. Findings.** The core content. Three to eight key findings, ordered by importance to the business decision (not by questionnaire order, not by the order the analyses were run). Each finding should have: a one-sentence headline in bold, supporting evidence (a chart or table with numbers), and an interpretive sentence explaining what the finding means in practice.

**6. Recommendations.** What the client should do. Recommendations must map to findings — every recommendation should have a corresponding piece of evidence in Section 5. Recommendations that are not grounded in the study's findings belong in the limitations section as future research directions.

**7. Limitations.** A section that names, rather than hides, the boundaries of the study's conclusions. A well-written limitations section increases, not decreases, the credibility of the report. Common limitations: sampling scope, response rate, single time point, self-report bias, inability to establish causation.

**8. Appendices.** Detailed tables, full statistical output, questionnaire, codebook, R scripts, and any material that supports the report but is too detailed for the main body.

Total report body: 10–15 pages is the appropriate range for a project of M731 scope. Longer is not better — it signals inability to prioritise.

## The Executive Summary Formula

Three paragraphs, one page:

**Paragraph 1 — Situation and complication.** What was the business question, and what tension or problem made the research necessary? One or two sentences.

**Paragraph 2 — Findings.** Two to three headline findings, quantified. Each statement should contain a number. "Maritime Roast Discovery subscribers are 35% more satisfied than Standard subscribers (4.8 vs. 3.5 on a 7-point scale). The primary satisfaction driver is coffee quality, which accounts for 28% of the explained variance in satisfaction. Approximately 22% of Standard-tier subscribers show churn risk indicators."

**Paragraph 3 — Recommendation.** What should the client do? Specific and actionable. "We recommend a Standard-tier product quality improvement programme targeting roast freshness and single-origin sourcing, piloted with a randomised test before full rollout. This addresses the primary satisfaction gap among the highest-risk segment."

If a decision-maker reads only this page, they should be able to act on it.

# The Pyramid Principle and SCQA

## Answer First

Barbara Minto's Pyramid Principle (1987), developed at McKinsey, is the most influential framework for executive business communication. Its core insight: place the answer or recommendation at the top of your communication, then provide supporting arguments, then provide supporting data.

This is the reverse of how most people write (and how academics always write). The instinct is to build up to the conclusion: here is the problem, here is the method, here is the analysis, and here, finally, is the answer. Executives read reports looking for the answer; they do not want to excavate for it.

The pyramid works because the structure matches executive reading behaviour. Executives read the top of the pyramid (the answer), decide if they need to know why (the supporting arguments), and decide if they need the evidence (the supporting data). At each level, they can stop if they have what they need. Most will stop after the first or second level.

## The SCQA Framework

The Situation-Complication-Question-Answer framework applies the Pyramid Principle to executive summaries and presentation openings:

**Situation:** A factual statement of the current state, accepted by both the writer and the reader as true. One sentence. "Maritime Roast currently offers three subscription tiers and serves 12,000 active subscribers across Atlantic Canada."

**Complication:** The change or problem that makes the situation unstable or decision-relevant. One sentence. "Subscriber churn has increased 15% year-over-year, and internal data suggests Standard-tier satisfaction may be driving the attrition."

**Question:** The specific question the decision-maker needs answered to resolve the complication. One sentence, usually implicit rather than explicit. "Should Maritime Roast invest in Standard-tier quality improvements to reduce churn?"

**Answer:** The conclusion of the research, in one or two sentences. "Yes. Standard-tier subscribers who rate coffee quality low are 2.3 times more likely to churn within 90 days. A focused quality improvement pilot is recommended before broad investment."

SCQA creates a natural narrative arc that motivates the research question before answering it, making both the analysis and the recommendation feel inevitable rather than arbitrary.

## The "So What?" Test

For every finding in the Findings section, apply the "so what?" test: does this finding change what the client should do, or is it interesting background information?

Before (raw data): "63% of subscribers rated delivery speed as important."

After ("so what?" applied): "Three in five subscribers care significantly about delivery speed — a potential vulnerability to competitors offering faster shipping, and an opportunity to differentiate on guaranteed 2-day delivery messaging."

The first version reports a number. The second version interprets it for a decision-maker. Every finding in every report should pass the "so what?" test before inclusion. If you cannot articulate what the finding implies for a business decision, either it belongs in an appendix or you have not finished thinking about it.

# Chart Design for Research Reports

## Tufte's Principles

Edward Tufte's principles of data visualization (1983, 2001) remain the best framework for decisions about chart design in research reports:

**Maximise the data-ink ratio.** The ratio of ink devoted to data to total ink in the figure should be as high as possible. Every mark on a chart that is not data should be eliminated or minimised: no decorative borders, no background gradients, no 3D effects, no chartjunk.

**Avoid chartjunk.** Any visual element that does not convey data or aid interpretation is chartjunk: decorative illustrations, gratuitous gridlines, unnecessary legends, empty whitespace with no function.

**Use small multiples.** For comparisons across groups or time periods, a series of small charts using the same scale is more effective than a single complex chart with many series.

**Label directly.** Place labels near the data they describe rather than in a separate legend. Eliminating the legend-lookup step reduces cognitive load.

## Chart Type by Data Type

| Question being answered | Appropriate chart type |
|------------------------|----------------------|
| One headline number | Big-number callout or text |
| Distribution of one variable | Histogram, density, or violin |
| Comparison of groups on one variable | Horizontal bar chart (sorted by value) |
| Trend over time | Line chart with or without area fill |
| Relationship between two continuous variables | Scatter plot |
| Composition as proportion of whole | Stacked bar (rarely pie) |
| Two variables by subgroup | Faceted chart (small multiples) |

Pie charts are almost always suboptimal. Bar charts show the same proportions more accurately (length is more perceptible than angle or area) and with direct value comparison. The rare exception where a pie is appropriate: a single yes/no proportion where the visual emphasis is on the dominant category (e.g., "93% of subscribers renewed").

## Before and After: Making a Chart Work

```{r}
#| label: chart-before

# Before: alphabetical order, no labels, default palette, no title story
library(ggplot2)
library(tidyverse)

survey <- read_csv("data/maritime_survey.csv")

ggplot(survey, aes(x = tier, fill = tier)) +
  geom_bar() +
  theme_gray()
```

```{r}
#| label: chart-after

# After: sorted by value, direct labels, minimal theme, story-first title
survey |>
  count(tier) |>
  mutate(tier = fct_reorder(tier, n)) |>    # sort by count
  ggplot(aes(x = n, y = tier)) +
  geom_col(fill = "#7A003C") +
  geom_text(aes(label = scales::comma(n)),
            hjust = -0.1, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Standard tier is the largest subscriber segment",
    subtitle = "Maritime Roast subscriber count by tier (n = 800)",
    x        = NULL, y = NULL,
    caption  = "Source: Maritime Roast subscriber survey, Sept. 2026"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.x        = element_blank()
  )
```

The improvements from the "before" to the "after" chart: horizontal bars (easier to read tier names); sorted by n (immediately communicates rank); value labels on bars (eliminates axis-lookup); story-first title (tells the reader what to conclude before they examine the chart); source citation (professional credibility); minimal gridlines (higher data-ink ratio).

## The Satisfaction Distribution Chart

```{r}
#| label: violin-boxplot

ggplot(survey, aes(x = tier, y = satisfaction, fill = tier, colour = tier)) +
  geom_violin(alpha = 0.3, draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_boxplot(width = 0.15, alpha = 0.7, outlier.alpha = 0.3) +
  scale_fill_manual(values  = c("#7A003C", "#FDBF57", "#444444")) +
  scale_colour_manual(values = c("#7A003C", "#FDBF57", "#444444")) +
  labs(
    title    = "Discovery subscribers are more satisfied — but distributions overlap",
    subtitle = "Satisfaction (1-7) by subscription tier",
    x        = NULL, y = "Satisfaction score (1-7)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The subtitle is doing important analytical work: "distributions overlap" is an honest statement about effect size that prevents over-interpretation of the group difference. Strong charts often have this dual function — showing the finding and calibrating its magnitude simultaneously.

# Advanced Quarto Reporting Features

## Tables with gt

The `gt` package produces publication-quality HTML tables from a tidyverse pipeline. It is the modern standard for presenting summary statistics, regression output, and descriptive tables in Quarto documents.

```{r}
#| label: gt-table

library(gt)
library(dplyr)

survey |>
  group_by(tier) |>
  summarise(
    n        = n(),
    mean_sat = mean(satisfaction, na.rm = TRUE),
    sd_sat   = sd(satisfaction,   na.rm = TRUE),
    med_sat  = median(satisfaction, na.rm = TRUE),
    pct_high = mean(satisfaction >= 6, na.rm = TRUE)
  ) |>
  gt() |>
  tab_header(
    title    = md("**Satisfaction by Subscription Tier**"),
    subtitle = "Maritime Roast Coffee Co. · Survey 2026"
  ) |>
  cols_label(
    tier     = "Tier",
    n        = "n",
    mean_sat = "Mean",
    sd_sat   = "SD",
    med_sat  = "Median",
    pct_high = "% Rating ≥6"
  ) |>
  fmt_number(columns = c(mean_sat, sd_sat, med_sat), decimals = 2) |>
  fmt_percent(columns = pct_high, decimals = 1) |>
  tab_spanner(
    label   = "Satisfaction (1-7)",
    columns = c(mean_sat, sd_sat, med_sat, pct_high)
  ) |>
  tab_source_note(
    md("*Note:* Satisfaction rated on a 7-point scale (1 = Very Dissatisfied, 7 = Very Satisfied).")
  )
```

The `gt` pipeline reads left to right: start with a data frame, pass to `gt()` to create a gt object, then chain formatting functions. Each function adds a layer of presentation: `tab_header()` for title/subtitle, `cols_label()` for column names, `fmt_number()` and `fmt_percent()` for cell formatting, `tab_spanner()` to group related columns, `tab_source_note()` for footnotes.

## Cross-References in Quarto

Quarto's cross-reference system automatically numbers and links figures and tables when you use the correct label prefix:

- Figures: `#| label: fig-<name>` — referenced as `@fig-<name>`
- Tables: `#| label: tbl-<name>` — referenced as `@tbl-<name>`
- Sections: Section headings referenced as `@sec-<name>` (with `{#sec-<name>}` on the heading)
- Equations: defined in a fenced equation environment, referenced as `@eq-<name>`

```{r}
#| label: fig-sat-by-tier
#| fig-cap: "Mean satisfaction by subscription tier with 95% confidence intervals"
#| fig-width: 7
#| fig-height: 4

# figure-producing code here
```

In the report text: "As shown in @fig-sat-by-tier, Discovery subscribers rate satisfaction significantly higher than Standard subscribers." Quarto resolves this reference to the auto-generated figure number and creates a hyperlink in HTML output.

Cross-references eliminate the error-prone process of manually maintaining figure and table numbers that fall out of sync when sections are reordered or new figures are added.

## Inline R Values: Living Documents

Instead of typing numbers into prose that become stale when data changes, Quarto lets you embed live R expressions directly in text. The syntax is a backtick followed by the letter r, then an R expression, then a closing backtick — for example, writing `round(mean(survey$satisfaction), 2)` as an inline expression means that value is computed and inserted at render time:

> "The mean satisfaction across all tiers was 4.2 (SD = 0.8)."

If the underlying data changes and you re-render, every inline statistic updates automatically, including those in the executive summary. This eliminates the copy-paste errors endemic to Excel-based report pipelines where numbers in prose fall out of sync with numbers in tables.

## Parameterised Reports

Parameterised reports extend the live-document concept to systematic variation across subgroups, time periods, or clients:

```yaml
---
title: "Maritime Roast Monthly Insights Report"
params:
  region: "Atlantic"
  month: "2026-09"
---
```

Within the document, `params$region` and `params$month` are available as R variables. A single render loop then produces one report per region:

```{r}
#| label: parameterised-render

regions <- c("Atlantic", "Ontario", "Quebec")

purrr::walk(regions, ~ quarto::quarto_render(
  input          = "monthly-report.qmd",
  execute_params = list(region = ., month = "2026-09"),
  output_file    = paste0("report-", ., "-2026-09.html")
))
```

This workflow is commercially significant. A monthly brand tracker producing per-region reports for five regions is five renders of one template rather than five separately maintained documents. Agencies that have implemented parameterised reporting have reported savings of hours per reporting cycle.

## The Professional Report YAML

```yaml
---
title: "Consumer Insights Report"
subtitle: "Maritime Roast Coffee Co. — Q3 2026"
author:
  - name: Research Team Name
    affiliation: DeGroote School of Business
date: "November 25, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  message: false
  warning: false
---
```

Key options: `toc-float: true` creates a floating sidebar table of contents; `code-fold: true` hides code by default but allows readers to expand it; `echo: false` produces a clean professional document without code visible; `number-sections: true` auto-numbers all headings; `fig-width` and `fig-height` ensure consistent figure sizing throughout.

# Presenting Research to Different Audiences

## Audience Calibration

The single most common error in research communication is presenting the same report to different audiences. Technical depth, terminology, level of qualification, and recommendation directness should all vary by audience:

| Audience | What they need | Presentation approach |
|---------|---------------|----------------------|
| Board / C-suite | Decision + 3 supporting data points | 5–7 slides; lead with recommendation; no methods |
| Marketing director | Finding + evidence + strategic implication | 12–15 slides; one chart per finding; brief methods |
| Research team | Complete methods + full output + caveats | Full report + appendices; statistical notation acceptable |
| Academic peer reviewer | Theory + design + analysis + limitations + replication | Journal article format; complete reporting |

The board version of your report is not a dumbed-down version — it is a precision instrument designed for a specific decision in a specific context. Creating it from the full report requires removing everything that does not serve the decision, not just shortening the full report.

## The One-Sentence Test

For every finding you plan to present, test whether you can state it in one sentence:

**Pass:** "Standard-tier subscribers are 30% more likely to churn in the first 90 days compared to Signature subscribers."

**Fail:** "There appear to be some differences between the tiers in terms of retention behaviours, particularly in the initial subscription period, which may be worth considering."

The failing version signals incomplete thinking. The one-sentence test forces clarity: what specifically did you find, and what does it imply? If you cannot state a finding in one sentence, you have not finished analysing it.

## Handling Difficult Client Situations

**The client who wants a different result.** A client asks: "Our satisfaction numbers should be higher — can you reanalyse the data?" The professional response:

1. Understand the concern: what specifically did they expect and why?
2. Check for legitimate data issues: is there a specific subgroup or question that might be interpreted differently?
3. Explain the finding: what factors might explain why satisfaction is lower than expected?
4. Do not reanalyse until significance: fishing for a different result by rerunning analyses with different specifications is p-hacking
5. Offer constructive additional analysis: "We could look at which specific items drive the low score, or examine whether recent subscribers rate differently from tenured ones."

**The Q&A question outside your scope.** Someone asks about a subgroup or comparison not in the pre-specified analysis:

1. "We looked at that — here's what we found." (If you did)
2. "Good question. That comparison wasn't pre-specified, so I'd treat any result as exploratory — I can run it and send a follow-up note."
3. "That's outside our sample — we can't address it from this data."

Response 2 is the most common and honest. It maintains methodological integrity while being genuinely helpful.

# Common Report Weaknesses and How to Fix Them

## Methodology Dumping

**Problem:** Three pages detailing sample size calculation formulas, questionnaire development theory, and statistical test mechanics before the findings begin.

**Fix:** Move methodological detail to appendices. The methodology section should be one to two paragraphs covering who was studied, how, and what tests were used — enough to assess credibility, not enough to reconstruct the study.

## Findings Without Interpretation

**Problem:** A table showing mean satisfaction by tier with no accompanying sentence explaining what the numbers mean for the business.

**Fix:** Every table and every chart must have an interpretive sentence in the surrounding text. Numbers without interpretation are raw data, not findings.

## Recommendations That Do Not Follow from Findings

**Problem:** "We recommend Maritime Roast expand into the Ontario market" when the study only measured Atlantic Canada subscribers and did not study Ontario.

**Fix:** Each recommendation should cite the specific finding that justifies it. Recommendations should stay within the evidential scope of the study.

## No Limitations Section

**Problem:** The report presents findings as if they apply universally, without acknowledging the constraints of the sample, the time period, the measurement approach, or the inability to establish causation.

**Fix:** A three-to-five-sentence limitations section naming the specific constraints on interpretation is professionally required. It does not undermine the findings — it contextualises them honestly. Include: sample scope (who was surveyed), temporal scope (when), measurement limitations (self-report, single-item measures), and causal limitations (observational design).

# References and Further Reading

**Minto, B. (1987).** *The Pyramid Principle: Logic in Writing and Thinking.* Pitman. The source of the SCQA framework and the argument for answer-first communication in business contexts.

**Tufte, E.R. (2001).** *The Visual Display of Quantitative Information* (2nd ed.). Graphics Press. The canonical reference on data visualization principles.

**Tufte, E.R. (1983).** *Envisioning Information.* Graphics Press. Introduces small multiples and other techniques for high-density information display.

**Cairo, A. (2016).** *The Truthful Art: Data, Charts, and Maps for Communication.* New Riders. A practical update of Tufte's principles for digital contexts, with extensive worked examples.

**Irizarry, R.A. (2019).** *Introduction to Data Science.* CRC Press (also freely available online). Chapter on data visualization provides excellent worked examples in ggplot2.

**Quarto documentation:** https://quarto.org — the authoritative reference for all Quarto features including parameterised reports, cross-references, and multi-format output.
