---
title: "MBA M731 - Marketing Research: Course Overview"
subtitle: "DeGroote School of Business, McMaster University - Fall 2026"
author: "Shay Abbasi (abbass44@mcmaster.ca)"
date: "2026-09-09"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: false
    code-tools: true
  pdf:
    toc: true
    number-sections: true
    documentclass: article
    pdf-engine: xelatex
execute:
  echo: true
  warning: false
  message: false
---

# Course information

**Course code and title.** BUSADMIN M731 - Marketing Research

**Term.** Fall 2026 (Wednesdays, September 9 - December 2, 2026)

**Location.** Ron Joyce Centre (RJC) 5100, Burlington campus

**Instructor.** Shay Abbasi - `abbass44@mcmaster.ca`
Office hours by appointment (Zoom); please book by email or scheduler in advance.

**Teaching assistant.** TBD

**Course website.** McMaster Avenue to Learn - Homepage BUSADMIN M731: Marketing Research

**Prerequisites.** M650 or equivalent, and I602 or equivalent.

**Credit value.** 3 units. Course elements flagged in the official outline:
Leadership, Global view, Ethics, Numeracy, Participation, Innovation, Group
work, Oral skills, Evidence-based, Guest speaker(s). No final exam.

# Course summary

MBA M731 is a graduate marketing research course that equips students to
**design, execute, and critically evaluate marketing research studies that
inform managerial decisions**. The course treats marketing research as a
decision-support discipline: every technique we cover is motivated by a
business question, and every result is interpreted in terms of managerial
and economic significance, not only statistical significance.

The course is delivered in **thirteen weekly three-hour sessions**: twelve
lecture weeks (Weeks 1-12) followed by a final presentation week (Week 13).
Each lecture week pairs conceptual foundations from the reading list with a
hands-on **R + Quarto** lab. All lectures, labs, and student deliverables
follow a **running case company** - Maritime Roast Coffee Co. - so that
techniques compose across weeks rather than sitting in isolation.

Students leave the course able to specify a research problem, design an
appropriate study (qualitative, survey, experimental, or secondary-data
based), implement the analysis in R with reproducible code, and communicate
findings to a decision-maker.

**Tools.** R (>= 4.3), RStudio / Posit, Quarto (or R Markdown). All lecture
materials, labs, and assignments are distributed as `.qmd` files that
students render locally.

# Course at a glance {#sec-glance}

| Wk | Date        | Lecture topic                                            | R lab / activity                    | Reading anchor           | Assessment              |
|----|-------------|----------------------------------------------------------|-------------------------------------|--------------------------|-------------------------|
| 1  | Sept. 9     | Research process and decision making                     | Getting started with R              | MG Ch. 1-2; CF Ch. 1-2   |                         |
| 2  | Sept. 16    | Secondary data and research designs                      | Data visualization                  | MG (design); CF Ch. 3    |                         |
| 3  | Sept. 23    | Measurement, data collection, sampling                   | Sampling and power in R             | MG; SIL Ch. 6-9          |                         |
| 4  | Sept. 30    | Survey design / measurement methods                      | Cleaning survey exports             | MG; SIL Ch. 8            | **Assignment 1 due**    |
| 5  | Oct. 7      | Qualitative marketing research                           | (Test day; short qual demo)         | MG; MAI Ch. 1, 2, 5, 6   | **Test 1 (L1-4)**       |
| 6  | Oct. 14     | Data analysis and hypothesis testing                     | Descriptives, t/prop/chi-square     | CF Ch. 3, 5, 6           |                         |
| 7  | Oct. 21     | Data analysis and hypothesis testing (cont.)             | ANOVA and post-hoc                  | CF Ch. 5-6               |                         |
| 8  | Oct. 28     | Marketing scales                                         | Reliability and alpha in R          | MG; CJ Ch. 10            | **Assignment 2 due**    |
| 9  | Nov. 4      | Factor analysis                                          | EFA in R                            | CF Ch. 8                 | **Project proposal + questionnaire due** |
| 10 | Nov. 11     | Experimental research                                    | A/B and factorial in R              | MG; CJ Ch. 7             |                         |
| 11 | Nov. 18     | Regression: drivers of marketing outcomes                | `lm()` and diagnostics              | CF Ch. 4, 7, 9           | **Test 2 (L5-10)**      |
| 12 | Nov. 25     | Research reports and presentations                       | Quarto reporting                    | CJ Ch. 14; guest speaker |                         |
| 13 | Dec. 2      | Team presentations                                       | Project presentations               | -                        | **Presentation + report due** |

::: {.callout-note title="How to read this document"}
The compressed schedule above is the fast reference. Full session details -
learning objectives, required and supplementary readings, R content, in-class
labs, ethics threads, project tie-ins, and classic-paper additions - appear
in @sec-sessions.
:::

# Learning outcomes

The official outline lists learning outcomes as `[TBD]` (Outcome 1, 2, 3,
XXXX). Pending the instructor's finalisation, the following outcomes align
with the course description and schedule; they should be reviewed and
revised before the outline is distributed to students.

By the end of the course, students will be able to:

1. Translate a managerial problem into a well-specified marketing research
   problem, and justify the choice of research design (exploratory,
   descriptive, or causal) given the decision context.
2. Design measurement instruments - questionnaires, rating scales, and
   experimental treatments - that satisfy standard reliability and
   validity criteria.
3. Draw and defend appropriate probability and non-probability samples,
   and compute sample-size requirements for common estimation and testing
   problems.
4. Evaluate and plan qualitative marketing research (focus groups, in-depth
   interviews, projective techniques, and observational / digital methods),
   and interpret systematically coded qualitative data.
5. Estimate and interpret core quantitative models used in marketing
   research - descriptive statistics, hypothesis tests, factor analysis,
   experimental designs, and linear and logistic regression - using R.
6. Reason about causal identification in marketing (randomised experiments
   and their threats to internal and external validity).
7. Produce a reproducible, decision-oriented research report using
   Quarto, and defend it in an oral presentation.
8. Apply core research-ethics principles (TCPS-2, PIPEDA, McMaster REB) to
   the design and execution of a marketing research study.

*[Instructor to finalise the learning-outcome list before term start.]*

# Required and optional readings

The official outline specifies the following texts and materials.

## Required

| Key      | Source                                                                                              | Access                                                                                          |
|----------|-----------------------------------------------------------------------------------------------------|-------------------------------------------------------------------------------------------------|
| **A2L**  | Avenue to Learn - registered course content, readings, and case materials                           | `http://avenue.mcmaster.ca` - FREE                                                              |
| **AV**   | Course videos, PowerPoint decks, R scripts, and Excel documents in Avenue                           | Avenue - FREE                                                                                   |
| **CF**   | Chapman, C. & Feit, E. M. (2019). *R for Marketing Research and Analytics*, 2nd ed. Springer.       | McMaster Springer e-book collection - FREE - `https://library.mcmaster.ca/databases/springer-ebook-collection` |
| -        | R / RStudio (Posit) / Quarto or R Markdown                                                          | `https://cran.rstudio.com/` and `https://posit.co/download/rstudio-desktop/` - FREE             |
| **MG**   | McDaniel Jr., C. & Gates, R. (2016). *Marketing Research Essentials*. John Wiley & Sons.            | McMaster Library or Amazon - **not in Sources folder; instructor to circulate PDF or excerpts** |

## Optional

| Key      | Source                                                                                                    | Access                                                                                                          |
|----------|-----------------------------------------------------------------------------------------------------------|-----------------------------------------------------------------------------------------------------------------|
| **CI**   | Churchill, G. A. & Iacobucci, D. *Marketing Research: Methodological Foundations*, 13th ed. (earlier editions acceptable). | Bookstore or Amazon - **not in Sources folder**                                                                 |
| **CDL**  | Chakrapani, C., Deal, K., & Levitin, J. (2020). *Practical Marketing Research: A Guidebook for Marketing Insights*. Standard Research Systems. | Bookstore or Amazon - **not in Sources folder**                                                                 |

## Instructor's working library (used for lecture prep and fallback readings)

These PDFs are already in the course `Sources/` folder. They are not on
the required list per the official outline, but every session below cites
them as a **fallback for MG** in case the McDaniel & Gates PDF is not
circulated in time. If MG is unavailable, the SIL / CJ / KOL fallbacks
substitute directly and the course still runs.

| Key    | Source (author, year)                                                                                        | Role                                                    |
|--------|--------------------------------------------------------------------------------------------------------------|---------------------------------------------------------|
| **SIL**| Silver, Stevens, Wrenn & Loudon (2013). *The Essentials of Marketing Research*, 3rd ed.                     | Primary MG fallback (process, sampling, measurement)    |
| **CJ** | Clow & James (2014). *Essentials of Marketing Research: Putting Research into Practice*. SAGE.               | Primary MG fallback (survey, scales, experiments)       |
| **KOL**| Kolb (2022). *Absolute Essentials of Marketing Research*. Routledge.                                         | Short-form supplementary reading                        |
| **MAI**| Maison (2026). *Qualitative Marketing Research*, 2nd ed. Routledge.                                          | Qualitative spine, Week 5                               |
| **LH** | Lawson & Hancock (2025). *Doing Marketing Research: Alternative Methods*, 2nd ed. Routledge.                 | Advanced qualitative (ZMET, ethnography)                |
| **MR** | Sontakki (2009). *Marketing Research*. Global Media.                                                         | Short treatments of process, design, scaling            |
| **PER**| Percy (Ed.) (1997). *Marketing Research That Pays Off*. Routledge / Haworth.                                 | Case bank for in-class discussion                       |
| **KRI**| Krishna (Ed.) (2010). *Sensory Marketing*. Routledge / Taylor & Francis.                                     | Optional deep-dive (not scheduled)                      |
| **MM** | Michael & Rossides (2025). *AI Powered Insight*. Routledge.                                                  | Optional deep-dive (not scheduled)                      |

**CF is the R spine.** All R code and quantitative walkthroughs draw from
Chapman & Feit's chapter code (per-chapter `.Rmd` decks are in the source
folder). Chapters used: 1, 2, 3, 4, 5, 6, 7, 8, 9 (§9.1-§9.2). CF Ch. 10-14
are not on this term's syllabus.

## Classic methodology papers (optional, graduate-level context)

For students who want the peer-reviewed methodological source for
techniques covered in class, one or two classic papers per week are
listed in @sec-sessions. All are available through McMaster Library
databases. None are required for tests, but they are the papers a
student going on to doctoral work or a senior insights role should
have read.

# Assessment structure

Per the official outline (Components and Weights table, confirmed):

| Component                                        | Weight | Notes                                                                                     |
|--------------------------------------------------|--------|-------------------------------------------------------------------------------------------|
| **Assignment 1** (individual)                    | 10%    | Instructions released Wk 2. Due at start of class in Week 4 (Sept. 30).                   |
| **Assignment 2** (individual)                    | 10%    | Instructions released Wk 6. Due at start of class in Week 8 (Oct. 28).                    |
| **Test 1** - MC and/or written (individual)      | 20%    | Week 5 (Oct. 7). Covers Lectures 1-4.                                                     |
| **Test 2** - MC and/or written (individual)      | 20%    | Week 11 (Nov. 18). Covers Lectures 5-10.                                                  |
| **Major project - Proposal + Questionnaire**     | 10%    | Group deliverable. Due at start of class in Week 9 (Nov. 4).                              |
| **Major project - Written report**               | 20%    | Group deliverable. Due at start of class in Week 13 (Dec. 2).                             |
| **Major project - Presentation**                 | 10%    | Delivered in Week 13 (Dec. 2).                                                            |
| **Total**                                        | 100%   | No final exam.                                                                            |

::: {.callout-important title="Outline ambiguities to resolve before student distribution"}
Three points in the official outline contradict themselves or are
ambiguous. The overview above uses the Components-and-Weights table as
authoritative. Instructor should confirm before students receive the
document.

1. **Test weight.** Components table = 20% each (40% total). Course
   Deliverables prose = 15% each (30% total). Overview uses 20% each.
2. **Test coverage.** Schedule table = Test 1 covers L1-4 and Test 2
   covers L5-10. Course Deliverables prose = "All content introduced in
   the course will potentially be asked for in each test." Overview uses
   the schedule.
3. **Preliminary report double-count.** The prose reads: "There will be
   a preliminary proposal and questionnaire design worth 10%... You will
   have a preliminary report due in the middle of the semester, worth
   10%." These two 10% items must be the same deliverable, or the
   project totals 50%. Overview treats them as one deliverable (proposal
   + questionnaire = the preliminary report) due Week 9.
:::

**Major project.** Groups of **5 or 6 students** pick a real or plausible
marketing decision, choose an appropriate research design, design a
questionnaire, analyse data (secondary or pre-collected), and deliver a
written report plus an in-class presentation. See @sec-project for the
project spec, the running-case template, and the human-subjects scope
declaration.

**Turnitin.** Written work may be submitted through Turnitin.com. Per
McMaster policy, students who do not wish their work to be checked
through the plagiarism-detection service must inform the instructor
before the assignment is due; no penalty applies.

**Course feedback.** An informal course review is conducted with
students by Week 4 to allow time for delivery modifications. Assignment
1 (10%, returned before Week 8) satisfies the McMaster requirement to
return >= 10% of the grade before Week 8.

**McMaster standard calculator** is permitted during examinations. See:
`www.mcmaster.ca/policy/Students-AcademicStudies/UndergraduateExaminationsPolicy.pdf`

**Grade conversion.** Standard MBA scale: A+ 90-100, A 85-89, A- 80-84,
B+ 77-79, B 73-76, B- 70-72, F 00-69.

# Course project and running case {#sec-project}

## The running case: Maritime Roast Coffee Co.

Every lecture, R lab, and in-class exercise uses a single running case
so that techniques compose across the term rather than sitting in
isolation.

**Maritime Roast Coffee Co.** (fictional) is a Halifax-based direct-to-
consumer specialty coffee subscription with roughly 15,000 subscribers
and a small but growing retail presence across Atlantic Canada. The
company is considering:

- Expanding into the Ontario market (new customer acquisition);
- Reducing churn among 6-month subscribers;
- Launching a "single-origin discovery" premium subscription tier;
- Revising subscription pricing.

Each week's lecture, R lab, and datasets return to Maritime Roast so
students see one company's problems addressed by successive techniques.
By Week 12, students have watched an end-to-end research programme
develop.

## The group project

Groups of **5 or 6 students** design a marketing research programme for
a **company and decision of the group's choice** (Maritime Roast may not
be used - the running case is the worked example only). Deliverables:

1. **Proposal + questionnaire** (10%) - Week 9 (Nov. 4).
   - 4-6 pages: business problem, research objectives, design
     justification, sampling plan, draft questionnaire.
2. **Written report** (20%) - Week 13 (Dec. 2).
   - 12-18 pages: full analysis, findings, decision-oriented
     recommendations, limitations. Rendered from `.qmd` source. Must
     include `sessionInfo()`.
3. **Presentation** (10%) - Week 13 (Dec. 2).
   - 15-20 minutes plus Q&A. All group members present.

::: {.callout-warning title="Project data scope (declared)"}
To keep every project deliverable inside the term timeline and outside
the McMaster REB approval window, **project analysis must use one of
the following**:

- **Secondary data only** (StatsCan, Kaggle, government open data,
  syndicated reports, published datasets);
- **Simulated / synthetic data** the group generates from stated
  assumptions;
- **A pre-collected dataset** already approved for use (instructor
  will provide two options if requested).

**Original human-subjects data collection is not permitted for the
term project** because a McMaster REB submission cannot realistically
be completed inside a 13-week term. Groups that want to field a real
questionnaire may do so post-course as an extension; this course
teaches the design and the analysis, not the fielding.
:::

**Team formation.** Teams are formed in Week 1 or 2, mixing background
(work function, industry, R experience, program). Self-selected teams
often disadvantage international students and part-timers; the
instructor will facilitate mixing. Report team-composition concerns to
the instructor by end of Week 3.

# Course datasets {#sec-datasets}

All datasets used across labs are pre-staged in the course A2L page and
mirrored in the shared course R project. External URLs are never
required mid-term. Every dataset ships with a `README.md` naming the
source, licence, size, and week of use.

**Recurring datasets (used across multiple weeks):**

| Dataset                | Origin                          | Rows   | Used in weeks       | Purpose                                             |
|------------------------|---------------------------------|--------|---------------------|-----------------------------------------------------|
| `maritime_survey.csv`  | Instructor-simulated            | ~800   | 4, 6, 7, 8, 9, 11   | Subscriber survey with demographics, brand-attachment battery, satisfaction items, WTP |
| `maritime_transactions.csv` | Instructor-simulated       | ~15,000 | 2, 8, 10           | Six months of subscription orders (tier, order value, churn flag, referral source) |

**One-off datasets:**

| Dataset                | Origin                                   | Rows   | Week | Purpose                                          |
|------------------------|------------------------------------------|--------|------|--------------------------------------------------|
| `statscan_coffee.csv`  | Statistics Canada open data              | -      | 2    | Household coffee spending by province            |
| `kaggle_survey_raw.csv`| Kaggle "Young People Survey" (Sabo, 2016)| 1,010  | 4    | Real messy survey export for cleaning practice   |
| `maritime_focus.txt`   | Instructor-simulated transcripts         | 4 FGs  | 5    | Focus-group transcripts for qualitative demo     |
| `maritime_abtest.csv`  | Instructor-simulated                     | ~2,000 | 10   | Randomised A/B test on subscription upgrade offer|
| `cf-*.csv`             | Chapman & Feit companion                 | varies | 1, 11 | CF chapter walkthroughs (mirrored, not fetched) |

**Reproducibility rule.** Every submission must render from source. The
`renv` package is introduced in Week 1 so groups can lock package
versions during the project. Every submission must include a chunk
producing `sessionInfo()` at the end.

# Assignment briefs (draft) {#sec-assignments}

The two individual assignments below are **draft briefs** for
instructor review. They map to specific learning outcomes and use the
Maritime Roast case.

## Assignment 1 - Research plan (10%)

- **Released:** Week 2 (Sept. 16)
- **Due:** Week 4 (Sept. 30), start of class
- **Format:** Individual. 4-6 pages including one page of secondary-data
  audit. Rendered from `.qmd` to PDF.
- **Learning outcomes assessed:** LO1 (problem translation), LO3
  (sampling logic), LO8 (research ethics).

**Task.** Maritime Roast Coffee Co. is considering entering the Ontario
market. The CEO has asked you to design (not yet execute) a research
plan. Deliver: (a) a one-paragraph business problem statement, (b) 3-5
research objectives, (c) a design justification (exploratory,
descriptive, or causal - and why), (d) a one-page secondary-data audit
of the Canadian coffee market with at least three cited sources, (e) a
sampling plan (frame, method, target n, rationale), and (f) an ethics
paragraph identifying human-subjects, privacy (PIPEDA), and consent
issues that would apply if the plan were executed.

**Rubric.** Problem framing 20% - Design justification 25% - Secondary
data use and citation 20% - Sampling plan 20% - Ethics 15%.

## Assignment 2 - Analytical memo (10%)

- **Released:** Week 6 (Oct. 14)
- **Due:** Week 8 (Oct. 28), start of class
- **Format:** Individual. Rendered `.qmd` (HTML or PDF), 3-5 pages plus
  code appendix. Must render from source.
- **Learning outcomes assessed:** LO2 (measurement), LO5 (quantitative
  models), LO7 (Quarto reproducibility).

**Task.** Using `maritime_survey.csv` (provided), produce an analytical
memo for the CEO addressing whether Ontario prospects differ from
current Atlantic subscribers on stated willingness-to-pay and brand
attachment. Deliver: (a) a brief data-inspection summary (shape,
missingness, dtypes), (b) reliability analysis for the brand-attachment
battery (Cronbach's alpha, item-total correlations, recommended drops),
(c) an appropriate hypothesis test for the WTP difference with effect
size, (d) one publication-quality figure, and (e) a plain-language
recommendation of no more than 200 words.

**Rubric.** R code correctness and clarity 25% - Statistical choices
and effect-size reporting 30% - Interpretation and managerial framing
25% - Reproducibility (renders cleanly, includes `sessionInfo()`) 20%.

# Weekly workload expectations

Per MBA norms (~6-8 hours outside class per 3-hour session), plan for
the following:

| Week  | Reading | R practice | Project/assignment | Total (hrs) | Notes                              |
|-------|---------|------------|--------------------|-------------|------------------------------------|
| 1     | 2       | 3          | 0                  | 5           | Environment setup; lighter start   |
| 2     | 3       | 3          | 2                  | 8           | Assignment 1 begins                |
| 3     | 3       | 3          | 2                  | 8           | Assignment 1                       |
| 4     | 3       | 2          | 4                  | 9           | Assignment 1 finish; project team formed |
| 5     | 4       | 1          | 4                  | 9           | Test 1 prep + project proposal work|
| 6     | 3       | 3          | 3                  | 9           | Assignment 2 begins                |
| 7     | 3       | 3          | 3                  | 9           | Project data planning              |
| 8     | 3       | 3          | 4                  | 10          | Assignment 2 due; questionnaire revise|
| 9     | 3       | 3          | 5                  | 11          | **Project proposal due** (peak week)|
| 10    | 2       | 3          | 4                  | 9           | Project analysis begins            |
| 11    | 4       | 2          | 4                  | 10          | Test 2 prep + project analysis     |
| 12    | 2       | 2          | 6                  | 10          | Report writing                     |
| 13    | 0       | 0          | 6                  | 6           | Presentation prep                  |

Peak workload is Week 9 (project proposal + questionnaire due, after
scales content in Wk 8). Wk 8 (Assignment 2 due + new content) and
Wk 11 (Test 2 + regression) also compress. If the term is genuinely
untenable, raise it at the Week 4 informal course review.

# Session-by-session plan {#sec-sessions}

Every lecture week is Wednesday, 3 hours, in RJC 5100. Default session
shape: ~75-min concept block, ~15-min break, ~75-min R lab, ~15-min
wrap-up and next-week briefing. **Test weeks (5 and 11)** and the
**presentation week (13)** deviate from this shape - see individual
sections.

Lecture topics in bold below are quoted from the official outline. MG
citations are marked `[MG Ch. ?]` because the McDaniel & Gates PDF is
not yet in the Sources folder; each session names a specific SIL/CJ
fallback so the course can run without MG.

## Week 1 (Sept. 9) - Class outline, schedule, the marketing research process and decision making {#sec-week-1}

**R Lab: basic statistics review and getting started with R**

**Learning objectives.**

- Understand course structure, evaluation, expectations, and the
  Maritime Roast running case.
- Define marketing research and locate it within the broader marketing
  decision process.
- Walk through the marketing research process end-to-end.
- Set up a working R + RStudio + Quarto environment; refresh basic
  descriptive statistics; initialise `renv`.

**Required readings.**

- **MG** [Ch. 1-2 on role and process of marketing research - to
  confirm].
- **CF** Ch. 1 "Welcome to R" and Ch. 2 "An Overview of the R
  Language".

**MG fallback (if MG unavailable):** SIL Ch. 1 "Introduction to
Marketing Research" and Ch. 2 "Ethics"; CJ Ch. 1-2.

**Supplementary.**

- **MR** "Introducing Marketing Research"; **KOL** Ch. 1.

**Classic paper (optional).** Kohli & Jaworski (1990), "Market
orientation: the construct, research propositions, and managerial
implications," *Journal of Marketing* 54(2). Grounds marketing
research in market-orientation strategy.

**R content.**

- Packages: base R, `tidyverse`, `knitr`, `quarto`, `renv`.
- Installing R and RStudio; creating an `.Rproj`; anatomy of a `.qmd`
  file; rendering to HTML/PDF.
- Initialising `renv::init()` on the course project.
- CF Chapter 2 walkthrough: vectors, data frames, indexing, reading a
  CSV, basic descriptives.

**In-class lab.**

- Live-code CF Chapter 2 on `maritime_transactions.csv`: dimensions,
  `summary()`, one plot. Students render their first `.qmd` report.

**Ethics thread (10 min).**

- Introduce TCPS-2 CORE (free online Tri-Council certification);
  students register during the week and complete before Week 5.
- PIPEDA in a marketing-research context.

**Industry context (10 min).** What an MR analyst does in industry:
brief -> plan -> field -> analyse -> report. The role of client-side
vs. supplier-side firms; ESOMAR and MRIA-ARIM.

**For your project.** Form teams and begin scoping a decision problem.
Assignment 1 will use the Maritime Roast case as a template.

**Assignment tie-in.** Ungraded diagnostic: verify every student has a
working R + Quarto toolchain before Week 2.

## Week 2 (Sept. 16) - Secondary data and research designs {#sec-week-2}

**R Lab: data visualization**

**Learning objectives.**

- Identify major sources of secondary data (syndicated, government,
  scanner/loyalty, digital) and evaluate their fitness for purpose.
- Distinguish exploratory, descriptive, and causal research designs and
  choose among them for a decision problem.
- Produce publication-quality graphics with `ggplot2`.

**Required readings.**

- **MG** [chapters on secondary data and research design - to confirm].
- **CF** Ch. 3 "Describing Data" (visualisation portions in particular).

**MG fallback:** SIL Ch. 3 "Secondary Data", Ch. 4 "Research Designs:
Exploratory and Qualitative", Ch. 5 "Descriptive and Causal Research";
CJ Ch. 3 "Secondary Data and Research".

**Supplementary.**

- **MR** "Research Design"; **KOL** Ch. 3.
- **PER** Part II Case (instructor selects specific case by term start).

**Classic paper (optional).** Golder (2000), "Historical method in
marketing research with new evidence on long-term market share
stability," *Journal of Marketing Research* 37(2). Model of what
serious secondary-data work looks like.

**R content.**

- Reading external data: `readr::read_csv()`, `readxl::read_excel()`,
  `haven::read_sav()` for SPSS, `jsonlite::fromJSON()` for APIs.
- `ggplot2` grammar: aesthetics, geoms, facets, themes.
- CF Ch. 3 idioms for histogram, boxplot, scatter, and grouped bar.

**In-class lab.**

- Load `statscan_coffee.csv` (pre-staged) and produce three ggplot
  figures answering: (a) how does household coffee spending vary by
  province, (b) has it changed since 2019, (c) does Atlantic Canada
  spend differently from Ontario? Write a one-page memo on the fitness
  of the StatsCan data for a Maritime Roast Ontario-entry decision.

**Ethics thread (5 min).** Ethical use of secondary data: attribution
requirements, terms of use for scraped or syndicated data, distinguishing
"public" from "reusable."

**Industry context (10 min).** The commercial secondary-data landscape:
Nielsen, IRI, Circana, Statista, Euromonitor, Mintel; StatsCan and open
government data as free alternatives.

**For your project.** Do the equivalent audit for your team's chosen
decision. Assignment 1 is due end of Week 4 and asks for exactly this
audit for Maritime Roast.

**Assignment tie-in.** Assignment 1 released; instructions live on A2L.

## Week 3 (Sept. 23) - Measurement, data collection, and sampling {#sec-week-3}

**R Lab: population and sampling statistics**

::: {.callout-warning title="This week combines three usually-separate topics"}
Measurement, data-collection modes, and sampling each often get their
own session. In the Fall 2026 schedule they share one 3-hour block.
Coverage below prioritises **measurement scales** (needed immediately
for questionnaires in Wk 4-6) and **sampling logic** (needed for
Assignment 1); data-collection modes are treated at overview level with
reading assigned for depth.
:::

**Learning objectives.**

- Distinguish nominal, ordinal, interval, and ratio scales and their
  analytic implications.
- Compare primary data-collection modes (mail, phone, in-person,
  online, mobile, panel) on cost, coverage, response bias, and speed.
- Distinguish probability from non-probability sampling and compute
  sample size for standard estimation and testing problems.

**Required readings.**

- **MG** [chapters on measurement, data collection, and sampling - to
  confirm].
- **CF** Ch. 3 (revisited for descriptive/sampling illustrations).

**MG fallback:** SIL Ch. 6 "Measurement", Ch. 7 "Primary Data
Collection", Ch. 9 "Sampling Methods and Sample Size"; CJ Ch. 6, Ch. 8,
Ch. 9.

**Supplementary.**

- **MR** "Process of Data Collection", "Attitude Measurement and
  Scaling"; **KOL** Ch. 4-5.

**Classic paper (optional).** Cochran, W. G. (1977), *Sampling
Techniques*, 3rd ed. (Ch. 1-4 for the theoretical foundation of the
sampling-size formulas the R code below wraps).

**R content.**

- Drawing samples: `dplyr::slice_sample()`, base `sample()`; simple
  random and stratified sampling via grouped `slice_sample()`.
- Sample-size calculations with `pwr::pwr.t.test()`,
  `pwr::pwr.p.test()`, `pwr::pwr.chisq.test()`.
- Weighting is mentioned as a pointer only; `survey::svydesign()` is
  left as a reference for interested students (no live demo).

**In-class lab.**

- Given a target Maritime Roast Ontario prospect population, a
  precision requirement, and an expected effect size, students compute
  required n for (a) estimating a proportion, (b) detecting a two-group
  mean difference, and (c) a stratified design. They then draw a
  simulated sample from a supplied population frame in R and inspect
  covariate balance.

**Ethics thread (10 min).** Sampling ethics: representative frames,
digital-divide exclusion, informed consent at recruitment, incidence
rate and respondent burden. Introduce panel providers (Dynata, Prolific,
MTurk) and quality issues (bots, straight-lining, attention checks).

**Industry context (10 min).** How panel procurement actually works in
practice: incidence rate, cost per complete ($5-25 typical),
respondent-quality flags.

**For your project.** Draft a sampling plan for your team's chosen
decision - target population, frame, method, target n, and a defence.
This is a required section of Assignment 1.

## Week 4 (Sept. 30) - Survey design / measurement methods {#sec-week-4}

**Assignment 1 due (start of class)**

**Learning objectives.**

- Move from constructs to items to a well-flowing questionnaire (order,
  screening, sensitive items, response-bias mitigation).
- Understand common sources of measurement error and how question
  wording and format contribute to them.
- Load, clean, and reverse-code survey exports in R.

**Required readings.**

- **MG** [chapter on questionnaire / survey design - to confirm].
- **CF** Ch. 3 (data-cleaning and recoding patterns).

**MG fallback:** SIL Ch. 8 "Designing the Data-gathering Instrument";
CJ Ch. 11 "Questionnaire Design".

**Supplementary.** **KOL** Ch. 5 "Writing quantitative research
surveys".

**Classic paper (optional).** Krosnick, J. A. (1999), "Survey
research," *Annual Review of Psychology* 50: 537-567. The canonical
survey-methodology review; catalogues response biases and their
mitigations.

**R content.**

- Loading Qualtrics / SurveyMonkey CSV exports; column typing.
- Reverse-coding with `dplyr::mutate(across(...))`; simple straight-
  line and speeder detection.
- Cleaning messy real-world exports with `janitor::clean_names()` and
  `readr` column-type overrides.
- Constructing composite scales from item batteries.

**In-class lab.**

- Load `kaggle_survey_raw.csv` (a deliberately messy real export from
  Kaggle's Young People Survey). Clean it: fix column names, coerce
  types, reverse-code the appropriate items, flag straight-liners,
  produce a clean tibble ready for analysis. Peer-review each team's
  cleaning script.

**Ethics thread (10 min).** Informed consent language, sensitive-item
handling (income, health, illegal behaviour), respondent burden
budgeting, deceptive question framing.

**Industry context (10 min).** Qualtrics as an industry standard; 15-
minute tour of a real Qualtrics project (survey flow, block logic,
distributions, data export).

**For your project.** Draft your team's questionnaire this week. It's
graded next Wednesday (Wk 5) informally and is a Wk 9 project
deliverable. It will be revised in Wk 8 (once scales are covered) - so
this is a draft, not a final.

**Assignment tie-in.** Assignment 1 due at start of session. Assignment
2 preview shared.

## Week 5 (Oct. 7) - Qualitative marketing research {#sec-week-5}

**Test 1** (individual, MC and/or written; covers Lectures 1-4)

::: {.callout-warning title="Session shape deviation"}
This week is not the default 75/15/75/15 split. Approximate shape:
Test 1 administration (~75 min), 15-min break, qualitative concept
block (~60 min), qualitative workshop (~30 min). No standalone R lab.
:::

**Learning objectives.**

- Choose among focus groups, in-depth interviews, projective
  techniques, and observational / ethnographic methods for a given
  research question.
- Understand the theoretical rationale for projective methods
  (unconscious motives, ZMET, System 1/2 thinking).
- Plan and evaluate qualitative data collection systematically (coding
  schemes, thematic analysis, quotes vs. inference).

**Required readings.**

- **MG** [qualitative research chapter - to confirm].
- **MAI** Ch. 1 "Why we need qualitative research methods", Ch. 2 "From
  consciousness to unconsciousness", Ch. 5 "Diversity in qualitative
  methods", Ch. 6 "Projective and enabling techniques".

**MG fallback:** CJ Ch. 4 "Qualitative Research"; SIL Ch. 4.

**Supplementary.**

- **KOL** Ch. 6-7; **LH** Ch. 3 (ZMET) *or* Ch. 9 (ethnography), one of
  the two, student's choice.

**Classic papers (optional).**

- Zaltman, G. (1997), "Rethinking market research: putting people back
  in," *JMR* 34(4). Origin paper for ZMET.
- Netzer, Feldman, Goldenberg & Fresko (2012), "Mining consumer
  dialogues in online forums," *Marketing Science* 31(3). Bridges
  qualitative and quantitative text mining.

**R content (illustrative only).**

- Test 1 occupies most of the class time. Any remaining R content is a
  short qualitative-data demonstration: reading `maritime_focus.txt`
  transcripts with `readtext`, tokenising with `tidytext`, producing a
  topic-frequency table.
- R is illustrative only for qualitative work; the primary tools remain
  moderator guides, coding schemes, and manual/CAQDAS analysis.

**In-class activity.**

- Test 1 administration.
- Focus-group discussion-guide workshop for the remaining class time:
  teams draft a 60-minute guide (warm-up, main sections, projective
  exercise, wrap-up) for a Maritime Roast churn investigation.

**Ethics thread (built into concept block).** Participant vulnerability
in focus groups; group-dynamics coercion; projective methods and the
line between insight and deception; debriefing.

**For your project.** Even if your project uses secondary data,
decision-makers usually want a qualitative "voice-of-customer"
supplement. Draft a 3-question interview guide you would use if you
could conduct 5 interviews for your project.

## Week 6 (Oct. 14) - Data analysis and hypothesis testing {#sec-week-6}

**Learning objectives.**

- Describe distributions in R (measures of centre and spread, skewness,
  outliers).
- Select the right hypothesis test for common marketing questions
  (one-sample, two-sample, paired, one-proportion, two-proportion,
  chi-square independence).
- Interpret p-values, confidence intervals, and effect sizes together;
  never in isolation.

**Required readings.**

- **CF** Ch. 3 "Describing Data" (in full), Ch. 5 "Comparing Groups:
  Tables and Visualizations", Ch. 6 "Comparing Groups: Statistical
  Tests".
- **MG** [data-analysis chapters, part 1 - to confirm].

**MG fallback:** SIL Ch. 11 "Analyzing and Interpreting Data for
Decisions"; CJ Ch. 12 "Fundamental Data Analysis".

**Classic paper (optional).** Simmons, J. P., Nelson, L. D. &
Simonsohn, U. (2011), "False-positive psychology," *Psychological
Science* 22(11). The p-hacking primer every graduate student should
read.

**R content.**

- Descriptives: `summary()`, `psych::describe()`, `dplyr::group_by() +
  summarise()`.
- Group comparisons: `t.test()`, `chisq.test()`, `binom.test()`,
  `prop.test()`, `wilcox.test()` where non-parametrics are warranted.
- Effect sizes via `effectsize::cohens_d()`,
  `effectsize::cramers_v()`.
- Tidying model output with `broom::tidy()` and `broom::glance()`.

**In-class lab.**

- Using `maritime_survey.csv`: descriptives on subscriber demographics;
  test whether stated WTP differs between current subscribers and
  lapsed subscribers; report effect size; write a plain-language
  recommendation for the CEO in <= 150 words.

**Ethics thread (15 min).** P-hacking, HARKing (Hypothesizing After
Results are Known), and selective reporting. How to write a
methods/results section that will withstand replication scrutiny.
Preregistration in commercial contexts (rarely done - discuss why).

**For your project.** Use today's tools on your project data over the
next fortnight; Assignment 2 (due Wk 8) will be a scaled-down version
of what your project analysis will do.

## Week 7 (Oct. 21) - Data analysis and hypothesis testing (continued) {#sec-week-7}

**Learning objectives.**

- Extend Week 6 to more than two groups: one-way and two-way ANOVA and
  their post-hoc tests.
- Read and interpret contingency tables and understand residual
  analysis for chi-square.
- Diagnose assumption violations (normality, equal variance,
  independence) and know robust alternatives.

**Required readings.**

- **CF** Ch. 5 and Ch. 6 revisited (the ANOVA and multi-group
  sections).
- **MG** [data-analysis chapters, part 2 - to confirm].

**MG fallback:** CJ Ch. 13 "Analysis of Differences and Regression"
(the ANOVA / differences half); SIL Ch. 11-12.

**Classic paper (optional).** Wilkinson, L. and the APA Task Force on
Statistical Inference (1999), "Statistical methods in psychology
journals: guidelines and explanations," *American Psychologist* 54(8).
Still the clearest statement of how to report and not report
statistical analyses.

**R content.**

- `aov()` for one-way and two-way ANOVA; `car::Anova()` for Type II/III
  sums of squares.
- Post-hoc tests: `TukeyHSD()`, `pairwise.t.test()` with p-value
  adjustment.
- Non-parametric alternatives: `kruskal.test()`, `friedman.test()`.
- Assumption checks: `car::leveneTest()`, `shapiro.test()`, diagnostic
  plots.

**In-class lab.**

- ANOVA-based analysis on `maritime_survey.csv`: is satisfaction
  different across the three subscription tiers? Check assumptions, run
  ANOVA, run appropriate post-hoc contrasts, report effect sizes,
  write a short managerial interpretation.

**Ethics thread (5 min).** Multiple-comparison inflation and how post-
hoc corrections address it; when a correction is genuinely appropriate
vs. when it hides a p-hacked analysis.

**For your project.** Team analytical work begins in earnest this week.
Data-cleaning should be complete; hypotheses locked; test choices
documented.

## Week 8 (Oct. 28) - Marketing scales {#sec-week-8}

**Assignment 2 due (start of class)**

**Learning objectives.**

- Design and evaluate rating scales (Likert, semantic differential,
  Stapel, constant-sum, MaxDiff) commonly used in marketing.
- Assess measurement reliability (test-retest, internal consistency /
  Cronbach's alpha) and validity (content, construct, criterion).
- Diagnose item-level performance and recommend items to drop or
  revise.

**Required readings.**

- **MG** [scales chapter - to confirm].
- **CJ** Ch. 10 "Marketing Scales".
- **CF** Ch. 4 sections on correlation (motivates inter-item
  correlation).

**MG fallback:** SIL Ch. 6 revisited (measurement fundamentals).

**Supplementary.** **MR** "Attitude Measurement and Scaling";
**KOL** relevant sections of Ch. 5.

**Classic papers (optional).**

- Churchill, G. A. (1979), "A paradigm for developing better measures
  of marketing constructs," *JMR* 16(1). The single most cited paper on
  scale construction.
- Rossiter, J. R. (2002), "The C-OAR-SE procedure for scale
  development in marketing," *IJRM* 19(4). Modern critique and
  alternative to Churchill.

**R content.**

- Simulating multi-item scale data with `MASS::mvrnorm()` on a designed
  covariance matrix (more transparent than a black-box simulator).
- Cronbach's alpha with `psych::alpha()`; item-total correlations;
  item-deletion recommendations.
- Correlation heatmap with `corrplot::corrplot()`.

**In-class lab.**

- Take the brand-attachment battery from `maritime_survey.csv`, map
  items to constructs, compute alpha and item statistics, and
  recommend items to drop or revise. Compare with a simulated 8-item
  battery of known factor structure.

**Ethics thread (5 min).** Culturally appropriate scales and item
translation; the ethics of importing US-developed scales into Canadian
or international populations without validation.

**Assignment tie-in.** Assignment 2 due at start of session.

**For your project.** **Revise the questionnaire you drafted in Wk 4**
using the reliability and validity criteria from today. This revised
questionnaire is the Wk 9 project deliverable.

## Week 9 (Nov. 4) - Factor analysis {#sec-week-9}

**Project proposal + questionnaire due (start of class)**

**Learning objectives.**

- Distinguish principal-component analysis (PCA) from exploratory
  factor analysis (EFA) and know when each is appropriate.
- Assess factorability (KMO, Bartlett's test of sphericity) and choose
  the number of factors (scree, parallel analysis, eigenvalue > 1).
- Rotate factors (varimax, oblimin) and interpret loadings and factor
  scores.
- Briefly contrast EFA with confirmatory factor analysis (CFA) - CFA
  itself is beyond the scope of the term.

**Required readings.**

- **CF** Ch. 8 "Reducing Data Complexity" (full).
- **MG** [multivariate / factor-analysis chapter, if present - to
  confirm].

**MG fallback:** CJ Ch. 13 sections that reference multivariate
context.

**Classic paper (optional).** Fornell, C. & Larcker, D. F. (1981),
"Evaluating structural equation models with unobservable variables and
measurement error," *JMR* 18(1). The convergent- and discriminant-
validity paper cited by every SEM paper since.

**R content.**

- Suitability tests: `psych::KMO()`, `psych::cortest.bartlett()`.
- Factor extraction and rotation: `psych::fa()` with rotation options,
  scree and parallel analysis via `psych::fa.parallel()`.
- PCA via base `prcomp()` and interpretation of loadings.
- Extracting factor scores and using them in downstream analysis
  (previews the regression payoff in Wk 11).

**In-class lab.**

- Run EFA on the multi-item needs battery in `maritime_survey.csv`:
  assess KMO and Bartlett; choose the number of factors via parallel
  analysis; rotate (varimax and oblimin) and interpret; produce a
  publication-style loading table with `gt` or `kableExtra`.

**Ethics thread.** None specific this week; factor analysis raises
interpretive-ethics questions (naming latent factors, cherry-picking
factor structures) covered under Wk 6-7 p-hacking thread.

**For your project.** If your project questionnaire includes a multi-
item battery, this technique will appear in your Wk 13 report. Factor
scores from EFA are the natural input to Wk 11 regression.

## Week 10 (Nov. 11) - Research design: experimental research {#sec-week-10}

**Learning objectives.**

- Distinguish laboratory, field, and quasi-experiments; understand
  internal vs. external validity threats.
- Implement between-subjects and within-subjects designs; reason about
  randomisation, blocking, and factorial structures.
- Analyse A/B tests and factorial experiments in R, and communicate
  what is - and is not - identified causally.

**Required readings.**

- **MG** [experimental research chapter - to confirm].
- **CJ** Ch. 7 "Experimental Research".

**MG fallback:** SIL Ch. 5 (causal research portion).

**Supplementary.**

- **CF** Ch. 5-6 revisited from an experimental-design perspective.
- **PER** any Part IV case for a commercial application of experimental
  logic (instructor confirms specific case by term start).

**Classic paper (optional).** Angrist, J. D. & Pischke, J.-S. (2010),
"The credibility revolution in empirical economics," *Journal of
Economic Perspectives* 24(2). The best short essay on why causal
identification is hard and why experiments matter.

**R content.**

- Randomisation and covariate-balance checks across arms.
- A/B analysis: `t.test()` or `prop.test()` on treatment vs. control,
  with effect sizes.
- 2x2 factorial ANOVA in R; interaction plots via `emmeans` or base
  interaction plots.

**In-class lab.**

- Analyse `maritime_abtest.csv` - a randomised A/B test on a
  subscription-upgrade offer. Descriptives, randomisation check,
  appropriate test, effect size, and a plain-language recommendation.
  Explicit treatment of what is *not* identified (novelty effects,
  selection, cross-contamination).

**Ethics thread (15 min).** Experimental ethics: informed consent in
online A/B testing (or its absence); the Facebook 2014 emotional-
contagion controversy; debrief protocols; when a "test" becomes an
IRB-requiring "experiment"; deception and its limits in projective /
priming studies.

**Industry context (10 min).** A/B testing at scale: Optimizely, Google
Optimize sunset, Amplitude, in-house experimentation platforms; the
overlap between MR and product analytics.

**For your project.** If your project design has any causal component,
this is the framework. Most project designs are descriptive - be
explicit in your report about what causal claims you are and are not
making.

## Week 11 (Nov. 18) - Regression: identifying drivers of marketing outcomes {#sec-week-11}

**Test 2** (individual, MC and/or written; covers Lectures 5-10)

::: {.callout-warning title="Session shape deviation"}
Test 2 (~75 min) occupies the first block. Regression content is
compressed into the remaining ~90 min; the deep treatment relies on
students having done the CF Ch. 4 and Ch. 7 reading in advance. Focus
in-class time on interpretation and diagnostics; the mechanics of
`lm()` are assumed from the pre-reading.
:::

**Learning objectives.**

- Specify, estimate, and interpret linear regression models used to
  identify drivers of marketing outcomes (satisfaction, purchase
  intent, spend).
- Check classical OLS assumptions and diagnose violations
  (heteroscedasticity, multicollinearity, non-linearity, influential
  observations).
- Extend to interactions and standardised coefficients for
  driver-importance interpretation.
- Introduce logistic regression for binary marketing outcomes.

**Required readings (do before class - session is compressed).**

- **CF** Ch. 4 "Relationships Between Continuous Variables", Ch. 7
  "Identifying Drivers of Outcomes: Linear Models", Ch. 9 §9.1-§9.2
  (collinearity and logistic regression).
- **MG** [regression chapter - to confirm].

**MG fallback:** CJ Ch. 13 "Analysis of Differences and Regression";
SIL Ch. 12 "Advanced Data Analysis".

**Supplementary.** **PER** any Part VI case for regression / modelling
in practice (specific case TBD by instructor).

**Classic paper (optional).** Gelman, A. & Hill, J. (2007), *Data
Analysis Using Regression and Multilevel/Hierarchical Models*, Ch. 3.
Best short treatment of regression interpretation in print.

**R content.**

- `lm()` and interpretation: `summary()`, `coef()`, `confint()`,
  `broom::tidy(model, conf.int = TRUE)`.
- Diagnostics: `plot(model)`, `car::vif()`, `lmtest::bptest()`; robust
  standard errors via `sandwich` + `lmtest::coeftest()`.
- Standardised coefficients via `lm.beta::lm.beta()`.
- `glm(..., family = binomial)` for a binary outcome; odds ratios via
  `exp(coef())`.

**In-class lab (compressed by Test 2).**

- Test 2 administration first.
- Short walkthrough on `maritime_survey.csv`: regress satisfaction on a
  small set of drivers (subscription tier, tenure, factor scores from
  Wk 7), diagnose, standardise, interpret for a general manager.

**Ethics thread (5 min).** Causal language in regression writeups:
"driver of" vs. "associated with"; when observational regressions get
sold as causal in client reports.

**For your project.** Regression is often the analytical centrepiece of
a project report. Diagnostics and honest interpretation matter more
than pursuing every possible model.

## Week 12 (Nov. 25) - Research reports and presentations {#sec-week-12}

**Guest speaker**

**Learning objectives.**

- Structure a decision-oriented research report: executive summary,
  findings, evidence, limitations, recommendations.
- Produce a reproducible, presentation-ready deliverable in Quarto
  (HTML, PDF, and slides from a single source).
- Present research findings to a non-technical audience.

**Required readings.**

- **MG** [reporting chapter - to confirm].
- **CJ** Ch. 14 "Research Reports and Presentations".

**MG fallback:** SIL Ch. 13 "The Research Report"; KOL Ch. 9.

**Classic reference (optional).** Tufte, E. R. (2001), *The Visual
Display of Quantitative Information*, 2nd ed. Selected chapters on
chart design.

**R content.**

- Quarto reporting: cross-references, tables via `gt` and
  `kableExtra`, figure captions and numbering, parameterised reports.
- Rendering the same `.qmd` source to HTML, PDF, and PPTX.

**In-class activity.**

- Guest speaker (industry MR practitioner or DeGroote faculty) on
  communicating research to decision-makers.
- Final-project report clinic: teams bring in-progress reports and
  receive peer + instructor feedback on structure and clarity.

**Ethics thread (10 min).** Reporting ethics: honest limitations
sections, avoiding cherry-picking, distinguishing findings from
recommendations, disclosing conflicts of interest.

**For your project.** Report drafts should be substantially complete
this week; the clinic is for revision, not first draft.

## Week 13 (Dec. 2) - Presentations {#sec-week-13}

**Learning objectives.**

- Deliver a professional 15-20 minute research presentation.
- Field methodological, analytical, and managerial questions.
- Critically evaluate other teams' work.

**Required readings.** None; deliverables focus.

**Activity.**

- Team presentations of the major project.
- Peer evaluation and instructor feedback.

**Assessment tie-in.**

- 10% presentation grade; 20% written report due at start of class.

# Ethics and research integrity {#sec-ethics}

Ethics is a flagged course element and is woven through every session
rather than delivered as a standalone unit. Threads by session:

| Week | Ethics thread                                                                 |
|------|-------------------------------------------------------------------------------|
| 1    | Introduction; TCPS-2 CORE registration; PIPEDA basics                         |
| 2    | Secondary-data attribution and terms of use                                   |
| 3    | Sampling ethics; representativeness; digital-divide exclusion                 |
| 4    | Informed consent; sensitive-item handling; respondent burden                  |
| 5    | Qualitative ethics; participant vulnerability; projective methods and deception|
| 6-7  | P-hacking, HARKing, selective reporting, preregistration                      |
| 8    | Culturally appropriate scales; item translation                               |
| 10   | Experimental ethics; consent in A/B testing; deception and debrief            |
| 11   | Causal-language ethics; correlation vs. causation in reports                  |
| 12   | Reporting ethics; honest limitations; conflict-of-interest disclosure         |

**External requirements.**

- **TCPS-2 CORE.** All students complete the Tri-Council's free online
  research-ethics tutorial (`https://tcps2core.ca`) by end of Week 4.
  Certificate uploaded to A2L.
- **PIPEDA.** Canadian federal privacy law governs personal information
  handling. Every project touching Canadian personal data must comply
  even if secondary.
- **McMaster REB.** Not required for this course because the project
  scope is limited to secondary or pre-collected data (see
  @sec-project). Students planning post-course original data collection
  should submit REB paperwork independently.
- **Client confidentiality.** Client-supplied information must be
  treated as confidential during and after the course (per outline).

# Contingencies {#sec-contingencies}

What breaks the course and how it is handled.

| If this happens                              | Then                                                                                                              |
|----------------------------------------------|-------------------------------------------------------------------------------------------------------------------|
| MG PDF is not circulated in time             | Every session lists a specific SIL/CJ fallback in place of MG. Course runs unaffected.                             |
| Guest speaker Wk 12 cancels                  | Instructor delivers a case teardown of a real published MR project + extended report clinic.                       |
| Student misses Test 1 (approved absence)     | Alternate written test administered Week 6 during office hours. Same coverage, same format.                        |
| Student misses Test 2 (approved absence)     | Alternate written test administered during December exam window. DSSAO coordinated.                                |
| Group falls apart mid-term                   | Notify instructor by Wk 6. Instructor reassigns students to remaining groups. Free-rider concerns raised in writing.|
| R install fails at Wk 1                      | Pre-recorded async setup video posted before Wk 1. Wk 1 evening drop-in office hour for installation problems.     |
| Project REB delay                            | Not applicable: project data scope is declared as secondary/pre-collected only.                                    |
| Test 2 + regression runs over Wk 11          | Regression material completes as pre-recorded video posted the same evening. Wk 12 opens with 15-min Q&A.          |
| Instructor cancels a class                   | Async recorded lecture posted to A2L within 48 hours; live make-up Q&A held in the next office hour.               |
| CF dataset URL breaks                        | All datasets mirrored in course repo; labs work offline once repo is cloned.                                       |
| A2L / Turnitin outage                        | Email submission to instructor accepted with same-day timestamp; no penalty.                                       |

# Course policies (summary from official outline)

**Attendance.** Punctuality and attendance are crucial. The instructor
may use Top Hat to record attendance via a 4-digit code displayed in
class. Students without a smartphone should notify the instructor for
an alternative check-in.

**Missed midterm / test.** Do NOT use the online McMaster Student
Absence Form (MSAF); the MBA program does not accept the MSAF. Missed
work >= 10% of the final grade for legitimate reasons requires
documentation submitted to the DSSAO within five working days.
Compensatory activities are set by the instructor. Maximum of three
Petitions for Missed Term Work per academic year. Missed work without
approved reason receives zero.

**Missed final examination.** Not applicable to M731 (no final exam).

**Academic accommodations (SAS).** Students with disabilities activate
accommodations via the mySAS portal within the first three weeks of
term. At the RJC campus, accommodations for tests are coordinated by
the MBA Faculty Office; email `DSBSAS@mcmaster.ca` at least two weeks
before each test.

**RISO (religious/indigenous/spiritual observances).** Submit to DSSAO
within 10 working days of the beginning of term.

**Academic integrity.** McMaster academic integrity policy applies:
`www.mcmaster.ca/academicintegrity`. Turnitin.com may be used;
students may opt out by informing the instructor before submission (no
penalty).

**Anonymous concerns channel.** Students uncomfortable directly
approaching the instructor may send a confidential anonymous email
via `http://mbastudent.degroote.mcmaster.ca/contact/anonymous/`.

**Generative AI.** Students may use generative AI tools for **learning,
reviewing course material, and studying only**. **Use of generative AI
for assessments is prohibited.** See:
`https://provost.mcmaster.ca/office-of-the-provost/generative-artificial-intelligence/`.

**Copyright and recording.** Lectures are copyright-protected. Students
may record for personal study only; recordings may not be distributed.

**Research using human subjects.** Any original research involving
human participants must comply with the Tri-Council Policy Statement
and McMaster REB requirements. This course's project is scoped to avoid
this (see @sec-project). Client-supplied information must be treated as
confidential during and after the course.

**Communication.** Correspondence with instructor and TA must use the
student's McMaster email account.

# Software setup

**R.** Version 4.3 or later. Install from CRAN.

**RStudio / Posit.** Latest release.

**Quarto.** Version 1.4 or later, from `https://quarto.org`.

**LaTeX (for PDF output).** Install TinyTeX via `quarto install
tinytex` after Quarto is installed.

**R packages.** Install on Day 1; the CF chapter decks list additional
dependencies as needed.

```{r}
#| eval: false
install.packages(c(
  # Core tidyverse and reporting
  "tidyverse", "readr", "readxl", "haven", "jsonlite",
  "knitr", "rmarkdown", "gt", "kableExtra", "quarto",
  "janitor", "broom",

  # Reproducibility
  "renv",

  # Statistics used across sessions
  "psych", "car", "lmtest", "sandwich", "effectsize", "pwr",
  "corrplot", "emmeans", "lm.beta",

  # Qualitative / text (illustrative in Week 5)
  "tidytext", "readtext",

  # CF-companion packages used in mapped chapters
  "MASS", "RColorBrewer"
))
```

**Environment lock.** Initialise `renv::init()` inside your course
project in Week 1 and `renv::snapshot()` before submitting anything.
Every submission must render from source and include a `sessionInfo()`
chunk at the end:

````
```{r}
sessionInfo()
```
````

# Notes and flagged items for instructor review

The following items are flagged for review before the outline is
finalised and distributed to students.

**Outline ambiguities (also flagged in Assessment section):**

- **Test weight contradiction.** Components table says 20% each; Course
  Deliverables prose says 15% each. Overview uses 20% each.
- **Test coverage contradiction.** Schedule specifies Test 1 = L1-4 and
  Test 2 = L5-10; Course Deliverables prose says both tests cover all
  content. Overview uses the schedule.
- **Preliminary report double-count.** Overview treats "proposal +
  questionnaire" and "preliminary report" as the same 10% deliverable
  due Week 7 (not two separate 10% items, which would total 50% for
  the project).

**Placeholders in the official outline:**

- **Learning outcomes** are `[TBD]` (Outcome 1-3, XXXX). Eight draft
  outcomes are proposed above.
- **Course objective / description**: `XX Brief Description XX` and
  `Full Course Description` are placeholders. The overview infers from
  the schedule.
- **Office hours** listed as `XX am/pm (Over Zoom)`.
- **Teaching assistant** listed as `TBD`.
- **Course-outline typo**: outline reads "Dava Visualization" for Wk 2
  R lab; overview silently corrects to "data visualization."

**Overview-level choices to confirm:**

- **Project data scope** declared as secondary / pre-collected /
  simulated only, to keep the project inside the term without REB
  paperwork. Confirm this fits the instructor's vision; if original
  data collection is required, project timeline and REB workflow need
  to be built in from Wk 1.
- **Project proposal + questionnaire due Wk 9** (not Wk 8) to give the
  Wk 8 scales content time to feed back into the questionnaire before
  submission.
- **Running case: Maritime Roast Coffee Co.** fictional Halifax-based
  D2C coffee subscription. Instructor should confirm this case fits or
  substitute a preferred one; all datasets, labs, and worked examples
  key off this choice.
- **PER case selection** committed at Part level only (Part II Wk 2,
  Part IV Wk 10, Part VI Wk 11). Specific case numbers to be chosen
  before term start.
- **MG chapter mappings** are placeholders (`[MG Ch. ?]`) throughout;
  every session names a specific SIL/CJ fallback that can substitute
  if MG remains unavailable.
- **Assignment briefs** in @sec-assignments are drafts requiring
  instructor approval.
- **Ontario-vs-Atlantic subscriber comparison** is baked into
  Assignment 2 and the Wk 6 lab; this assumes `maritime_survey.csv`
  will include a `region` variable. Confirm when the dataset is
  generated.

**Sequencing note (design choice, not error).** The Fall 2026 outline
places qualitative (Wk 5) and analytical techniques (Wk 6-7 hypothesis
testing) *before* scales (Wk 8) and factor analysis (Wk 9). This is
atypical - measurement (scales / factor analysis) usually sits together
in a block that precedes analysis. Overview respects the outline
sequence but the questionnaire timeline reflects the friction: Wk 4
first draft, Wk 8 revision after scales, Wk 9 submission.

**Not scheduled this term (kept as supplementary references only).**
CF Ch. 10 (CFA), Ch. 11 (segmentation / clustering), Ch. 12
(association rules), Ch. 13 (choice modelling / conjoint), Ch. 14
(Markov chains). Sensory marketing (KRI) and AI-in-MR (MM) are not
scheduled. All remain useful for advanced final-project directions.
