---
title: "Session 10 — Experimental Research"
subtitle: "M731 Week 10 · November 11, 2026"
author: "Shay Abbasi"
date: "2026-11-11"
---

## Session 10 {background-color="#7A003C"}

Experimental research: causation, not just correlation

::: {.notes}
Big conceptual shift today - from association (Wks 6-7) to
causation. Frame at the start: everything we've done was descriptive
or inferential about associations. Today we ask "does X cause Y?"
:::

## Today's agenda

1. Correlation vs. causation — the setup
2. Lab, field, and quasi-experiments
3. Internal vs. external validity
4. Randomisation and blocking
5. Factorial designs and interactions
6. **Break (15 min)**
7. R lab: A/B analysis on `maritime_abtest.csv`
8. Ethics: consent in online A/B testing
9. Industry: A/B testing platforms
10. Wrap-up

::: {.notes}
Ethics thread is 15 min today - the Facebook 2014 emotional-contagion
case is the anchor discussion. Prepare a slide with the details.
:::

## Learning objectives

By end of session 10 you will be able to:

- Distinguish lab, field, and quasi-experiments
- Understand internal vs. external validity threats
- Design between-subjects and within-subjects experiments
- Analyse A/B tests and 2×2 factorials in R
- Explicitly reason about what is (and isn't) identified causally

::: {.notes}
"What isn't identified" is a critical LO. Every causal claim has
threats. Naming them honestly is graduate-level thinking.
:::

# Part 1 — Correlation vs. causation {background-color="#7A003C"}

## The classic problem

Observation: customers who receive our marketing email spend 20% more than those who don't.

Does the email *cause* higher spend? Or:

- Loyal customers were more likely to subscribe to the email list
- Higher-spending customers had different attributes to begin with
- Something else caused both

Without randomisation, you can't separate these.

::: {.notes}
This example is real and recurring in marketing. The "$20% lift from
email" is almost always inflated by selection. Randomised holdout
tests reveal the true effect (typically much smaller).
:::

## The counterfactual

Causal question: what would have happened to person A if they had *not* received the email?

Fundamental problem: you never observe both counterfactuals for the same person at the same time.

**Randomisation** creates comparable groups so the observed difference approximates the counterfactual difference.

::: {.notes}
Rubin causal model / potential outcomes framework. Don't dwell -
students who go on will encounter this in more depth. The takeaway:
random assignment is the tool for causal identification.
:::

## When correlation is enough

- Descriptive research (who's spending most?)
- Predictive analytics (who's likely to churn?)
- Forecasting (what will Q4 look like?)

Correlation is fine when the goal is prediction, not intervention.

::: {.notes}
Machine-learning models are correlational. That's fine - they're
built to predict, not intervene. A predictive model of churn tells
you "these customers are at risk" - not "if you take action X, churn
will change by Y."
:::

## When causation matters

- **Any decision to act:** launch, spend, price, redesign
- **Any counterfactual claim:** "if we hadn't done X, we would have Y"
- **Any policy intervention:** government, HR, health

Marketing decisions almost always require causal thinking.

::: {.notes}
This is why experiments matter. Every ad-test, price-test, offer-test,
UX-test in industry is asking a causal question. Not doing an
experiment doesn't remove the causal question - it just leaves it
unanswered.
:::

# Part 2 — Types of experiments {background-color="#7A003C"}

## Laboratory experiments

- Controlled setting (usually a lab or research facility)
- High internal validity (you control everything)
- Low external validity (does behaviour generalise to the real world?)

Common in academic MR; rare in commercial.

::: {.notes}
Academic consumer-research labs run lab experiments constantly.
Commercial MR uses field or quasi-experiments because the market
context matters. Both have their place.
:::

## Field experiments

- Real-world setting (website, store, mailing)
- Real subjects, real behaviour, real stakes
- Better external validity than lab
- Harder to control confounds

The A/B test is the paradigm field experiment.

::: {.notes}
Field experiments are the dominant commercial paradigm now. Google,
Meta, Amazon run thousands of A/B tests concurrently. Airbnb, Netflix,
Shopify are known for their experimentation cultures.
:::

## Quasi-experiments

- Some form of manipulation, but without full random assignment
- Uses "natural experiments" or non-random treatment allocation
- Requires stronger assumptions for causal claims

Examples: DID, RDD, IV, synthetic control.

::: {.notes}
Won't cover these in depth - Wk 10 focuses on true randomised
experiments. Mention that econometrics offers a rich toolkit for
observational causal inference. Interested students should take an
econometrics elective.
:::

# Part 3 — Validity {background-color="#7A003C"}

## Internal validity

Do the results of the study reflect the causal effect within the study?

**Threats:**

- Selection (groups differ before treatment)
- History (external events during the study)
- Maturation (subjects change over time)
- Testing / instrumentation (measurement changes)
- Attrition (differential drop-out)
- Regression to the mean

::: {.notes}
Cook & Campbell 1979 lists these classically. Randomisation defends
against selection - the biggest threat. Others require thoughtful
design.
:::

## External validity

Do results generalise beyond the study?

**Threats:**

- Non-representative sample
- Artificial setting
- Novelty / Hawthorne effect
- Specific stimuli tested

Every experiment sacrifices some external validity for internal validity.

::: {.notes}
The trade-off is real. A tightly-controlled lab study has high
internal validity but the effect may not survive in the wild. A
messy field test has more external validity but confounds. Design
depends on the decision at stake.
:::

## Between-subjects vs. within-subjects

- **Between:** each subject gets one treatment; compare across
- **Within:** each subject gets all treatments; compare within

Within is more powerful (same-person control) but has order effects.

::: {.notes}
Within-subjects requires counter-balancing to control for order
effects. Latin squares are the classical design. For online A/B tests
this is impractical - between-subjects is default.
:::

## Blocking

Stratify random assignment by a variable known to affect the outcome.

Example: block on region (assign 50/50 within each region separately).

Reduces variance → more powerful test.

::: {.notes}
Blocking is under-used in practice. If you know region matters,
block on it. Then treatment effect estimates are more precise.
Stratified randomisation for e-commerce A/B tests should block on
prior spend.
:::

## Factorial designs

Manipulate multiple factors simultaneously:

- 2×2 factorial: 2 levels of factor A × 2 levels of factor B → 4 groups
- Examines main effects + interaction

More efficient than testing one factor at a time.

::: {.notes}
Classic MR example: 2×2 factorial of price (low/high) × endorsement
(present/absent) on purchase intent. Four cells, main effects of each,
and the interaction. All from n=200-400 total.
:::

## Break — 15 minutes {background-color="#FDBF57"}

::: {.notes}
Have `maritime_abtest.csv` loaded. Check that `treatment` and
`upgraded` columns are present.
:::

# Part 4 — R lab: A/B analysis {background-color="#7A003C"}

## The dataset

`maritime_abtest.csv`: ~2,000 subscribers randomised to see either the current subscription-upgrade offer (control) or a new offer (treatment).

Outcome: did they upgrade in the following 30 days? (binary)

::: {.notes}
Frame as a decision: Maritime Roast is choosing between two offers.
Data comes from a randomised test. What should the CEO do?
:::

## First step: descriptives + randomisation check

```{r}
#| eval: false
library(tidyverse)

ab <- read_csv("data/maritime_abtest.csv")

# Overall structure
count(ab, treatment)

# Randomisation check: do covariates balance across arms?
ab |>
  group_by(treatment) |>
  summarise(
    n = n(),
    mean_age = mean(age),
    prop_female = mean(gender == "F"),
    mean_tenure = mean(tenure_months)
  )
```

::: {.notes}
Live-code. If randomisation worked, covariates should be balanced
across arms. Any large imbalance is a red flag - investigate. In this
dataset, randomisation should be clean (it's simulated).
:::

## Two-proportion test

```{r}
#| eval: false
ab |>
  group_by(treatment) |>
  summarise(
    n = n(),
    upgraded = sum(upgraded),
    upgrade_rate = mean(upgraded)
  )

prop.test(
  x = c(sum(ab$upgraded[ab$treatment == "control"]),
        sum(ab$upgraded[ab$treatment == "new_offer"])),
  n = c(sum(ab$treatment == "control"),
        sum(ab$treatment == "new_offer"))
)
```

Reports difference, CI, p-value.

::: {.notes}
Live-code. Interpret: control has X% upgrade rate, treatment has Y%.
Difference of Z pp, 95% CI [L, U]. Effect size = difference expressed
as absolute or relative.
:::

## Relative vs. absolute effect

- **Absolute lift:** treatment rate - control rate (5 pp)
- **Relative lift:** absolute / control (50% lift on a 10% baseline)

Both matter. CEOs often want relative; statisticians prefer absolute.

::: {.notes}
Report both. "Upgrade rate rose from 10% to 15%, a 5 percentage-point
absolute lift (50% relative)." Ambiguity between absolute and relative
is a common source of miscommunication.
:::

## Regression-based analysis

Adjust for covariates for more precision:

```{r}
#| eval: false
fit <- glm(upgraded ~ treatment + age + tenure_months + gender,
           family = binomial, data = ab)
summary(fit)
exp(coef(fit))  # odds ratios
```

Gives you the treatment effect controlling for covariates.

::: {.notes}
Logistic regression is the tool for binary outcomes. Adjusting for
pre-treatment covariates increases power. In a well-randomised
experiment, the estimate should be similar to the unadjusted -
covariates just reduce variance.
:::

## Factorial designs in R

If the design had also varied "message tone" (formal vs. casual):

```{r}
#| eval: false
fit2 <- aov(upgraded ~ offer * tone, data = ab)
Anova(fit2, type = 2)

library(emmeans)
emm <- emmeans(fit2, ~ offer | tone)
plot(emm)
```

::: {.notes}
Factorials let you see interactions - does the new offer work better
with casual tone? Only informative if the design was factorial from
the start. Post-hoc splitting a two-arm test into subgroups is
p-hacking.
:::

## In-lab exercise

Using `maritime_abtest.csv`:

1. Descriptives by arm
2. Randomisation check (do covariates balance?)
3. Two-proportion test on the upgrade rate
4. Effect size and 95% CI
5. Adjusted analysis via logistic regression
6. Plain-language recommendation for the CEO (150 words) including one **threat to identification** to flag

A fuller worked version of this A/B analysis is in
**`solutions-10.qmd`** in this session's folder.

::: {.notes}
Give 40 min. The "threat to identification" bullet is the
graduate-level rigour move - forces students to name a specific
threat, not just claim causation.
:::

# Part 5 — Ethics thread (15 min) {background-color="#7A003C"}

## Facebook 2014 emotional-contagion study

- FB ran a covert experiment on 689,003 users
- Manipulated the emotional tone of news feed content
- Measured whether users' own posts became more positive or negative
- **Published in PNAS. Consent implicit via ToS.**
- Massive public backlash

::: {.notes}
Kramer, Guillory & Hancock 2014. The methodological paper is
straightforward - the ethical firestorm was about consent. FB argued
its ToS covered research; ethicists disagreed. This case shaped
industry thinking.
:::

## What went wrong ethically

- No explicit informed consent for the experiment
- No debriefing after
- Potential harm (emotional manipulation) not disclosed
- Vulnerable populations (those with depression) not screened out
- Academic co-authors published without IRB scrutiny

::: {.notes}
Not just "consent" - the whole informed-consent framework was
short-circuited. Users didn't know they were in a study. They
couldn't opt out. Debriefing didn't happen. IRB review was minimal.
:::

## The industry response

- Facebook created an internal ethics review process
- Many tech companies now have internal IRB-like boards
- Academic-industry collaborations require IRB approval for both sides
- "A/B testing" is legally distinct from "experimentation" (murky)

::: {.notes}
The murky legal distinction is important. If you're "optimising the
product," some argue no consent is needed. If you're "doing research,"
consent is required. Same technical manipulation, different framing.
Ethically the distinction is dubious.
:::

## Where the line sits

For your career:

- Optimising a product feature at your company: internal ethics review, no formal IRB usually
- Publishing academic research using company data: full IRB required
- Deceptive designs (concealment, misleading instructions): debrief required
- Vulnerable populations (children, medical): heightened protection

::: {.notes}
Practical guidance. If in doubt, ask an ethics review. McMaster REB
exists exactly for this reason. Better to over-review than under.
:::

# Part 6 — Industry {background-color="#7A003C"}

## A/B testing platforms

- **Optimizely** — enterprise, mature
- **Google Optimize** — sunset 2023 (replaced by GA4 experiments, less capable)
- **VWO (Visual Website Optimizer)** — mid-market
- **Amplitude Experiment** — product analytics + experimentation
- **Split.io, LaunchDarkly** — feature flags with experimentation
- **In-house platforms** — big companies (Netflix, Uber, Airbnb) build their own

Choice depends on integration with your product stack.

::: {.notes}
Enterprise clients often use custom internal platforms. Startups use
VWO or Amplitude. Middle-market a mix. Ask if students have used any
of these.
:::

## MR and product analytics — the overlap

- Traditional MR: pre-launch, custom studies, structured methodology
- Product analytics: post-launch, dashboards, always-on
- **Experimentation bridges both** — designed studies + always-on data

The insights profession is converging.

::: {.notes}
Job titles reflect this - "Consumer Insights" and "Product Insights"
increasingly overlap. Career paths cross. R + SQL + Python is the
skill triangle for the modern insights analyst.
:::

# Part 7 — For your project {background-color="#7A003C"}

## Does your project have a causal component?

If yes:

- Be explicit about the causal claim
- Name identification threats in your report
- Consider whether an experiment could have answered the question

If no (most projects are descriptive):

- Use associative language ("associated with", not "drives")
- Don't over-claim in the executive summary

::: {.notes}
Most course projects are descriptive - that's fine. Push students to
use appropriately hedged language. Report writing quality separates
strong from mediocre projects.
:::

## Week 11 preview

- **Concept:** Regression - identifying drivers of outcomes
- **Test 2** — covers Lectures 5-10
- **R lab compressed** — regression walkthrough post-test

Prep reading: CF Ch. 4, 7, 9 §9.1-9.2. Read *before* class since session is compressed.

::: {.notes}
Warn: Wk 11 is compressed by Test 2. Pre-reading is essential -
the R lab time is short. Test 2 review sheet on A2L by Friday.
:::

## Thank you — questions?

`abbass44@mcmaster.ca`

::: {.notes}
Stay for questions on Test 2 prep and project analysis plans.
:::
