---
title: "Solutions — Session 10: A/B Test on the Subscription-Upgrade Offer"
subtitle: "M731 · Full worked walkthrough (companion to session-10.qmd)"
author: "Shay Abbasi"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Purpose

Full worked version of the Session 10 in-lab exercise: randomisation
checks, a two-proportion test, effect sizes, and a covariate-adjusted
logistic regression on the subscription-upgrade A/B test.

## Setup

```{r}
#| label: setup
library(tidyverse)

ab <- read_csv("data/maritime_abtest.csv")
glimpse(ab)
count(ab, treatment)
```

## Step 1 — Randomisation check

```{r}
#| label: randomisation-check
ab |>
  group_by(treatment) |>
  summarise(
    n = n(),
    mean_age = mean(age),
    prop_female = mean(gender == "F"),
    mean_tenure = mean(tenure_months),
    .groups = "drop"
  )
```

**Interpretation.** Age, gender split, and tenure should look
essentially identical across arms — this dataset was generated with
clean random assignment, so any imbalance you see is sampling noise,
not a design flaw. In real data, a meaningful imbalance here would be
a red flag worth investigating before trusting the treatment-effect
estimate.

## Step 2 — Two-proportion test

```{r}
#| label: proportion-test
arm_summary <- ab |>
  group_by(treatment) |>
  summarise(n = n(), upgraded = sum(upgraded), upgrade_rate = mean(upgraded), .groups = "drop")

arm_summary

prop.test(
  x = arm_summary$upgraded,
  n = arm_summary$n
)
```

**Interpretation.** The confidence interval on the difference should
exclude zero, with the new offer showing a higher upgrade rate. Report
both:

```{r}
#| label: lift
control_rate <- arm_summary$upgrade_rate[arm_summary$treatment == "control"]
treat_rate   <- arm_summary$upgrade_rate[arm_summary$treatment == "new_offer"]

abs_lift <- treat_rate - control_rate
rel_lift <- abs_lift / control_rate

cat("Absolute lift:", round(100 * abs_lift, 1), "pp\n")
cat("Relative lift:", round(100 * rel_lift, 1), "%\n")
```

**Reporting sentence.** The new offer lifted the upgrade rate from
about `r scales::percent(control_rate, accuracy=0.1)` to
`r scales::percent(treat_rate, accuracy=0.1)` — an absolute lift of
roughly `r round(100*abs_lift,1)` percentage points
(`r round(100*rel_lift,0)`% relative), a statistically significant
and managerially meaningful improvement.

## Step 3 — Covariate-adjusted logistic regression

```{r}
#| label: logistic
fit <- glm(upgraded ~ treatment + age + tenure_months + gender, family = binomial, data = ab)
summary(fit)
exp(coef(fit))
```

**Interpretation.** The adjusted treatment odds ratio should closely
match the unadjusted proportion-test result — expected in a
well-randomised experiment, where covariates mainly tighten precision
rather than change the point estimate. If the adjusted and unadjusted
estimates diverged substantially, that itself would be a signal to
recheck randomisation.

## Step 4 — CEO recommendation (150 words) + one threat to flag

> The new upgrade offer increased the 30-day upgrade rate by roughly
> [X] percentage points ([Y]% relative) relative to the current offer,
> a difference unlikely to be due to chance (p < .05) and adjusted
> estimates from a covariate-controlled logistic regression confirm the
> same direction and magnitude. **Recommendation**: roll out the new
> offer broadly. **Threat to identification to flag**: this test
> measures a 30-day upgrade outcome only — it does not tell us whether
> subscribers who upgraded under the new offer are more likely to
> downgrade or churn later (a novelty effect). Before a permanent
> rollout, extend monitoring to 90-180 days to confirm the lift is
> durable, not just a short-term response to a new offer.

## Take-away for students

A clean randomised experiment identifies the *causal* effect of the
offer on 30-day upgrade — but "identified" doesn't mean
"unconditionally actionable." Always name at least one thing the
experiment's design can't tell you.
