---
title: "Session 3 — Measurement, Data Collection, Sampling"
subtitle: "M731 Week 3 · September 23, 2026"
author: "Shay Abbasi"
date: "2026-09-23"
---

## Session 3 {background-color="#7A003C"}

Measurement · Data collection · Sampling

Three topics, one 3-hour block.

::: {.notes}
Be honest with the room: this session combines three usually-separate
topics because that's how the outline sequences them. We'll go wide
not deep on data-collection modes (they'll read the depth) and prioritise
measurement scales and sampling calculations.
:::

## Today's agenda

1. Recap + housekeeping
2. Measurement scales (nominal / ordinal / interval / ratio)
3. Data collection modes (overview level)
4. Sampling: probability vs non-probability
5. Sample-size calculations
6. **Break (15 min)**
7. R lab: sampling and power in R
8. Ethics: representativeness, digital divide, panel quality
9. Industry: how panels actually work
10. Wrap-up

::: {.notes}
Keep the concept block moving - three topics in one session means
pace matters. Leave a few minutes at the end for open questions.
:::

## Learning objectives

By end of session 3 you will be able to:

- Distinguish nominal / ordinal / interval / ratio scales and their analytic implications
- Compare primary data-collection modes on cost, coverage, response bias, and speed
- Distinguish probability from non-probability sampling
- Compute sample sizes for standard estimation and testing problems

::: {.notes}
Objective 4 is the R lab focus. Objectives 1-3 are the concept block.
:::

# Part 1 — Measurement {background-color="#7A003C"}

## Why measurement matters

- Analysis is only as good as measurement
- Wrong scale = wrong statistic = wrong conclusion
- Measurement error is silent — you often can't detect it in the analysis

Get this right at design time, not analysis time.

::: {.notes}
Illustrate: computing a mean of an ordinal variable ("How satisfied
are you? 1-5") is technically wrong. Everyone does it. But the
gymnastics we go through when a client asks "why is the mean higher
this quarter" are avoidable if you understood scale types at
questionnaire design time.
:::

## The four levels of measurement

| Level    | Distinguishes | Order | Equal intervals | Meaningful zero |
|----------|---------------|-------|-----------------|-----------------|
| Nominal  | Categories    | No    | No              | No              |
| Ordinal  | Rank order    | Yes   | No              | No              |
| Interval | Distances     | Yes   | Yes             | No              |
| Ratio    | Multiples     | Yes   | Yes             | Yes             |

::: {.notes}
Stevens (1946) framework. Some statisticians (Velleman) have argued
these are more of a spectrum than a hard hierarchy. For practical
purposes, the framework is fine.
:::

## Examples

- **Nominal:** province of residence, subscription tier name, brand chosen
- **Ordinal:** ranking of brands, "small/medium/large", Likert (arguably)
- **Interval:** temperature (°C), calendar year, IQ score
- **Ratio:** age, income, purchase amount, time to purchase

Nominal + ordinal = categorical. Interval + ratio = continuous.

::: {.notes}
Likert is famously contested - some treat as ordinal, some as
interval. Most commercial MR treats Likert as interval and computes
means; academic MR splits. We'll be pragmatic - treat as interval when
convenient, disclose the assumption.
:::

## Legal statistics by scale

| Scale    | Central tendency | Variability | Association |
|----------|------------------|-------------|-------------|
| Nominal  | Mode             | Count       | Chi-square  |
| Ordinal  | Median           | Range       | Spearman ρ  |
| Interval | Mean             | SD          | Pearson r   |
| Ratio    | Mean, geo mean   | SD, CV      | Pearson r   |

Higher scales inherit lower-scale statistics.

::: {.notes}
Table is a reference. Ratio can use geometric mean (useful for
percentages and growth rates); interval can't. Coefficient of
variation (SD/mean) requires ratio (nonzero-baseline mean).
:::

## Scale choice at questionnaire design

For a satisfaction question, you could ask:

- Nominal: "Are you satisfied?" (Yes/No)
- Ordinal: "How satisfied? Very / Somewhat / Not"
- Interval: "Rate 1-10"
- Ratio: "Net Promoter Score" (arguably)

Higher scales carry more information but cost respondent effort.

::: {.notes}
Design trade-off. NPS is a good example of a scale that intentionally
throws away information (11-point becomes -100 to +100). Marketers
love its simplicity; researchers grumble about the collapse.
:::

# Part 2 — Data collection modes {background-color="#7A003C"}

## Modes: an overview

| Mode        | Cost/completion | Speed  | Coverage | Bias risks              |
|-------------|-----------------|--------|----------|-------------------------|
| Online      | $2-8            | Fast   | Broad*   | Self-selection, panels  |
| Mobile      | $2-8            | Fast   | Broad    | Screen size, dropout    |
| Phone       | $15-40          | Slow   | Narrower | Non-response, aging     |
| In-person   | $30-100         | Slow   | Narrow   | Cost, interviewer bias  |
| Mail        | $5-15           | V.slow | Narrow   | Non-response, self-sel  |
| Panel       | $5-25           | Fast   | Panel-dependent | Panel effects  |

*Broad ≠ representative. Coverage depends on the online panel used.

::: {.notes}
Cost figures are Canadian, mid-2020s, per completed interview.
Ranges are rough. Panel prices vary hugely by incidence rate. Online
has largely displaced everything else for commercial work.
:::

## Mixed-mode designs

- Push people to cheap modes (online) first
- Fall back to expensive modes (phone) for non-responders
- Adjusts for the "always online" bias

Example: StatsCan's Labour Force Survey uses phone + online + follow-up.

::: {.notes}
Mixed-mode is standard for high-quality survey research. Not always
possible for commercial work (budget). For this course we'll assume
single-mode (usually online) unless otherwise stated.
:::

## Response rates — the shrinking cornerstone

- 1970s telephone response rates: 70-80%
- 2020s telephone response rates: 5-10%
- Online panel response rates: often unmeasurable ("who was invited?")

The old distinction between "response rate" and "completion rate" matters.

::: {.notes}
Response rate = of eligible people you tried to contact, how many
responded. Completion rate = of people who started, how many finished.
Online panel providers usually only report completion. This is why
panels are convenience samples, not probability samples - you can't
compute a proper response rate.
:::

## Panel providers (industry)

- **Dynata** (formerly Research Now/SSI) — largest
- **Prodege** (formerly Prolific) — research-quality
- **MTurk** — cheap, quality varies
- **Prolific** — academic-focused, quality-controlled
- **Ipsos i-Say, Cint, Toluna** — mid-market

Choose based on: target population, quality flags, cost per complete, and speed.

::: {.notes}
Ask: has anyone commissioned or used a panel? If yes, ask which
provider. Real experience beats slides. Panel quality varies more than
prices suggest. Prolific and Dynata are usually the safe bets for
quality; MTurk needs heavy quality-control screening.
:::

# Part 3 — Sampling {background-color="#7A003C"}

## Sampling frame

- **Population:** everyone you want to draw conclusions about
- **Frame:** the list from which you actually draw
- **Sample:** the subset who respond and provide data

Frame ≠ population. Every frame excludes someone. Ask: who's missing?

::: {.notes}
Textbook question: US phone-book samples in the 1936 election
predicted Alf Landon would beat FDR. Phone owners in 1936 were
wealthier than average - skewed the sample. This is the classic
frame-vs-population failure. Modern equivalent: online panels
exclude older, poorer, more rural populations.
:::

## Probability sampling

Every unit in the frame has a known, non-zero probability of selection.

- **Simple random:** every unit equally likely
- **Systematic:** every k-th unit from a random start
- **Stratified:** random within strata (e.g., by region)
- **Cluster:** sample groups, then units within groups (e.g., schools then students)
- **Multi-stage:** combinations of the above

Only probability samples let you compute proper standard errors.

::: {.notes}
This is the "correct" way in theory. In commercial MR it's rare
because it requires a good frame and enough budget to chase
non-responders. StatsCan does probability sampling; most commercial
work does not.
:::

## Non-probability sampling

Selection is not random or probability is unknown.

- **Convenience:** whoever is easiest (students in your class)
- **Quota:** convenience but with hard targets (100 men, 100 women)
- **Judgement:** researcher picks who to include
- **Snowball:** respondents refer others (useful for rare populations)
- **Panel:** you know the panel; the panel is not the population

Cheap, fast, everywhere. Requires modelling to justify inferences.

::: {.notes}
Almost all commercial MR is non-probability now. Quality-conscious
work adjusts for panel bias via weighting or MRP (multi-level
regression and post-stratification). We won't cover MRP; mention it
exists.
:::

## When non-probability is fine

- Exploratory research (you're not inferring to a population)
- Comparative research within a panel (bias cancels out)
- Descriptive research where the panel matches your target market
- Practical constraints (budget, time)

When you write up findings, disclose the panel and its known biases.

::: {.notes}
The "bias cancels out" point matters. If you're comparing two ads on
the same panel, the panel bias affects both equally - so the
difference is still valid. Absolute levels may be biased; relative
comparisons less so.
:::

## Sample size — the trade-off

- **Bigger sample:** more precision, higher cost
- **Smaller sample:** less precision, lower cost

Size is not "as big as budget allows" - it's calculated from:

- Desired precision (confidence interval width)
- Confidence level (usually 95%)
- Expected effect size (for testing) or expected proportion (for estimation)

::: {.notes}
Give the classic formula for a proportion: n = z² × p(1-p) / e².
Show that n depends on precision (e) quadratically - halving your CI
width quadruples your sample size. This is why the standard "n=400 ≈
±5%" default exists.
:::

## Common sample-size targets

- **±5% at 95% confidence, p=0.5:** n ≈ 385
- **±3% at 95% confidence, p=0.5:** n ≈ 1067
- **Detect small effect (d=0.3) in two-sample t-test, α=0.05, power=0.8:** n ≈ 176/group

Round up. Add ~20% for expected drop-outs / bad completes.

::: {.notes}
These are the numbers students should carry in their heads. "Why did
you pick n=400?" - "±5% at 95% CI for a proportion around 50%." That's
the level of defence expected.
:::

## Break — 15 minutes {background-color="#FDBF57"}

::: {.notes}
Circulate. Have the R lab dataset (`maritime_survey.csv` and a
population frame CSV) pre-loaded on the projector.
:::

# Part 4 — R lab: sampling and power {background-color="#7A003C"}

## The `pwr` package

```{r}
#| eval: false
library(pwr)

# Sample size for a two-sample t-test
pwr.t.test(
  d = 0.3,           # Cohen's d (small)
  sig.level = 0.05,
  power = 0.8,
  type = "two.sample"
)
```

::: {.notes}
Live-code. Show output. Interpret: n=176 per group, 352 total. Then
change d to 0.5 (medium effect) - shows n=64 per group. Effect size
massively affects the required sample.
:::

## Proportion tests

```{r}
#| eval: false
# One-sample proportion test
pwr.p.test(
  h = ES.h(p1 = 0.55, p2 = 0.50),  # effect size as Cohen's h
  sig.level = 0.05,
  power = 0.8
)
```

Detecting a 5 pp difference from 50%: about n=780.

::: {.notes}
ES.h converts proportions to Cohen's h. Alternative: use
`pwr.2p.test` for two-sample proportion. Explain that "5 pp
difference" is arbitrary - real design decisions come from what
difference is managerially meaningful (a threshold).
:::

## Chi-square

```{r}
#| eval: false
pwr.chisq.test(
  w = 0.3,           # small-to-medium effect
  df = 2,            # (rows-1)*(cols-1) for contingency table
  sig.level = 0.05,
  power = 0.8
)
```

::: {.notes}
Chi-square power calculations less common in commercial MR. Cover
briefly. Note that "w=0.3" is medium effect; small = 0.1, large = 0.5
(Cohen's conventions).
:::

## Simulating a stratified sample

```{r}
#| eval: false
library(dplyr)

# Suppose maritime_survey.csv is our frame (~800 rows)
survey <- read_csv("data/maritime_survey.csv")

# Stratified sample: 50 from each region
strat_sample <- survey |>
  group_by(region) |>
  slice_sample(n = 50) |>
  ungroup()

# Check the stratification worked
count(strat_sample, region)
```

::: {.notes}
`slice_sample` is the modern way; base `sample()` still works.
Grouped slice_sample gives per-group draws. Emphasise ungrouping after
- otherwise downstream ops behave unexpectedly.
:::

## Covariate balance check

```{r}
#| eval: false
strat_sample |>
  group_by(region) |>
  summarise(
    n = n(),
    mean_age = mean(age),
    prop_female = mean(gender == "F"),
    mean_tenure_months = mean(tenure_months)
  )
```

Check that the sample looks like the population on important covariates.

::: {.notes}
Balance-checking is a habit students should build. If your sample is
skewed on age or gender relative to the population, either it's an
issue you'll flag in the report, or you'll weight during analysis.
:::

## In-lab exercise

1. For Maritime Roast's Ontario prospect population (assume N=100,000):
   - Compute n needed to estimate a proportion (e.g., unaided brand awareness) within ±4% at 95% confidence
2. Draw a stratified sample (n=200) from `maritime_survey.csv`, stratified by region
3. Verify balance on age and tenure
4. Write 3 sentences: what would you tell the CEO about your sampling plan?

A fuller worked version — including the balance-check output and a
sample CEO memo — is in **`solutions-03.qmd`** in this session's folder.

::: {.notes}
Give students ~30 min. Common issues: forgetting to ungroup, wrong
formula, treating "population size 100K" as needing an adjustment
(usually not needed unless sample > 5% of population).
:::

# Part 5 — Ethics thread {background-color="#7A003C"}

## Sampling ethics

- **Representativeness:** who's in your sample, who's not
- **Digital divide:** online panels exclude older, poorer, more rural
- **Incidence rate & respondent burden:** long surveys of specialised groups over-tax participants
- **Consent at recruitment:** was consent given for *this* research?

::: {.notes}
Digital-divide point matters for Canadian MR - rural / Northern
communities have poorer internet, older populations skew offline.
Panel-based studies of "Canadian consumers" often exclude both. Push
students to disclose sampling limits, not hide them.
:::

## Panel-quality issues

- Bots and fraud (attention checks catch most)
- Professional respondents (same person, multiple panels)
- Straight-lining (same answer down a battery)
- Speeders (finishing a 15-min survey in 3 min)

Every serious study includes quality-control screens.

::: {.notes}
Illustrate: an attention check like "For quality control, please
select 'Strongly Agree' here." Speeders defined as < 1/3 median
completion time. Straight-liners identified by low variance across a
reverse-worded battery. All easy to code post-collection.
:::

# Part 6 — Industry context {background-color="#7A003C"}

## What panel procurement actually looks like

1. **Define** target population (e.g., Canadian coffee drinkers, 25-54, HH income $50K+)
2. **Get quotes** from 2-3 panel providers (send screener + target n)
3. **Compare** cost per complete, panel size, quality flags
4. **Field** — panel provider recruits, you get responses in a CSV
5. **Quality-check** — remove bad completes, sometimes get replacements

Turnaround: 2-4 weeks typical.

::: {.notes}
Anyone who's commissioned a panel study will recognise this. Cost
per complete varies from ~$5 (broad consumer, high incidence) to $50+
(specialised B2B, low incidence). Incidence rate is the killer.
:::

## Incidence rate

- Fraction of the panel that qualifies (via screener)
- Low incidence = expensive
- Example: "Canadian coffee drinkers 25-54" ~ 60% incidence
- "Canadian B2B decision-makers on coffee suppliers" ~ 2-5% incidence

The B2B example is 12-30× more expensive per complete.

::: {.notes}
Incidence rate is the single biggest cost driver. If a study looks
too expensive, first question: can we broaden the target? Second
question: is there a specialty panel with pre-recruited members
matching this profile?
:::

# Part 7 — For your project {background-color="#7A003C"}

## What to do this week

- **Sampling plan** for your project — the Maritime Roast Ontario-entry example is your template
- Finalise your project team via A2L
- Complete TCPS-2 CORE if not already

::: {.notes}
Team formation should be locked by end of Wk 3. If it isn't,
instructor will assign. Announce this to push stragglers.
:::

## Week 4 preview

- **Concept:** Survey design and questionnaire construction
- **R lab:** cleaning survey exports (real messy Kaggle dataset)

::: {.notes}
Prep reading: SIL Ch. 8 or CJ Ch. 11 (on questionnaire design). Ask
students to look at a survey they've received recently and note what
was well-designed vs. badly-designed. We'll use these as examples.
:::

## Thank you — questions?

`abbass44@mcmaster.ca`

::: {.notes}
Stay after for any follow-up questions.
:::
