---
title: "Session 3 — Measurement, Data Collection, and Sampling"
subtitle: "M731 Week 3 · September 23, 2026 · Comprehensive Reading Notes"
author: "Shay Abbasi"
date: "2026-09-23"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    self-contained: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

# Overview

Session 3 covers three related topics in one block: measurement scales, data-collection modes, and sampling. They are related because measurement decisions (what scale to use for each question) constrain analysis options, data-collection mode decisions (how to reach respondents) affect which biases appear in the data, and sampling decisions (who is in the study and how many) determine whether findings generalise beyond the sample.

The central theme connecting all three topics is **measurement error**: the gap between what a study measures and what it intends to measure. Understanding where that gap comes from — and how to minimise it — is the most important practical skill in survey-based marketing research.

---

# Measurement

## Why Measurement Is the Foundation

Analysis is only as good as the measurement that feeds it. A regression model with perfectly specified predictors still produces biased estimates if the dependent variable is measured with systematic error. A factor analysis with excellent theoretical motivation still yields uninterpretable factors if the items were worded ambiguously. A segmentation study cannot distinguish segments that differ on constructs measured too crudely to detect the difference.

Measurement problems are particularly dangerous because they are often invisible: you cannot see measurement error in the data. A variable labeled "satisfaction" might be measuring social desirability (respondents giving answers they think are expected), acquiescence (respondents agreeing with everything regardless of content), or genuine satisfaction — and the data itself cannot tell you which. Measurement validity is established at design time, not salvaged in analysis.

## The Taxonomy of Measurement Error

Classical test theory (Novick, 1966) partitions total measurement error into two fundamentally different types:

**Systematic error (bias)** is constant, directional error that does not average out with larger samples. Its sources in survey research include:

- *Poorly worded questions*: leading phrasing, loaded language, and double-barrelled items push respondents toward particular answers regardless of their true attitude
- *Social desirability bias*: respondents report what they believe is socially acceptable rather than what is true (income is typically overreported; sensitive behaviours are underreported)
- *Acquiescence bias*: a general tendency to agree with agreement-type items regardless of content; more pronounced in self-administered surveys
- *Mode effects*: the same respondent gives systematically different answers depending on whether they are completing a survey online, by phone, or in person

**Random error** is unpredictable variation around the true value. Its sources include respondent inattention and fatigue, ambiguous question wording interpreted differently by different people, and transient mood effects at the time of measurement. Random error *does* average out with larger samples — which is why reliability (the ratio of true variance to total variance) increases with scale length. Systematic error does not average out, ever. A question with acquiescence bias at n=100 is equally biased at n=10,000.

This distinction drives a key course principle: **instrument design is a validity problem, not a statistical problem**. No amount of analytic sophistication compensates for systematic measurement error baked into the questionnaire.

## The Four Levels of Measurement

Stevens (1946) proposed four levels of measurement that differ in the mathematical operations they permit:

| Level | Distinguishes | Order | Equal intervals | Meaningful zero | Permitted statistics |
|---|---|---|---|---|---|
| Nominal | Categories | No | No | No | Mode, chi-square |
| Ordinal | Rank order | Yes | No | No | Median, Spearman |
| Interval | Equal distances | Yes | Yes | No | Mean, SD, Pearson |
| Ratio | True multiples | Yes | Yes | Yes | All, including geometric mean |

The most consequential distinction for marketing research is the difference between **ordinal** and **interval** scales, because most survey measurement instruments produce ordinal data and most analytical routines assume interval data.

### The Meaningful Zero Distinction

The difference between interval and ratio scales turns on whether the zero point represents a true absence of the property being measured.

Temperature in Celsius is interval: 0°C does not mean "no temperature" — it is an arbitrary point (the freezing point of water). Consequently, 20°C is not twice as warm as 10°C. Temperature in Kelvin is ratio: 0K is absolute zero (complete absence of thermal motion), so 200K is genuinely twice as warm as 100K.

Income is ratio: $0 means no income, and $100K is genuinely twice $50K. Calendar year is interval: Year 0 is an arbitrary historical reference, and Year 2000 is not twice Year 1000.

The practical implication: ratio variables can be log-transformed, expressed as multiples, and used in coefficient-of-variation comparisons. Interval variables cannot. You can say "revenue tripled" (ratio); you cannot say "satisfaction doubled from 3 to 6 on a 7-point scale" (interval at best).

### Common Marketing Research Examples

- **Nominal**: subscription tier name (Standard/Signature/Discovery), province of residence, brand chosen in a choice task, gender identity
- **Ordinal**: preference ranking, perceived quality (poor/fair/good/excellent), Likert item (arguably)
- **Interval**: temperature in Celsius, calendar year, most psychological scale scores
- **Ratio**: age in years, income in dollars, purchase amount, time since last purchase, number of prior visits

## The Likert Scale Debate

The treatment of Likert-type items (e.g., "I feel connected to Maritime Roast as a brand" rated 1–5 from Strongly Disagree to Strongly Agree) is one of the most debated methodological questions in applied social science.

**The ordinal position** (Jamieson, 2004) argues that the distances between adjacent scale points are unknown and likely unequal. The psychological distance between "Somewhat agree" and "Agree" may be very different from the distance between "Agree" and "Strongly agree." Computing a mean of an ordinal variable is technically impermissible — only the median is appropriate.

**The interval position** (Norman, 2010) argues that simulation studies consistently show that parametric tests (t-test, ANOVA, regression) are robust to violations of the ordinal assumption when scale points are 5 or more. Forcing non-parametric analysis loses statistical power with no practical benefit to the conclusions.

**The pragmatic resolution used in this course** follows the dominant convention in applied marketing research:

- Treat *single* Likert items as ordinal for descriptive reporting: report median and IQR, not mean and SD
- Treat *composite scales* (averages of 5+ items on the same construct) as interval for analysis: the averaging process creates a more continuous distribution, reducing the ordinal concern
- Always disclose the assumption: "composite scores from the 5-item brand-attachment battery are treated as approximately interval for regression analysis"

The disclosure matters. It is not dishonest to treat a composite scale as interval; it is dishonest to do so without acknowledging the assumption.

## Scale Sensitivity and Ceiling Effects

A common design mistake is choosing a scale that is too coarse to detect the differences you care about. Ceiling effects occur when most respondents cluster at the top of the scale, leaving no room to discriminate among the majority.

For example, if 85% of Maritime Roast subscribers rate overall satisfaction as "Very satisfied" on a 5-point scale, you cannot distinguish the happily-satisfied from the ecstatically-satisfied — or identify early warning signs of churn risk among the "merely satisfied." The fix is either a more granular scale (7-point or 10-point), a more discriminating outcome variable (e.g., repurchase intent or advocacy likelihood, which tend to have more spread than global satisfaction), or analysis at the attribute level rather than the overall-satisfaction level.

**Rule of thumb**: for concrete evaluative questions (satisfaction, quality ratings), 5-point scales are usually sufficient. For abstract attitudinal or psychological constructs (brand attachment, identity connection, authenticity perceptions), 7-point scales provide more resolution.

---

# Data Collection Modes

## The Mode Landscape

| Mode | Approx. cost/complete | Speed | Coverage | Bias risks |
|---|---|---|---|---|
| Online (panel) | $2–8 | Fast | Broad but skewed | Self-selection, professional respondents |
| Mobile | $2–8 | Fast | Broad | Screen size, dropout |
| Phone (CATI) | $15–40 | Slow | Narrower (aging) | Social desirability, non-response |
| In-person (CAPI) | $30–100 | Slow | Narrow | Cost, interviewer effects |
| Mail | $5–15 | Very slow | Narrow | Non-response, literacy |
| Panel (online) | $5–25 | Fast | Panel-dependent | Panel conditioning |

*Cost figures are Canadian mid-market, 2025, per completed interview. Ranges vary substantially with incidence rate.*

Online modes have largely displaced all others for commercial marketing research. But "broad coverage" does not mean "representative." An online panel excludes older, rural, and lower-income Canadians at rates that matter for many research questions.

## Mode Effects on Measured Attitudes

A subtle but important point: different data-collection modes produce systematically different responses from the same respondents — not just different samples, but different *answers* from the same people:

| Mode | Effect on responses |
|---|---|
| Self-administered (online, mail) | Less social-desirability bias; more honest answers on sensitive items |
| Interviewer-administered (phone, in-person) | More social-desirability bias; interviewer effects on ambiguous items |
| Mobile | Shorter responses, higher dropout on complex question formats |
| In-person CAPI | Best for showing stimuli (pack concepts, shelf displays); most expensive |

This has a critical implication for longitudinal tracking studies: if you switch data-collection mode mid-study (e.g., from phone to online), you introduce a structural break in the trend that may be entirely an artefact of the mode change rather than a genuine attitude shift. Clients who see a satisfaction score drop after a mode switch often attribute it to their marketing — a costly misinterpretation that a careful MR analyst will flag and model out.

## Response Rates: A Shrinking Cornerstone

Response rates in survey research have declined dramatically over decades:

- 1970s telephone response rates: 70–80%
- 2020s telephone response rates: 5–10%
- Online panel response rates: often unmeasurable (the denominator — who was invited — is typically not disclosed by panel providers)

The American Association for Public Opinion Research (AAPOR) defines several response rate measures, of which the most commonly cited are:

| Measure | Formula | What it means |
|---|---|---|
| Response Rate 1 (RR1) | Completes / (Completes + Refusals + Non-contacts + Unknown eligibility) | Most conservative |
| Cooperation Rate | Completes / (Completes + Refusals) | Excludes non-contacts |
| Completion Rate | Completes / Starters | What panels typically report |

Panel providers typically report *completion rates* (85–95% of those who start a survey finish it) rather than *response rates* (which would require knowing how many were invited and did not start). This is why online panels are technically non-probability samples: without a known denominator of eligible invitees, you cannot compute a defensible response rate or sampling probability.

---

# Sampling

## The Sampling Framework

Three concepts define the scope of a sampling exercise:

**Population**: everyone you want to draw conclusions about. For Maritime Roast's Ontario-entry study, the population might be "adults aged 25–54 living in Ontario who purchase coffee at least once per week."

**Frame**: the list or mechanism from which you actually draw your sample. In commercial research, the frame is typically an online panel provider's database. The frame never perfectly matches the population — some members of the population are not in the frame (excluded), and some members of the frame may not qualify for the study (screened out).

**Sample**: the subset of the frame that responds to the study and provides usable data.

The gap between frame and population is the source of **frame bias** — systematic differences between the sample you can reach and the population you want to study.

## The Literary Digest Disaster: A Canonical Cautionary Tale

The 1936 US presidential election produced the most-cited sampling failure in history. *Literary Digest* magazine mailed 10 million ballots and received 2.4 million responses (a 24% response rate — remarkable by modern standards). Their prediction: Alf Landon would win the presidency with 57% of the popular vote. Franklin Roosevelt won with 62%.

Two simultaneous errors explained the failure:

1. **Frame bias**: the mailing list was drawn from telephone directories and automobile registrations — both disproportionately wealthy and Republican-leaning populations in the middle of the Great Depression. Poor, Democratic-leaning voters who could not afford telephones or cars were systematically excluded.

2. **Non-response bias**: even within the biased frame, non-respondents (likely the poorer, more Democratic subset) differed systematically from respondents.

Meanwhile, George Gallup predicted the correct outcome using a probability sample of only 3,000 people.

The enduring lesson: **sample size does not compensate for sampling bias**. A biased frame of 10 million produces less accurate results than an unbiased sample of 3,000. Modern equivalents abound: a Twitter poll of 500,000 respondents tells you nothing about general public opinion because Twitter users are demographically unrepresentative of the population on nearly every dimension.

## Probability Sampling Methods

In probability sampling, every unit in the frame has a known, non-zero probability of selection. This is required for:

- Computing design-based standard errors (confidence intervals)
- Making valid statistical inferences from sample to population
- Satisfying the methodological standards of academic and government research

**Simple random sampling (SRS)**: every unit in the frame has an equal probability of selection. Conceptually simple; practically difficult at scale without a complete, enumerable frame.

**Systematic sampling**: select every $k$th unit from a randomly-chosen starting point. Efficient when the frame is ordered; susceptible to bias if the frame has periodic patterns.

**Stratified random sampling**: divide the frame into non-overlapping strata (e.g., by province, age group, or tier) and draw independent random samples within each stratum. The key advantage is guaranteed representation of all strata and potentially higher precision if strata are internally homogeneous.

- *Proportional stratification*: each stratum is sampled in proportion to its share of the population. Produces a self-weighting sample — no post-hoc weights are needed for overall estimates.
- *Disproportional stratification*: over-sample small strata to ensure enough cases for separate sub-group analysis. Weights are applied during analysis to restore population proportions for overall estimates.

For Maritime Roast, disproportional stratification by region is appropriate: Atlantic Canada represents only ~6% of the Canadian population, but the company needs enough Atlantic respondents to analyse the home market separately. Proportional sampling from a national frame would yield only ~24 Atlantic respondents in n=400 — too few for any meaningful sub-group analysis.

**Cluster sampling**: sample groups (clusters) first, then sample individuals within clusters. Common in educational research (sample schools, then students within schools) where individual-level frames do not exist but cluster-level frames do. Less precise than SRS for a given sample size (because within-cluster individuals tend to be similar to each other), but much cheaper when travel is required.

## Non-Probability Sampling Methods

In non-probability sampling, the probability of any unit being selected is unknown. Most commercial marketing research uses non-probability methods:

**Convenience sampling**: select whoever is easiest to reach. Low cost; high bias; appropriate only for exploratory research where population inference is not claimed.

**Quota sampling**: convenience sampling with hard caps by demographic category (e.g., 200 women, 200 men; or proportional to province shares). More representative than pure convenience; still non-probability.

**Snowball sampling**: existing participants recruit additional participants from their networks. Useful for hard-to-reach or rare populations (e.g., craft coffee enthusiasts, heavy subscription-box users) where no frame exists.

**Panel sampling**: participants are drawn from a pre-recruited panel of volunteers. The panel is a convenience sample of people willing to take surveys, which is itself a self-selected group. Panel data supports relative comparisons within the panel better than absolute level estimates.

### When Non-Probability Is Acceptable

Non-probability sampling is not inherently wrong — it depends on the research purpose:

- For **exploratory research**, you are not inferring to a population; any convenient sample serves
- For **within-panel comparisons** (e.g., comparing two ad concepts shown to the same panel), the panel bias affects both conditions equally and largely cancels out in the comparison
- For **descriptive research** where the panel's known demographic profile closely matches the target population, results can be interpreted with appropriate caveats
- For **practical constraints**: budget and timeline realities mean most commercial research cannot achieve probability sampling; the obligation is to disclose the limitation, not to pretend it away

## Weighting: Adjusting for Sampling Imbalance

When a sample is demographically unbalanced relative to the known population, **post-stratification weighting** can partially correct this. If your sample is 60% female but the target population is 50% female, you down-weight female respondents and up-weight male respondents:

```{r}
#| label: weighting-example
#| eval: false
library(tidyverse)
library(survey)

# Create post-stratification weights
survey_data <- survey_data |>
  mutate(
    weight = case_when(
      gender == "F" ~ 0.50 / 0.60,   # population share / sample share
      gender == "M" ~ 0.50 / 0.40,
      TRUE          ~ 1.0
    )
  )

# Weighted mean using base R
weighted.mean(survey_data$wtp, survey_data$weight, na.rm = TRUE)

# Proper variance estimation using the survey package
svy_design <- svydesign(ids = ~1, weights = ~weight, data = survey_data)
svymean(~wtp, svy_design, na.rm = TRUE)
```

Weighting corrects for *measurable* imbalances — demographics you can observe and compare to a known population. It cannot correct for *unmeasured* biases, such as the fact that panel members have self-selected into survey-taking, which is itself a selection effect that likely correlates with engagement, tech-savviness, and response style.

## Sample Size Derivation

Sample size is not "as many as the budget allows." It is calculated from the analytic requirements of the study. Two derivations matter most for this course:

### For Estimating a Proportion

$$n = \frac{z^2 \cdot p(1-p)}{e^2}$$

Where:
- $z$ = z-score for the desired confidence level (1.96 for 95%)
- $p$ = expected proportion (use 0.50 if unknown — this maximises the required n and is therefore conservative)
- $e$ = desired margin of error (e.g., 0.05 for ±5%)

**Worked example**: Maritime Roast wants to estimate unaided brand awareness among Ontario coffee drinkers within ±4 percentage points at 95% confidence.

$$n = \frac{1.96^2 \times 0.5 \times 0.5}{0.04^2} = \frac{3.8416 \times 0.25}{0.0016} = \frac{0.9604}{0.0016} \approx 600$$

Round up to 630 and add 25% for expected poor-quality completes and drop-outs: total fielded sample of approximately 785. This is the minimum; if the study also needs to analyse regional sub-groups separately, multiply the minimum by the number of sub-groups.

**An important insight**: for large populations, sample size does *not* depend on population size. Whether the population is 100,000 or 10,000,000, the required n for ±5% precision at 95% confidence is approximately 385. The precision of a sample is determined by its absolute size, not its proportion of the population. (The finite population correction applies only when the sample exceeds ~5% of the population — rarely relevant in commercial MR.)

### For Detecting an Effect in a Hypothesis Test

```{r}
#| label: power-analysis
#| eval: false
library(pwr)

# How many respondents per group to detect a 4pp conversion lift?
# Control rate: 10%, Treatment rate: 14%, alpha = 0.05, power = 0.80
effect_h <- ES.h(p1 = 0.14, p2 = 0.10)  # Cohen's h for two proportions

pwr.2p.test(
  h           = effect_h,
  sig.level   = 0.05,
  power       = 0.80,
  alternative = "two.sided"
)

# For a two-sample t-test with a medium effect (d = 0.5):
pwr.t.test(
  d         = 0.5,
  sig.level = 0.05,
  power     = 0.80,
  type      = "two.sample"
)
```

Cohen (1988) proposed conventional effect size thresholds for different test statistics:

| Measure | Small | Medium | Large |
|---|---|---|---|
| Cohen's d (two-group mean difference) | 0.2 | 0.5 | 0.8 |
| f (ANOVA, multi-group) | 0.1 | 0.25 | 0.4 |
| w (chi-square) | 0.1 | 0.3 | 0.5 |
| r (correlation) | 0.1 | 0.3 | 0.5 |

These are **conventions**, not laws. The effect size that matters for a decision depends on the decision context, not on Cohen's table. A d=0.2 improvement in subscriber satisfaction affecting 15,000 subscribers may matter enormously from a CLV perspective; a d=0.8 difference in a construct that drives no actionable recommendation does not matter at all. Always justify your effect size choice from the business context.

## The Sub-Group Analysis Problem

The most common sample-size planning mistake is designing for the total sample without planning for sub-group comparisons. If you plan n=400 for a study but want to compare WTP across four regions (Atlantic, Ontario, Prairie, BC), you need approximately n=100 per region for standard parametric tests. If Atlantic Canada is 6% of Canada's population and you use proportional sampling, you get only ~24 Atlantic respondents from n=400 — far too few.

The solution is **disproportional stratified sampling**: over-sample small strata to reach minimum cell sizes, then weight to the population for overall estimates. This is why research proposals should always specify the analytic plan *before* finalising the sample design — the required cell sizes drive the sampling strategy.

---

# R Lab: Sampling and Power Calculations

## The Complete Sampling Workflow

```{r}
#| label: sampling-workflow
#| eval: false
library(tidyverse)
set.seed(42)   # Always set seed for reproducibility

# 1. Load the population frame
frame <- read_csv("data/maritime_survey.csv")
cat("Frame size:", nrow(frame), "\n")

# 2. Define stratified allocation
target_n <- 200

strata_allocation <- frame |>
  count(region) |>
  mutate(
    prop_population = n / sum(n),
    target_n        = round(target_n * prop_population)
  )

strata_allocation

# 3. Draw proportional stratified sample
stratified_sample <- frame |>
  group_by(region) |>
  group_split() |>
  purrr::map2(
    strata_allocation$target_n,
    ~ slice_sample(.x, n = .y)
  ) |>
  bind_rows()

# 4. Verify stratification
count(stratified_sample, region)

# 5. Balance check: compare sample demographics to frame
bind_rows(
  frame            |> summarise(source = "Frame",  mean_age = mean(age), prop_female = mean(gender == "F")),
  stratified_sample |> summarise(source = "Sample", mean_age = mean(age), prop_female = mean(gender == "F"))
)
```

## Power Analysis Examples

```{r}
#| label: power-examples
#| eval: false
library(pwr)

# 1. Two-sample t-test: comparing mean WTP between tiers
#    Effect size d = 0.3 (small-to-medium for a pricing study)
pwr.t.test(
  d         = 0.3,
  sig.level = 0.05,
  power     = 0.80,
  type      = "two.sample"
)
# Result: n ≈ 176 per group, 352 total

# 2. Two-proportion test: A/B conversion rate test
#    Control = 10%, Treatment = 14% (4pp lift is managerially meaningful)
effect_h <- ES.h(p1 = 0.14, p2 = 0.10)
pwr.2p.test(
  h           = effect_h,
  sig.level   = 0.05,
  power       = 0.80,
  alternative = "two.sided"
)

# 3. What power do we achieve given a fixed budget (n = 500 per arm)?
pwr.2p.test(
  h           = effect_h,
  sig.level   = 0.05,
  n           = 500,
  alternative = "two.sided"
)
```

## Panel Quality Control Screens

Every panel dataset should pass through a quality control workflow before analysis. Three screens catch the most common data quality problems:

```{r}
#| label: qc-screens
#| eval: false
library(tidyverse)

raw <- read_csv("data/maritime_survey.csv")

# 1. Speeder screen: flag anyone finishing in less than 1/3 the median time
median_time <- median(raw$survey_duration_sec, na.rm = TRUE)
raw <- raw |>
  mutate(flag_speeder = survey_duration_sec < (median_time / 3))

# 2. Straight-liner screen: low SD across a Likert battery
#    True respondents vary their answers; straight-liners do not
battery_cols <- select(raw, starts_with("needs_"))
raw$item_sd  <- apply(battery_cols, 1, sd, na.rm = TRUE)
raw <- raw |>
  mutate(flag_straightliner = item_sd < 0.5)

# 3. Attention check: did they follow the embedded instruction item?
raw <- raw |>
  mutate(flag_attention = (attention_check != "Strongly agree"))

# 4. Summarise flags
raw |>
  summarise(
    pct_speeders      = mean(flag_speeder, na.rm = TRUE),
    pct_straightline  = mean(flag_straightliner, na.rm = TRUE),
    pct_attention_fail = mean(flag_attention, na.rm = TRUE),
    pct_any_flag      = mean(flag_speeder | flag_straightliner | flag_attention, na.rm = TRUE)
  )

# 5. Remove flagged respondents
clean <- raw |>
  filter(!flag_speeder, !flag_straightliner, !flag_attention)

cat("Removed:", nrow(raw) - nrow(clean), "respondents\n")
cat("Retained:", nrow(clean), "(", round(100 * nrow(clean)/nrow(raw), 1), "%)\n")
```

Typical removal rates: 5–10% speeders, 2–5% straight-liners, 3–8% attention check failures. A total removal rate of 10–20% is normal and should be factored into initial sample size planning (hence the "add 20% to required n" rule of thumb).

---

# Ethics: Sampling and Representation

## The Digital Divide

Online panel research excludes Canadians who are not online or not members of a panel. Per Statistics Canada's Survey on Canadians' Safety in the Digital Space (2021), internet use rates vary substantially by age and geography:

| Group | Internet use rate |
|---|---|
| Canadians aged 15–34 | 99% |
| Canadians aged 65–74 | 82% |
| Canadians aged 75+ | 57% |
| Urban households | 93% |
| Rural households | 79% |
| Lowest income quintile households | 80% |

For Maritime Roast's Ontario-entry study, the digital divide is less of a concern because the target market (specialty coffee subscription buyers) skews younger, urban, and higher-income — groups with near-universal internet access. If studying Atlantic Canada broadly, rural and older exclusions matter more.

The professional obligation is always to disclose rather than conceal. A good sampling limitation disclosure reads: "This study was conducted online via [panel name] and may underrepresent respondents aged 65+, rural residents, and lower-income households. Findings should be interpreted with this limitation in mind."

## Panel Ethics: Respondent Rights and Quality Obligations

Commercial panel research involves a reciprocal obligation: panel providers recruit participants with the understanding that they will be surveyed occasionally, typically with a small incentive (cash, points, or gift cards). Researchers in turn have obligations to those participants:

- **Disclosed purpose**: participants should know broadly what the research is about, even if the specific sponsor is blinded
- **Reasonable burden**: surveys should be as short as consistent with the research objectives — the "just one more question" tendency should be actively resisted
- **Data security**: responses should be stored securely and accessed only by authorised personnel
- **Consent renewal**: panel membership consent should be refreshed periodically; consent given years ago for general survey participation does not constitute consent for any specific sensitive study

## Who's Missing from Your Sample?

Every sampling decision excludes someone. The professional obligation is to identify who is missing and whether their exclusion materially affects the conclusions. A strong sampling plan includes a "limitations" section that identifies:

- *Frame exclusions*: who cannot appear in the sample (non-panel members, non-internet users, non-English/French speakers if the survey is not translated)
- *Screener exclusions*: who was explicitly screened out (non-coffee drinkers, outside the age range, outside the geography)
- *Non-response*: what we know about people who started the survey but did not complete it (drop-off patterns, if available)
- *Population implications*: which segments of the target population are under- or over-represented and how this might bias the findings

---

# How Commercial Panels Work

## The Procurement Workflow

When a firm commissions a commercial panel study, the typical workflow is:

1. **Define the target population**: demographic and behavioural screener criteria
2. **Get quotes** from 2–3 panel providers: send the screener and target n; receive a cost per complete and estimated timeline
3. **Select a provider** based on cost, panel quality certifications (Insights Association standards), and historical performance with similar studies
4. **Field the study**: the provider recruits matching panellists; responses arrive in a CSV or via API
5. **Quality-check the data**: remove bad completes (speeders, straight-liners, attention-check failures) — the provider may offer replacement completes for flagged records
6. **Analyse and report**

Turnaround for a standard consumer panel study is 2–4 weeks from brief to data delivery. More complex studies with low incidence rates or specialist populations take 4–8 weeks.

## Incidence Rate: The Primary Cost Driver

Incidence rate — the fraction of the general panel population that qualifies for the study through the screener — is the single most important cost driver in commercial primary research.

If 60% of panel members qualify (e.g., "adult Ontarians who drink coffee"), the panel provider needs to screen roughly 1.7 people per complete. If 8% qualify (e.g., "Ontarians who spend $50+ per month on coffee"), the provider needs to screen roughly 12.5 people per complete — approximately 7× the cost.

Practical implications:

- If a study looks too expensive, the first question is: "Can we broaden the target definition without losing decision validity?"
- Low-incidence studies require larger upfront fielding budgets and longer timelines
- Some specialist populations (B2B decision-makers, rare health conditions) have such low incidence in general panels that a specialty panel with pre-recruited members is more efficient

---

# References

Chapman, C., & Feit, E. M. (2023). *R for marketing research and analytics* (2nd ed.). Springer.

Cohen, J. (1988). *Statistical power analysis for the behavioral sciences* (2nd ed.). Lawrence Erlbaum.

Jamieson, S. (2004). Likert scales: How to (ab)use them. *Medical Education, 38*, 1217–1218.

Norman, G. (2010). Likert scales, levels of measurement and the "laws" of statistics. *Advances in Health Sciences Education, 15*, 625–632.

Novick, M. R. (1966). The axioms and principal results of classical test theory. *Journal of Mathematical Psychology, 3*(1), 1–18.

Silver, L., Stevens, R., Wrenn, B., & Loudon, D. (2021). *The essentials of marketing research*. SAGE Publications.

Stevens, S. S. (1946). On the theory of scales of measurement. *Science, 103*(2684), 677–680.
