---
title: "Solutions — Session 3: Sampling and Power"
subtitle: "M731 · Full worked walkthrough (companion to session-03.qmd)"
author: "Shay Abbasi"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Purpose

Full worked version of the Session 3 in-lab exercise: sample-size
calculation with `pwr`, a stratified sample drawn from
`maritime_survey.csv`, and a balance check.

## Setup

```{r}
#| label: setup
library(tidyverse)
library(pwr)

survey <- read_csv("data/maritime_survey.csv")
glimpse(survey)
```

## Step 1 — Sample size for a population proportion

Maritime Roast's Ontario prospect population is assumed at
N = 100,000. We want unaided brand awareness estimated to within ±4%
at 95% confidence.

```{r}
#| label: sample-size
# n = z^2 * p(1-p) / e^2, p=0.5 (most conservative) unless you have a prior estimate
z <- 1.96
e <- 0.04
p <- 0.5

n_needed <- (z^2 * p * (1 - p)) / e^2
n_needed
```

**Interpretation.** We need roughly `r ceiling(n_needed)` completes.
Because the sample (~601) is well under 5% of the 100,000-person
population, no finite-population correction is needed — a common
point of confusion students raise in class.

We can cross-check with `pwr` framing (`pwr.p.test` is built for
comparing a proportion to a benchmark, not simple estimation, but it's
useful to see how `ES.h` effect sizes translate to sample size for a
hypothesis-testing version of this same question):

```{r}
#| label: pwr-check
pwr.p.test(h = ES.h(p1 = 0.54, p2 = 0.50), sig.level = 0.05, power = 0.8)
```

## Step 2 — Stratified sample by region

```{r}
#| label: stratified-sample
set.seed(731)

strat_sample <- survey |>
  group_by(region) |>
  slice_sample(n = 200) |>
  ungroup()

count(strat_sample, region)
```

**Interpretation.** 200 respondents were drawn from each region
(Atlantic and Ontario_prospect), giving 400 total — comfortably above
the ~385 needed for ±5% precision, and close to what a ±4% target
would require if we'd stratified proportionally instead.

## Step 3 — Balance check

```{r}
#| label: balance-check
strat_sample |>
  group_by(region) |>
  summarise(
    n = n(),
    mean_age = mean(age),
    prop_female = mean(gender == "F"),
    mean_tenure_months = mean(tenure_months),
    .groups = "drop"
  )
```

**Interpretation.** Age, gender split, and tenure look broadly similar
across the two strata — no red flags. If one region's mean age were,
say, 10+ years older than the other's, that would be worth flagging in
the sampling section of a report, since it could confound any
region-level comparison downstream (e.g., the Week 6 WTP-by-region
test).

## Step 4 — What to tell the CEO

> Our sampling plan draws 200 Atlantic subscribers and 200 Ontario
> prospects, stratified by region so both groups are represented with
> equal precision. At n=400, we can estimate any single proportion
> (e.g., unaided awareness) to within about ±5% at 95% confidence — well
> within budget and timeline for this stage of the Ontario-entry
> decision. A larger n=601-per-group sample would tighten that to ±4%
> if the board wants sharper precision before committing a full
> marketing budget.

## Take-away for students

Sample size is never "as big as the budget allows" — it's derived from
a stated precision target, and that target should map to a real
managerial threshold (how close does an estimate need to be before the
CEO acts differently on it?).
